---
title: "Aim 2: SILA-Derived p-tau217 Positivity Risk Models"
format:
  html:
    embed-resources: true
    self-contained: true
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
execute:
  warning: false
  message: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = here::here())
```

## Overview

This analysis evaluates predictors of earlier **SILA-derived p-tau217 positivity** in PREVENT-AD using a set of Cox models built around **time-invariant genetic and risk-burden variables**. The script includes a primary APOE-focused model family, additional models incorporating **family history**, **CogDrisk**, and **LIBRA2**, and cumulative-burden models that summarize multiple risk components into a single score.

The analysis is intentionally restricted to predictors that are fixed at baseline or treated as baseline summaries so that participants can contribute SILA-estimated crossing ages that may occur before their first observed plasma visit.

**Predictor domains included:**

- APOE ε4 carrier status
- Polygenic risk score (`GRS`)
- Sex
- Education
- Family history burden
- CogDrisk score
- LIBRA2 score
- Cumulative burden scores derived from APOE, family history, high GRS, and high CogDrisk or LIBRA2

**Outcome construction:**

- **Event**: last observed p-tau217 value is above the p-tau217 positivity threshold (`val >= valt0`)
- **Event age**: SILA-estimated crossing age (`estaget0`) when available
- **Censoring age**: age at last observed p-tau217 measurement (`age_last`)

Three time-to-event specifications are used across the model set:

1. **Standard**: `Surv(surv_age, event)`
2. **Left-truncated**: `Surv(age_first, surv_age_lt, event)`
3. **Time-since-baseline**: `Surv(fu_time, event)` with baseline age included as a covariate

For the left-truncated analysis, participants with SILA-estimated event ages before their first observed plasma visit are retained and approximated as converting at study entry (`age_first + ε`) so they remain in the analysis cohort while satisfying counting-process survival requirements. The time-since-baseline models instead exclude these pre-baseline SILA events and model follow-up from the first observed plasma visit. Results are presented as descriptive tables and survival plots, adjusted Cox model summaries, and cumulative-burden visualizations.

```{r packages}
library(survival)
library(patchwork)
library(gt)
library(gtsummary)
library(survminer)
library(ComplexUpset)
library(broom)
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)

source("workflow/scripts/config.R")
```

## Load Data

```{r load-data}
sila_fits <- read_csv(
  here("results", "Aim1", "tables", "PREVENTAD_SILA_individual_fits.csv"),
  show_col_types = FALSE
) %>%
  filter(biomarker == "ptau217")

baseline_dat <- readRDS(file.path(DATA_CLEANED_PATH, "PREVENTAD_baseline.rds"))

cat("p-tau217 SILA rows loaded:", nrow(sila_fits), "\n")
cat("Participants in p-tau217 SILA data:", n_distinct(sila_fits$subid), "\n")
```

## Build Analysis Dataset

```{r build-analysis-data}
collapse_burden_factor <- function(x) {
  factor(
    if_else(x >= 3L, "3+", as.character(x)),
    levels = c("0", "1", "2", "3+")
  )
}

collapse_burden_binary <- function(x) {
  factor(
    if_else(x >= 2L, "2+", "0-1"),
    levels = c("0-1", "2+")
  )
}

# Baseline-derived burden cutoffs, aligned with the original Aim 2 Cox script.
grs_q4_cutoff <- quantile(baseline_dat$GRS, probs = 0.75, na.rm = TRUE)
cogdrisk_cutoff_1sd <- mean(baseline_dat$score_cogd, na.rm = TRUE) +
                       sd(baseline_dat$score_cogd, na.rm = TRUE)
libra2_cutoff_1sd <- mean(baseline_dat$score_libra2, na.rm = TRUE) +
                     sd(baseline_dat$score_libra2, na.rm = TRUE)

# Study entry: first observed p-tau217 visit
sila_first <- sila_fits %>%
  group_by(subid) %>%
  slice_min(age, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  transmute(
    CONP_ID = subid,
    age_first = age
  )

# Event/censoring info: last observed p-tau217 visit
sila_last <- sila_fits %>%
  group_by(subid) %>%
  slice_max(age, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  transmute(
    CONP_ID = subid,
    age_last = age,
    val = val,
    valt0 = valt0,
    estaget0 = estaget0
  ) %>%
  left_join(sila_first, by = "CONP_ID")

predictors_apoe <- baseline_dat %>%
  transmute(
    CONP_ID,
    baseline_age = age,
    apoe,
    apoe_e4 = factor(if_else(apoe == "e4+", "e4+", "non-e4+"),
                     levels = c("non-e4+", "e4+")),
    apoe_burden = if_else(apoe == "e4+", 1L, 0L),
    GRS,
    grs_burden = if_else(GRS >= grs_q4_cutoff, 1L, 0L),
    family_history,
    fhx_2plus = factor(
      if_else(grepl("2\\+", family_history), "2+ FDR", "<2 FDR"),
      levels = c("<2 FDR", "2+ FDR")
    ),
    fhx_burden = if_else(grepl("2\\+", family_history), 1L, 0L),
    score_cogd,
    cogdrisk_burden = if_else(score_cogd > cogdrisk_cutoff_1sd, 1L, 0L),
    score_libra2,
    libra2_burden = if_else(score_libra2 > libra2_cutoff_1sd, 1L, 0L),
    sex,
    educ
  )

cox_data_full <- sila_last %>%
  left_join(predictors_apoe, by = "CONP_ID") %>%
  mutate(
    event = if_else(val >= valt0, 1L, 0L),
    surv_age = if_else(event == 1L & !is.na(estaget0), estaget0, age_last),
    prebaseline_event = event == 1L & !is.na(surv_age) & surv_age < age_first
  )

# Use only variables needed for the APOE-focused models.
cox_data <- cox_data_full %>%
  drop_na(surv_age, age_first, event, apoe, GRS, sex, educ, baseline_age) %>%
  mutate(
    # Counting-process Surv(start, stop, event) requires stop > start.
    # Approximate pre-baseline SILA events as occurring at study entry.
    surv_age_lt = if_else(
      event == 1L & surv_age <= age_first,
      age_first + 1e-3,
      surv_age
    )
  )

cox_data_since_baseline <- cox_data %>%
  filter(!prebaseline_event) %>%
  mutate(
    fu_time = surv_age - age_first,
    event_group = factor(if_else(event == 1L, "A+", "A-"), levels = c("A-", "A+")),
    risk_score_cogd = apoe_burden + fhx_burden + grs_burden + cogdrisk_burden,
    risk_score_libra2 = apoe_burden + fhx_burden + grs_burden + libra2_burden,
    risk_score_cogd_f = collapse_burden_factor(risk_score_cogd),
    risk_score_libra2_f = collapse_burden_factor(risk_score_libra2),
    risk_score_cogd_bin = collapse_burden_binary(risk_score_cogd),
    risk_score_libra2_bin = collapse_burden_binary(risk_score_libra2)
  )

cox_data_risk <- cox_data %>%
  drop_na(fhx_2plus, score_cogd)

cox_data_since_baseline_risk <- cox_data_since_baseline %>%
  drop_na(fhx_2plus, score_cogd, risk_score_cogd, risk_score_cogd_f, risk_score_cogd_bin)

cox_data_since_baseline_libra <- cox_data_since_baseline %>%
  drop_na(fhx_2plus, score_libra2, risk_score_libra2, risk_score_libra2_f, risk_score_libra2_bin)

table1_data <- cox_data_since_baseline %>%
  drop_na(
    fhx_2plus,
    score_cogd,
    score_libra2,
    risk_score_cogd_f,
    risk_score_libra2_f
  )

cat("Participants in p-tau217 SILA data:", n_distinct(cox_data_full$CONP_ID), "\n")
cat("Participants after complete-case filtering:", n_distinct(cox_data$CONP_ID), "\n")
cat("Events:", sum(cox_data$event), "\n")
cat("Censored:", sum(cox_data$event == 0), "\n")
cat("Pre-baseline SILA events retained:", sum(cox_data$prebaseline_event), "\n")

cox_data %>%
  summarise(
    min_entry = round(min(age_first, na.rm = TRUE), 1),
    median_entry = round(median(age_first, na.rm = TRUE), 1),
    max_entry = round(max(age_first, na.rm = TRUE), 1),
    min_surv_age = round(min(surv_age, na.rm = TRUE), 1),
    max_surv_age = round(max(surv_age, na.rm = TRUE), 1)
  ) %>%
  print()

cat("Participants in Table 1 descriptive cohort:", n_distinct(table1_data$CONP_ID), "\n")
```

## Pre-Baseline Event Loss Table

```{r prebaseline-loss}
loss_table <- cox_data %>%
  summarise(
    analysis_set_n = n(),
    events = sum(event),
    censored = sum(event == 0),
    prebaseline_events = sum(prebaseline_event),
    retained_after_excluding_prebaseline = sum(!prebaseline_event),
    retained_events = sum(event == 1L & !prebaseline_event),
    retained_censored = sum(event == 0L & !prebaseline_event),
    baseline_followup_n = nrow(cox_data_since_baseline)
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "metric",
    values_to = "n"
  ) %>%
  mutate(
    metric = recode(
      metric,
      analysis_set_n = "Participants in APOE analysis set",
      events = "Events in APOE analysis set",
      censored = "Censored in APOE analysis set",
      prebaseline_events = "Pre-baseline SILA events",
      retained_after_excluding_prebaseline = "Participants retained after excluding pre-baseline events",
      retained_events = "Events retained after exclusion",
      retained_censored = "Censored retained after exclusion",
      baseline_followup_n = "Participants in time-since-baseline model"
    )
  )

loss_table %>%
  gt() %>%
  tab_header(
    title = "Impact of Excluding Pre-Baseline SILA Events",
    subtitle = "Counts are based on the complete-case APOE analysis set"
  )
```

## Fit Cox Models

```{r fit-models}
surv_std <- Surv(cox_data$surv_age, cox_data$event)
surv_lt  <- Surv(cox_data$age_first, cox_data$surv_age_lt, cox_data$event)
surv_baseline <- Surv(cox_data_since_baseline$fu_time, cox_data_since_baseline$event)

model_standard <- coxph(
  surv_std ~ apoe_e4 + GRS + sex + educ,
  data = cox_data,
  ties = "efron"
)

model_left_truncated <- coxph(
  surv_lt ~ apoe_e4 + GRS + sex + educ,
  data = cox_data,
  ties = "efron"
)

model_since_baseline <- coxph(
  surv_baseline ~ baseline_age + apoe_e4 + GRS + sex + educ,
  data = cox_data_since_baseline,
  ties = "efron"
)

model_since_baseline_risk <- coxph(
  Surv(fu_time, event) ~ apoe_e4 + GRS + fhx_2plus + score_cogd,
  data = cox_data_since_baseline_risk,
  ties = "efron"
)

model_since_baseline_risk_interaction <- coxph(
  Surv(fu_time, event) ~ apoe_e4 * score_cogd,
  data = cox_data_since_baseline_risk,
  ties = "efron"
)

model_since_baseline_cogd_burden <- coxph(
  Surv(fu_time, event) ~ risk_score_cogd,
  data = cox_data_since_baseline_risk,
  ties = "efron"
)

model_since_baseline_cogd_burden_factor <- coxph(
  Surv(fu_time, event) ~ risk_score_cogd_f,
  data = cox_data_since_baseline_risk,
  ties = "efron"
)

model_since_baseline_cogd_burden_binary <- coxph(
  Surv(fu_time, event) ~ risk_score_cogd_bin,
  data = cox_data_since_baseline_risk,
  ties = "efron"
)

model_since_baseline_libra <- coxph(
  Surv(fu_time, event) ~ baseline_age + sex + educ + apoe_e4 + GRS + fhx_2plus + score_libra2,
  data = cox_data_since_baseline_libra,
  ties = "efron"
)

model_since_baseline_libra_interaction <- coxph(
  Surv(fu_time, event) ~ baseline_age + sex + educ + apoe_e4 * score_libra2,
  data = cox_data_since_baseline_libra,
  ties = "efron"
)

model_since_baseline_libra_burden <- coxph(
  Surv(fu_time, event) ~ baseline_age + sex + educ + risk_score_libra2,
  data = cox_data_since_baseline_libra,
  ties = "efron"
)

model_since_baseline_libra_burden_factor <- coxph(
  Surv(fu_time, event) ~ baseline_age + sex + educ + risk_score_libra2_f,
  data = cox_data_since_baseline_libra,
  ties = "efron"
)

model_since_baseline_libra_burden_binary <- coxph(
  Surv(fu_time, event) ~ baseline_age + sex + educ + risk_score_libra2_bin,
  data = cox_data_since_baseline_libra,
  ties = "efron"
)

tidy_model <- function(model, model_label, analysis_label) {
  tidy(model, exponentiate = TRUE, conf.int = TRUE) %>%
    transmute(
      analysis = analysis_label,
      model = model_label,
      term,
      HR = estimate,
      conf.low,
      conf.high,
      p_value = p.value
    )
}

results_primary <- bind_rows(
  tidy_model(model_standard, "Standard", "Primary: APOE + GRS + sex + educ"),
  tidy_model(model_left_truncated, "Left-truncated", "Primary: APOE + GRS + sex + educ"),
  tidy_model(model_since_baseline, "Time-since-baseline", "Primary: APOE + GRS + sex + educ")
)

results_risk <- bind_rows(
  tidy_model(model_since_baseline_risk, "Time-since-baseline", "CogDrisk: APOE + GRS + family history + CogDrisk"),
  tidy_model(model_since_baseline_risk_interaction, "Time-since-baseline", "CogDrisk: APOE x CogDrisk interaction"),
  tidy_model(model_since_baseline_cogd_burden, "Time-since-baseline", "CogDrisk cumulative burden"),
  tidy_model(model_since_baseline_cogd_burden_factor, "Time-since-baseline", "CogDrisk cumulative burden factor"),
  tidy_model(model_since_baseline_cogd_burden_binary, "Time-since-baseline", "CogDrisk cumulative burden binary")
)

results_libra <- bind_rows(
  tidy_model(model_since_baseline_libra, "Time-since-baseline", "LIBRA2: APOE + GRS + family history + LIBRA2 + age + sex + educ"),
  tidy_model(model_since_baseline_libra_interaction, "Time-since-baseline", "LIBRA2: APOE x LIBRA2 interaction + age + sex + educ"),
  tidy_model(model_since_baseline_libra_burden, "Time-since-baseline", "LIBRA2 cumulative burden"),
  tidy_model(model_since_baseline_libra_burden_factor, "Time-since-baseline", "LIBRA2 cumulative burden factor"),
  tidy_model(model_since_baseline_libra_burden_binary, "Time-since-baseline", "LIBRA2 cumulative burden binary")
)

all_results <- bind_rows(results_primary, results_risk, results_libra)

results_primary
```

## Table 1

```{r table1}
table1_data %>%
  transmute(
    `p-tau217 status` = event_group,
    Age = baseline_age,
    Sex = sex,
    `Family history` = fhx_2plus,
    APOE = apoe_e4,
    GRS = GRS,
    CogDrisk = score_cogd,
    LIBRA2 = score_libra2,
    `CogDrisk cumulative burden` = risk_score_cogd_f,
    `LIBRA2 cumulative burden` = risk_score_libra2_f
  ) %>%
  tbl_summary(
    by = `p-tau217 status`,
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    digits = list(all_continuous() ~ 2),
    missing = "no"
  ) %>%
  modify_header(label ~ "**Variable**") %>%
  bold_labels() %>%
  as_gt() %>%
  tab_header(
    title = "Table 1. Time-Since-Baseline Cohort Characteristics",
    subtitle = "Columns reflect participants classified as A- versus A+ in the time-since-baseline analysis cohort"
  )
```

## UpSet Plot

```{r upset-plot}
#| fig-width: 11
#| fig-height: 6

upset_df <- table1_data %>%
  transmute(
    `p-tau217 status` = event_group,
    APOE4 = apoe_burden,
    `2+ FHx` = fhx_burden,
    `High GRS` = grs_burden,
    `High CogDrisk` = cogdrisk_burden,
    `High LIBRA2` = libra2_burden
  ) %>%
  as.data.frame()

upset(
  upset_df,
  c("APOE4", "2+ FHx", "High GRS", "High CogDrisk", "High LIBRA2"),
  name = "Risk Indicator",
  width_ratio = 0.12,
  min_degree = 0,
  set_sizes = (
    upset_set_size() +
      ylab("Set Size") +
      theme(
        text = element_text(size = 9),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
      )
  ),
  base_annotations = list(
    "Intersection size" = (
      intersection_size(
        mapping = aes(fill = `p-tau217 status`)
      ) +
        scale_fill_manual(values = c("A-" = "#2166AC", "A+" = "#B2182B")) +
        ylab("Intersection size") +
        theme(
          text = element_text(size = 9),
          legend.position = c(0.88, 0.82)
        )
    )
  )
) +
  ggtitle("Overlap of Binary Risk Indicators in the Time-Since-Baseline Cohort") +
  theme(text = element_text(size = 9))
```

## APOE-Stratified Survival Plots

```{r survival-figure}
#| fig-width: 12
#| fig-height: 5

plot_survival <- function(fit, model_label) {
  broom::tidy(fit) %>%
    mutate(
      strata = gsub("^apoe_e4=", "", strata),
      strata = factor(strata, levels = c("non-e4+", "e4+"))
    ) %>%
    ggplot(aes(x = time, y = estimate, colour = strata, fill = strata)) +
    geom_step(linewidth = 0.8) +
    geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.15, colour = NA) +
    scale_colour_manual(values = c("non-e4+" = "#2166AC", "e4+" = "#B2182B"),
                        name = "APOE ε4") +
    scale_fill_manual(values = c("non-e4+" = "#2166AC", "e4+" = "#B2182B"),
                      name = "APOE ε4") +
    labs(
      title = model_label,
      x = "Age (years)",
      y = "Probability of Remaining p-tau217 Negative"
    ) +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom")
}

sf_standard <- survfit(Surv(surv_age, event) ~ apoe_e4, data = cox_data)
sf_left_truncated <- survfit(Surv(age_first, surv_age_lt, event) ~ apoe_e4, data = cox_data)
sf_since_baseline <- survfit(Surv(fu_time, event) ~ apoe_e4, data = cox_data_since_baseline)

plot_standard <- plot_survival(sf_standard, "Standard")
plot_lt <- plot_survival(sf_left_truncated, "Left-truncated")
plot_baseline <- plot_survival(sf_since_baseline, "Time-since-baseline")

combined_plot <- (plot_standard | plot_lt) +
  plot_annotation(
    title = "APOE ε4 and SILA-Derived p-tau217 Positivity",
    subtitle = "Left-truncated panel approximates pre-baseline SILA events at study entry"
  ) &
  theme(legend.position = "bottom")

combined_plot
```

```{r survival-figure-baseline}
#| fig-width: 6
#| fig-height: 5

plot_baseline +
  plot_annotation(
    title = "APOE ε4 and p-tau217 Positivity: Time Since Baseline",
    subtitle = "Pre-baseline SILA events excluded"
  ) &
  theme(legend.position = "bottom")
```

## Results Table

```{r results-table}
results_display <- results_primary %>%
  mutate(
    term = recode(
      term,
      "apoe_e4e4+" = "APOE ε4 carrier (vs non-carrier)",
      "baseline_age" = "Baseline age (years)",
      "GRS" = "GRS (continuous)",
      "sexMale" = "Male sex (vs female)",
      "educ" = "Education (years)"
    ),
    hr_ci = sprintf("%.2f (%.2f-%.2f)", HR, conf.low, conf.high),
    p_fmt = format.pval(p_value, digits = 3, eps = 0.001)
  ) %>%
  select(model, term, hr_ci, p_fmt) %>%
  pivot_wider(
    names_from = model,
    values_from = c(hr_ci, p_fmt),
    names_glue = "{model}_{.value}"
  ) %>%
  rename(
    Term = term,
    `Standard HR (95% CI)` = `Standard_hr_ci`,
    `Standard p` = `Standard_p_fmt`,
    `LT HR (95% CI)` = `Left-truncated_hr_ci`,
    `LT p` = `Left-truncated_p_fmt`,
    `Baseline-time HR (95% CI)` = `Time-since-baseline_hr_ci`,
    `Baseline-time p` = `Time-since-baseline_p_fmt`
  )

results_display %>%
  gt() %>%
  tab_header(
    title = "APOE, GRS, Sex, and Education: Cox Models for p-tau217 Positivity",
    subtitle = "Standard, left-truncated, and time-since-baseline Cox models"
  ) %>%
  tab_spanner(
    label = "Standard",
    columns = c(`Standard HR (95% CI)`, `Standard p`)
  ) %>%
  tab_spanner(
    label = "Left-truncated",
    columns = c(`LT HR (95% CI)`, `LT p`)
  ) %>%
  tab_spanner(
    label = "Time-since-baseline",
    columns = c(`Baseline-time HR (95% CI)`, `Baseline-time p`)
  ) %>%
  tab_footnote(
    footnote = "HR > 1 indicates faster progression to estimated p-tau217 positivity on the corresponding timescale.",
    locations = cells_title()
  )
```

## CogDrisk Model

```{r results-table-risk}
risk_results_display <- results_risk %>%
  filter(analysis %in% c(
    "CogDrisk: APOE + GRS + family history + CogDrisk",
    "CogDrisk: APOE x CogDrisk interaction"
  )) %>%
  mutate(
    analysis = recode(
      analysis,
      "CogDrisk: APOE + GRS + family history + CogDrisk" = "Main effects",
      "CogDrisk: APOE x CogDrisk interaction" = "APOE x CogDrisk interaction"
    ),
    term = recode(
      term,
      "apoe_e4e4+" = "APOE ε4 carrier (vs non-carrier)",
      "GRS" = "GRS (continuous)",
      "fhx_2plus2+ FDR" = "Family history 2+ FDR (vs <2 FDR)",
      "score_cogd" = "CogDrisk score (continuous)",
      "apoe_e4e4+:score_cogd" = "APOE ε4 x CogDrisk interaction"
    ),
    hr_ci = sprintf("%.2f (%.2f-%.2f)", HR, conf.low, conf.high),
    p_fmt = format.pval(p_value, digits = 3, eps = 0.001)
  ) %>%
  select(Model = analysis, Term = term, `HR (95% CI)` = hr_ci, p = p_fmt)

risk_results_display %>%
  gt() %>%
  tab_header(
    title = "CogDrisk Models",
    subtitle = "Time-since-baseline Cox models"
  ) %>%
  tab_footnote(
    footnote = "Family history is coded as 2+ FDR versus <2 FDR. Baseline age is omitted from these models because age is embedded in CogDrisk. The interaction model excludes GRS and family history by design. HR > 1 indicates faster progression to estimated p-tau217 positivity.",
    locations = cells_title()
  )
```

```{r results-table-risk-burden}
risk_burden_display <- results_risk %>%
  filter(analysis == "CogDrisk cumulative burden") %>%
  mutate(
    term = recode(
      term,
      "risk_score_cogd" = "Cumulative burden score (APOE + family history + high GRS + high CogDrisk)"
    ),
    `HR (95% CI)` = sprintf("%.2f (%.2f-%.2f)", HR, conf.low, conf.high),
    p = format.pval(p_value, digits = 3, eps = 0.001)
  ) %>%
  select(Term = term, `HR (95% CI)`, p)

risk_burden_display %>%
  gt() %>%
  tab_header(
    title = "CogDrisk Cumulative Burden",
    subtitle = "Time-since-baseline Cox model"
  ) %>%
  tab_footnote(
    footnote = "High GRS is defined as the top quartile. High CogDrisk is defined as >1 SD above the baseline mean.",
    locations = cells_title()
  )
```

```{r results-table-risk-burden-factor}
risk_burden_factor_display <- results_risk %>%
  filter(analysis == "CogDrisk cumulative burden factor") %>%
  mutate(
    Term = case_when(
      term == "risk_score_cogd_f1" ~ "Cumulative burden = 1 (vs 0)",
      term == "risk_score_cogd_f2" ~ "Cumulative burden = 2 (vs 0)",
      term == "risk_score_cogd_f3+" ~ "Cumulative burden = 3+ (vs 0)",
      TRUE ~ term
    ),
    `HR (95% CI)` = sprintf("%.2f (%.2f-%.2f)", HR, conf.low, conf.high),
    p = format.pval(p_value, digits = 3, eps = 0.001)
  ) %>%
  select(Term, `HR (95% CI)`, p)

risk_burden_factor_display %>%
  gt() %>%
  tab_header(
    title = "CogDrisk Cumulative Burden as a Factor",
    subtitle = "Time-since-baseline Cox model"
  ) %>%
  tab_footnote(
    footnote = "Reference group is burden score 0. High GRS is defined as the top quartile. High CogDrisk is defined as >1 SD above the baseline mean. Scores of 3 and 4 are combined as 3+.",
    locations = cells_title()
  )
```

```{r results-table-risk-burden-binary}
risk_burden_binary_display <- results_risk %>%
  filter(analysis == "CogDrisk cumulative burden binary") %>%
  mutate(
    Term = case_when(
      term == "risk_score_cogd_bin2+" ~ "Cumulative burden 2+ (vs 0-1)",
      TRUE ~ term
    ),
    `HR (95% CI)` = sprintf("%.2f (%.2f-%.2f)", HR, conf.low, conf.high),
    p = format.pval(p_value, digits = 3, eps = 0.001)
  ) %>%
  select(Term, `HR (95% CI)`, p)

risk_burden_binary_display %>%
  gt() %>%
  tab_header(
    title = "CogDrisk Cumulative Burden Dichotomized",
    subtitle = "Time-since-baseline Cox model"
  ) %>%
  tab_footnote(
    footnote = "Reference group is burden score 0-1 and the exposed group is burden score 2+.",
    locations = cells_title()
  )
```

## LIBRA2 Model

```{r results-table-libra}
libra_results_display <- results_libra %>%
  filter(analysis %in% c(
    "LIBRA2: APOE + GRS + family history + LIBRA2 + age + sex + educ",
    "LIBRA2: APOE x LIBRA2 interaction + age + sex + educ"
  )) %>%
  mutate(
    analysis = recode(
      analysis,
      "LIBRA2: APOE + GRS + family history + LIBRA2 + age + sex + educ" = "Main effects",
      "LIBRA2: APOE x LIBRA2 interaction + age + sex + educ" = "APOE x LIBRA2 interaction"
    ),
    term = recode(
      term,
      "baseline_age" = "Baseline age (years)",
      "sexMale" = "Male sex (vs female)",
      "educ" = "Education (years)",
      "apoe_e4e4+" = "APOE ε4 carrier (vs non-carrier)",
      "GRS" = "GRS (continuous)",
      "fhx_2plus2+ FDR" = "Family history 2+ FDR (vs <2 FDR)",
      "score_libra2" = "LIBRA2 score (continuous)",
      "apoe_e4e4+:score_libra2" = "APOE ε4 x LIBRA2 interaction"
    ),
    `HR (95% CI)` = sprintf("%.2f (%.2f-%.2f)", HR, conf.low, conf.high),
    p = format.pval(p_value, digits = 3, eps = 0.001)
  ) %>%
  select(Model = analysis, Term = term, `HR (95% CI)`, p)

libra_results_display %>%
  gt() %>%
  tab_header(
    title = "LIBRA2 Models",
    subtitle = "Time-since-baseline Cox models"
  ) %>%
  tab_footnote(
    footnote = "Family history is coded as 2+ FDR versus <2 FDR. Baseline age, sex, and education are included as covariates because they are not embedded in LIBRA2. The interaction model excludes GRS and family history by design.",
    locations = cells_title()
  )
```

```{r results-table-libra-burden}
libra_burden_display <- results_libra %>%
  filter(analysis == "LIBRA2 cumulative burden") %>%
  mutate(
    term = recode(
      term,
      "baseline_age" = "Baseline age (years)",
      "sexMale" = "Male sex (vs female)",
      "educ" = "Education (years)",
      "risk_score_libra2" = "Cumulative burden score (APOE + family history + high GRS + high LIBRA2)"
    ),
    `HR (95% CI)` = sprintf("%.2f (%.2f-%.2f)", HR, conf.low, conf.high),
    p = format.pval(p_value, digits = 3, eps = 0.001)
  ) %>%
  select(Term = term, `HR (95% CI)`, p)

libra_burden_display %>%
  gt() %>%
  tab_header(
    title = "LIBRA2 Cumulative Burden",
    subtitle = "Time-since-baseline Cox model"
  ) %>%
  tab_footnote(
    footnote = "High GRS is defined as the top quartile. High LIBRA2 is defined as >1 SD above the baseline mean.",
    locations = cells_title()
  )
```

```{r results-table-libra-burden-factor}
libra_burden_factor_display <- results_libra %>%
  filter(analysis == "LIBRA2 cumulative burden factor") %>%
  mutate(
    Term = case_when(
      term == "baseline_age" ~ "Baseline age (years)",
      term == "sexMale" ~ "Male sex (vs female)",
      term == "educ" ~ "Education (years)",
      term == "risk_score_libra2_f1" ~ "Cumulative burden = 1 (vs 0)",
      term == "risk_score_libra2_f2" ~ "Cumulative burden = 2 (vs 0)",
      term == "risk_score_libra2_f3+" ~ "Cumulative burden = 3+ (vs 0)",
      TRUE ~ term
    ),
    `HR (95% CI)` = sprintf("%.2f (%.2f-%.2f)", HR, conf.low, conf.high),
    p = format.pval(p_value, digits = 3, eps = 0.001)
  ) %>%
  select(Term, `HR (95% CI)`, p)

libra_burden_factor_display %>%
  gt() %>%
  tab_header(
    title = "LIBRA2 Cumulative Burden as a Factor",
    subtitle = "Time-since-baseline Cox model"
  ) %>%
  tab_footnote(
    footnote = "Reference group is burden score 0. High GRS is defined as the top quartile. High LIBRA2 is defined as >1 SD above the baseline mean. Scores of 3 and 4 are combined as 3+.",
    locations = cells_title()
  )
```

```{r results-table-libra-burden-binary}
libra_burden_binary_display <- results_libra %>%
  filter(analysis == "LIBRA2 cumulative burden binary") %>%
  mutate(
    Term = case_when(
      term == "baseline_age" ~ "Baseline age (years)",
      term == "sexMale" ~ "Male sex (vs female)",
      term == "educ" ~ "Education (years)",
      term == "risk_score_libra2_bin2+" ~ "Cumulative burden 2+ (vs 0-1)",
      TRUE ~ term
    ),
    `HR (95% CI)` = sprintf("%.2f (%.2f-%.2f)", HR, conf.low, conf.high),
    p = format.pval(p_value, digits = 3, eps = 0.001)
  ) %>%
  select(Term, `HR (95% CI)`, p)

libra_burden_binary_display %>%
  gt() %>%
  tab_header(
    title = "LIBRA2 Cumulative Burden Dichotomized",
    subtitle = "Time-since-baseline Cox model"
  ) %>%
  tab_footnote(
    footnote = "Reference group is burden score 0-1 and the exposed group is burden score 2+.",
    locations = cells_title()
  )
```

## Cumulative Burden Visualizations

```{r cumulative-burden-forest}
#| fig-width: 11
#| fig-height: 5.5

burden_forest_data <- bind_rows(
  results_risk %>%
    filter(analysis == "CogDrisk cumulative burden factor") %>%
    mutate(
      framework = "CogDrisk",
      burden = case_when(
        term == "risk_score_cogd_f1" ~ "1",
        term == "risk_score_cogd_f2" ~ "2",
        term == "risk_score_cogd_f3+" ~ "3+",
        TRUE ~ NA_character_
      )
    ),
  results_libra %>%
    filter(analysis == "LIBRA2 cumulative burden factor") %>%
    mutate(
      framework = "LIBRA2",
      burden = case_when(
        term == "risk_score_libra2_f1" ~ "1",
        term == "risk_score_libra2_f2" ~ "2",
        term == "risk_score_libra2_f3+" ~ "3+",
        TRUE ~ NA_character_
      )
    )
) %>%
  filter(!is.na(burden)) %>%
  mutate(
    burden = factor(burden, levels = c("1", "2", "3+")),
    label = factor(
      paste0("Burden ", burden),
      levels = c("Burden 3+", "Burden 2", "Burden 1", "Burden 0 (ref)")
    )
  )

burden_reference_data <- tibble(
  framework = c("CogDrisk", "LIBRA2"),
  burden = factor("0", levels = c("1", "2", "3+")),
  label = factor("Burden 0 (ref)", levels = c("Burden 3+", "Burden 2", "Burden 1", "Burden 0 (ref)")),
  HR = 1,
  conf.low = 1,
  conf.high = 1
)

bind_rows(
  burden_forest_data %>% select(framework, label, HR, conf.low, conf.high),
  burden_reference_data
) %>%
  mutate(
    burden_level = factor(
      case_when(
        label == "Burden 0 (ref)" ~ "0",
        label == "Burden 1"       ~ "1",
        label == "Burden 2"       ~ "2",
        label == "Burden 3+"      ~ "3+",
        TRUE ~ NA_character_
      ),
      levels = c("0", "1", "2", "3+")
    )
  ) %>%
  ggplot(aes(x = HR, y = label, color = burden_level)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey50") +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.15) +
  geom_point(size = 2.6) +
  scale_color_viridis_d(option = "plasma", begin = 0.2, end = 0.9,
                        name = "Cumulative Burden") +
  facet_wrap(~ framework, scales = "free_y") +
  scale_x_log10() +
  labs(
    title = "Adjusted Hazard Ratios by Cumulative Burden Category",
    subtitle = "Reference group is cumulative burden score 0",
    x = "Hazard ratio (log scale)",
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold")
  )
```

```{r}

# --- LIBRA2 cumulative burden only ---
libra2_forest_data <- results_libra %>%
  filter(analysis == "LIBRA2 cumulative burden factor") %>%
  mutate(
    burden = case_when(
      term == "risk_score_libra2_f1"  ~ "1",
      term == "risk_score_libra2_f2"  ~ "2",
      term == "risk_score_libra2_f3+" ~ "3+",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(burden)) %>%
  transmute(
    label = factor(
      burden,
      levels = c("3+", "2", "1", "0")
    ),
    HR,
    conf.low,
    conf.high
  )

libra2_reference <- tibble(
  label = factor("0", levels = c("3+", "2", "1", "0")),
  HR = 1,
  conf.low = 1,
  conf.high = 1
)

plot_df <- bind_rows(libra2_forest_data, libra2_reference)

p_libra2 <- ggplot(plot_df, aes(x = HR, y = label)) +
  geom_vline(xintercept = 1, linetype = "dashed", linewidth = 0.3, color = "grey50") +
  geom_errorbarh(
    aes(xmin = conf.low, xmax = conf.high),
    height = 0.12,
    linewidth = 0.3,
    color = "black"
  ) +
  geom_point(size = 1.6, color = "#1B5E8A") +
  scale_x_log10(
    breaks = c(1, 3, 10),
    labels = c("1", "3", "10")
  ) +
  labs(
    x = "Hazard ratio",
    y = NULL
  ) +
  coord_cartesian(xlim = c(0.8, 18)) +
  theme_minimal(base_size = 8) +
  theme(
    axis.title.x = element_text(size = 8),
    axis.text.x = element_text(size = 7),
    axis.text.y = element_text(size = 7),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.major.x = element_line(linewidth = 0.2, color = "grey85"),
    plot.margin = margin(2, 2, 2, 2)
  )

p_libra2

ggsave(
  filename = "libra2_cumulative_burden_prelim.png",
  plot = p_libra2,
  width = 2,
  height = 1.6,
  units = "in",
  dpi = 600
)
```


```{r cumulative-burden-km}
#| fig-width: 11
#| fig-height: 5.5

plot_burden_survival <- function(fit, title_text) {
  broom::tidy(fit) %>%
    mutate(
      strata = sub("^.*=", "", strata),
      strata = factor(strata, levels = c("0", "1", "2", "3+"))
    ) %>%
    ggplot(aes(x = time, y = estimate, color = strata, fill = strata)) +
    geom_step(linewidth = 0.85) +
    geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.12, colour = NA) +
    scale_color_manual(
      values = c("0" = "#2166AC", "1" = "#67A9CF", "2" = "#EF8A62", "3+" = "#B2182B"),
      name = "Burden"
    ) +
    scale_fill_manual(
      values = c("0" = "#2166AC", "1" = "#67A9CF", "2" = "#EF8A62", "3+" = "#B2182B"),
      name = "Burden"
    ) +
    labs(
      title = title_text,
      x = "Years since baseline",
      y = "Probability of Remaining p-tau217 Negative"
    ) +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom")
}

sf_cogd_burden <- survfit(Surv(fu_time, event) ~ risk_score_cogd_f, data = cox_data_since_baseline_risk)
sf_libra_burden <- survfit(Surv(fu_time, event) ~ risk_score_libra2_f, data = cox_data_since_baseline_libra)

(plot_burden_survival(sf_cogd_burden, "CogDrisk cumulative burden") |
  plot_burden_survival(sf_libra_burden, "LIBRA2 cumulative burden")) +
  plot_annotation(
    title = "Time-Since-Baseline Survival by Cumulative Burden Group",
    subtitle = "Curves are descriptive and unadjusted; factor-based Cox models provide adjusted hazard ratios"
  ) &
  theme(legend.position = "bottom")
```

```{r cumulative-burden-adjusted}
#| fig-width: 11
#| fig-height: 5.5

burden_palette <- c("0" = "#2166AC", "1" = "#67A9CF", "2" = "#EF8A62", "3+" = "#B2182B")

mode_value <- function(x) {
  ux <- na.omit(unique(x))
  ux[which.max(tabulate(match(x, ux)))]
}

cogd_newdata <- tibble(
  risk_score_cogd_f = factor(c("0", "1", "2", "3+"), levels = c("0", "1", "2", "3+"))
)

libra_newdata <- tibble(
  baseline_age = mean(cox_data_since_baseline_libra$baseline_age, na.rm = TRUE),
  sex = mode_value(cox_data_since_baseline_libra$sex),
  educ = mean(cox_data_since_baseline_libra$educ, na.rm = TRUE),
  risk_score_libra2_f = factor(c("0", "1", "2", "3+"), levels = c("0", "1", "2", "3+"))
)

sf_cogd_adjusted <- survfit(model_since_baseline_cogd_burden_factor, newdata = cogd_newdata)
sf_libra_adjusted <- survfit(model_since_baseline_libra_burden_factor, newdata = libra_newdata)

adjusted_cogd_plot <- ggsurvplot(
  sf_cogd_adjusted,
  data = as.data.frame(cox_data_since_baseline_risk),
  conf.int = TRUE,
  censor = FALSE,
  legend.title = "Burden",
  legend.labs = c("0", "1", "2", "3+"),
  palette = burden_palette,
  ggtheme = theme_minimal(base_size = 12),
  xlab = "Years since baseline",
  ylab = "Adjusted probability of remaining p-tau217 negative",
  title = "CogDrisk cumulative burden"
)$plot +
  theme(legend.position = "bottom")

adjusted_libra_plot <- ggsurvplot(
  sf_libra_adjusted,
  data = as.data.frame(cox_data_since_baseline_libra),
  conf.int = TRUE,
  censor = FALSE,
  legend.title = "Burden",
  legend.labs = c("0", "1", "2", "3+"),
  palette = burden_palette,
  ggtheme = theme_minimal(base_size = 12),
  xlab = "Years since baseline",
  ylab = "Adjusted probability of remaining p-tau217 negative",
  title = "LIBRA2 cumulative burden"
)$plot +
  theme(legend.position = "bottom")

(adjusted_cogd_plot | adjusted_libra_plot) +
  plot_annotation(
    title = "Adjusted Survival Curves by Cumulative Burden Group",
    subtitle = "Curves are model-based predictions with 95% confidence bands from survfit() and survminer::ggsurvplot()"
  ) &
  theme(legend.position = "bottom")
```

```{r cumulative-burden-adjusted-binary}
#| fig-width: 11
#| fig-height: 5.5

cogd_bin_newdata <- tibble(
  risk_score_cogd_bin = factor(c("0-1", "2+"), levels = c("0-1", "2+"))
)

libra_bin_newdata <- tibble(
  baseline_age = mean(cox_data_since_baseline_libra$baseline_age, na.rm = TRUE),
  sex = mode_value(cox_data_since_baseline_libra$sex),
  educ = mean(cox_data_since_baseline_libra$educ, na.rm = TRUE),
  risk_score_libra2_bin = factor(c("0-1", "2+"), levels = c("0-1", "2+"))
)

sf_cogd_adjusted_bin <- survfit(model_since_baseline_cogd_burden_binary, newdata = cogd_bin_newdata)
sf_libra_adjusted_bin <- survfit(model_since_baseline_libra_burden_binary, newdata = libra_bin_newdata)

binary_palette <- c("0-1" = "#2166AC", "2+" = "#B2182B")

adjusted_cogd_plot_bin <- ggsurvplot(
  sf_cogd_adjusted_bin,
  data = as.data.frame(cox_data_since_baseline_risk),
  conf.int = TRUE,
  censor = FALSE,
  legend.title = "Burden",
  legend.labs = c("0-1", "2+"),
  palette = binary_palette,
  ggtheme = theme_minimal(base_size = 12),
  xlab = "Years since baseline",
  ylab = "Adjusted probability of remaining p-tau217 negative",
  title = "CogDrisk burden dichotomized"
)$plot +
  theme(legend.position = "bottom")

adjusted_libra_plot_bin <- ggsurvplot(
  sf_libra_adjusted_bin,
  data = as.data.frame(cox_data_since_baseline_libra),
  conf.int = TRUE,
  censor = FALSE,
  legend.title = "Burden",
  legend.labs = c("0-1", "2+"),
  palette = binary_palette,
  ggtheme = theme_minimal(base_size = 12),
  xlab = "Years since baseline",
  ylab = "Adjusted probability of remaining p-tau217 negative",
  title = "LIBRA2 burden dichotomized"
)$plot +
  theme(legend.position = "bottom")

(adjusted_cogd_plot_bin | adjusted_libra_plot_bin) +
  plot_annotation(
    title = "Adjusted Survival Curves by Dichotomized Cumulative Burden",
    subtitle = "Curves are model-based predictions with 95% confidence bands for burden 0-1 versus 2+"
  ) &
  theme(legend.position = "bottom")
```

## Save Results

```{r save-results}
dir.create(here("results", "Aim2", "tables"), recursive = TRUE, showWarnings = FALSE)
dir.create(here("results", "Aim2", "figures"), recursive = TRUE, showWarnings = FALSE)

write_csv(
  all_results,
  here("results", "Aim2", "tables", "PREVENTAD_cox_sila_apoe_results.csv")
)

write_csv(
  results_primary,
  here("results", "Aim2", "tables", "PREVENTAD_cox_sila_apoe_primary.csv")
)

ggsave(
  filename = here("results", "Aim2", "figures", "PREVENTAD_cox_sila_apoe_survival.png"),
  plot = combined_plot,
  width = 12,
  height = 5,
  dpi = 300
)

ggsave(
  filename = here("results", "Aim2", "figures", "PREVENTAD_cox_sila_apoe_survival_baseline_time.png"),
  plot = plot_baseline,
  width = 6,
  height = 5,
  dpi = 300
)

cat("Results saved to:\n")
cat("- results/Aim2/tables/PREVENTAD_cox_sila_apoe_results.csv\n")
cat("- results/Aim2/tables/PREVENTAD_cox_sila_apoe_primary.csv\n")
cat("- results/Aim2/figures/PREVENTAD_cox_sila_apoe_survival.png\n")
cat("- results/Aim2/figures/PREVENTAD_cox_sila_apoe_survival_baseline_time.png\n")
```
