---
title: "Aim 2: Cox Models with SILA Bootstrap Uncertainty"
format:
  html:
    embed-resources: true
    self-contained: true
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
execute:
  warning: false
  message: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = here::here())
```

## Overview

This analysis extends `aim2_cox_sila.qmd` with two key additions:

**1. Re-inclusion of single-visit participants.**
The main SILA pipeline requires ≥2 longitudinal biomarker observations to fit individual trajectories and was filtered accordingly. However, participants with exactly one GWAS-eligible plasma measurement are still informative for Cox models: their single observed value relative to the Youden threshold (p-tau181: 6.42 pg/mL; p-tau217: 3.06 pg/mL) defines their event status directly. They are included here as observed events or right-censored observations.

**2. Bootstrap propagation of SILA estimation uncertainty.**
SILA fits are treated as fixed point estimates in `aim2_cox_sila.qmd`. 10 bootstrap samples of multi-visit participants are drawn with replacement, SILA is refit on each sample, and the full Cox + Kaplan-Meier pipeline is re-estimated. The spread across 10 iterations provides confidence intervals that incorporate both sampling variability and SILA fitting uncertainty.

**Primary bootstrap output:**
Survival probability at each age with 95% bootstrap CI bands from the 10 KM curves per group.


```{r packages}
library(survival)
library(broom)
library(patchwork)
library(gt)
library(furrr)
library(silaR)
library(pROC)
library(readxl)

source('workflow/scripts/config.R')
```

## Load and Prepare Data

```{r load-data}
# Multi-visit participants (≥2 biomarker visits + GWAS) 
all_dat      <- readRDS(file.path(DATA_CLEANED_PATH, "PREVENTAD_longitudinal.rds"))
baseline_dat <- readRDS(file.path(DATA_CLEANED_PATH, "PREVENTAD_baseline.rds"))

mv_ids <- unique(all_dat$CONP_ID)

cat("Multi-visit participants (≥2 visits + GWAS):", length(mv_ids), "\n")
cat("Total longitudinal observations:", nrow(all_dat), "\n")

# GWAS-only biomarker data (all GWAS-eligible, including single-visit) 
biomarkers_gwas_all <- readRDS(
  file.path(DATA_INTERMEDIATE_PATH, "PREVENTAD_biomarkers_gwas_all.rds")
)

all_gwas_ids <- unique(biomarkers_gwas_all$CONP_ID)
sv_ids       <- setdiff(all_gwas_ids, mv_ids)

cat("\nTotal GWAS-eligible participants:", length(all_gwas_ids), "\n")
cat("Single-visit participants:", length(sv_ids), "\n")

# Single-visit biomarker observations (one row per participant per biomarker)
single_visit_wide <- biomarkers_gwas_all %>%
  filter(CONP_ID %in% sv_ids) %>%
  group_by(CONP_ID) %>%
  slice_max(age, n = 1, with_ties = FALSE) %>%
  ungroup()

# Predictors for single-visit participants
PREVENTAD_raw <- readRDS(file.path(DATA_INTERMEDIATE_PATH, "PREVENTAD_raw.rds"))

# Computed risk scores (score_cogd, score_libra2, family_history)
sv_scores <- readRDS(file.path(DATA_INTERMEDIATE_PATH, "PREVENTAD_singlevisi_scores.rds"))

# APOE genotype
sv_apoe <- PREVENTAD_raw$genetics %>%
  filter(CONP_ID %in% sv_ids) %>%
  mutate(
    genotype = str_replace_all(APOE, " ", ""),
    allele1  = as.numeric(str_extract(genotype, "^\\d")),
    allele2  = as.numeric(str_extract(genotype, "\\d$")),
    apoe     = case_when(
      APOE == "3 2"                             ~ "e2+",
      APOE == "3 3"                             ~ "e3/e3",
      APOE %in% c("4 2", "4 3", "4 4")         ~ "e4+",
      TRUE                                      ~ NA_character_
    ),
    apoe = factor(apoe, levels = c("e3/e3", "e2+", "e4+"))
  ) %>%
  select(CONP_ID, apoe)

# GRS for single-visit participants
bellenguez_results <- readxl::read_xlsx(
  "data/raw/41588_2022_1024_MOESM4_ESM.xlsx",
  sheet = "Supplementary Table 32",
  skip  = 2
)
GRSweights <- data.frame(
  snp           = paste0(bellenguez_results$rsid, "_", bellenguez_results$`Risk allele`),
  beta          = bellenguez_results$Beta,
  effect_allele = bellenguez_results$`Risk allele`
)

calculate_grs <- function(geno_matrix, weights_df) {
  common_snps    <- intersect(colnames(geno_matrix), weights_df$snp)
  geno_subset    <- geno_matrix[, common_snps, drop = FALSE]
  weights_subset <- weights_df[match(common_snps, weights_df$snp), ]
  as.numeric(geno_subset %*% weights_subset$beta)
}

gwas_raw_all  <- PREVENTAD_raw$GWAS
id_cols       <- c("CONP_ID", "CONP_CandID")
geno_matrix   <- as.matrix(gwas_raw_all[, !names(gwas_raw_all) %in% id_cols])
mode(geno_matrix) <- "integer"
grs_all <- data.frame(
  CONP_ID = gwas_raw_all$CONP_ID,
  GRS     = calculate_grs(geno_matrix, GRSweights)
)

sv_grs <- grs_all %>% filter(CONP_ID %in% sv_ids)

# Sex and education from demographics
sv_demo <- PREVENTAD_raw$demographics %>%
  filter(CONP_ID %in% sv_ids) %>%
  mutate(sex = factor(Sex, levels = c("Female", "Male"))) %>%
  select(CONP_ID, sex, educ = Education_years)

# Combine single-visit predictors
sv_predictors <- sv_apoe %>%
  left_join(sv_grs,                                                          by = "CONP_ID") %>%
  left_join(sv_scores %>% select(CONP_ID, score_cogd, score_libra2,
                                  family_history),                           by = "CONP_ID") %>%
  left_join(sv_demo,                                                         by = "CONP_ID")

cat("\nSingle-visit predictor coverage:\n")
sv_predictors %>%
  summarise(
    n             = n(),
    has_apoe      = sum(!is.na(apoe)),
    has_grs       = sum(!is.na(GRS)),
    has_fhx       = sum(!is.na(family_history)),
    has_sex       = sum(!is.na(sex)),
    has_cogd      = sum(!is.na(score_cogd)),
    has_libra2    = sum(!is.na(score_libra2))
  ) %>%
  print()

# Amyloid status from PET data
THRESHOLD_AMYLOID <- 22.32 

pet_nav <- read_csv(
  file.path(PREVENTAD_DATA_PATH,
            "PET_NAV_SUVR_ref-wholeCerebellum_Registered_PREVENTAD_2025-03-21.csv"),
  show_col_types = FALSE
)

amyloid_lookup <- pet_nav %>%
  filter(PET_tracer == "NAV") %>%
  select(CONP_ID, centiloid = WhlCbl_centiloid) %>%
  distinct() %>%
  group_by(CONP_ID) %>%
  slice_max(centiloid, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(
    amyloid_status = factor(
      if_else(centiloid >= THRESHOLD_AMYLOID, "A+", "A-"),
      levels = c("A-", "A+")
    )
  ) %>%
  select(CONP_ID, amyloid_status)
```

```{r create-risk-burden}
# Biomarker thresholds 
THRESHOLD_PTAU181 <- 6.42   # pg/mL
THRESHOLD_PTAU217 <- 3.06   # pg/mL

# Baseline risk burden cutoffs 
baseline_only_mv <- all_dat %>%
  group_by(CONP_ID) %>%
  slice_min(years, n = 1, with_ties = FALSE) %>%
  ungroup()

CUTOFF_COGDRISK <- 7.52
CUTOFF_LIBRA2 <- 7.35

# Multi-visit predictors
predictors_mv <- baseline_dat %>%
  mutate(
    grs_quartile    = ntile(GRS, 4),
    grs_burden      = if_else(grs_quartile == 4, 1L, 0L),
    fhx_burden      = case_when(str_detect(family_history, "2\\+") ~ 1L, TRUE ~ 0L),
    apoe_burden     = if_else(apoe == "e4+", 1L, 0L),
    cogdrisk_burden = if_else(score_cogd   > CUTOFF_COGDRISK, 1L, 0L),
    libra2_burden   = if_else(score_libra2 > CUTOFF_LIBRA2,   1L, 0L),
    risk_score_cogd   = fhx_burden + apoe_burden + cogdrisk_burden + grs_burden,
    risk_score_libra2 = fhx_burden + apoe_burden + libra2_burden   + grs_burden,
  ) %>%
  select(CONP_ID, age, sex, educ, apoe, apoe_burden, family_history, fhx_burden,
         GRS, grs_burden, score_cogd, cogdrisk_burden,
         score_libra2, libra2_burden,
         risk_score_cogd, risk_score_libra2)

# Single-visit predictors
grs_q4_cutoff <- quantile(baseline_dat$GRS, 0.75, na.rm = TRUE)

predictors_sv <- sv_predictors %>%
  mutate(
    grs_burden  = if_else(!is.na(GRS) & GRS >= grs_q4_cutoff, 1L, 0L),
    fhx_burden  = case_when(str_detect(as.character(family_history), "2\\+") ~ 1L,
                             !is.na(family_history)                          ~ 0L,
                             TRUE                                            ~ NA_integer_),
    apoe_burden = if_else(apoe == "e4+", 1L, 0L),
    cogdrisk_burden = if_else(!is.na(score_cogd) & score_cogd > CUTOFF_COGDRISK, 1L, 0L),
    libra2_burden   = if_else(!is.na(score_libra2) & score_libra2 > CUTOFF_LIBRA2, 1L, 0L),
    risk_score_cogd   = if_else(
      !is.na(cogdrisk_burden),
      fhx_burden + apoe_burden + cogdrisk_burden + grs_burden,
      NA_integer_
    ),
    risk_score_libra2 = if_else(
      !is.na(libra2_burden),
      fhx_burden + apoe_burden + libra2_burden + grs_burden,
      NA_integer_
    )
  ) %>%
  # Add age from their single biomarker visit
  left_join(single_visit_wide %>% select(CONP_ID, age), by = "CONP_ID") %>%
  select(CONP_ID, age, sex, educ, apoe, apoe_burden, family_history, fhx_burden,
         GRS, grs_burden, score_cogd, cogdrisk_burden,
         score_libra2, libra2_burden,
         risk_score_cogd, risk_score_libra2)

# Combined predictor table (multi-visit + single-visit)
predictors_all <- bind_rows(predictors_mv, predictors_sv)
```

```{r build-cox-data}
# Multi-visit: derive event status from existing SILA fits
# Use stored SILA output (the reference fit from the full cohort)
sila_fits <- read_csv(
  here("results", "Aim1", "tables", "PREVENTAD_SILA_individual_fits.csv"),
  show_col_types = FALSE
)

thresholds <- list(ptau181 = THRESHOLD_PTAU181, ptau217 = THRESHOLD_PTAU217)

sila_first_mv <- sila_fits %>%
  group_by(subid, biomarker) %>%
  slice_min(age, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  rename(CONP_ID = subid, age_first = age) %>%
  select(CONP_ID, biomarker, age_first)

sila_last_mv <- sila_fits %>%
  group_by(subid, biomarker) %>%
  slice_max(age, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  rename(CONP_ID = subid, age_last = age) %>%
  left_join(sila_first_mv, by = c("CONP_ID", "biomarker")) %>%
  mutate(event    = if_else(val >= valt0, 1L, 0L),
         # Events --> SILA-estimated crossing age; censored --> last observed visit age
         surv_age = if_else(val >= valt0 & !is.na(estaget0), estaget0, age_last),
         source   = "multi-visit")

# ingle-visit: event from observed value vs threshold
sila_last_sv <- single_visit_wide %>%
  select(CONP_ID, age, ptau181, ptau217) %>%
  pivot_longer(cols = c(ptau181, ptau217), names_to = "biomarker", values_to = "val") %>%
  filter(!is.na(val), is.finite(val)) %>%
  mutate(
    valt0    = if_else(biomarker == "ptau181", THRESHOLD_PTAU181, THRESHOLD_PTAU217),
    event    = if_else(val >= valt0, 1L, 0L),
    age_last = age,
    # Use age_first = age_last - 0.5 for left-truncation (single visit = no earlier obs)
    age_first = age_last - 0.5,
    surv_age  = age_last,
    source    = "single-visit"
  ) %>%
  select(CONP_ID, biomarker, val, valt0, event, age_last, age_first, surv_age, source)

# Join with predictors + amyloid
cox_data_full <- bind_rows(
  sila_last_mv %>% select(CONP_ID, biomarker, val, valt0, event, age_last, age_first, surv_age, source),
  sila_last_sv
) %>%
  left_join(predictors_all, by = "CONP_ID") %>%
  left_join(amyloid_lookup,  by = "CONP_ID")

cat("\nCox data summary:\n")
cox_data_full %>%
  group_by(biomarker, source) %>%
  summarise(
    n          = n(),
    n_events   = sum(event, na.rm = TRUE),
    pct_events = round(mean(event, na.rm = TRUE) * 100, 1),
    .groups    = "drop"
  ) %>%
  print()
```

## Kaplan-Meier Survival Curves (Observed Data)

```{r km-apoe-observed}
#| fig-width: 12
#| fig-height: 5

# KM by APOE e4 carrier status 
plot_km_apoe <- function(data, biomarker_label) {
  dat <- data %>%
    drop_na(apoe_burden, surv_age, event) %>%
    mutate(apoe_group = factor(if_else(apoe_burden == 1L, "e4+", "e4-"),
                               levels = c("e4-", "e4+")))

  km_formula <- as.formula("Surv(surv_age, event) ~ apoe_group")
  km_fit     <- survfit(km_formula, data = dat)
  km_tidy    <- broom::tidy(km_fit) %>%
    mutate(strata = gsub("^apoe_group=", "", strata))

  sd_fit     <- survdiff(km_formula, data = dat)
  lr_p       <- pchisq(sd_fit$chisq, df = 1, lower.tail = FALSE)
  pval_label <- if (lr_p < 0.001) "p < 0.001" else paste0("p = ", round(lr_p, 3))

  ggplot(km_tidy, aes(x = time, y = estimate, colour = strata, fill = strata)) +
    geom_step(linewidth = 0.8) +
    geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2, colour = NA) +
    scale_colour_manual(values = c("e4-" = "#2166AC", "e4+" = "#B2182B"),
                        name = "APOE \u03b54 status") +
    scale_fill_manual(values   = c("e4-" = "#2166AC", "e4+" = "#B2182B"),
                      name = "APOE \u03b54 status") +
    annotate("text", x = Inf, y = Inf, hjust = 1.05, vjust = 1.5,
             label = pval_label, size = 3.5) +
    labs(title = paste(biomarker_label, "\u2013 APOE \u03b54 Status (observed data)"),
         x = "Age (years)", y = "Probability of ptau Negativity") +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom")
}

cox_ptau181 <- cox_data_full %>% filter(biomarker == "ptau181") %>%
  drop_na(surv_age, event)
cox_ptau217 <- cox_data_full %>% filter(biomarker == "ptau217") %>%
  drop_na(surv_age, event)

print(plot_km_apoe(cox_ptau181, "p-tau181"))
print(plot_km_apoe(cox_ptau217, "p-tau217"))
```

```{r km-risk-score}
#| fig-width: 14
#| fig-height: 6

# Collapse top level to "3+" (only 1 participant has score = 4)
collapse_risk <- function(x) {
  factor(if_else(x >= 3, "3+", as.character(x)), levels = c("0", "1", "2", "3+"))
}

plot_km_risk <- function(data, biomarker_label, group_var, group_label,
                          palette = c("0" = "#2166AC", "1" = "#74ADD1",
                                      "2" = "#F46D43", "3+" = "#B2182B")) {
  km_formula <- as.formula(paste("Surv(surv_age, event) ~", group_var))
  km_fit     <- survfit(km_formula, data = data)
  km_tidy    <- broom::tidy(km_fit) %>%
    mutate(strata = gsub(paste0("^", group_var, "="), "", strata))

  sd_fit     <- survdiff(km_formula, data = data)
  lr_p       <- pchisq(sd_fit$chisq, df = length(sd_fit$n) - 1, lower.tail = FALSE)
  pval_label <- if (lr_p < 0.001) "p < 0.001" else paste0("p = ", round(lr_p, 3))

  ggplot(km_tidy, aes(x = time, y = estimate, colour = strata, fill = strata)) +
    geom_step(linewidth = 0.8) +
    geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.15, colour = NA) +
    scale_colour_manual(values = palette, name = group_label) +
    scale_fill_manual(values   = palette, name = group_label) +
    annotate("text", x = Inf, y = Inf, hjust = 1.05, vjust = 1.5,
             label = pval_label, size = 3.5) +
    labs(title = paste(biomarker_label, "\u2013", group_label),
         x = "Age (years)", y = "Probability of ptau Negativity") +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom")
}

cox_ptau181 <- cox_ptau181 %>%
  mutate(risk_score_cogd_f   = collapse_risk(risk_score_cogd),
         risk_score_libra2_f = collapse_risk(risk_score_libra2))

cox_ptau217 <- cox_ptau217 %>%
  mutate(risk_score_cogd_f   = collapse_risk(risk_score_cogd),
         risk_score_libra2_f = collapse_risk(risk_score_libra2))

print(plot_km_risk(cox_ptau181, "p-tau181", "risk_score_cogd_f",
                   "No. CogDrisk indicators (0\u20133+)"))
print(plot_km_risk(cox_ptau181, "p-tau181", "risk_score_libra2_f",
                   "No. LIBRA2 indicators (0\u20133+)"))
print(plot_km_risk(cox_ptau217, "p-tau217", "risk_score_cogd_f",
                   "No. CogDrisk indicators (0\u20133+)"))
print(plot_km_risk(cox_ptau217, "p-tau217", "risk_score_libra2_f",
                   "No. LIBRA2 indicators (0\u20133+)"))
```

## Bootstrap: SILA Uncertainty Propagation

### Method

For each of 10 bootstrap iterations:  
1. **Stratified bootstrap**: Multi-visit participants are resampled with replacement, stratified by biomarker positivity status at last visit (within each biomarker), preserving the event rate.  
2. **SILA refitting**: SILA is refit on the bootstrap sample using the fixed Youden thresholds. Duplicated participants receive unique subids (`CONP_ID_N`) so SILA treats them as independent.  
3. **Single-visit participants**: Added to each iteration with fixed event status (not bootstrapped, as SILA requires ≥2 observations).  
4. **Cox + KM**: Cox models (M1–M4) and KM curves stratified by amyloid status are re-estimated on the combined data.  
5. **Stored outputs**: Cox HRs and KM survival estimates at each age in `age_grid`.  

After 10 iterations, 95% bootstrap CI bands are the 2.5th/97.5th percentile across iterations.  

```{r define-bootstrap-function}
# Pre-compute single-visit data in long format for efficient use in bootstrap
single_visit_long <- single_visit_wide %>%
  select(CONP_ID, age, ptau181, ptau217) %>%
  pivot_longer(cols = c(ptau181, ptau217), names_to = "biomarker", values_to = "val") %>%
  filter(!is.na(val), is.finite(val)) %>%
  mutate(
    valt0     = if_else(biomarker == "ptau181", THRESHOLD_PTAU181, THRESHOLD_PTAU217),
    event     = if_else(val >= valt0, 1L, 0L),
    age_last  = age,
    surv_age  = age,   # single-visit: no estaget0 available; always use observed age
    age_first = age - 0.5
  )

# Pre-pivot multi-visit data for SILA bootstrap
multi_visit_long <- all_dat %>%
  select(CONP_ID, age, ptau181, ptau217) %>%
  pivot_longer(cols = c(ptau181, ptau217), names_to = "biomarker", values_to = "val") %>%
  filter(!is.na(val), is.finite(val)) %>%
  arrange(CONP_ID, biomarker, age)

# Age grid for survival estimates
age_grid <- seq(55, 90, by = 0.5)

# Quartile boundaries (fixed from first-visit cross-sectional slice) 
quartile_cuts <- purrr::map(c("ptau181", "ptau217"), function(bm) {
  first_vals <- multi_visit_long %>%
    filter(biomarker == bm) %>%
    group_by(CONP_ID) %>%
    slice_min(age, n = 1, with_ties = FALSE) %>%
    ungroup()
  list(
    q25 = quantile(first_vals$val, 0.25, na.rm = TRUE),
    q50 = quantile(first_vals$val, 0.50, na.rm = TRUE),
    q75 = quantile(first_vals$val, 0.75, na.rm = TRUE)
  )
}) %>% setNames(c("ptau181", "ptau217"))

# Bootstrap iteration function
run_bootstrap_iter <- function(iter,
                                multi_visit_long,
                                single_visit_long,
                                predictors_all,
                                amyloid_lookup,
                                thresholds,
                                quartile_cuts,
                                age_grid) {
  set.seed(iter)

  bm_results <- purrr::map(c("ptau181", "ptau217"), function(bm) {
    thr <- thresholds[[bm]]
    mv  <- multi_visit_long %>% filter(biomarker == bm)

    # Assign first-visit quartile stratum using fixed boundaries
    cuts <- quartile_cuts[[bm]]
    first_obs <- mv %>%
      group_by(CONP_ID) %>%
      slice_min(age, n = 1, with_ties = FALSE) %>%
      ungroup() %>%
      mutate(
        quartile = case_when(
          val <= cuts$q25 ~ "Q1",
          val <= cuts$q50 ~ "Q2",
          val <= cuts$q75 ~ "Q3",
          TRUE            ~ "Q4"
        )
      )

    # Stratified bootstrap: sample with replacement within each quartile
    boot_ids <- first_obs %>%
      group_by(quartile) %>%
      slice_sample(prop = 1, replace = TRUE) %>%
      mutate(boot_row = row_number()) %>%
      ungroup() %>%
      select(CONP_ID, boot_row)

    # All observations for bootstrapped participants, with unique subids
    boot_sila <- mv %>%
      right_join(boot_ids, by = "CONP_ID") %>%
      mutate(subid = paste0(CONP_ID, "_", boot_row)) %>%
      # IQR outlier removal (mirrors SILA QMD preprocessing)
      {
        q1 <- quantile(.$val, 0.25, na.rm = TRUE)
        q3 <- quantile(.$val, 0.75, na.rm = TRUE)
        iqr <- q3 - q1
        filter(., val >= (q1 - 3 * iqr) & val <= (q3 + 3 * iqr))
      } %>%
      select(subid, age, val) %>%
      arrange(subid, age)

    # Fit SILA wrapped in tryCatch; returns NULL on convergence failure
    tryCatch({
      # Verify threshold is within data range
      thr_adj <- if (thr < min(boot_sila$val) || thr > max(boot_sila$val)) {
        median(boot_sila$val)
      } else {
        thr
      }

      res    <- sila(boot_sila, dt = 0.25, val0 = thr_adj, maxi = 200)
      resfit <- sila_estimate(res$tsila, boot_sila)

      # Recover original CONP_ID by stripping boot suffix
      sila_last <- resfit %>%
        mutate(CONP_ID = sub("_[0-9]+$", "", subid)) %>%
        group_by(subid) %>%
        slice_max(age, n = 1, with_ties = FALSE) %>%
        ungroup() %>%
        mutate(event    = as.integer(val >= valt0),
               # Events --> SILA-estimated crossing age; censored --> last observed visit age
               surv_age = if_else(val >= valt0 & !is.na(estaget0), estaget0, age),
               age_last = age)

      sila_first <- resfit %>%
        mutate(CONP_ID = sub("_[0-9]+$", "", subid)) %>%
        group_by(subid) %>%
        slice_min(age, n = 1, with_ties = FALSE) %>%
        ungroup() %>%
        select(subid, age_first = age)

      sila_last <- sila_last %>%
        left_join(sila_first, by = "subid")

      # Single-visit participants: fixed event status, appended to bootstrap data
      sv <- single_visit_long %>%
        filter(biomarker == bm) %>%
        mutate(subid = CONP_ID)

      # Combine multi + single visit
      cox_dat <- bind_rows(
        sila_last %>% select(CONP_ID, surv_age, age_last, age_first, event),
        sv        %>% select(CONP_ID, surv_age, age_last, age_first, event)
      ) %>%
        left_join(predictors_all, by = "CONP_ID") %>%
        left_join(amyloid_lookup,  by = "CONP_ID") %>%
        drop_na(event, surv_age, age_first, apoe_burden, fhx_burden, grs_burden)

      # Standard and left-truncated survival objects
      # surv_age = estaget0 for events, age_last for censored 
      surv_std <- Surv(cox_dat$surv_age, cox_dat$event)
      surv_lt  <- Surv(cox_dat$age_first, cox_dat$surv_age, cox_dat$event)

      # Cox models (standard + left-truncated) 
      cox_models <- list(
        m1_std = tryCatch(
          coxph(surv_std ~ risk_score_cogd,
                data = cox_dat %>% drop_na(risk_score_cogd), ties = "efron"),
          error = function(e) NULL),
        m2_std = tryCatch(
          coxph(surv_std ~ age_last + sex + educ + risk_score_libra2,
                data = cox_dat %>% drop_na(risk_score_libra2, sex, educ), ties = "efron"),
          error = function(e) NULL),
        m3_std = tryCatch(
          coxph(surv_std ~ apoe_burden + fhx_burden + grs_burden,
                data = cox_dat, ties = "efron"),
          error = function(e) NULL),
        m4_std = tryCatch(
          coxph(surv_std ~ age_last + sex + educ + apoe_burden + fhx_burden +
                            grs_burden + libra2_burden,
                data = cox_dat %>% drop_na(libra2_burden, sex, educ), ties = "efron"),
          error = function(e) NULL),
        m1_lt = tryCatch(
          coxph(surv_lt ~ risk_score_cogd,
                data = cox_dat %>% drop_na(risk_score_cogd), ties = "efron"),
          error = function(e) NULL),
        m2_lt = tryCatch(
          coxph(surv_lt ~ age_last + sex + educ + risk_score_libra2,
                data = cox_dat %>% drop_na(risk_score_libra2, sex, educ), ties = "efron"),
          error = function(e) NULL),
        m3_lt = tryCatch(
          coxph(surv_lt ~ apoe_burden + fhx_burden + grs_burden,
                data = cox_dat, ties = "efron"),
          error = function(e) NULL),
        m4_lt = tryCatch(
          coxph(surv_lt ~ age_last + sex + educ + apoe_burden + fhx_burden +
                           grs_burden + libra2_burden,
                data = cox_dat %>% drop_na(libra2_burden, sex, educ), ties = "efron"),
          error = function(e) NULL)
      )

      hrs <- purrr::imap_dfr(cox_models, function(m, nm) {
        if (is.null(m)) return(NULL)
        broom::tidy(m, exponentiate = TRUE, conf.int = FALSE) %>%
          mutate(model = nm, iter = iter, biomarker = bm)
      })

      # KM survival by amyloid status
      km_dat <- cox_dat %>% drop_na(amyloid_status)

      km_amyloid_tbl <- tryCatch({
        km <- survfit(Surv(surv_age, event) ~ amyloid_status, data = km_dat)
        s   <- summary(km, times = age_grid, extend = TRUE)
        tibble(
          iter   = iter, biomarker = bm,
          group  = sub("amyloid_status=", "", as.character(s$strata)),
          age    = s$time, survival = s$surv
        )
      }, error = function(e) NULL)

      # KM survival by APOE e4 carrier status
      km_apoe_tbl <- tryCatch({
        dat_apoe <- cox_dat %>%
          drop_na(apoe_burden) %>%
          mutate(apoe_group = factor(if_else(apoe_burden == 1L, "e4+", "e4-"),
                                     levels = c("e4-", "e4+")))
        km <- survfit(Surv(surv_age, event) ~ apoe_group, data = dat_apoe)
        s  <- summary(km, times = age_grid, extend = TRUE)
        tibble(
          iter = iter, biomarker = bm,
          group = sub("apoe_group=", "", as.character(s$strata)),
          age = s$time, survival = s$surv
        )
      }, error = function(e) NULL)

      # KM survival by CogDrisk risk score (0 / 1 / 2 / 3+) 
      km_risk_cogd_tbl <- tryCatch({
        dat_rc <- cox_dat %>%
          drop_na(risk_score_cogd) %>%
          mutate(risk_group = factor(
            if_else(risk_score_cogd >= 3L, "3+", as.character(risk_score_cogd)),
            levels = c("0", "1", "2", "3+")
          ))
        km <- survfit(Surv(surv_age, event) ~ risk_group, data = dat_rc)
        s  <- summary(km, times = age_grid, extend = TRUE)
        tibble(
          iter = iter, biomarker = bm,
          group = sub("risk_group=", "", as.character(s$strata)),
          age = s$time, survival = s$surv
        )
      }, error = function(e) NULL)

      # KM survival by LIBRA2 risk score (0 / 1 / 2 / 3+) 
      km_risk_libra2_tbl <- tryCatch({
        dat_rl <- cox_dat %>%
          drop_na(risk_score_libra2) %>%
          mutate(risk_group = factor(
            if_else(risk_score_libra2 >= 3L, "3+", as.character(risk_score_libra2)),
            levels = c("0", "1", "2", "3+")
          ))
        km <- survfit(Surv(surv_age, event) ~ risk_group, data = dat_rl)
        s  <- summary(km, times = age_grid, extend = TRUE)
        tibble(
          iter = iter, biomarker = bm,
          group = sub("risk_group=", "", as.character(s$strata)),
          age = s$time, survival = s$surv
        )
      }, error = function(e) NULL)

      list(hrs = hrs, km_amyloid = km_amyloid_tbl, km_apoe = km_apoe_tbl,
           km_risk_cogd = km_risk_cogd_tbl, km_risk_libra2 = km_risk_libra2_tbl)

    }, error = function(e) NULL)  # SILA convergence failure
  })

  bm_results_clean <- purrr::compact(bm_results)
  if (length(bm_results_clean) == 0) return(NULL)

  list(
    hrs            = dplyr::bind_rows(purrr::map(bm_results_clean, "hrs")),
    km_amyloid     = dplyr::bind_rows(purrr::map(bm_results_clean, "km_amyloid")),
    km_apoe        = dplyr::bind_rows(purrr::map(bm_results_clean, "km_apoe")),
    km_risk_cogd   = dplyr::bind_rows(purrr::map(bm_results_clean, "km_risk_cogd")),
    km_risk_libra2 = dplyr::bind_rows(purrr::map(bm_results_clean, "km_risk_libra2"))
  )
}
```

```{r run-bootstrap}
#| cache: false

N_BOOT     <- 10
BOOT_CACHE <- here("results", "Aim2", "bootstrap_cache.rds")

dir.create(here("results", "Aim2"), recursive = TRUE, showWarnings = FALSE)

if (file.exists(BOOT_CACHE)) {
  cat("Loading cached bootstrap results from:", BOOT_CACHE, "\n")
  boot_results <- readRDS(BOOT_CACHE)
} else {
  cat("Running", N_BOOT, "bootstrap iterations: \n")
  plan(multisession, workers = max(1L, parallel::detectCores() - 1L))

  boot_results <- future_map(
    seq_len(N_BOOT),
    run_bootstrap_iter,
    multi_visit_long = multi_visit_long,
    single_visit_long = single_visit_long,
    predictors_all   = predictors_all,
    amyloid_lookup   = amyloid_lookup,
    thresholds       = thresholds,
    quartile_cuts    = quartile_cuts,
    age_grid         = age_grid,
    .options = furrr_options(seed = TRUE),
    .progress = TRUE
  )

  plan(sequential)
  saveRDS(boot_results, BOOT_CACHE)
  cat("Bootstrap complete. Results saved to:", BOOT_CACHE, "\n")
}

# Check convergence rate
n_success <- sum(!sapply(boot_results, is.null))
cat("Successful iterations:", n_success, "/", N_BOOT,
    sprintf("(%.1f%%)\n", 100 * n_success / N_BOOT))
```

```{r aggregate-bootstrap}
# Flatten results from all successful iterations
boot_valid <- purrr::compact(boot_results)

boot_hrs_all        <- dplyr::bind_rows(purrr::map(boot_valid, "hrs"))
boot_km_amyloid     <- dplyr::bind_rows(purrr::map(boot_valid, "km_amyloid"))
boot_km_apoe        <- dplyr::bind_rows(purrr::map(boot_valid, "km_apoe"))
boot_km_risk_cogd   <- dplyr::bind_rows(purrr::map(boot_valid, "km_risk_cogd"))
boot_km_risk_libra2 <- dplyr::bind_rows(purrr::map(boot_valid, "km_risk_libra2"))

# Helper: summarise bootstrap KM draws to median + 95% CI band
summarise_km_boot <- function(df) {
  df %>%
    group_by(biomarker, group, age) %>%
    summarise(
      surv_median = median(survival,          na.rm = TRUE),
      surv_lo     = quantile(survival, 0.025, na.rm = TRUE),
      surv_hi     = quantile(survival, 0.975, na.rm = TRUE),
      n_iters     = sum(!is.na(survival)),
      .groups     = "drop"
    )
}

km_summary_amyloid     <- summarise_km_boot(boot_km_amyloid)
km_summary_apoe        <- summarise_km_boot(boot_km_apoe)
km_summary_risk_cogd   <- summarise_km_boot(boot_km_risk_cogd)
km_summary_risk_libra2 <- summarise_km_boot(boot_km_risk_libra2)

# ── Bootstrap HR summaries ────────────────────────────────────────────────────
hr_boot_summary <- boot_hrs_all %>%
  group_by(biomarker, model, term) %>%
  summarise(
    HR_median = median(estimate,             na.rm = TRUE),
    HR_lo     = quantile(estimate, 0.025,    na.rm = TRUE),
    HR_hi     = quantile(estimate, 0.975,    na.rm = TRUE),
    n_iters   = sum(!is.na(estimate)),
    .groups   = "drop"
  )

cat("Bootstrap amyloid KM rows:", nrow(km_summary_amyloid), "\n")
cat("Bootstrap APOE KM rows:",    nrow(km_summary_apoe), "\n")
cat("Bootstrap HR summary rows:", nrow(hr_boot_summary), "\n")
```


## Bootstrap Survival Curves by APOE Status

```{r km-apoe-bootstrap-plot}
#| fig-width: 12
#| fig-height: 5

plot_km_bootstrap_apoe <- function(km_summary_df, biomarker_label) {
  bm_key <- sub("p-tau", "ptau", tolower(biomarker_label))
  km_summary_df %>%
    filter(biomarker == bm_key, !is.na(group)) %>%
    ggplot(aes(x = age, y = surv_median, colour = group, fill = group)) +
    geom_step(linewidth = 0.9) +
    geom_ribbon(aes(ymin = surv_lo, ymax = surv_hi), alpha = 0.20, colour = NA) +
    scale_colour_manual(values = c("e4-" = "#2166AC", "e4+" = "#B2182B"),
                        name = "APOE status") +
    scale_fill_manual(values   = c("e4-" = "#2166AC", "e4+" = "#B2182B"),
                      name = "APOE status") +
    labs(
      title    = paste(biomarker_label, "\u2013 ptau Negativity by APOE \u03b54 Status"),
      subtitle = "Bootstrap median survival (95% CI from 10 SILA bootstrap iterations)",
      x        = "Age (years)",
      y        = "Probability of ptau Negativity"
    ) +
    coord_cartesian(ylim = c(0, 1)) +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom")
}

print(plot_km_bootstrap_apoe(km_summary_apoe, "p-tau181"))
print(plot_km_bootstrap_apoe(km_summary_apoe, "p-tau217"))
```

## Bootstrap Survival Curves by Risk Score

```{r km-risk-bootstrap-plot}
#| fig-width: 14
#| fig-height: 10

risk_palette <- c("0" = "#2166AC", "1" = "#74ADD1", "2" = "#F46D43", "3+" = "#B2182B")

plot_km_bootstrap_risk <- function(km_summary_df, biomarker_label, score_label) {
  bm_key <- sub("p-tau", "ptau", tolower(biomarker_label))
  km_summary_df %>%
    filter(biomarker == bm_key, !is.na(group)) %>%
    mutate(group = factor(group, levels = c("0", "1", "2", "3+"))) %>%
    ggplot(aes(x = age, y = surv_median, colour = group, fill = group)) +
    geom_step(linewidth = 0.9) +
    geom_ribbon(aes(ymin = surv_lo, ymax = surv_hi), alpha = 0.15, colour = NA) +
    scale_colour_manual(values = risk_palette, name = score_label) +
    scale_fill_manual(values   = risk_palette, name = score_label) +
    labs(
      title    = paste(biomarker_label, "\u2013 ptau Negativity by", score_label),
      subtitle = "Bootstrap median survival (95% CI from 10 SILA bootstrap iterations)",
      x        = "Age (years)",
      y        = "Probability of ptau Negativity"
    ) +
    coord_cartesian(ylim = c(0, 1)) +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom")
}

print(plot_km_bootstrap_risk(km_summary_risk_cogd,   "p-tau181",
                              "No. CogDrisk indicators (0\u20133+)"))
print(plot_km_bootstrap_risk(km_summary_risk_libra2,  "p-tau181",
                              "No. LIBRA2 indicators (0\u20133+)"))
print(plot_km_bootstrap_risk(km_summary_risk_cogd,   "p-tau217",
                              "No. CogDrisk indicators (0\u20133+)"))
print(plot_km_bootstrap_risk(km_summary_risk_libra2,  "p-tau217",
                              "No. LIBRA2 indicators (0\u20133+)"))
```


## Cox Model Results (Full Observed Data + Bootstrap CI)

```{r fit-cox-full}
# Fit M1–M4 (standard + left-truncated) on the full observed dataset.
# Point estimates and p-values come from these models; CIs come from bootstrap.
fit_cox_models_full <- function(data) {
  dat      <- data %>% drop_na(surv_age, age_first, event, sex, educ,
                                apoe_burden, fhx_burden, grs_burden)
  surv_std <- Surv(dat$surv_age, dat$event)
  surv_lt  <- Surv(dat$age_first, dat$surv_age, dat$event)

  list(
    m1_std = coxph(surv_std ~ risk_score_cogd,
                   data = dat %>% drop_na(risk_score_cogd), ties = "efron"),
    m2_std = coxph(surv_std ~ age_last + sex + educ + risk_score_libra2,
                   data = dat %>% drop_na(risk_score_libra2), ties = "efron"),
    m3_std = coxph(surv_std ~ apoe_burden + fhx_burden + grs_burden,
                   data = dat, ties = "efron"),
    m4_std = coxph(surv_std ~ age_last + sex + educ + apoe_burden + fhx_burden +
                              grs_burden + libra2_burden,
                   data = dat %>% drop_na(libra2_burden), ties = "efron"),
    m1_lt  = coxph(surv_lt ~ risk_score_cogd,
                   data = dat %>% drop_na(risk_score_cogd), ties = "efron"),
    m2_lt  = coxph(surv_lt ~ age_last + sex + educ + risk_score_libra2,
                   data = dat %>% drop_na(risk_score_libra2), ties = "efron"),
    m3_lt  = coxph(surv_lt ~ apoe_burden + fhx_burden + grs_burden,
                   data = dat, ties = "efron"),
    m4_lt  = coxph(surv_lt ~ age_last + sex + educ + apoe_burden + fhx_burden +
                              grs_burden + libra2_burden,
                   data = dat %>% drop_na(libra2_burden), ties = "efron")
  )
}

models_ptau181 <- fit_cox_models_full(cox_ptau181)
models_ptau217 <- fit_cox_models_full(cox_ptau217)
```

```{r extract-cox-results-full}
# Tidy a Cox model into a one-row-per-term tibble with observed HR + p-value.
# Bootstrap CIs are joined separately via add_boot_ci().
tidy_cox_full <- function(model, biomarker_label, model_label) {
  if (is.null(model)) return(NULL)
  broom::tidy(model, exponentiate = TRUE, conf.int = FALSE) %>%
    mutate(
      biomarker = biomarker_label,
      model     = model_label,
      p_value   = p.value,
      sig       = case_when(
        p.value < 0.001 ~ "***",
        p.value < 0.01  ~ "**",
        p.value < 0.05  ~ "*",
        TRUE            ~ ""
      ),
      HR = estimate
    ) %>%
    select(biomarker, model, term, HR, p_value, sig)
}

results_obs_ptau181 <- bind_rows(
  tidy_cox_full(models_ptau181$m1_std, "p-tau181", "M1: CogDrisk Composite"),
  tidy_cox_full(models_ptau181$m2_std, "p-tau181", "M2: LIBRA2 Composite"),
  tidy_cox_full(models_ptau181$m3_std, "p-tau181", "M3: CogDrisk Components"),
  tidy_cox_full(models_ptau181$m4_std, "p-tau181", "M4: LIBRA2 Components")
)

results_obs_ptau217 <- bind_rows(
  tidy_cox_full(models_ptau217$m1_std, "p-tau217", "M1: CogDrisk Composite"),
  tidy_cox_full(models_ptau217$m2_std, "p-tau217", "M2: LIBRA2 Composite"),
  tidy_cox_full(models_ptau217$m3_std, "p-tau217", "M3: CogDrisk Components"),
  tidy_cox_full(models_ptau217$m4_std, "p-tau217", "M4: LIBRA2 Components")
)

# Join bootstrap CIs onto observed results.
# Normalises biomarker names and model labels to match observed results.
add_boot_ci <- function(obs_results, boot_summary) {
  boot_ci <- boot_summary %>%
    filter(str_ends(model, "_std")) %>%
    mutate(
      biomarker = case_match(biomarker,
        "ptau181" ~ "p-tau181", "ptau217" ~ "p-tau217", .default = biomarker),
      model     = case_match(model,
        "m1_std" ~ "M1: CogDrisk Composite",
        "m2_std" ~ "M2: LIBRA2 Composite",
        "m3_std" ~ "M3: CogDrisk Components",
        "m4_std" ~ "M4: LIBRA2 Components",
        .default = model)
    ) %>%
    select(biomarker, model, term,
           boot_lo = HR_lo, boot_hi = HR_hi, n_boot_iters = n_iters)

  obs_results %>%
    left_join(boot_ci, by = c("biomarker", "model", "term"))
}

all_results <- bind_rows(
  add_boot_ci(results_obs_ptau181, hr_boot_summary),
  add_boot_ci(results_obs_ptau217, hr_boot_summary)
)
```

### Results: p-tau181

```{r table-ptau181}
# Format observed HR with bootstrap CI band and significance stars.
fmt_boot <- function(hr_obs, lo, hi, p) {
  sig <- dplyr::case_when(
    p < 0.001 ~ "***", p < 0.01 ~ "**", p < 0.05 ~ "*", TRUE ~ ""
  )
  if (!is.na(lo) && !is.na(hi)) {
    sprintf("%.2f (%.2f\u2013%.2f)%s", hr_obs, lo, hi, sig)
  } else {
    sprintf("%.2f (no boot CI)%s", hr_obs, sig)
  }
}

make_results_table <- function(results_df, biomarker_label) {
  results_df %>%
    filter(!term %in% c("age_last", "sex", "educ")) %>%
    rowwise() %>%
    mutate(
      hr_fmt = fmt_boot(HR, boot_lo, boot_hi, p_value),
      p_fmt  = format.pval(p_value, digits = 3, eps = 0.001)
    ) %>%
    ungroup() %>%
    select(Model = model, Term = term, `HR (95% boot CI)` = hr_fmt, p = p_fmt) %>%
    gt() %>%
    tab_header(
      title    = paste("Cox Models:", biomarker_label, "Time to Threshold (Bootstrap CI)"),
      subtitle = "Observed HR with 95% bootstrap CI from 10 SILA iterations"
    ) %>%
    tab_style(
      style     = cell_text(weight = "bold"),
      locations = cells_column_labels()
    ) %>%
    tab_footnote(
      footnote  = paste(
        "HR > 1 indicates faster tau positivity onset (earlier age).",
        "Bootstrap CI incorporates SILA fitting uncertainty (10 iterations).",
        "Events use SILA-estimated crossing age (estaget0); censored at last observed visit.",
        "* p<0.05; ** p<0.01; *** p<0.001"
      ),
      locations = cells_title()
    )
}

make_results_table(all_results %>% filter(biomarker == "p-tau181"), "p-tau181")
```

### Results: p-tau217

```{r table-ptau217}
make_results_table(all_results %>% filter(biomarker == "p-tau217"), "p-tau217")
```
