---
title: "Aim 2: Predicting Tau PET Positivity from Plasma Biomarkers"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
execute:
  warning: false
  message: false
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = here::here())
```


```{r loading}
library(pROC)

# Load configuration
source('workflow/scripts/config.R')
```

## Overview

This analysis predicts amyloid and tau PET positivity from plasma biomarkers (ptau181, ptau217) using logistic regression. We calculate:

- ROC curves and AUC
- Optimal cutoffs using Youden index (sensitivity + specificity - 1)
- Sensitivity and specificity at optimal cutoffs

**Tau positivity thresholds: (PREVENT-AD cohort-specific)**

- Tau: SUVR ≥ 1.28

## Load and Prepare Data

### Load Pre-Processed Biomarker Data

Using biomarker data already filtered by inclusion criteria (02_clean_PREVENT-AD.R):

- CONP_ID with GWAS data
- CONP_ID with at least 2 biomarker visits

```{r load-biomarkers-processed}
# Load pre-processed biomarker data (already filtered by inclusion criteria)
biomarkers <- readRDS(here("data", "intermediate", "PREVENTAD_biomarkers.rds"))

# Load demographics for covariates (Sex, BMI)
# demographics <- readRDS(here("data", "intermediate", "PREVENTAD_demographics.rds"))
crs_factors <- readRDS(here("data", "intermediate", "PREVENTAD_crsfactors_imputed.rds")) %>%
  select(CONP_ID, Sex, Age, Weight, Height) %>%
  mutate(BMI = Weight / (Height / 100)^2) %>%
  select(CONP_ID, BMI, Sex, Age)

# Get list of participants meeting inclusion criteria
criteria_met <- biomarkers %>%
  distinct(CONP_ID)

cat("=== INCLUSION CRITERIA (from pre-processed data) ===\n")
cat("Participants meeting criteria:", nrow(criteria_met), "\n")
cat("Total biomarker observations:", nrow(biomarkers), "\n")
```

### Load PET Data

```{r load-pet}
# Load raw PET data
pet_ftp <- read_csv(
  file.path(PREVENTAD_DATA_PATH,
            "PET_FTP_SUVR_ref-infCerebellarGray_Registered_PREVENTAD_2024-08-01.csv"),
  show_col_types = FALSE
)
pet_info <- read_csv(
  file.path(PREVENTAD_DATA_PATH,
            "PET_info_participants_Registered_PREVENTAD_2024-08-01.csv"),
  show_col_types = FALSE
)

# Process PET data - apply inclusion criteria
pet_tau <- pet_ftp %>%
  filter(CONP_ID %in% criteria_met$CONP_ID) %>%
  select(CONP_ID, SUVR=`meta-roi_SUVR`) 

# Get PET dates
pet_tau_dates <- pet_info %>%
  filter(PET_tracer == "FTP") %>%
  select(CONP_ID, Date_PET_visit, Age_PET_visit) %>%
  distinct() %>%
  rename(
    pet_date = Date_PET_visit,
    age_at_pet = Age_PET_visit
  )

pet_tau <- pet_tau %>%
  left_join(pet_tau_dates, by = c("CONP_ID"))

# PREVENT-AD positivity thresholds
SUVR_THRESHOLD <- 1.28

pet_tau <- pet_tau %>%
  mutate(
    tau_positive = as.integer(SUVR >= SUVR_THRESHOLD),
    tau_status = factor(
      ifelse(SUVR >= SUVR_THRESHOLD, "T+", "T-"),
      levels = c("T-", "T+")
    )
  )

cat("PET participants:", n_distinct(pet_tau$CONP_ID), "\n")
cat("Tau positive:", sum(pet_tau$tau_positive),
    "(", round(mean(pet_tau$tau_positive) * 100, 1), "%)\n")
```

### Match PET with Closest Biomarker Timepoint
```{r prepare-biomarkers}
# Extract ptau181 data from pre-processed biomarkers
biomarkers_ptau181 <- biomarkers %>%
  filter(!is.na(ptau181)) %>%
  select(CONP_ID, biomarker_visit = Visit_label, biomarker_date = Date_taken, ptau181)

# Extract ptau217 data from pre-processed biomarkers
biomarkers_ptau217 <- biomarkers %>%
  filter(!is.na(ptau217)) %>%
  select(CONP_ID, biomarker_visit = Visit_label, biomarker_date = Date_taken, ptau217)

cat("Participants with ptau181:", n_distinct(biomarkers_ptau181$CONP_ID), "\n")
cat("Participants with ptau217:", n_distinct(biomarkers_ptau217$CONP_ID), "\n")

# Function to find closest biomarker measurement to PET scan
find_closest_biomarker <- function(pet_data, biomarker_data, biomarker_col) {

  # Convert dates to Date objects
  pet_data <- pet_data %>%
    mutate(pet_date_parsed = ymd(paste0(pet_date, "-01")))

  biomarker_data <- biomarker_data %>%
    mutate(biomarker_date_parsed = ymd(paste0(biomarker_date, "-01")))

  # For each PET scan, find closest biomarker

  matched_data <- pet_data %>%
    left_join(biomarker_data, by = "CONP_ID", relationship = "many-to-many") %>%
    mutate(
      date_diff_months = abs(interval(biomarker_date_parsed, pet_date_parsed) / months(1))
    ) %>%
    group_by(CONP_ID) %>%
    slice_min(date_diff_months, n = 1, with_ties = FALSE) %>%
    ungroup()

  return(matched_data)
}

# Match ptau181
pet_tau_ptau181 <- find_closest_biomarker(pet_tau, biomarkers_ptau181, "ptau181") %>%
  filter(!is.na(ptau181)) %>%
  left_join(crs_factors, by = "CONP_ID")

# Match ptau217
pet_tau_ptau217 <- find_closest_biomarker(pet_tau, biomarkers_ptau217, "ptau217") %>%
  filter(!is.na(ptau217)) %>%
  left_join(crs_factors, by = "CONP_ID")
```

## Descriptive Statistics

```{r descriptive-tau-ptau181}
# Summary table for ptau181 analysis cohort
pet_tau_ptau181 %>%
  select(tau_status, SUVR, ptau181, age_at_pet) %>%
  tbl_summary(
    by = tau_status,
    statistic = list(
      all_continuous() ~ "{mean} ({sd})"
    ),
    label = list(
      SUVR ~ "SUVR",
      ptau181 ~ "p-tau181 (pg/mL)",
      age_at_pet ~ "Age at PET"
    )
  ) %>%
  add_overall() %>%
  add_p() %>%
  as_gt() %>%
  tab_header(title = "Cohort Characteristics by Tau Status (ptau181 analysis)")
```

```{r descriptive-tau-ptau217}
# Summary table for ptau217 analysis cohort
pet_tau_ptau217 %>%
  select(tau_status, SUVR, ptau217, age_at_pet) %>%
  tbl_summary(
    by = tau_status,
    statistic = list(
      all_continuous() ~ "{mean} ({sd})"
    ),
    label = list(
      SUVR ~ "SUVR",
      ptau217 ~ "p-tau217 (pg/mL)",
      age_at_pet ~ "Age at PET"
    )
  ) %>%
  add_overall() %>%
  add_p() %>%
  as_gt() %>%
  tab_header(title = "Cohort Characteristics by Tau Status (ptau217 analysis)")
```

## Biomarker Distributions

```{r distributions-tau}
#| fig-width: 10
#| fig-height: 4

# ptau181 distribution
p1 <- pet_tau_ptau181 %>%
  ggplot(aes(x = ptau181, fill = tau_status)) +
  geom_histogram(bins = 30, alpha = 0.7, position = "identity") +
  scale_fill_manual(values = c("T-" = "#2166AC", "T+" = "#B2182B")) +
  labs(
    title = "p-tau181 Distribution by Tau Status",
    x = "p-tau181 (pg/mL)",
    y = "Count",
    fill = "Tau Status"
  ) +
  theme_minimal()

# ptau217 distribution
p2 <- pet_tau_ptau217 %>%
  ggplot(aes(x = ptau217, fill = tau_status)) +
  geom_histogram(bins = 30, alpha = 0.7, position = "identity") +
  scale_fill_manual(values = c("T-" = "#2166AC", "T+" = "#B2182B")) +
  labs(
    title = "p-tau217 Distribution by Tau Status",
    x = "p-tau217 (pg/mL)",
    y = "Count",
    fill = "Tau Status"
  ) +
  theme_minimal()

library(patchwork)
p1 + p2
```

## Logistic Regression Models

### p-tau181 Models

```{r model-ptau181}
# Model 1: Biomarker only
model_ptau181_bio <- glm(
  tau_positive ~ ptau181,
  data = pet_tau_ptau181,
  family = binomial(link = "logit")
)

# Model results
cat("=== Model Results ===\n\n")
summary(model_ptau181_bio)
cat("\nOdds Ratios (95% CI):\n")
exp(cbind(OR = coef(model_ptau181_bio), confint(model_ptau181_bio)))
```

### p-tau217 Models

```{r model-ptau217}
# Model 1: Biomarker only
model_ptau217_bio <- glm(
  tau_positive ~ ptau217,
  data = pet_tau_ptau217,
  family = binomial(link = "logit")
)

# Model results
cat("=== Model 1Results ===\n\n")
summary(model_ptau217_bio)
cat("\nOdds Ratios (95% CI):\n")
exp(cbind(OR = coef(model_ptau217_bio), confint(model_ptau217_bio)))
```

## ROC Analysis and Optimal Cutoffs

### p-tau181 ROC

```{r roc-ptau181}
#| fig-width: 6
#| fig-height: 6

# ROC curve - ptau181
roc_ptau181 <- roc(
  pet_tau_ptau181$tau_positive,
  pet_tau_ptau181$ptau181,
  levels = c(0, 1),
  direction = "<"
)

# AUC with confidence interval
auc_ptau181 <- auc(roc_ptau181)
ci_auc_ptau181 <- ci.auc(roc_ptau181)

cat("=== p-tau181 ROC Analysis ===\n")
cat("AUC:", round(auc_ptau181, 3), "\n")
cat("95% CI:", round(ci_auc_ptau181[1], 3), "-", round(ci_auc_ptau181[3], 3), "\n")

# Youden index optimal cutoff
coords_ptau181 <- coords(roc_ptau181, "best", best.method = "youden", ret = "all")
cat("\n=== Optimal Cutoff (Youden Index) ===\n")
cat("Cutoff:", round(coords_ptau181$threshold, 2), "pg/mL\n")
cat("Sensitivity:", round(coords_ptau181$sensitivity * 100, 1), "%\n")
cat("Specificity:", round(coords_ptau181$specificity * 100, 1), "%\n")
cat("PPV:", round(coords_ptau181$ppv * 100, 1), "%\n")
cat("NPV:", round(coords_ptau181$npv * 100, 1), "%\n")
cat("Youden Index:", round(coords_ptau181$sensitivity + coords_ptau181$specificity - 1, 3), "\n")

# Plot ROC
plot(roc_ptau181,
     main = "ROC Curve: p-tau181 Predicting Tau Positivity",
     print.auc = TRUE,
     print.auc.x = 0.4,
     print.auc.y = 0.2,
     col = "#2166AC",
     lwd = 2)

# Add optimal cutoff point
points(coords_ptau181$specificity, coords_ptau181$sensitivity,
       pch = 19, col = "#B2182B", cex = 1.5)
text(coords_ptau181$specificity - 0.1, coords_ptau181$sensitivity + 0.05,
     paste0("Cutoff: ", round(coords_ptau181$threshold, 2), " pg/mL"),
     col = "#B2182B", cex = 0.8)
```

### p-tau217 ROC

```{r roc-ptau217}
#| fig-width: 6
#| fig-height: 6

# ROC curve - ptau217
roc_ptau217 <- roc(
  pet_tau_ptau217$tau_positive,
  pet_tau_ptau217$ptau217,
  levels = c(0, 1),
  direction = "<"
)

# AUC with confidence interval
auc_ptau217 <- auc(roc_ptau217)
ci_auc_ptau217 <- ci.auc(roc_ptau217)

cat("=== p-tau217 ROC Analysis ===\n")
cat("AUC:", round(auc_ptau217, 3), "\n")
cat("95% CI:", round(ci_auc_ptau217[1], 3), "-", round(ci_auc_ptau217[3], 3), "\n")

# Youden index optimal cutoff
coords_ptau217 <- coords(roc_ptau217, "best", best.method = "youden", ret = "all")
cat("\n=== Optimal Cutoff (Youden Index) ===\n")
cat("Cutoff:", round(coords_ptau217$threshold, 2), "pg/mL\n")
cat("Sensitivity:", round(coords_ptau217$sensitivity * 100, 1), "%\n")
cat("Specificity:", round(coords_ptau217$specificity * 100, 1), "%\n")
cat("PPV:", round(coords_ptau217$ppv * 100, 1), "%\n")
cat("NPV:", round(coords_ptau217$npv * 100, 1), "%\n")
cat("Youden Index:", round(coords_ptau217$sensitivity + coords_ptau217$specificity - 1, 3), "\n")

# Plot ROC
plot(roc_ptau217,
     main = "ROC Curve: p-tau217 Predicting Amyloid Positivity",
     print.auc = TRUE,
     print.auc.x = 0.4,
     print.auc.y = 0.2,
     col = "#2166AC",
     lwd = 2)

# Add optimal cutoff point
points(coords_ptau217$specificity, coords_ptau217$sensitivity,
       pch = 19, col = "#B2182B", cex = 1.5)
text(coords_ptau217$specificity + 0.1, coords_ptau217$sensitivity - 0.05,
     paste0("Cutoff: ", round(coords_ptau217$threshold, 2), " pg/mL"),
     col = "#B2182B", cex = 0.8)
```


## Summary Results

```{r summary-table}

# Create summary table with biomarker-only models
results_summary <- tibble(
  Biomarker = c("p-tau181", "p-tau217"),
  N = c(nrow(pet_tau_ptau181), nrow(pet_tau_ptau217)),
  `N Tau+` = c(sum(pet_tau_ptau181$tau_positive), sum(pet_tau_ptau181$tau_positive)),
  `AUC (95% CI)` = c(
    paste0(round(auc(roc_ptau181), 3), " (", round(ci_auc_ptau181[1], 3), "-", round(ci_auc_ptau181[3], 3), ")"),
    paste0(round(auc(roc_ptau217), 3), " (", round(ci_auc_ptau217[1], 3), "-", round(ci_auc_ptau217[3], 3), ")")
  ),
  `Cutoff (pg/mL)` = c(
    round(coords_ptau181$threshold, 2),
    round(coords_ptau217$threshold, 2)
  ),
  `Sensitivity` = c(
    paste0(round(coords_ptau181$sensitivity * 100, 1), "%"),
    paste0(round(coords_ptau217$sensitivity * 100, 1), "%")
  ),
  `Specificity` = c(
    paste0(round(coords_ptau181$specificity * 100, 1), "%"),
    paste0(round(coords_ptau217$specificity * 100, 1), "%")
  ),
  `PPV` = c(
    paste0(round(coords_ptau181$ppv * 100, 1), "%"),
    paste0(round(coords_ptau217$ppv * 100, 1), "%")
  ),
  `NPV` = c(
    paste0(round(coords_ptau181$npv * 100, 1), "%"),
    paste0(round(coords_ptau217$npv * 100, 1), "%")
  ),
  `Youden Index` = c(
    round(coords_ptau181$sensitivity + coords_ptau181$specificity - 1, 3),
    round(coords_ptau217$sensitivity + coords_ptau217$specificity - 1, 3)
  )
)

results_summary %>%
  gt() %>%
  tab_header(
    title = "Summary: Plasma Biomarkers Predicting Tau PET Positivity",
    subtitle = paste0("Tau positivity defined as SUVR ≥ ", SUVR_THRESHOLD, " (PREVENT-AD specific)")
  )
```

```{r save-results}

# Save primary results
write_csv(results_summary,
          file.path(AIM1_OUTPUT_PATH$tables, "PREVENTAD_tau_prediction_primary.csv"))

cat("\nResults saved to:\n")
cat("- ", file.path(AIM1_OUTPUT_PATH$tables, "PREVENTAD_tau_prediction_primary.csv"), "\n")
```

