---
title: "Aim 2: SILA Analysis"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
execute:
  warning: false
  message: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = here::here())
```

## Overview

This analysis uses SILA to estimate time to amyloid positivity threshold from longitudinal plasma biomarker data.

**Biomarkers analyzed:**

- p-tau181
- p-tau217

**Amyloid positivity thresholds:**

- Centiloid ≥ 22.32 (PREVENT-AD cohort-specific)

**Biomarker cutoffs:** Derived from ROC analysis (Youden index optimal cutoffs) in `PREVENT-AD_PET.qmd`
- p-tau181: 6.42 pg/mL
- p-tau217: 3.06 pg/mL

```{r packages}
library(silaR)
library(patchwork)
library(pROC)

source('workflow/scripts/config.R')
```

## Load and Prepare Data

```{r load-data}

# Load pre-processed longitudinal biomarker data
biomarkers_long <- readRDS(here("data", "intermediate", "PREVENTAD_biomarkers.rds"))

# Get list of participants meeting inclusion criteria
criteria_met <- biomarkers_long %>%
  distinct(CONP_ID)

cat("=== INCLUSION CRITERIA (from pre-processed data) ===\n")
cat("Participants meeting criteria:", nrow(criteria_met), "\n")
cat("\nLongitudinal data summary:\n")
cat("Total observations:", nrow(biomarkers_long), "\n")
cat("Unique participants:", n_distinct(biomarkers_long$CONP_ID), "\n")
cat("Mean visits per participant:", round(nrow(biomarkers_long) / n_distinct(biomarkers_long$CONP_ID), 1), "\n")
```


```{r load-pet-derive-cutoffs}

# Load PET data
pet_nav <- read_csv(
  file.path(PREVENTAD_DATA_PATH,
            "PET_NAV_SUVR_ref-wholeCerebellum_Registered_PREVENTAD_2025-03-21.csv"),
  show_col_types = FALSE
)

pet_info <- read_csv(
  file.path(PREVENTAD_DATA_PATH,
            "PET_info_participants_Registered_PREVENTAD_2024-08-01.csv"),
  show_col_types = FALSE
)

pet_amyloid <- pet_nav %>%
  filter(PET_tracer == "NAV") %>%
  filter(CONP_ID %in% criteria_met$CONP_ID) %>%
  select(CONP_ID, session, amyloid_index_SUVR, WhlCbl_centiloid) %>%
  rename(
    pet_session = session,
    amyloid_suvr = amyloid_index_SUVR,
    centiloid = WhlCbl_centiloid
  )

# Get PET dates
pet_dates <- pet_info %>%
  filter(PET_tracer == "NAV") %>%
  select(CONP_ID, PET_session, Date_PET_visit, Age_PET_visit) %>%
  distinct() %>%
  rename(pet_session = PET_session, pet_date = Date_PET_visit, age_at_pet = Age_PET_visit)

pet_amyloid <- pet_amyloid %>%
  left_join(pet_dates, by = c("CONP_ID", "pet_session"))

cat("\nPET data summary:\n")
cat("Participants with PET:", n_distinct(pet_amyloid$CONP_ID), "\n")
cat("Centiloid range:", round(min(pet_amyloid$centiloid, na.rm = TRUE), 1),
    "to", round(max(pet_amyloid$centiloid, na.rm = TRUE), 1), "\n")
```


```{r derive-cutoffs}
# Amyloid positivity thresholds (Centiloid)
THRESHOLD_PRIMARY <- 22.32  

# Function to find closest biomarker measurement to PET scan
find_closest_biomarker <- function(pet_data, biomarker_data, biomarker_col) {

  # Convert dates to Date objects
  pet_data <- pet_data %>%
    mutate(pet_date_parsed = ymd(paste0(pet_date, "-01")))

  biomarker_data <- biomarker_data %>%
    mutate(biomarker_date_parsed = ymd(paste0(Date_taken, "-01")))

  # For each PET scan, find closest biomarker
  matched_data <- pet_data %>%
    left_join(biomarker_data, by = "CONP_ID", relationship = "many-to-many") %>%
    mutate(
      date_diff_months = abs(interval(biomarker_date_parsed, pet_date_parsed) / months(1))
    ) %>%
    group_by(CONP_ID, pet_session) %>%
    slice_min(date_diff_months, n = 1, with_ties = FALSE) %>%
    ungroup()

  return(matched_data)
}

# Prepare biomarker data for matching 
biomarkers_ptau181 <- biomarkers_long %>%
  filter(!is.na(ptau181)) %>%
  filter(ptau181 > 0) 

biomarkers_ptau217 <- biomarkers_long %>%
  filter(!is.na(ptau217)) %>%
  filter(ptau217 > 0) 

# Match PET with biomarkers
pet_ptau181 <- find_closest_biomarker(pet_amyloid, biomarkers_ptau181, "ptau181") %>%
  filter(!is.na(ptau181))

pet_ptau217 <- find_closest_biomarker(pet_amyloid, biomarkers_ptau217, "ptau217") %>%
  filter(!is.na(ptau217))

# Derive optimal cutoffs using ROC/Youden index for PRIMARY threshold (log-transformed)
pet_ptau181_primary <- pet_ptau181 %>%
  mutate(amyloid_positive = as.integer(centiloid >= THRESHOLD_PRIMARY))

pet_ptau217_primary <- pet_ptau217 %>%
  mutate(amyloid_positive = as.integer(centiloid >= THRESHOLD_PRIMARY))

roc_ptau181_primary <- roc(pet_ptau181_primary$amyloid_positive, pet_ptau181_primary$ptau181,
                           levels = c(0, 1), direction = "<")
roc_ptau217_primary <- roc(pet_ptau217_primary$amyloid_positive, pet_ptau217_primary$ptau217,
                           levels = c(0, 1), direction = "<")

cutoff_ptau181_primary <- coords(roc_ptau181_primary, "best", best.method = "youden", ret = "threshold")$threshold
cutoff_ptau217_primary <- coords(roc_ptau217_primary, "best", best.method = "youden", ret = "threshold")$threshold


cat("\n=== OPTIMAL BIOMARKER CUTOFFS (Youden Index) ===\n")
cat("\nPrimary threshold (CL >=", THRESHOLD_PRIMARY, "):\n")
cat("  p-tau181:", round(cutoff_ptau181_primary, 2), "pg/mL\n")
cat("  p-tau217:", round(cutoff_ptau217_primary, 2), "pg/mL\n")

# Create amyloid status lookup for plotting
amyloid_status_lookup <- pet_amyloid %>%
  select(CONP_ID, centiloid) %>%
  distinct() %>%
  group_by(CONP_ID) %>%
  slice_max(centiloid, n = 1) %>%  # Use max centiloid if multiple PET scans

  ungroup() %>%
  mutate(
    amyloid_status = factor(
      ifelse(centiloid >= THRESHOLD_PRIMARY, "A+", "A-"),
      levels = c("A-", "A+")
    )
  ) %>%
  select(CONP_ID, amyloid_status)

cat("\nAmyloid status for plotting:\n")
cat("  A+:", sum(amyloid_status_lookup$amyloid_status == "A+"), "\n")
cat("  A-:", sum(amyloid_status_lookup$amyloid_status == "A-"), "\n")
```

## Longitudinal Biomarker Trajectories

```{r biomarker-trajectories}
#| fig-width: 12
#| fig-height: 10

# Join amyloid status to longitudinal biomarker data
biomarkers_with_status <- biomarkers_long %>%
  left_join(amyloid_status_lookup, by = "CONP_ID") %>%
  mutate(amyloid_status = ifelse(is.na(amyloid_status), "No PET", as.character(amyloid_status)),
         amyloid_status = factor(amyloid_status, levels = c("A-", "A+", "No PET")))

# Combine BL00 and EL00 into single baseline category and create proper visit ordering
biomarkers_with_status <- biomarkers_with_status %>%
  mutate(
    Visit_combined = case_when(
      Visit_label %in% c("BL00", "EL00") ~ "BL/EL00",
      TRUE ~ Visit_label
    ),
    # Create numeric ordering for visits
    Visit_order = case_when(
      Visit_combined == "BL/EL00" ~ 0,
      grepl("^FU", Visit_combined) ~ as.numeric(gsub("FU", "", Visit_combined)),
      TRUE ~ NA_real_
    ),
    Visit_combined = factor(Visit_combined,
                            levels = c("BL/EL00", paste0("FU", sprintf("%02d", c(3, 6, 9, 12, 15, 18, 21, 24, 36, 48, 60, 72, 84, 96, 108)))))
  )

# p-tau181 boxplots by visit and amyloid status
p_box_181 <- biomarkers_with_status %>%
  filter(!is.na(ptau181), amyloid_status != "No PET") %>%
  ggplot(aes(x = Visit_combined, y = ptau181, fill = amyloid_status)) +
  geom_boxplot(outlier.shape = 21, outlier.size = 1.5, position = position_dodge(width = 0.8)) +
  scale_fill_manual(
    values = c("A-" = "#2166AC", "A+" = "#B2182B"),
    name = "Amyloid Status"
  ) +
  labs(
    title = "p-tau181 by Visit and Amyloid Status",
    x = "Visit",
    y = "p-tau181 (pg/mL)"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1))

# p-tau217 boxplots by visit and amyloid status
p_box_217 <- biomarkers_with_status %>%
  filter(!is.na(ptau217), amyloid_status != "No PET") %>%
  ggplot(aes(x = Visit_combined, y = ptau217, fill = amyloid_status)) +
  geom_boxplot(outlier.shape = 21, outlier.size = 1.5, position = position_dodge(width = 0.8)) +
  scale_fill_manual(
    values = c("A-" = "#2166AC", "A+" = "#B2182B"),
    name = "Amyloid Status"
  ) +
  labs(
    title = "p-tau217 by Visit and Amyloid Status",
    x = "Visit",
    y = "p-tau217 (pg/mL)"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1))

# Combine plots
p_box_181 / p_box_217
```

## SILA Analysis Functions

```{r sila-functions}

# Function to run SILA for a single biomarker
run_sila_biomarker <- function(data, biomarker_col, threshold_val, biomarker_name) {

  cat("\n--- Running SILA for", biomarker_name, "---\n")
  cat("Threshold:", threshold_val, "\n")

  # Prepare data in SILA format
  sila_data <- data %>%
    filter(!is.na(.data[[biomarker_col]])) %>%
    filter(!is.na(age)) %>%
    filter(is.finite(.data[[biomarker_col]])) %>%
    filter(is.finite(age)) %>%
    select(subid = CONP_ID, age, val = all_of(biomarker_col)) %>%
    arrange(subid, age)

  # Remove extreme outliers (beyond 3 SD from mean, or use IQR method)
  q1 <- quantile(sila_data$val, 0.25, na.rm = TRUE)
  q3 <- quantile(sila_data$val, 0.75, na.rm = TRUE)
  iqr <- q3 - q1
  lower_bound <- q1 - 3 * iqr
  upper_bound <- q3 + 3 * iqr

  n_before <- nrow(sila_data)
  sila_data <- sila_data %>%
    filter(val >= lower_bound & val <= upper_bound)
  n_after <- nrow(sila_data)

  if (n_before != n_after) {
    cat("Removed", n_before - n_after, "extreme outliers\n")
  }

  # Check minimum requirements
  n_subjects <- n_distinct(sila_data$subid)
  n_obs <- nrow(sila_data)

  cat("N subjects:", n_subjects, "\n")
  cat("N observations:", n_obs, "\n")
  cat("Value range:", round(min(sila_data$val), 2), "-", round(max(sila_data$val), 2), "\n")

  # Check for any remaining issues
  if (any(!is.finite(sila_data$val)) || any(!is.finite(sila_data$age))) {
    cat("ERROR: Non-finite values detected after cleaning\n")
    return(NULL)
  }

  if (n_subjects < 10) {
    warning(paste("Insufficient subjects for", biomarker_name, ":", n_subjects))
    return(NULL)
  }

  # Ensure threshold is within data range
  if (threshold_val < min(sila_data$val) || threshold_val > max(sila_data$val)) {
    cat("WARNING: Threshold", threshold_val, "is outside data range\n")
    cat("Adjusting to median value:", round(median(sila_data$val), 2), "\n")
    threshold_val <- median(sila_data$val)
  }

  # Run SILA
  tryCatch({
    cat("Running sila()...\n")
    res <- sila(sila_data, dt = 0.25, val0 = threshold_val, maxi = 200)
    cat("Running sila_estimate()...\n")
    resfit <- sila_estimate(res$tsila, sila_data)
    cat("SILA completed successfully\n")

    list(
      biomarker = biomarker_name,
      threshold = threshold_val,
      sila_result = res,
      sila_fit = resfit,
      n_subjects = n_subjects,
      n_obs = n_obs
    )
  }, error = function(e) {
    cat("ERROR:", e$message, "\n")
    warning(paste("SILA failed for", biomarker_name, ":", e$message))
    return(NULL)
  })
}

# Function to create SILA plots
plot_sila_results <- function(sila_output, title_suffix = "", amyloid_lookup = NULL) {

  if (is.null(sila_output)) return(NULL)

  res <- sila_output$sila_result
  resfit <- sila_output$sila_fit
  biomarker <- sila_output$biomarker
  threshold <- sila_output$threshold

  # Join amyloid status if provided
  if (!is.null(amyloid_lookup)) {
    resfit <- resfit %>%
      left_join(amyloid_lookup, by = c("subid" = "CONP_ID")) %>%
      mutate(amyloid_status = ifelse(is.na(amyloid_status), "Unknown", as.character(amyloid_status)))
  }

  # DRS curve
  p1 <- ggplot(res$tdrs, aes(x = val, y = rate)) +
    geom_ribbon(aes(ymin = rate - ci, ymax = rate + ci), alpha = 0.3, fill = "#2166AC") +
    geom_line(color = "#2166AC", linewidth = 1) +
    geom_vline(xintercept = threshold, linetype = "dashed", color = "red") +
    labs(
      title = paste(biomarker, "- Discrete Rate Sampling"),
      subtitle = title_suffix,
      x = biomarker,
      y = expression(Delta * " per Year")
    ) +
    theme_minimal()

  # SILA modeled curve
  p2 <- ggplot(res$tsila, aes(x = adtime, y = val)) +
    geom_line(color = "#2166AC", linewidth = 1) +
    geom_hline(yintercept = threshold, linetype = "dashed", color = "red") +
    labs(
      title = paste(biomarker, "- SILA Modeled Curve"),
      subtitle = title_suffix,
      x = "Time from Threshold (years)",
      y = biomarker
    ) +
    theme_minimal()

  # Individual trajectories aligned to threshold - colored by amyloid status
  if (!is.null(amyloid_lookup) && "amyloid_status" %in% names(resfit)) {
    p3 <- ggplot(resfit, aes(x = estdtt0, y = val, group = subid, color = amyloid_status)) +
      geom_line(alpha = 0.4) +
      geom_point(alpha = 0.5, size = 0.8) +
      geom_line(data = res$tsila, aes(x = adtime, y = val, group = NULL, color = NULL),
                color = "black", linewidth = 1.2) +
      geom_hline(yintercept = threshold, linetype = "dashed", color = "red") +
      scale_color_manual(
        values = c("A-" = "#2166AC", "A+" = "#B2182B", "Unknown" = "gray50"),
        name = "Amyloid Status"
      ) +
      labs(
        title = paste(biomarker, "- Aligned to Threshold"),
        subtitle = paste0(title_suffix, " (n=", sila_output$n_subjects, ")"),
        x = "Estimated Time to Threshold (years)",
        y = biomarker
      ) +
      theme_minimal() +
      theme(legend.position = "bottom")
  } else {
    p3 <- ggplot(resfit, aes(x = estdtt0, y = val, group = subid)) +
      geom_line(alpha = 0.3, color = "gray50") +
      geom_point(alpha = 0.3, size = 0.5) +
      geom_line(data = res$tsila, aes(x = adtime, y = val, group = NULL),
                color = "#B2182B", linewidth = 1.2) +
      geom_hline(yintercept = threshold, linetype = "dashed", color = "red") +
      labs(
        title = paste(biomarker, "- Aligned to Threshold"),
        subtitle = paste0(title_suffix, " (n=", sila_output$n_subjects, ")"),
        x = "Estimated Time to Threshold (years)",
        y = biomarker
      ) +
      theme_minimal()
  }

  list(drs = p1, sila = p2, aligned = p3)
}

# Function to extract SILA summary statistics
extract_sila_summary <- function(sila_output) {

  if (is.null(sila_output)) {
    return(tibble(
      biomarker = NA, threshold = NA, n_subjects = NA, n_obs = NA,
      mean_time_to_threshold = NA, sd_time_to_threshold = NA,
      median_time_to_threshold = NA, pct_past_threshold = NA
    ))
  }

  resfit <- sila_output$sila_fit

  # Get baseline (first observation) estimates per subject
  baseline_estimates <- resfit %>%
    group_by(subid) %>%
    slice_min(age, n = 1) %>%
    ungroup()

  tibble(
    biomarker = sila_output$biomarker,
    threshold = sila_output$threshold,
    n_subjects = sila_output$n_subjects,
    n_obs = sila_output$n_obs,
    mean_time_to_threshold = mean(baseline_estimates$estdtt0, na.rm = TRUE),
    sd_time_to_threshold = sd(baseline_estimates$estdtt0, na.rm = TRUE),
    median_time_to_threshold = median(baseline_estimates$estdtt0, na.rm = TRUE),
    pct_past_threshold = mean(baseline_estimates$estdtt0 < 0, na.rm = TRUE) * 100
  )
}
```

## Run SILA Analysis

### Primary Analysis (PREVENT-AD Threshold: CL ≥ 22.32)

```{r sila-primary}
#| fig-width: 12
#| fig-height: 8

# Run SILA for each biomarker - Primary threshold (log-transformed)
sila_results_primary <- list()

# ptau181 (log)
sila_results_primary$ptau181 <- run_sila_biomarker(
  biomarkers_long, "ptau181", cutoff_ptau181_primary, "p-tau181"
)

# ptau217 (log)
sila_results_primary$ptau217 <- run_sila_biomarker(
  biomarkers_long, "ptau217", cutoff_ptau217_primary, "p-tau217"
)

# Create plots for primary analysis (with amyloid status coloring)
plots_primary <- lapply(sila_results_primary, function(x) {
  plot_sila_results(x, paste0("ptau181 ≥ 6.42; ptau217 ≥ 3.06"),
                    amyloid_lookup = amyloid_status_lookup)
})

# Display aligned trajectory plots
aligned_plots <- compact(lapply(plots_primary, function(x) x$aligned))
if (length(aligned_plots) > 0) {
  wrap_plots(aligned_plots, ncol = 2)
}
```

## Summary Results

### Primary Threshold (PREVENT-AD Specific: CL ≥ 22.32)

```{r summary-primary}

summary_primary <- bind_rows(lapply(sila_results_primary, extract_sila_summary)) %>%
  filter(!is.na(biomarker)) %>%
  mutate(analysis = "Primary (CL ≥ 22.32)")

summary_primary %>%
  select(-analysis) %>%
  gt() %>%
  tab_header(
    title = "SILA Results - Primary Analysis",
    subtitle = paste0("Amyloid positivity: Centiloid ≥ ", THRESHOLD_PRIMARY, " (PREVENT-AD specific)")
  ) %>%
  fmt_number(
    columns = c(threshold, mean_time_to_threshold, sd_time_to_threshold, median_time_to_threshold),
    decimals = 2
  ) %>%
  fmt_number(columns = pct_past_threshold, decimals = 1) %>%
  cols_label(
    biomarker = "Biomarker",
    threshold = "Cutoff (pg/mL)",
    n_subjects = "N Subjects",
    n_obs = "N Observations",
    mean_time_to_threshold = "Mean Time (years)",
    sd_time_to_threshold = "SD",
    median_time_to_threshold = "Median",
    pct_past_threshold = "% Past Threshold"
  )
```

- Individuals on average are already biomarker positive



## Individual SILA Curves

### Discrete Rate Sampling Curves

```{r drs-curves}
#| fig-width: 10
#| fig-height: 4

# DRS plots - Primary
drs_plots_primary <- compact(lapply(plots_primary, function(x) x$drs))
if (length(drs_plots_primary) > 0) {
  wrap_plots(drs_plots_primary, ncol = 2) +
    plot_annotation(title = "Discrete Rate Sampling Curves - Primary Threshold")
}
```

### SILA Modeled Curves

```{r sila-curves}
#| fig-width: 10
#| fig-height: 4

# SILA modeled curves - Primary
sila_plots_primary <- compact(lapply(plots_primary, function(x) x$sila))
if (length(sila_plots_primary) > 0) {
  wrap_plots(sila_plots_primary, ncol = 2) +
    plot_annotation(title = "SILA Modeled Curves - Primary Threshold")
}
```

## Save Results

```{r save-results}
# Save summary table
write_csv(summary_primary,
          here("results", "Aim1", "tables", "PREVENTAD_SILA_summary.csv"))

# Save individual SILA fit data for further analysis
sila_fits_primary <- lapply(names(sila_results_primary), function(nm) {
  if (!is.null(sila_results_primary[[nm]]) && !is.null(sila_results_primary[[nm]]$sila_fit)) {
    sila_results_primary[[nm]]$sila_fit %>%
      mutate(biomarker = nm, threshold_type = "primary")
  } else {
    NULL
  }
}) %>%
  compact() %>%
  bind_rows()

if (nrow(sila_fits_primary) > 0) {
  write_csv(sila_fits_primary,
            here("results", "Aim1", "tables", "PREVENTAD_SILA_individual_fits.csv"))
  saveRDS(sila_fits_primary,
          here("data", "intermediate", "PREVENTAD_SILA_individual_fits.rds"))
  cat("- results/Aim1/tables/PREVENTAD_SILA_individual_fits.csv\n")
  cat("- data/intermediate/PREVENTAD_SILA_individual_fits.rds\n")
} else {
  cat("Warning: No SILA fit data to save\n")
}

cat("Results saved to:\n")
cat("- results/Aim1/tables/PREVENTAD_SILA_summary.csv\n")
```

