---
title: "Aim 2: Predicting Amyloid PET Positivity from Plasma Biomarkers"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
execute:
  warning: false
  message: false
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = here::here())
```


```{r loading}
library(pROC)

# Load configuration
source('workflow/scripts/config.R')
```

## Overview

This analysis predicts amyloid PET positivity from plasma biomarkers (ptau181, ptau217) using logistic regression. We calculate:

- ROC curves and AUC
- Optimal cutoffs using Youden index (sensitivity + specificity - 1)
- Sensitivity and specificity at optimal cutoffs

**Amyloid positivity thresholds:**

- Centiloid ≥ 22.32 (PREVENT-AD cohort-specific)
- Centiloid ≥ 18 (Sensitivity analysis)

## Load and Prepare Data

### Load Pre-Processed Biomarker Data

Using biomarker data already filtered by inclusion criteria (02_clean_PREVENT-AD.R):

- CONP_ID with GWAS data
- CONP_ID with at least 2 biomarker visits

```{r load-biomarkers-processed}
# Exclusion criteria
exclude_ids <- readRDS(here("data", "intermediate", "PREVENTAD_missingcrs.rds"))

# Load pre-processed biomarker data (already filtered by inclusion criteria)
biomarkers <- readRDS(here("data", "intermediate", "PREVENTAD_biomarkers.rds")) %>% 
  filter(!CONP_ID %in% exclude_ids$CONP_ID)

# Load demographics for covariates (age, sex, BMI)
crs_factors <- readRDS(here("data", "intermediate", "PREVENTAD_crsfactors_imputed.rds")) %>%
  filter(!CONP_ID %in% exclude_ids$CONP_ID) %>%
  select(CONP_ID, Sex, Age, Weight, Height) %>%
  mutate(BMI = Weight / (Height / 100)^2) %>%
  select(CONP_ID, BMI, Sex, Age)



cat("=== INCLUSION CRITERIA (from pre-processed data) ===\n")
cat("Participants meeting criteria:", nrow(crs_factors$CONP_ID), "\n")
cat("Total biomarker observations:", nrow(biomarkers), "\n")
```

### Load PET Data

```{r load-pet}
# Load raw PET data
pet_nav <- read_csv(
  file.path(PREVENTAD_DATA_PATH,
            "PET_NAV_SUVR_ref-wholeCerebellum_Registered_PREVENTAD_2025-03-21.csv"),
  show_col_types = FALSE
) 

pet_info <- read_csv(
  file.path(PREVENTAD_DATA_PATH,
            "PET_info_participants_Registered_PREVENTAD_2024-08-01.csv"),
  show_col_types = FALSE
)

# Process PET data - apply exclusion criteria
pet_amyloid <- pet_nav %>%
  filter(PET_tracer == "NAV") %>% 
  filter(!CONP_ID %in% exclude_ids$CONP_ID) %>%
  select(CONP_ID, session, amyloid_index_SUVR, WhlCbl_centiloid) %>%
  rename(
    pet_session = session,
    amyloid_suvr = amyloid_index_SUVR,
    centiloid = WhlCbl_centiloid
  )

# Get PET dates
pet_dates <- pet_info %>%
  filter(PET_tracer == "NAV") %>%
  select(CONP_ID, Date_PET_visit, Age_PET_visit) %>%
  distinct() %>%
  rename(
    pet_date = Date_PET_visit,
    age_at_pet = Age_PET_visit
  )

pet_amyloid <- pet_amyloid %>%
  left_join(pet_dates, by = c("CONP_ID"))

# PREVENT-AD amyloid positivity threshold
PRIMARY_CENTILOID_THRESHOLD <- 22.32  
SECONDARY_CENTILOID_THRESHOLD <- 18

pet_amyloid <- pet_amyloid %>%
  mutate(
    amyloid_positive_primary = as.integer(centiloid >= PRIMARY_CENTILOID_THRESHOLD),
    amyloid_status_primary = factor(
      ifelse(centiloid >= PRIMARY_CENTILOID_THRESHOLD, "A+", "A-"),
      levels = c("A-", "A+")
    ),
    amyloid_positive_secondary = as.integer(centiloid >= SECONDARY_CENTILOID_THRESHOLD),
    amyloid_status_secondary = factor(
      ifelse(centiloid >= SECONDARY_CENTILOID_THRESHOLD, "A+", "A-"),
      levels = c("A-", "A+")
    )
  )

cat("PET participants:", n_distinct(pet_amyloid$CONP_ID), "\n")
cat("\nPrimary threshold (CL >=", PRIMARY_CENTILOID_THRESHOLD, "):\n")
cat("Amyloid positive:", sum(pet_amyloid$amyloid_positive_primary),
    "(", round(mean(pet_amyloid$amyloid_positive_primary) * 100, 1), "%)\n")
cat("\nSecondary threshold (CL >=", SECONDARY_CENTILOID_THRESHOLD, "):\n")
cat("Amyloid positive:", sum(pet_amyloid$amyloid_positive_secondary),
    "(", round(mean(pet_amyloid$amyloid_positive_secondary) * 100, 1), "%)\n")
```

### Match PET with Closest Biomarker Timepoint

```{r prepare-biomarkers}
# Extract ptau181 data from pre-processed biomarkers
biomarkers_ptau181 <- biomarkers %>%
  filter(!is.na(ptau181)) %>%
  select(CONP_ID, biomarker_visit = Visit_label, biomarker_date = Date_taken, ptau181)

# Extract ptau217 data from pre-processed biomarkers
biomarkers_ptau217 <- biomarkers %>%
  filter(!is.na(ptau217)) %>%
  select(CONP_ID, biomarker_visit = Visit_label, biomarker_date = Date_taken, ptau217)

cat("Participants with ptau181:", n_distinct(biomarkers_ptau181$CONP_ID), "\n")
cat("Participants with ptau217:", n_distinct(biomarkers_ptau217$CONP_ID), "\n")

# Function to find closest biomarker measurement to PET scan
find_closest_biomarker <- function(pet_data, biomarker_data, biomarker_col) {

  # Convert to date 
  pet_data <- pet_data %>%
    mutate(pet_date_parsed = ymd(paste0(pet_date, "-01")))

  biomarker_data <- biomarker_data %>%
    mutate(biomarker_date_parsed = ymd(paste0(biomarker_date, "-01")))

  # For each PET scan, find closest biomarker
  matched_data <- pet_data %>%
    left_join(biomarker_data, by = "CONP_ID", relationship = "many-to-many") %>%
    mutate(
      date_diff_months = abs(interval(biomarker_date_parsed, pet_date_parsed) / months(1))
    ) %>%
    group_by(CONP_ID, pet_session) %>%
    slice_min(date_diff_months, n = 1, with_ties = FALSE) %>%
    ungroup()

  return(matched_data)
}

# Match ptau181
pet_ptau181 <- find_closest_biomarker(pet_amyloid, biomarkers_ptau181, "ptau181") %>%
  filter(!is.na(ptau181)) %>%
  left_join(crs_factors, by = "CONP_ID")

# Match ptau217
pet_ptau217 <- find_closest_biomarker(pet_amyloid, biomarkers_ptau217, "ptau217") %>%
  filter(!is.na(ptau217)) %>%
  left_join(crs_factors, by = "CONP_ID")
```

## Descriptive Statistics

```{r descriptive-ptau181-primary}
# Summary table for ptau181 analysis cohort
pet_ptau181 %>%
  select(amyloid_status_primary, centiloid, ptau181, age_at_pet) %>%
  tbl_summary(
    by = amyloid_status_primary,
    statistic = list(
      all_continuous() ~ "{mean} ({sd})"
    ),
    label = list(
      centiloid ~ "Centiloid",
      ptau181 ~ "p-tau181 (pg/mL)",
      age_at_pet ~ "Age at PET"
    )
  ) %>%
  add_overall() %>%
  add_p() %>%
  as_gt() %>%
  tab_header(title = "Cohort Characteristics by Amyloid Status (ptau181)",
             subtitle = "Centiloid ≥ 22.32")
```

```{r descriptive-ptau181-secondary}
# Summary table for ptau181 analysis cohort
pet_ptau181 %>%
  select(amyloid_status_secondary, centiloid, ptau181, age_at_pet) %>%
  tbl_summary(
    by = amyloid_status_secondary,
    statistic = list(
      all_continuous() ~ "{mean} ({sd})"
    ),
    label = list(
      centiloid ~ "Centiloid",
      ptau181 ~ "p-tau181 (pg/mL)",
      age_at_pet ~ "Age at PET"
    )
  ) %>%
  add_overall() %>%
  add_p() %>%
  as_gt() %>%
  tab_header(title = "Cohort Characteristics by Amyloid Status (ptau181)",
             subtitle = "Centiloid ≥ 18")
```

```{r descriptive-ptau217-primary}
# Summary table for ptau217 analysis cohort
pet_ptau217 %>%
  select(amyloid_status_primary, centiloid, ptau217, age_at_pet) %>%
  tbl_summary(
    by = amyloid_status_primary,
    statistic = list(
      all_continuous() ~ "{mean} ({sd})"
    ),
    label = list(
      centiloid ~ "Centiloid",
      ptau217 ~ "p-tau217 (pg/mL)",
      age_at_pet ~ "Age at PET"
    )
  ) %>%
  add_overall() %>%
  add_p() %>%
  as_gt() %>%
  tab_header(title = "Cohort Characteristics by Amyloid Status (ptau217)",
             subtitle = "Centiloid ≥ 22.32")
```

```{r descriptive-ptau217-secondary}
# Summary table for ptau217 analysis cohort
pet_ptau217 %>%
  select(amyloid_status_secondary, centiloid, ptau217, age_at_pet) %>%
  tbl_summary(
    by = amyloid_status_secondary,
    statistic = list(
      all_continuous() ~ "{mean} ({sd})"
    ),
    label = list(
      centiloid ~ "Centiloid",
      ptau217 ~ "p-tau217 (pg/mL)",
      age_at_pet ~ "Age at PET"
    )
  ) %>%
  add_overall() %>%
  add_p() %>%
  as_gt() %>%
  tab_header(title = "Cohort Characteristics by Amyloid Status (ptau217)",
             subtitle = "Centiloid ≥ 18")
```

## Biomarker Distributions

```{r distributions}
#| fig-width: 10
#| fig-height: 4

# ptau181 distribution
p1 <- pet_ptau181 %>%
  ggplot(aes(x = ptau181, fill = amyloid_status_primary)) +
  geom_histogram(bins = 30, alpha = 0.7, position = "identity") +
  scale_fill_manual(values = c("A-" = "#2166AC", "A+" = "#B2182B")) +
  labs(
    title = "p-tau181 Distribution by Amyloid Status",
    subtitle = "Centiloid ≥ 22.32",
    x = "p-tau181 (pg/mL)",
    y = "Count",
    fill = "Amyloid Status"
  ) +
  theme_minimal()

# ptau217 distribution
p2 <- pet_ptau217 %>%
  ggplot(aes(x = ptau217, fill = amyloid_status_primary)) +
  geom_histogram(bins = 30, alpha = 0.7, position = "identity") +
  scale_fill_manual(values = c("A-" = "#2166AC", "A+" = "#B2182B")) +
  labs(
    title = "p-tau217 Distribution by Amyloid Status",
    subtitle = "Centiloid ≥ 22.32",
    x = "p-tau217 (pg/mL)",
    y = "Count",
    fill = "Amyloid Status"
  ) +
  theme_minimal()

library(patchwork)
p1 + p2
```

## Logistic Regression Models

### p-tau181 Models

```{r model-ptau181}
# Model 1: Biomarker only
model_ptau181_bio <- glm(
  amyloid_positive_primary ~ ptau181,
  data = pet_ptau181,
  family = binomial(link = "logit")
)

# Model 1s: Biomarker only (Secondary threshold)
model_ptau181_bio_s <- glm(
  amyloid_positive_secondary ~ ptau181,
  data = pet_ptau181,
  family = binomial(link = "logit")
)

# Model 2: Covariates only (age, sex, BMI)
model_ptau181_cov <- glm(
  amyloid_positive_primary ~ age_at_pet + Sex + BMI,
  data = pet_ptau181,
  family = binomial(link = "logit")
)

# Model 3: Biomarker + covariates
model_ptau181_full <- glm(
  amyloid_positive_primary ~ ptau181 + age_at_pet + Sex + BMI,
  data = pet_ptau181,
  family = binomial(link = "logit")
)

# Model 1: Biomarker only
cat("=== Model 1: p-tau181 Only ===\n\n")
summary(model_ptau181_bio)
cat("\nOdds Ratios (95% CI):\n")
exp(cbind(OR = coef(model_ptau181_bio), confint(model_ptau181_bio)))

# Model 1s: Biomarker only (Secondary threshold)
cat("=== Model 1s: p-tau181 Only (Secondary threshold) ===\n\n")
summary(model_ptau181_bio_s)
cat("\nOdds Ratios (95% CI):\n")
exp(cbind(OR = coef(model_ptau181_bio_s), confint(model_ptau181_bio_s)))

# Model 2: Covariates only
cat("\n\n=== Model 2: Covariates Only (Age, Sex, BMI) ===\n\n")
summary(model_ptau181_cov)
cat("\nOdds Ratios (95% CI):\n")
exp(cbind(OR = coef(model_ptau181_cov), confint(model_ptau181_cov)))

# Model 3: Full model
cat("\n\n=== Model 3: p-tau181 + Covariates ===\n\n")
summary(model_ptau181_full)
cat("\nOdds Ratios (95% CI):\n")
exp(cbind(OR = coef(model_ptau181_full), confint(model_ptau181_full)))
```

### p-tau217 Models

```{r model-ptau217}
# Model 1: Biomarker only
model_ptau217_bio <- glm(
  amyloid_positive_primary ~ ptau217,
  data = pet_ptau217,
  family = binomial(link = "logit")
)

# Model 1s: Biomarker only (Secondary threshold)
model_ptau217_bio_s <- glm(
  amyloid_positive_secondary ~ ptau217,
  data = pet_ptau217,
  family = binomial(link = "logit")
)

# Model 2: Covariates only (age, sex, BMI)
model_ptau217_cov <- glm(
  amyloid_positive_primary ~ age_at_pet + Sex + BMI,
  data = pet_ptau217,
  family = binomial(link = "logit")
)

# Model 3: Biomarker + covariates
model_ptau217_full <- glm(
  amyloid_positive_primary ~ ptau217 + age_at_pet + Sex + BMI,
  data = pet_ptau217,
  family = binomial(link = "logit")
)

# Model 1: Biomarker only
cat("=== Model 1: p-tau217 Only ===\n\n")
summary(model_ptau217_bio)
cat("\nOdds Ratios (95% CI):\n")
exp(cbind(OR = coef(model_ptau217_bio), confint(model_ptau217_bio)))

# Model 1s: Biomarker only (Secondary threshold)
cat("=== Model 1: p-tau217 Only (Secondary threshold) ===\n\n")
summary(model_ptau217_bio_s)
cat("\nOdds Ratios (95% CI):\n")
exp(cbind(OR = coef(model_ptau217_bio_s), confint(model_ptau217_bio_s)))

# Model 2: Covariates only
cat("\n\n=== Model 2: Covariates Only (Age, Sex, BMI) ===\n\n")
summary(model_ptau217_cov)
cat("\nOdds Ratios (95% CI):\n")
exp(cbind(OR = coef(model_ptau217_cov), confint(model_ptau217_cov)))

# Model 3: Full model
cat("\n\n=== Model 3: p-tau217 + Covariates ===\n\n")
summary(model_ptau217_full)
cat("\nOdds Ratios (95% CI):\n")
exp(cbind(OR = coef(model_ptau217_full), confint(model_ptau217_full)))
```

## ROC Analysis and Optimal Cutoffs

### p-tau181 ROC

```{r roc-ptau181}
#| fig-width: 6
#| fig-height: 6

# ROC curve - ptau181
roc_ptau181 <- roc(
  pet_ptau181$amyloid_positive_primary,
  pet_ptau181$ptau181,
  levels = c(0, 1),
  direction = "<"
)

# AUC with confidence interval
auc_ptau181 <- auc(roc_ptau181)
ci_auc_ptau181 <- ci.auc(roc_ptau181)

cat("=== p-tau181 ROC Analysis ===\n")
cat("AUC:", round(auc_ptau181, 3), "\n")
cat("95% CI:", round(ci_auc_ptau181[1], 3), "-", round(ci_auc_ptau181[3], 3), "\n")

# Youden index optimal cutoff
coords_ptau181 <- coords(roc_ptau181, "best", best.method = "youden", ret = "all")
cat("\n=== Optimal Cutoff (Youden Index) ===\n")
cat("Cutoff:", round(coords_ptau181$threshold, 2), "pg/mL\n")
cat("Sensitivity:", round(coords_ptau181$sensitivity * 100, 1), "%\n")
cat("Specificity:", round(coords_ptau181$specificity * 100, 1), "%\n")
cat("PPV:", round(coords_ptau181$ppv * 100, 1), "%\n")
cat("NPV:", round(coords_ptau181$npv * 100, 1), "%\n")
cat("Youden Index:", round(coords_ptau181$sensitivity + coords_ptau181$specificity - 1, 3), "\n")

# Plot ROC
plot(roc_ptau181,
     main = "ROC Curve: p-tau181 Predicting Amyloid Positivity",
     print.auc = TRUE,
     print.auc.x = 0.4,
     print.auc.y = 0.2,
     col = "#2166AC",
     lwd = 2)

# Add optimal cutoff point
points(coords_ptau181$specificity, coords_ptau181$sensitivity,
       pch = 19, col = "#B2182B", cex = 1.5)
text(coords_ptau181$specificity - 0.1, coords_ptau181$sensitivity + 0.05,
     paste0("Cutoff: ", round(coords_ptau181$threshold, 2), " pg/mL"),
     col = "#B2182B", cex = 0.8)
```

### p-tau217 ROC

```{r roc-ptau217}
#| fig-width: 6
#| fig-height: 6

# ROC curve - ptau217
roc_ptau217 <- roc(
  pet_ptau217$amyloid_positive_primary,
  pet_ptau217$ptau217,
  levels = c(0, 1),
  direction = "<"
)

# AUC with confidence interval
auc_ptau217 <- auc(roc_ptau217)
ci_auc_ptau217 <- ci.auc(roc_ptau217)

cat("=== p-tau217 ROC Analysis ===\n")
cat("AUC:", round(auc_ptau217, 3), "\n")
cat("95% CI:", round(ci_auc_ptau217[1], 3), "-", round(ci_auc_ptau217[3], 3), "\n")

# Youden index optimal cutoff
coords_ptau217 <- coords(roc_ptau217, "best", best.method = "youden", ret = "all")
cat("\n=== Optimal Cutoff (Youden Index) ===\n")
cat("Cutoff:", round(coords_ptau217$threshold, 2), "pg/mL\n")
cat("Sensitivity:", round(coords_ptau217$sensitivity * 100, 1), "%\n")
cat("Specificity:", round(coords_ptau217$specificity * 100, 1), "%\n")
cat("PPV:", round(coords_ptau217$ppv * 100, 1), "%\n")
cat("NPV:", round(coords_ptau217$npv * 100, 1), "%\n")
cat("Youden Index:", round(coords_ptau217$sensitivity + coords_ptau217$specificity - 1, 3), "\n")

# Plot ROC
plot(roc_ptau217,
     main = "ROC Curve: p-tau217 Predicting Amyloid Positivity",
     print.auc = TRUE,
     print.auc.x = 0.4,
     print.auc.y = 0.2,
     col = "#2166AC",
     lwd = 2)

# Add optimal cutoff point
points(coords_ptau217$specificity, coords_ptau217$sensitivity,
       pch = 19, col = "#B2182B", cex = 1.5)
text(coords_ptau217$specificity - 0.1, coords_ptau217$sensitivity + 0.05,
     paste0("Cutoff: ", round(coords_ptau217$threshold, 2), " pg/mL"),
     col = "#B2182B", cex = 0.8)
```


## Summary Results

```{r summary-table}
# Create summary table with biomarker-only models
results_summary <- tibble(
  Biomarker = c("p-tau181", "p-tau217"),
  N = c(nrow(pet_ptau181), nrow(pet_ptau217)),
  `N Amyloid+` = c(sum(pet_ptau181$amyloid_positive_primary), sum(pet_ptau217$amyloid_positive_primary)),
  `AUC (95% CI)` = c(
    paste0(round(auc(roc_ptau181), 3), " (", round(ci_auc_ptau181[1], 3), "-", round(ci_auc_ptau181[3], 3), ")"),
    paste0(round(auc(roc_ptau217), 3), " (", round(ci_auc_ptau217[1], 3), "-", round(ci_auc_ptau217[3], 3), ")")
  ),
  `Cutoff (pg/mL)` = c(
    round(coords_ptau181$threshold, 2),
    round(coords_ptau217$threshold, 2)
  ),
  `Sensitivity` = c(
    paste0(round(coords_ptau181$sensitivity * 100, 1), "%"),
    paste0(round(coords_ptau217$sensitivity * 100, 1), "%")
  ),
  `Specificity` = c(
    paste0(round(coords_ptau181$specificity * 100, 1), "%"),
    paste0(round(coords_ptau217$specificity * 100, 1), "%")
  ),
  `PPV` = c(
    paste0(round(coords_ptau181$ppv * 100, 1), "%"),
    paste0(round(coords_ptau217$ppv * 100, 1), "%")
  ),
  `NPV` = c(
    paste0(round(coords_ptau181$npv * 100, 1), "%"),
    paste0(round(coords_ptau217$npv * 100, 1), "%")
  ),
  `Youden Index` = c(
    round(coords_ptau181$sensitivity + coords_ptau181$specificity - 1, 3),
    round(coords_ptau217$sensitivity + coords_ptau217$specificity - 1, 3)
  )
)

results_summary %>%
  gt() %>%
  tab_header(
    title = "Summary: Plasma Biomarkers Predicting Amyloid PET Positivity",
    subtitle = paste0("Amyloid positivity defined as Centiloid ≥ ", PRIMARY_CENTILOID_THRESHOLD, " (PREVENT-AD specific)")
  )
```

```{r save-results}
# Save primary results
write_csv(results_summary,
          file.path(AIM1_OUTPUT_PATH$tables, "PREVENTAD_amyloid_prediction_primary.csv"))

cat("\nResults saved to:\n")
cat("- ", file.path(AIM1_OUTPUT_PATH$tables, "PREVENTAD_amyloid_prediction_primary.csv"), "\n")
```

