---
title: "EDA: Bivariate Analysis"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
execute:
  warning: false
  message: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warnings = FALSE)
knitr::opts_knit$set(root.dir = here::here())
```

```{r setting}
# Load packages and data
library(broom)
library(car)
library(patchwork)

source('workflow/scripts/config.R')

predictors_dat <- readRDS(file.path(DATA_CLEANED_PATH, "PREVENTAD_predictors.rds"))
lastvisit_dat <- readRDS(file.path(DATA_CLEANED_PATH, "PREVENTAD_lastvisit.rds"))
cogd_dat <- readRDS(file.path(DATA_INTERMEDIATE_PATH, "PREVENTAD_cogd_scored.rds"))
libra2_dat <- readRDS(file.path(DATA_INTERMEDIATE_PATH, "PREVENTAD_libra2_scored.rds"))
all_dat <- readRDS(file.path(DATA_CLEANED_PATH, "PREVENTAD_longitudinal.rds"))

lastvisit_longer <- lastvisit_dat %>%
  pivot_longer(
    cols = all_of(BIOMARKERS_TRANS),
    names_to = "biomarker",
    values_to = "measure"
  )
```

## Overview

This analysis examines associations between AD blood-based biomarkers (ptau-217, ptau-181, NfL, GFAP, Aβ42/40 ratio, and ptau-217/Aβ42 ratio) and key risk indicators including family history, genetic risk (GRS, APOE), clinical risk scores (CogDrisk, LIBRA2), and demographic factors. All biomarkers except Aβ42/40 ratio were log-transformed to normalize distributions.

## AD risk indicator associations with biomarker levels (measured at last visit)

### Genetic Risk Score (GRS)

```{r grs}
grs_plot <- lastvisit_longer %>%
  ggplot(aes(x = GRS, y = measure)) +
  geom_point(alpha = 0.5, size = 1, na.rm=TRUE) +
  geom_smooth(method = "loess", se = TRUE, linewidth = 0.5, na.rm=TRUE, color="blue") +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.5, na.rm=TRUE, color = "red", linetype = "dashed") +
  stat_cor(label.x.npc=0.6,
           label.y.npc=0.8,
           size=2.75,
           geom = "label",
           color="red",
           label.sep = "\n",
           na.rm=TRUE) +
  facet_wrap(~biomarker, scales="free") + 
  labs(x = "GRS",
       y = "Biomarker Level",
       title = "GRS vs AD Blood-based Biomarker Levels",
       subtitle = "Biomarker level taken at last visit") +
  theme_minimal() +
  theme(legend.position = "bottom") 
grs_plot

# Save plot
# ggsave(file.path(AIM1_OUTPUT_PATH$figures, "PREVENTAD_GRS_vs_BBM.png"), grs_plot)
```

### CogDrisk 

```{r cogdrisk}
cogd_plot <- lastvisit_longer %>%
  ggplot(aes(x = score_cogd, y = measure)) +
  geom_point(alpha = 0.5, size = 1, na.rm=TRUE) +
  geom_smooth(method = "loess", se = TRUE, linewidth = 0.5, color = "blue", na.rm=TRUE) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.5, na.rm=TRUE, color = "red", linetype = "dashed") +
  stat_cor(label.x.npc=0.6,
           label.y.npc=0.2,
           size=2.75,
           geom = "label",
           color="red",
           label.sep = "\n",
           na.rm=TRUE) +
  facet_wrap(~biomarker, scales="free") + 
  labs(x = "CogDrisk",
       y = "Biomarker Level",
       title = "CogDrisk vs Biomarker Levels",
       subtitle = "Biomarker level taken at last visit") +
  theme_minimal() +
  theme(legend.position = "bottom") 
cogd_plot
# Save plot
# ggsave(file.path(AIM1_OUTPUT_PATH$figures, "PREVENTAD_CogD_vs_BBM.png"), cogd_plot)
```

### LIBRA2

```{r libra2}
libra2_plot <- lastvisit_longer %>%
  ggplot(aes(x = score_libra2, y = measure)) +
  geom_point(alpha = 0.5, size = 1, na.rm=TRUE) +
  geom_smooth(method = "loess", se = TRUE, linewidth = 0.5, color = "blue", na.rm=TRUE) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.5, na.rm=TRUE, color = "red", linetype = "dashed") +
  stat_cor(label.x.npc=0.6,
           label.y.npc=0.2,
           size=2.75,
           geom = "label",
           #position="jitter",
           color="red",
           label.sep = "\n",
           na.rm=TRUE) +
  facet_wrap(~biomarker, scales="free") + 
  labs(x = "LIBRA2",
       y = "Biomarker Level",
       title = "LIBRA2 vs AD Blood-based Biomarker Levels",
       subtitle = "Biomarker level taken at last visit") +
  theme_minimal() +
  theme(legend.position = "bottom") 
libra2_plot
# Save plot
# ggsave(file.path(AIM1_OUTPUT_PATH$figures, "PREVENTAD_LIBRA2_vs_BBM.png"), libra2_plot)
```

### APOE

```{r apoe}
apoe_plot <- lastvisit_longer %>%
  ggplot(aes(x=apoe, y=measure, fill=apoe)) +
  geom_violin(alpha=0.7) +
  facet_wrap(~ biomarker, scales="free") +
  geom_boxplot(width=0.1, fill="white") +
  stat_compare_means(method = "kruskal.test",  
                     label.y.npc = 0.9,
                     geom = "label",
                     #fill = "white",
                     size=3) +
  labs(x = "APOE",
       y = "Biomarker Level",
       title = "Biomarker Levels by APOE Status",
       subtitle = "Biomarker level taken at last visit") +
  theme(legend.position="none",
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
apoe_plot
# Save plot
# ggsave(file.path(AIM1_OUTPUT_PATH$figures, "PREVENTAD_apoe_vs_BBM.png"), plot2)
```

### Family history

```{r fhx}
fhx_plot <- lastvisit_longer %>%
  ggplot(aes(x=family_history, y=measure, fill=family_history)) +
  geom_violin(alpha=0.7) +
  facet_wrap(~ biomarker, scales="free") +
  geom_boxplot(width=0.1, fill="white") +
  stat_compare_means(method = "kruskal.test",  
                     label.y.npc = 0.8,
                     geom = "label",
                     #fill = "white",
                     size=3) +
  labs(x = "Family History",
       y = "Biomarker Level",
       title = "Biomarker Levels by Family History",
       subtitle = "Biomarker level taken at last visit") +
  theme(legend.position="none",
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
fhx_plot
# Save plot
# ggsave(file.path(AIM1_OUTPUT_PATH$figures, "PREVENTAD_apoe_vs_BBM.png"), plot2)
```

## Demographic risk factor associations with biomarker levels (last visit)

### Age
```{r age}
dem_dat_longer <- lastvisit_longer %>%
  select(age, sex, educ, biomarker, measure)

dem_age_plot <- dem_dat_longer %>%
  ggplot(aes(x = age, y = measure)) +
  geom_point(alpha = 0.5, size = 1, na.rm=TRUE) +
  geom_smooth(method = "loess", se = TRUE, linewidth = 0.5, color = "blue", na.rm=TRUE) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.5, na.rm=TRUE, color = "red", linetype = "dashed") +
  stat_cor(label.x.npc=0.6,
           label.y.npc=0.2,
           size=2.75,
           geom = "label",
           color="red",
           label.sep = "\n",
           na.rm=TRUE) +
  facet_wrap(~biomarker, scales="free") + 
  labs(x = "Age",
       y = "Biomarker Level (pg/mL or AB ratio)",
       title = "Age vs AD Blood-based Biomarker Levels",
       subtitle = "Biomarker level taken at last visit") +
  theme_minimal() +
  theme(legend.position = "bottom") 
dem_age_plot
```
### Sex

```{r sex}
dem_sex_plot <- dem_dat_longer %>%
  ggplot(aes(x=sex, y=measure, fill=sex)) +
  geom_violin(alpha=0.7) +
  facet_wrap(~ biomarker, scales="free") +
  geom_boxplot(width=0.1, fill="white") +
  stat_compare_means(method = "wilcox.test",  
                     label.y.npc = 0.8,
                     geom = "label",
                     size=3) +
  labs(x = "Sex",
       y = "Biomarker Level (pg/mL or AB ratio)",
       title = "Biomarker Levels by Sex",
       subtitle = "Biomarker level taken at last visit") +
  theme(legend.position="none",
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
dem_sex_plot
```

### Education

```{r educ}
dem_educ_plot <- dem_dat_longer %>%
  ggplot(aes(x = educ, y = measure)) +
  geom_point(alpha = 0.5, size = 1, na.rm=TRUE) +
  geom_smooth(method = "loess", se = TRUE, linewidth = 0.5, color = "blue", na.rm=TRUE) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.5, na.rm=TRUE, color = "red", linetype = "dashed") +
  stat_cor(label.x.npc=0.6,
           label.y.npc=0.2,
           size=2.75,
           geom = "label",
           #position="jitter",
           color="red",
           label.sep = "\n",
           na.rm=TRUE) +
  facet_wrap(~biomarker, scales="free") + 
  labs(x = "Education (years)",
       y = "Biomarker Level (pg/mL or AB ratio)",
       title = "Education vs AD Blood-based Biomarker Levels",
       subtitle = "Biomarker level taken at last visit") +
  theme_minimal() +
  theme(legend.position = "bottom") 
dem_educ_plot

# Save plots
# ggsave(file.path(AIM1_OUTPUT_PATH$figures, "PREVENTAD_age_vs_BBM.png"), dem_age_plot)
# ggsave(file.path(AIM1_OUTPUT_PATH$figures, "PREVENTAD_sex_vs_BBM.png"), dem_sex_plot)
# ggsave(file.path(AIM1_OUTPUT_PATH$figures, "PREVENTAD_educ_vs_BBM.png"), dem_educ_plot)
```

## Individual CRS factor associations with biomarker levels (last visit)

### CogDrisk

```{r cogd-factors-setup}
cogd_bm_dat <- lastvisit_dat %>%
  left_join(cogd_dat, by="CONP_ID") 

cogd_preventad <- expand_grid(outcome = BIOMARKERS_TRANS) %>%
  mutate(
    eq = glue("{outcome} ~ Sex + Age + Education_level + BMI_category + Hypertension + High_cholesterol + Depression + Atrial_fibrillation + Diabetes + Insomnia + TBI + Smoking + Cognitive_engagement + Social_engagement + Physical_inactivity"),
    res = map(eq, ~lm(as.formula(.), data = cogd_bm_dat)),
    dataf = map(res, tidy),
    mod = map(res, glance)
  )
```

```{r cogd-factors-model}
res_cogd_preventad <- bind_rows(select(cogd_preventad, outcome, dataf)) %>%
  unnest(dataf) %>%
  filter(
    str_detect(term, '^Sex|^Age|^Education_level|^BMI_category|^Hypertension|^High_cholesterol|^Depression|^Atrial_fibrillation|^Diabetes|^Insomnia|^TBI|^Smoking|^Cognitive_engagement|^Social_engagement|^Physical_inactivity')
  ) %>%
  arrange(outcome, term, estimate) %>%
  mutate(
    # Significance flag
    sig = ifelse(p.value < 0.05, TRUE, FALSE),
    
    # Confidence intervals
    conf.low = estimate - (std.error * 1.96),
    conf.high = estimate + (std.error * 1.96),
    
    # Exponentiated values for log-transformed outcomes 
    exp_est = case_when(
      str_detect(outcome, "_log") ~ exp(estimate),
      TRUE ~ NA_real_
    ),
    exp_lci = case_when(
      str_detect(outcome, "_log") ~ exp(conf.low),
      TRUE ~ NA_real_
    ),
    exp_uci = case_when(
      str_detect(outcome, "_log") ~ exp(conf.high),
      TRUE ~ NA_real_
    ),
    pct_change = case_when(
      str_detect(outcome, "_log") ~ (exp_est - 1) * 100,
      TRUE ~ NA_real_
    ),
    
    # Result category
    res = case_when(
      sig == TRUE ~ "Significant",
      sig == FALSE ~ "Not Significant"
    ),
    stars = case_when(
      p.value > 0.05 ~ "", 
      between(p.value,  0.01, 0.05) ~ "*",
      between(p.value, 0.001, 0.01) ~ "**",
      p.value < 0.001 ~ "***"
    ),
    
    term = fct_recode(term,
                      "Atrial fibrillation" = "Atrial_fibrillation",
                      "BMI (Obese)" = "BMI_categoryObese",
                      "BMI (Overweight)" = "BMI_categoryOverweight",
                      "BMI (Underweight)" = "BMI_categoryUnderweight",
                      "Cognitive engagement (Middle)" = "Cognitive_engagement2",
                      "Cognitive engagement (High)" = "Cognitive_engagement3",
                      "Depression" = "Depression",
                      "Diabetes" = "Diabetes",
                      "Education level (Low)" = "Education_levelLow",
                      "Education level (Middle)" = "Education_levelMiddle",
                      "High Cholesterol" = "High_cholesterol",
                      "Hypertension" = "Hypertension",
                      "Insomnia" = "Insomnia",
                      "Physical inactivity" = "Physical_inactivity",
                      "Sex (Male)" = "SexMale",
                      "Social Engagement (Lonely)" = "Social_engagement1",
                      "TBI" = "TBI")) %>%
  # Round numeric values
  mutate(across(c(estimate, std.error, statistic, p.value, 
                  conf.low, conf.high, exp_est, exp_lci, exp_uci, pct_change), 
                ~round(., digits = 3)))
```

```{r cogd-factors-table}
cogd_res_table <- res_cogd_preventad %>% 
  select(outcome, term, estimate, std.error, statistic, p.value, stars) %>% 
  pivot_wider(
    names_from = outcome,
    names_sep = "_",
    values_from = c(estimate, std.error, statistic, p.value, stars)
    ) %>%
  gt() %>% 
  tab_spanner(
    label = md("**AB42/40**"),
    columns = ends_with("ab_ratio")
  ) %>%
  tab_spanner(
    label = md("**GFAP**"),
    columns = ends_with("gfap_log")
  ) %>%
  tab_spanner(
    label = md("**NfL**"),
    columns = ends_with("nfl_log")
  ) %>%
  tab_spanner(
    label = md("**ptau181**"),
    columns = ends_with("ptau181_log")
  ) %>%
  tab_spanner(
    label = md("**ptau217**"),
    columns = ends_with("ptau217_log")
  ) %>%
  tab_spanner(
    label = md("**ptau217/AB42**"),
    columns = ends_with("ab42_ratio_log")
  ) %>%
  cols_label(
    starts_with("estimate")~ md("*Beta*"),
    starts_with("std.error")~ md("*SE*"),
    starts_with("statistic")~md("*Statistic*"),
    starts_with("p.value")~ md("*p.value*"),
    starts_with("stars")~ ""
  ) %>%
  tab_header(title = md("**Biomarker ~ CogDrisk Factors**")) 
cogd_res_table
```

```{r cogd-factors-plot}
plot_data <- res_cogd_preventad %>%
    mutate(sig = factor(sig, levels = c(TRUE, FALSE)))
  
res_plot <- ggplot(plot_data,
                   aes(y = term, x = estimate, xmin = conf.low, xmax = conf.high, color = term, alpha = sig)) +
  geom_point(size = 3, position = position_dodge(width = 0.5)) +  
  geom_errorbarh(linewidth = 0.8, 
                position = position_dodge(width = 0.5), 
                height = 0.2) +  
  geom_vline(xintercept = 0, linetype = 2) +
  scale_alpha_manual(
    values = c(`FALSE` = 0.2, `TRUE` = 1),  # Not sig = transparent, sig = opaque
    labels = c(`FALSE` = "Not Significant", `TRUE` = "Significant"),
    name = "Significance"
  ) +
  facet_wrap(~outcome, scales = "free_x", ncol = 2) +
  labs(
    x = "Beta (95% CI)",
    y = "CogDrisk Factors",
    title = "Biomarker ~ CogDrisk Factors"
  ) +
  theme_bw() +
  theme(
    axis.text.y = element_text(size = 10, angle = 0, hjust = 1),
    axis.text.x = element_text(size = 10),
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 14, hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(size = 11, hjust = 0.5),
    strip.background = element_rect(fill = "grey90"),
    strip.text = element_text(face = "bold", size = 11),
    legend.position = "none"
  )
res_plot

# Save plot
# ggsave(file.path(AIM1_OUTPUT_PATH$figures, "PREVENTAD_CogDfactors_vs_BBM.png"), res_plot, width = 14, height = 14, dpi = 300)
```


### LIBRA2

```{r libra2-factors-setup}
libra_bm_dat <- lastvisit_dat %>%
  left_join(libra2_dat, by="CONP_ID")

libra_preventad <- expand_grid(outcome = BIOMARKERS_TRANS) %>%
  mutate(
    eq = glue("{outcome} ~ Sleep_disturbances + Hypertension + Hypercholesterolemia + Obesity + Hearing_impairment + Diabetes + Depression + Smoking + Cognitive_activity + Social_participation + Physical_activity"),
    res = map(eq, ~lm(as.formula(.), data = libra_bm_dat)),
    dataf = map(res, tidy),
    mod = map(res, glance)
  )
```

```{r libra2-factors-model}
res_libra_preventad <- bind_rows(select(libra_preventad, outcome, dataf)) %>%
  unnest(dataf) %>%
  filter(
    str_detect(term, '^Sleep_disturbances|^Hypertension|^Hypercholesterolemia|^Obesity|^Hearing_impairment|^Diabetes|^Depression|^Smoking|^Cognitive_activity|^Social_participation|^Physical_activity')
  ) %>%
  arrange(outcome, term, estimate) %>%
  mutate(
    # Significance flag
    sig = ifelse(p.value < 0.05, TRUE, FALSE),
    
    # Confidence intervals
    conf.low = estimate - (std.error * 1.96),
    conf.high = estimate + (std.error * 1.96),
    
    # Exponentiated values for log-transformed outcomes 
    exp_est = case_when(
      str_detect(outcome, "_log") ~ exp(estimate),
      TRUE ~ NA_real_
    ),
    exp_lci = case_when(
      str_detect(outcome, "_log") ~ exp(conf.low),
      TRUE ~ NA_real_
    ),
    exp_uci = case_when(
      str_detect(outcome, "_log") ~ exp(conf.high),
      TRUE ~ NA_real_
    ),
    pct_change = case_when(
      str_detect(outcome, "_log") ~ (exp_est - 1) * 100,
      TRUE ~ NA_real_
    ),
    
    # Result category
    res = case_when(
      sig == TRUE ~ "Significant",
      sig == FALSE ~ "Not Significant"
    ),
    stars = case_when(
      p.value > 0.05 ~ "", 
      between(p.value,  0.01, 0.05) ~ "*",
      between(p.value, 0.001, 0.01) ~ "**",
      p.value < 0.001 ~ "***"
    ),
    
    term = fct_recode(term,
                      "Sleep Disturbances" = "Sleep_disturbances",
                      "Hearing Impairment" = "Hearing_impairment",
                      "Cognitive Activity" = "Cognitive_activity",
                      "Social Participation" = "Social_participation",
                      "Physical Activity" = "Physical_activity")) %>%
  # Round numeric values
  mutate(across(c(estimate, std.error, statistic, p.value, 
                  conf.low, conf.high, exp_est, exp_lci, exp_uci, pct_change), 
                ~round(., digits = 3)))
```

```{r libra2-factors-table}
libra_res_table <- res_libra_preventad %>% 
    select(outcome, term, estimate, std.error, statistic, p.value, stars) %>% 
    pivot_wider(
      names_from = outcome,
      names_sep = "_",
      values_from = c(estimate, std.error, statistic, p.value, stars)
    ) %>%
    gt() %>% 
    tab_spanner(
      label = md("**AB42/40**"),
      columns = ends_with("ab_ratio")
    ) %>%
    tab_spanner(
      label = md("**GFAP**"),
      columns = ends_with("gfap_log")
    ) %>%
    tab_spanner(
      label = md("**NfL**"),
      columns = ends_with("nfl_log")
    ) %>%
    tab_spanner(
      label = md("**ptau181**"),
      columns = ends_with("ptau181_log")
    ) %>%
    tab_spanner(
      label = md("**ptau217**"),
      columns = ends_with("ptau217_log")
    ) %>%
    tab_spanner(
      label = md("**ptau217/AB42**"),
      columns = ends_with("ab42_ratio_log")
    ) %>%
    cols_label(
      starts_with("estimate")~ md("*Beta*"),
      starts_with("std.error")~ md("*SE*"),
      starts_with("statistic")~md("*Statistic*"),
      starts_with("p.value")~ md("*p.value*"),
      starts_with("stars")~ ""
    ) %>%
    tab_header(title = md("**Biomarker ~ LIBRA2 Factors**")) 
libra_res_table
```

```{r libra2-factors-plot}
plot_data <- res_libra_preventad %>%
    mutate(sig = factor(sig, levels = c(TRUE, FALSE)))
  
res_plot <- ggplot(plot_data,
                   aes(y = term, x = estimate, xmin = conf.low, xmax = conf.high, color = term, alpha = sig) 
  ) +
  geom_point(size = 3, position = position_dodge(width = 0.5)) +  
  geom_errorbarh(linewidth = 0.8, position = position_dodge(width = 0.5), height = 0.2) + 
  geom_vline(xintercept = 0, linetype = 2) +
  scale_alpha_manual(
    values = c(`FALSE` = 0.2, `TRUE` = 1),  # Not sig = transparent, sig = opaque
    labels = c(`FALSE` = "Not Significant", `TRUE` = "Significant"),
    name = "Significance"
  ) +
  facet_wrap(~outcome, scales = "free_x", ncol = 2) +
  labs(
    x = "Beta (95% CI)",
    y = "LIBRA2 Factors",
    title = "Biomarker ~ LIBRA2 Factors"
  ) +
  theme_bw() +
  theme(
    axis.text.y = element_text(size = 10, angle = 0, hjust = 1),
    axis.text.x = element_text(size = 10),
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 14, hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(size = 11, hjust = 0.5),
    strip.background = element_rect(fill = "grey90"),
    strip.text = element_text(face = "bold", size = 11),
    legend.position = "none"
  )
res_plot

# Save plot
# ggsave(file.path(AIM1_OUTPUT_PATH$figures, "PREVENTAD_LIBRA2factors_vs_BBM.png"), res_plot, width = 14, height = 14, dpi = 300)
```
