---
title: "PREVENT-AD CRS Missing Data Analysis"
format: html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = here::here())
```

```{r load-packages}
source('workflow/scripts/config.R')
source(COGD_RECODE_FN)
source(LIBRA_RECODE_FN)
```

```{r load-data}
# Raw CRS factors — used for true missingness assessment
crs_factors <- readRDS(file.path(DATA_INTERMEDIATE_PATH, "PREVENTAD_crsfactors.rds"))

# Prepared (pre-imputation) and imputed scored data
cogd_original_scored  <- readRDS(file.path(DATA_INTERMEDIATE_PATH, "PREVENTAD_cogd_original_scored.rds"))
cogd_scored           <- readRDS(file.path(DATA_INTERMEDIATE_PATH, "PREVENTAD_cogd_scored.rds"))
libra_original_scored <- readRDS(file.path(DATA_INTERMEDIATE_PATH, "PREVENTAD_libra2_scored_original.rds"))
libra_scored          <- readRDS(file.path(DATA_INTERMEDIATE_PATH, "PREVENTAD_libra2_scored.rds"))

# Demographics and biomarkers for comparison table
risk_baseline <- readRDS(file.path(DATA_INTERMEDIATE_PATH, "PREVENTAD_risk_dat_baseline.rds"))
biomarkers    <- readRDS(file.path(DATA_INTERMEDIATE_PATH, "PREVENTAD_biomarkers.rds"))

bm_baseline <- biomarkers %>%
  group_by(CONP_ID) %>%
  slice_min(years, n = 1) %>%
  ungroup() %>%
  select(CONP_ID, ab_ratio, gfap, nfl, ptau181, ptau217)
```

```{r variable-lists}
cogdrisk_vars <- c("Sex", "Education_level", "Hypertension", "BMI_category",
                   "High_cholesterol", "Depression", "Atrial_fibrillation",
                   "Diabetes", "TBI", "Smoking", "Cognitive_engagement",
                   "Social_engagement", "Physical_inactivity")

libra_vars <- c("Hypertension", "Hypercholesterolemia", "Obesity",
                "Diabetes", "Depression", "Smoking", "Cognitive_activity",
                "Physical_activity")
```

# Per-Person Missingness

Note: derived variables (e.g. `Physical_inactivity`) default to non-NA values when
input data are missing, so missingness is assessed from the raw `crs_factors` data.
Exercise is considered complete if `exer_curr_act1_intensity` is non-NA — when
intensity is missing, days and hours are always missing too.

```{r per-person-missingness}
person_missing <- crs_factors %>%
  transmute(
    CONP_ID,
    # --- CogDrisk components ---
    cogd_Sex              = is.na(Sex),
    cogd_Education        = is.na(Education_years),
    cogd_Hypertension     = is.na(Systolic_blood_pressure) & is.na(Diastolic_blood_pressure),
    cogd_BMI              = is.na(Weight) & is.na(Height),
    cogd_High_cholesterol = is.na(total_cholesterol_value),
    cogd_Depression       = is.na(past_depression),
    cogd_AF               = is.na(past_atrial_fibrillation),
    cogd_Diabetes         = is.na(treatment_diabetes) & is.na(hba1c_value),
    cogd_TBI              = is.na(head_injury_hospitalized) & is.na(head_injury_severe),
    cogd_Smoking          = is.na(smoking_present),
    cogd_Cognitive        = is.na(epoch_score_currently),
    cogd_Social           = is.na(social_life_frequency_activities) &
                            is.na(social_life_frequency_visitors) &
                            is.na(social_life_frequency_visits) &
                            is.na(social_life_frequency_phone_calls),
    cogd_Exercise         = is.na(exer_curr_act1_intensity),
    # --- LIBRA2 components ---
    libra_Hypertension    = is.na(Systolic_blood_pressure) & is.na(Diastolic_blood_pressure),
    libra_Hypercholest    = is.na(total_cholesterol_value) & is.na(LDL_value),
    libra_Obesity         = is.na(Weight) & is.na(Height),
    libra_Diabetes        = is.na(treatment_diabetes) & is.na(hba1c_value),
    libra_Depression      = is.na(past_depression),
    libra_Smoking         = is.na(smoking_present),
    libra_Cognitive       = is.na(epoch_score_currently),
    libra_Exercise        = is.na(exer_curr_act1_intensity)
  ) %>%
  mutate(
    cogd_n_missing    = rowSums(across(starts_with("cogd_"))),
    cogd_pct_missing  = round(cogd_n_missing / length(cogdrisk_vars) * 100, 1),
    libra_n_missing   = rowSums(across(starts_with("libra_"))),
    libra_pct_missing = round(libra_n_missing / length(libra_vars) * 100, 1)
  )

cat("=== CogDrisk: missing variables per person ===\n")
print(table(person_missing$cogd_n_missing))

cat("\n=== LIBRA2: missing variables per person ===\n")
print(table(person_missing$libra_n_missing))
```

```{r per-person-viz, fig.width=10, fig.height=4}
p_cogd_miss <- ggplot(person_missing, aes(x = cogd_pct_missing)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "white") +
  labs(title = "CogDrisk: % Variables Missing per Person",
       x = "% Missing", y = "Count") +
  theme_minimal()

p_libra_miss <- ggplot(person_missing, aes(x = libra_pct_missing)) +
  geom_histogram(binwidth = 5, fill = "darkgreen", color = "white") +
  labs(title = "LIBRA2: % Variables Missing per Person",
       x = "% Missing", y = "Count") +
  theme_minimal()

p_cogd_miss + p_libra_miss
```

# Characterization by Missingness Group

Participants are grouped by their CogDrisk component missingness to assess whether
those with more missing data differ systematically in demographics or biomarkers.
Thresholds shown: complete (0%), low (1–25%), high (>25%).

```{r missingness-groups}
# Attach demographics, scores, and biomarkers to person_missing
person_char <- person_missing %>%
  select(CONP_ID, cogd_n_missing, cogd_pct_missing, libra_n_missing, libra_pct_missing) %>%
  left_join(risk_baseline %>% select(CONP_ID, baseline_age, sex, educ), by = "CONP_ID") %>%
  left_join(cogd_original_scored %>% select(CONP_ID, score_cogd), by = "CONP_ID") %>%
  left_join(libra_original_scored %>% select(CONP_ID, score_libra2), by = "CONP_ID") %>%
  left_join(bm_baseline, by = "CONP_ID") %>%
  mutate(
    cogd_miss_group = factor(
      case_when(
        cogd_pct_missing == 0         ~ "Complete (0%)",
        cogd_pct_missing <= 25        ~ "Low (1-25%)",
        TRUE                          ~ "High (>25%)"
      ),
      levels = c("Complete (0%)", "Low (1-25%)", "High (>25%)")
    ),
    gfap_log   = log(gfap),
    nfl_log    = log(nfl),
    ptau181_log = log(ptau181),
    ptau217_log = log(ptau217)
  )
```

# Sensitivity Analysis: Exclusion Thresholds

How many participants are retained and how do score distributions shift at
three candidate exclusion thresholds (complete cases only, ≤25% missing, ≤50% missing).

```{r threshold-counts}
thresholds <- c(0, 25, 50)

thresh_summary <- purrr::map_dfr(thresholds, function(thresh) {
  tibble(
    Threshold        = paste0("<= ", thresh, "%"),
    CogDrisk_n      = sum(person_char$cogd_pct_missing <= thresh),
    CogDrisk_pct    = round(mean(person_char$cogd_pct_missing <= thresh) * 100, 1),
    LIBRA2_n        = sum(person_char$libra_pct_missing <= thresh),
    LIBRA2_pct      = round(mean(person_char$libra_pct_missing <= thresh) * 100, 1)
  )
})

thresh_summary %>%
  gt() %>%
  cols_label(
    Threshold     ~ "Exclusion threshold",
    CogDrisk_n    ~ "N retained",
    CogDrisk_pct  ~ "% retained",
    LIBRA2_n      ~ "N retained",
    LIBRA2_pct    ~ "% retained"
  ) %>%
  tab_spanner(label = "CogDrisk", columns = c(CogDrisk_n, CogDrisk_pct)) %>%
  tab_spanner(label = "LIBRA2",   columns = c(LIBRA2_n, LIBRA2_pct)) %>%
  tab_caption("Sample size at each exclusion threshold")
```

```{r threshold-scores, fig.width=12, fig.height=5}
# Build long data for each threshold x score combination
score_at_threshold <- purrr::map_dfr(thresholds, function(thresh) {
  person_char %>%
    filter(cogd_pct_missing <= thresh) %>%
    select(CONP_ID, score_cogd, score_libra2) %>%
    pivot_longer(c(score_cogd, score_libra2),
                 names_to = "score_type", values_to = "score") %>%
    mutate(
      threshold = paste0("<= ", thresh, "% missing")#,
      # score_type = recode(score_type,
                          # "score_cogd"   = "CogDrisk",
                          # "score_libra2" = "LIBRA2")
    )
})

score_at_threshold <- score_at_threshold %>%
  mutate(threshold = factor(threshold,
                            levels = paste0("<= ", thresholds, "% missing")))

ggplot(score_at_threshold, aes(x = threshold, y = score, fill = threshold)) +
  geom_violin(alpha = 0.7, trim = FALSE) +
  geom_boxplot(width = 0.1, fill = "white", outlier.size = 0.8) +
  facet_wrap(~ score_type, scales = "free_y") +
  labs(title = "Score distributions at each exclusion threshold",
       x = "Threshold applied to CogDrisk % missing",
       y = "Score") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 20, hjust = 1))
```

```{r threshold-biomarkers, fig.width=12, fig.height=8}
# Same idea for biomarkers
bm_at_threshold <- purrr::map_dfr(thresholds, function(thresh) {
  person_char %>%
    filter(cogd_pct_missing <= thresh) %>%
    select(CONP_ID, ab_ratio, gfap_log, nfl_log, ptau181_log, ptau217_log) %>%
    pivot_longer(-CONP_ID, names_to = "biomarker", values_to = "value") %>%
    mutate(
      threshold = paste0("<= ", thresh, "% missing")#,
      # biomarker = recode(biomarker,
      #                    "ab_ratio"    = "Aβ42/40",
      #                    "gfap_log"    = "GFAP (log)",
      #                    "nfl_log"     = "NfL (log)",
      #                    "ptau181_log" = "p-tau181 (log)",
      #                    "ptau217_log" = "p-tau217 (log)")
    )
})

bm_at_threshold <- bm_at_threshold %>%
  mutate(threshold = factor(threshold,
                            levels = paste0("<= ", thresholds, "% missing")))

ggplot(bm_at_threshold, aes(x = threshold, y = value, fill = threshold)) +
  geom_violin(alpha = 0.7, trim = FALSE) +
  geom_boxplot(width = 0.1, fill = "white", outlier.size = 0.8) +
  facet_wrap(~ biomarker, scales = "free_y", ncol = 3) +
  stat_compare_means(method = "kruskal.test",
                     label.x.npc = 0.05,
                     label.y.npc = 0.8,
                     geom = "label",
                     size = 5) +
  labs(title = "Biomarker distributions at each exclusion threshold",
       x = "Threshold", y = NULL) +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 20, hjust = 1))
```

# CogDrisk Analysis

## Missing Data by Variable

```{r cogdrisk-missing-viz, fig.width=10, fig.height=6}
cogd_missing_summary <- person_missing %>%
  select(starts_with("cogd_"), -cogd_n_missing, -cogd_pct_missing) %>%
  summarise(across(everything(), sum)) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "N_Missing") %>%
  mutate(
    Variable = sub("^cogd_", "", Variable),
    Percent_Missing = round(N_Missing / nrow(crs_factors) * 100, 1)
  ) %>%
  arrange(desc(N_Missing))

p1 <- ggplot(cogd_missing_summary, aes(x = reorder(Variable, N_Missing), y = N_Missing)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_text(aes(label = paste0(N_Missing, " (", Percent_Missing, "%)")),
            hjust = -0.1, size = 3) +
  coord_flip() +
  labs(title = "CogDrisk: Missing Data by Variable",
       x = "Variable", y = "Number of Missing Values") +
  theme_minimal() +
  ylim(0, max(cogd_missing_summary$N_Missing) * 1.2)

print(p1)
# ggsave(file.path(EDA_OUTPUT_PATH$figures, "PREVENTAD_CogDrisk_missing.png"), p1)

# Heatmap of missing data patterns
cogd_heatmap <- person_missing %>%
  select(CONP_ID, starts_with("cogd_"), -cogd_n_missing, -cogd_pct_missing) %>%
  mutate(participant_id = row_number()) %>%
  pivot_longer(cols = -c(CONP_ID, participant_id),
               names_to = "Variable", values_to = "Missing") %>%
  mutate(Variable = sub("^cogd_", "", Variable)) %>%
  filter(Missing)

if (nrow(cogd_heatmap) > 0) {
  p2 <- ggplot(cogd_heatmap, aes(x = Variable, y = participant_id)) +
    geom_tile(fill = "red", alpha = 0.7) +
    coord_flip() +
    labs(title = "CogDrisk: Missing Data Pattern Heatmap",
         x = "Variable", y = "Participant") +
    theme_minimal() +
    theme(axis.text.y = element_text(size = 8))
  print(p2)
}

cat("\nCogDrisk Missing Data Summary:\n")
print(cogd_missing_summary)
```

## Original vs Imputed Data

```{r cogdrisk-comparison}
df_cogd <- cogd_original_scored %>%
  select(CONP_ID, original_score = score_cogd) %>%
  #filter(!CONP_ID %in% person_missing_ids$CONP_ID) %>%
  left_join(cogd_scored %>% select(CONP_ID, imputed_score = score_cogd), by = "CONP_ID")

df_cogd_longer <- df_cogd %>%
  pivot_longer(cols = c(original_score, imputed_score),
               names_to = "method", values_to = "score_value")

var_ratio <- sd(df_cogd$imputed_score, na.rm = TRUE) / sd(df_cogd$original_score, na.rm = TRUE)
wilcox_result <- wilcox.test(df_cogd$original_score, df_cogd$imputed_score, paired = FALSE)

cat(sprintf("SD ratio (imputed vs complete): %.3f\n", var_ratio))
cat(sprintf("Wilcoxon test p-value: %.4f\n", wilcox_result$p.value))

plot1 <- ggplot(df_cogd_longer, aes(x = method, y = score_value, fill = method)) +
  geom_violin(alpha = 0.7) +
  geom_boxplot(width = 0.1, fill = "white") +
  stat_compare_means(comparisons = list(c("original_score", "imputed_score")),
                     method = "wilcox.test") +
  labs(subtitle = paste0("SD ratio (imputed vs complete) = ", round(var_ratio, 3)),
       x = "Method", y = "CogDrisk Score") +
  ggtitle("CogDrisk: Complete Cases vs Imputed Data") +
  theme_minimal()

print(plot1)
# ggsave(file.path(EDA_OUTPUT_PATH$figures, "CogD_imputation_comparison.png"), plot1, width = 8, height = 6)
```

# LIBRA2 Analysis

## Missing Data by Variable

```{r libra-missing-viz, fig.width=10, fig.height=6}
libra_missing_summary <- person_missing %>%
  select(starts_with("libra_"), -libra_n_missing, -libra_pct_missing) %>%
  summarise(across(everything(), sum)) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "N_Missing") %>%
  mutate(
    Variable = sub("^libra_", "", Variable),
    Percent_Missing = round(N_Missing / nrow(crs_factors) * 100, 1)
  ) %>%
  arrange(desc(N_Missing))

p3 <- ggplot(libra_missing_summary, aes(x = reorder(Variable, N_Missing), y = N_Missing)) +
  geom_bar(stat = "identity", fill = "darkgreen") +
  geom_text(aes(label = paste0(N_Missing, " (", Percent_Missing, "%)")),
            hjust = -0.1, size = 3) +
  coord_flip() +
  labs(title = "LIBRA2: Missing Data by Variable",
       x = "Variable", y = "Number of Missing Values") +
  theme_minimal() +
  ylim(0, max(libra_missing_summary$N_Missing) * 1.2)

print(p3)

libra_heatmap <- person_missing %>%
  select(CONP_ID, starts_with("libra_"), -libra_n_missing, -libra_pct_missing) %>%
  mutate(participant_id = row_number()) %>%
  pivot_longer(cols = -c(CONP_ID, participant_id),
               names_to = "Variable", values_to = "Missing") %>%
  mutate(Variable = sub("^libra_", "", Variable)) %>%
  filter(Missing)

if (nrow(libra_heatmap) > 0) {
  p4 <- ggplot(libra_heatmap, aes(x = Variable, y = participant_id)) +
    geom_tile(fill = "darkred", alpha = 0.7) +
    coord_flip() +
    labs(title = "LIBRA2: Missing Data Pattern Heatmap",
         x = "Variable", y = "Participant") +
    theme_minimal() +
    theme(axis.text.y = element_text(size = 8))
  print(p4)
}

cat("\nLIBRA2 Missing Data Summary:\n")
print(libra_missing_summary)
```

## Complete Cases

```{r complete-libra}
n_complete_libra <- sum(person_missing$libra_n_missing == 0)

cat(sprintf("Complete cases: %d / %d (%.1f%%)\n",
            n_complete_libra,
            nrow(person_missing),
            n_complete_libra / nrow(person_missing) * 100))
```

## Complete Cases vs Imputed Data

```{r libra-comparison}
df_libra <- libra_original_scored %>%
  select(CONP_ID, original_score = score_libra2) %>%
  #filter(!CONP_ID %in% person_missing_ids$CONP_ID) %>%
  left_join(libra_scored %>% select(CONP_ID, imputed_score = score_libra2), by = "CONP_ID")

df_libra_longer <- df_libra %>%
  pivot_longer(cols = c(original_score, imputed_score),
               names_to = "method", values_to = "score_value")

var_ratio_libra <- sd(df_libra$imputed_score, na.rm = TRUE) / sd(df_libra$original_score, na.rm = TRUE)
wilcox_result_libra <- wilcox.test(df_libra$original_score, df_libra$imputed_score, paired = FALSE)

cat(sprintf("SD ratio (imputed vs complete): %.3f\n", var_ratio_libra))
cat(sprintf("Wilcoxon test p-value: %.4f\n", wilcox_result_libra$p.value))

plot_libra <- ggplot(df_libra_longer, aes(x = method, y = score_value, fill = method)) +
  geom_violin(alpha = 0.7) +
  geom_boxplot(width = 0.1, fill = "white") +
  stat_compare_means(comparisons = list(c("original_score", "imputed_score")),
                     method = "wilcox.test") +
  labs(subtitle = paste0("SD ratio (imputed vs complete) = ", round(var_ratio_libra, 3)),
       x = "Method", y = "LIBRA2 Score") +
  ggtitle("LIBRA2: Complete Cases vs Imputed Data") +
  theme_minimal()

print(plot_libra)
# ggsave(file.path(EDA_OUTPUT_PATH$figures, "LIBRA_imputation_comparison.png"), plot_libra, width = 8, height = 6)
```

# Summary and save results

```{r summary-stats}
cat("=== Summary ===\n")
cat(sprintf("CogDrisk  — Total: %d, Complete cases: %d (%.1f%%)\n",
            nrow(person_missing), n_complete_cogd,
            n_complete_cogd / nrow(person_missing) * 100))
cat(sprintf("LIBRA2    — Total: %d, Complete cases: %d (%.1f%%)\n",
            nrow(person_missing), n_complete_libra,
            n_complete_libra / nrow(person_missing) * 100))

# Save IDs of individuals with >25% missing data
person_missing_ids <- person_missing %>%
  filter(cogd_pct_missing > 25 & libra_pct_missing > 25) %>%
  select(CONP_ID)
saveRDS(person_missing_ids, file.path(DATA_INTERMEDIATE_PATH, "PREVENTAD_missingcrs.rds"))
```
