---
title: "EDA: Univariate Analysis"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
execute:
  warning: false
  message: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warnings = FALSE)
knitr::opts_knit$set(root.dir = here::here())
```

```{r setting}
# Load packages and data
library(broom)
library(car)
library(patchwork)

source('workflow/scripts/config.R')

biomarkers <- readRDS(file.path(DATA_INTERMEDIATE_PATH, 
                                "PREVENTAD_biomarkers.rds"))
predictors <- readRDS(file.path(DATA_CLEANED_PATH, "PREVENTAD_predictors.rds"))
baseline_dat <- readRDS(file.path(DATA_CLEANED_PATH, "PREVENTAD_baseline.rds"))
lastvisit_dat <- readRDS(file.path(DATA_CLEANED_PATH, "PREVENTAD_lastvisit.rds"))
all_dat <- readRDS(file.path(DATA_CLEANED_PATH, "PREVENTAD_longitudinal.rds"))

GRS <- readRDS(file.path(DATA_INTERMEDIATE_PATH, "PREVENTAD_GRS.rds"))
```

## Overview

This analysis examines the properties and basic features of our analysis data: AD blood-based biomarkers (ptau-217, ptau-181, NfL, GFAP, Aβ42/40 ratio, and ptau-217/Aβ42 ratio) and key risk indicators including family history, genetic risk (GRS, APOE), clinical risk scores (CogDrisk, LIBRA2), and demographic factors. 

### APOE

```{r apoe}
apoe_hist <- predictors %>%
  ggplot(aes(x=apoe, fill=apoe)) +
  geom_bar(stat="count", alpha=0.7) + 
  geom_text(
    stat = "count",  
    aes(label = after_stat(count)),  
    vjust = -0.3,  
    hjust = 0.5,   
    size = 3.5     
  ) +
  labs(x = "APOE Status",
       y = "Count") + 
  theme(legend.position="none")
apoe_hist

ggsave(
  file.path(EDA_OUTPUT_PATH$figures, "APOE_dist.png"),
  apoe_hist, width = 14, height = 10, dpi = 300
)
```

### GRS

```{r grs}
grs_plot <- GRS %>%
  ggplot(aes(x = GRS)) +
  geom_density(col="red") +
  labs(x = "GRS",
       y = "Density")
grs_plot

ggsave(
  file.path(EDA_OUTPUT_PATH$figures, "GRS_plot.png"),
  grs_plot, width = 14, height = 10, dpi = 300
)
```

### Family history

```{r fhx}
fhx_hist <- predictors %>%
  ggplot(aes(x=family_history, fill=family_history)) +
  geom_bar(stat="count", alpha=0.7) + 
  geom_text(
    stat = "count",  
    aes(label = after_stat(count)),  
    vjust = -0.3,  
    hjust = 0.5,   
    size = 3.5     
  ) +
  labs(x = "Family History",
       y = "Count") + 
  theme(legend.position="none")
fhx_hist
```

### CogDrisk

```{r cogd}
cogd_plot <- predictors %>%
  ggplot(aes(x = score_cogd)) +
  geom_histogram(aes(y = ..density..), binwidth=2, fill="steelblue", alpha=0.7) +
  geom_density(col="red") +
  labs(x = "CogDrisk Score",
       y = "Density")
cogd_plot
print(summary(predictors$score_cogd))
```

### LIBRA2

```{r libra2}
libra2_plot <- predictors %>%
  ggplot(aes(x = score_libra2)) +
  geom_histogram(aes(y = ..density..), binwidth=2, fill="steelblue", alpha=0.7) +
  geom_density(col="red") +
  labs(x = "LIBRA2 Score",
       y = "Density")
libra2_plot
print(summary(predictors$score_libra2))
```

```{r}
grs_plot <- predictors %>%
  ggplot(aes(x = GRS)) +
  geom_histogram(aes(y = ..density..), bins=20, fill="steelblue", alpha=0.7) +
  geom_density(col="red") +
  labs(title = "Genetic Risk Scores",
       subtitle = "(GRS)",
       x = "GRS",
       y = "Density")

cont_plot <- cogd_plot | libra2_plot | grs_plot
cont_plot <- cont_plot + plot_annotation(title = "PREVENT-AD Continuous Predictors")
cont_plot

# Save plots
# ggsave(file.path(EDA_OUTPUT_PATH$figures, "PREVENTAD_predictors_categorical_hist.png"), cat_hist, width = 12, height = 8, dpi = 300)
# ggsave(file.path(EDA_OUTPUT_PATH$figures, "PREVENTAD_predictors_continuous_hist.png"), cont_plot, width = 12, height = 8, dpi = 300)
```
- Continuous predictors (CogDrisk, LIBRA2, GRS) are not normally distributed.
  
  
# Biomarker distributions 

```{r biomarker-dist}
biomarkers_longer <- all_dat %>%
  select(CONP_ID, Visit_label, all_of(BIOMARKERS)) %>%
  pivot_longer(
    cols = all_of(BIOMARKERS),
    names_to = "biomarker",
    values_to = "measure"
  )

biomarker_plot <- biomarkers_longer %>%
  ggplot(aes(x=measure)) +
  geom_density() +
  facet_wrap(~biomarker, scales="free") +
  labs(title = "Biomarker Distributions",
       x = "Biomarker Level",
       y = "Density")

biomarker_plot

biomarkers_transform_longer <- all_dat %>%
  select(CONP_ID, Visit_label, all_of(BIOMARKERS_TRANS)) %>%
  pivot_longer(
    cols = all_of(BIOMARKERS_TRANS),
    names_to = "biomarker",
    values_to = "measure"
  )

biomarker_transformed_plot <- biomarkers_transform_longer %>%
  ggplot(aes(x=measure)) +
  geom_density() +
  facet_wrap(~biomarker, scales="free")

biomarker_transformed_plot
```

## Biomarker QQ plot

```{r biomarker-qq}
ggplot(biomarkers_longer, aes(sample = measure)) +
    stat_qq(size = 0.8, alpha = 0.6) +
    stat_qq_line(color = "red", linewidth = 1) +
    facet_wrap(~biomarker, scales = "free", ncol = 3) +
    labs(title = "Q-Q Plot",
         x = "Theoretical Quantiles", y = "Sample Quantiles") +
    theme_minimal() +
    theme(
      strip.text = element_text(face = "bold", size = 10),
      strip.background = element_rect(fill = "grey90", color = NA),
      plot.title = element_text(size = 12, face = "bold", hjust = 0.5)
    )

ggplot(biomarkers_transform_longer, aes(sample = measure)) +
    stat_qq(size = 0.8, alpha = 0.6) +
    stat_qq_line(color = "red", linewidth = 1) +
    facet_wrap(~biomarker, scales = "free", ncol = 3) +
    labs(title = "Q-Q Plot (Transformed)",
         x = "Theoretical Quantiles", y = "Sample Quantiles") +
    theme_minimal() +
    theme(
      strip.text = element_text(face = "bold", size = 10),
      strip.background = element_rect(fill = "grey90", color = NA),
      plot.title = element_text(size = 12, face = "bold", hjust = 0.5)
    )
```
- Biomarkers (before and after log transformation) still do not look normally distributed.


## Biomarker outliers

```{r outlier-summary}
# Function to detect outliers using IQR 
detect_outliers_iqr <- function(x) {
  q1 <- quantile(x, 0.25, na.rm = TRUE)
  q3 <- quantile(x, 0.75, na.rm = TRUE)
  iqr <- q3 - q1
  lower_bound <- q1 - 1.5 * iqr
  upper_bound <- q3 + 1.5 * iqr
  return(list(
    is_outlier = x < lower_bound | x > upper_bound,
    lower = lower_bound,
    upper = upper_bound
  ))
}

# Original ----
outlier_summary <- biomarkers_longer %>%
  group_by(biomarker) %>%
  summarise(
    n_total = sum(!is.na(measure)),
    n_outliers = sum(detect_outliers_iqr(measure)$is_outlier, na.rm = TRUE),
    pct_outliers = round(n_outliers / n_total * 100, 2),
    lower_bound = detect_outliers_iqr(measure)$lower,
    upper_bound = detect_outliers_iqr(measure)$upper,
    .groups = "drop"
  )

outlier_summary %>%
  gt() %>%
  tab_header(
    title = "Biomarker Outliers",
    subtitle = "IQR Method: 1.5 × IQR"
  ) %>%
  cols_label(
    biomarker = "Biomarker",
    n_total = "N Total",
    n_outliers = "N Outliers",
    pct_outliers = "% Outliers",
    lower_bound = "Lower Bound",
    upper_bound = "Upper Bound"
  ) %>%
  fmt_number(columns = c(lower_bound, upper_bound), decimals = 2) %>%
  tab_style(
    style = cell_fill(color = "#fff3cd"),
    locations = cells_body(columns = pct_outliers, rows = pct_outliers > 5)
  ) 

# Transformed ----
outlier_summary_transform <- biomarkers_transform_longer %>%
  group_by(biomarker) %>%
  summarise(
    n_total = sum(!is.na(measure)),
    n_outliers = sum(detect_outliers_iqr(measure)$is_outlier, na.rm = TRUE),
    pct_outliers = round(n_outliers / n_total * 100, 2),
    lower_bound = detect_outliers_iqr(measure)$lower,
    upper_bound = detect_outliers_iqr(measure)$upper,
    .groups = "drop"
  )

outlier_summary_transform %>%
  gt() %>%
  tab_header(
    title = "Biomarker Outliers",
    subtitle = "IQR Method: 1.5 × IQR"
  ) %>%
  cols_label(
    biomarker = "Biomarker",
    n_total = "N Total",
    n_outliers = "N Outliers",
    pct_outliers = "% Outliers",
    lower_bound = "Lower Bound",
    upper_bound = "Upper Bound"
  ) %>%
  fmt_number(columns = c(lower_bound, upper_bound), decimals = 2) %>%
  tab_style(
    style = cell_fill(color = "#fff3cd"),
    locations = cells_body(columns = pct_outliers, rows = pct_outliers > 5)
  ) 
```

```{r outlier-by-visit}
# Original ----
baseline_biomarkers <- baseline_dat %>%
  select(CONP_ID, all_of(BIOMARKERS)) %>%
  pivot_longer(cols = all_of(BIOMARKERS), names_to = "biomarker", values_to = "measure") %>%
  mutate(visit_type = "Baseline")

last_biomarkers <- lastvisit_dat %>%
  select(CONP_ID, all_of(BIOMARKERS)) %>%
  pivot_longer(cols = all_of(BIOMARKERS), names_to = "biomarker", values_to = "measure") %>%
  mutate(visit_type = "LastVisit")

visit_biomarkers <- bind_rows(baseline_biomarkers, last_biomarkers)

outlier_by_visit <- visit_biomarkers %>%
  group_by(visit_type, biomarker) %>%
  summarise(
    n_total = sum(!is.na(measure)),
    n_outliers = sum(detect_outliers_iqr(measure)$is_outlier, na.rm = TRUE),
    pct_outliers = round(n_outliers / n_total * 100, 2),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = visit_type,
    values_from = c(n_total, n_outliers, pct_outliers),
    names_glue = "{visit_type}_{.value}"
  ) %>%
  select(biomarker,
         Baseline_n_total, Baseline_n_outliers, Baseline_pct_outliers,
         LastVisit_n_total, LastVisit_n_outliers, LastVisit_pct_outliers)

outlier_by_visit %>%
  gt() %>%
  tab_header(
    title = "Outlier Counts: Baseline vs Last Visit"
  ) %>%
  cols_label(
    biomarker = "Biomarker",
    Baseline_n_total = "N",
    Baseline_n_outliers = "N Outliers",
    Baseline_pct_outliers = "%",
    LastVisit_n_total = "N",
    LastVisit_n_outliers = "N Outliers",
    LastVisit_pct_outliers = "%"
  ) %>%
  tab_spanner(
    label = "Baseline",
    columns = c(Baseline_n_total, Baseline_n_outliers, Baseline_pct_outliers)
  ) %>%
  tab_spanner(
    label = "Last Visit",
    columns = c(LastVisit_n_total, LastVisit_n_outliers, LastVisit_pct_outliers)
  ) %>%
  fmt_number(columns = c(Baseline_pct_outliers, LastVisit_pct_outliers), decimals = 1) %>%
  tab_style(
    style = cell_fill(color = "#fff3cd"),
    locations = cells_body(
      columns = c(Baseline_pct_outliers, LastVisit_pct_outliers),
      rows = Baseline_pct_outliers > 5 | LastVisit_pct_outliers > 5
    )
  )

# Transformed ----
baseline_biomarkers_transformed <- baseline_dat %>%
  select(CONP_ID, all_of(BIOMARKERS_TRANS)) %>%
  pivot_longer(cols = all_of(BIOMARKERS_TRANS), names_to = "biomarker", values_to = "measure") %>%
  mutate(visit_type = "Baseline")

last_biomarkers_transformed <- lastvisit_dat %>%
  select(CONP_ID, all_of(BIOMARKERS_TRANS)) %>%
  pivot_longer(cols = all_of(BIOMARKERS_TRANS), names_to = "biomarker", values_to = "measure") %>%
  mutate(visit_type = "LastVisit")

visit_biomarkers_transformed <- bind_rows(baseline_biomarkers_transformed, last_biomarkers_transformed)

outlier_by_visit_transformed <- visit_biomarkers_transformed %>%
  group_by(visit_type, biomarker) %>%
  summarise(
    n_total = sum(!is.na(measure)),
    n_outliers = sum(detect_outliers_iqr(measure)$is_outlier, na.rm = TRUE),
    pct_outliers = round(n_outliers / n_total * 100, 2),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = visit_type,
    values_from = c(n_total, n_outliers, pct_outliers),
    names_glue = "{visit_type}_{.value}"
  ) %>%
  select(biomarker,
         Baseline_n_total, Baseline_n_outliers, Baseline_pct_outliers,
         LastVisit_n_total, LastVisit_n_outliers, LastVisit_pct_outliers)

outlier_by_visit_transformed %>%
  gt() %>%
  tab_header(
    title = "Outlier Counts: Baseline vs Last Visit",
    subtitle = "After log transforming"
  ) %>%
  cols_label(
    biomarker = "Biomarker",
    Baseline_n_total = "N",
    Baseline_n_outliers = "N Outliers",
    Baseline_pct_outliers = "%",
    LastVisit_n_total = "N",
    LastVisit_n_outliers = "N Outliers",
    LastVisit_pct_outliers = "%"
  ) %>%
  tab_spanner(
    label = "Baseline",
    columns = c(Baseline_n_total, Baseline_n_outliers, Baseline_pct_outliers)
  ) %>%
  tab_spanner(
    label = "Last Visit",
    columns = c(LastVisit_n_total, LastVisit_n_outliers, LastVisit_pct_outliers)
  ) %>%
  fmt_number(columns = c(Baseline_pct_outliers, LastVisit_pct_outliers), decimals = 1) %>%
  tab_style(
    style = cell_fill(color = "#fff3cd"),
    locations = cells_body(
      columns = c(Baseline_pct_outliers, LastVisit_pct_outliers),
      rows = Baseline_pct_outliers > 5 | LastVisit_pct_outliers > 5
    )
  )
```

```{r outlier-individuals}
# Identify individuals with outliers after transformation
outlier_flags <- all_dat %>%
  select(CONP_ID, Visit_label, all_of(BIOMARKERS_TRANS)) %>%
  mutate(across(all_of(BIOMARKERS_TRANS),
                ~detect_outliers_iqr(.)$is_outlier,
                .names = "{.col}_outlier"))

# Count outliers per individual
individual_outlier_counts <- outlier_flags %>%
  group_by(CONP_ID) %>%
  summarise(
    n_visits = n(),
    n_outlier_observations = sum(c_across(ends_with("_outlier")), na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(n_outlier_observations > 0)

cat("Individuals with at least one outlier biomarker value:", nrow(individual_outlier_counts), "\n")
cat("Total individuals:", n_distinct(all_dat$CONP_ID), "\n")
cat("Percentage:", round(nrow(individual_outlier_counts) / n_distinct(all_dat$CONP_ID) * 100, 1), "%\n")
```
