---
title: "aim1_crosssectional"
format: html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warnings = FALSE)
knitr::opts_knit$set(root.dir = here::here())
```

# Load packages and data
```{r setting}
library(broom)
library(car)

source('workflow/scripts/config.R')

biomarkers <- readRDS(file.path(DATA_CLEANED_PATH$filtered, 
                                "PREVENTAD_biomarkers_filtered.rds"))

predictors_dat <- readRDS(file.path(DATA_INTERMEDIATE_PATH, "PREVENTAD_predictors.rds"))
```


# Get last measured biomarkers
```{r}
ptau217_last <- biomarkers %>%
  group_by(CONP_ID) %>%
  filter(!is.na(ptau217)) %>%
  select(CONP_ID, Date_taken, age, ptau217) %>%
  slice_tail() 

other_biomarkers_last <- biomarkers %>%
  group_by(CONP_ID) %>%
  select(CONP_ID, all_of(BIOMARKERS)) %>%
  select(-ptau217) %>%
  filter_at(vars(-CONP_ID), all_vars(!is.na(.))) %>%
  slice_tail() 

all_dat <- predictors_dat %>% 
  left_join((ptau217_last %>%
               select(CONP_ID, ptau217)), by="CONP_ID") %>%
  left_join(other_biomarkers_last, by="CONP_ID") %>%
  column_to_rownames(var="CONP_ID")
```




# Defining models
```{r}
summary_list <- c()
model_comparisons <- c()

# Loop through biomarkers
for (biomarker in BIOMARKERS){
  cat("\n=== Processing:", biomarker, "===\n")
  
  data_temp <- all_dat %>%
    mutate(log_biomarker = log(all_dat[[biomarker]]))
  
  # Model 1: demographics only
  model1 <- lm(log_biomarker ~ Candidate_Age + Sex + BMI, data=data_temp)
  
  # Model 2: + genetics
  model2 <- lm(log_biomarker ~ Candidate_Age + Sex + Education_years + apoe_category + GRS, data=data_temp)

  # Model 3: + family history
  model3 <- lm(log_biomarker~ Candidate_Age + Sex + Education_years + apoe_category + GRS + FDR_ext_AD, data=data_temp)

  # Model 4: + LIBRA2
  model4 <- lm(log_biomarker ~ Candidate_Age + Sex + Education_years + apoe_category + GRS + FDR_ext_AD + score_libra2, data=data_temp)

  # Model 5: genetics + family history + CogDrisk
  model5 <- lm(log_biomarker ~ apoe_category + GRS + FDR_ext_AD + score_cogd, data=data_temp)

  # Model 6: genetics + family history + CogDrisk + LIBRA2
  model6 <- lm(log_biomarker ~ apoe_category + GRS + FDR_ext_AD + score_cogd + score_libra2, data=data_temp)

  # Model 7: demographics + genetics + family history + CogDrisk + LIBRA2
  model7 <- lm(log_biomarker ~ Candidate_Age + Sex + Education_years + apoe_category + GRS + FDR_ext_AD + score_cogd + score_libra2, data=data_temp)

  
  results_temp <- bind_rows(
    tidy(model1) %>% mutate(model = "model1"),
    tidy(model2) %>% mutate(model = "model2"),
    tidy(model3) %>% mutate(model = "model3"),
    tidy(model4) %>% mutate(model = "model4"),
    tidy(model5) %>% mutate(model = "model5"),
    tidy(model6) %>% mutate(model = "model6")
  ) %>% mutate(biomarker = biomarker)
  
  summary_list[[biomarker]] <- results_temp
  print(results_temp)
  
  # comparison_temp <- tibble(
  #   biomarker = biomarker,
  #   model = c("model1", "model2", "model3", "model4", "model5", "model6"),
  #   R2 = c(summary(model1)$r.squared, summary(model2)$r.squared, 
  #          summary(model3)$r.squared, summary(model4)$r.squared, 
  #          summary(model5)$r.squared, summary(model6)$r.squared),
  #   Adj_R2 = c(summary(model1)$adj.r.squared, summary(model2)$adj.r.squared,
  #              summary(model3)$adj.r.squared, summary(model4)$adj.r.squared,
  #              summary(model5)$adj.r.squared, summary(model6)$adj.r.squared),
  #   AIC = c(AIC(model1), AIC(model2), AIC(model3), AIC(model4), AIC(model5), AIC(model6)),
  #   BIC = c(BIC(model1), BIC(model2), BIC(model3), BIC(model4), BIC(model5), BIC(model6))
  # ) %>%
  #   mutate(
  #     delta_R2 = R2 - lag(R2),
  #     delta_AIC = AIC - lag(AIC)
  #   )
  # 
  # model_comparisons[[biomarker]] <- comparison_temp
  # 
  # print(comparison_temp)
}
```

## Model 1: demographics
- ptau231: adj r2 = 0.028, AIC = 593.17, BIC = 611.33

## Model 2: demographics + genetics
- AB ratio: adj r2 = 0.06, AIC = -51.45, BIC = -22.40
- GFAP: adj r2 = 0.24, AIC = 310.49, BIC = 339.54
- NFL: adj r2 = 0.21, AIC = 242.67, BIC = 271.71
- ptau181: adj r2 = 0.09, AIC = 431.84, BIC = 460.89

## Model 3: demographics + genetics + family history
- ptau217: adj r2 = 0.1654, AIC = 392.56, BIC = 433.04


## Stratified analyses by APOE category
### e3/e3
```{r}
data_e3e3 <- all_dat %>% filter(apoe_category == "e3/e3")

summary_list <- c()
model_comparisons <- c()

# Loop through biomarkers
for (biomarker in BIOMARKERS){
  cat("\n=== Processing:", biomarker, "===\n")
  
  data_temp <- data_e3e3 %>%
    mutate(log_biomarker = log(data_e3e3[[biomarker]]))
  
  # Model 1: demographics only
  model1 <- lm(log_biomarker ~ Candidate_Age + Sex + Education_years, data=data_temp)
  # summary_list[[i]][1] <- summary(model1)
  
  # Model 2: + genetics
  model2 <- lm(log_biomarker ~ Candidate_Age + Sex + Education_years + GRS, data=data_temp)
  # summary_list[[i]][2] <- summary(model2)
  
  # Model 3: + family history
  model3 <- lm(log_biomarker~ Candidate_Age + Sex + Education_years + GRS + FDR_ext_AD, data=data_temp)
  # summary_list[[i]][3] <- summary(model3)
  
  # Model 4: + LIBRA2
  model4 <- lm(log_biomarker ~ Candidate_Age + Sex + Education_years + GRS + FDR_ext_AD + score_libra2, data=data_temp)
  # summary_list[[i]][4] <- summary(model4)
  
  # Model 5: genetics + family history + CogDrisk
  model5 <- lm(log_biomarker ~ GRS + FDR_ext_AD + score_cogd, data=data_temp)
  # summary_list[[i]][5] <- summary(model5)
  
  # Model 6: genetics + family history + CogDrisk + LIBRA2
  model6 <- lm(log_biomarker ~ GRS + FDR_ext_AD + score_cogd + score_libra2, data=data_temp)
  # summary_list[[i]][6] <- summary(model6)
  
  # Model 7: demographics + genetics + family history + CogDrisk + LIBRA2
  model7 <- lm(log_biomarker ~ Candidate_Age + Sex + Education_years + GRS + FDR_ext_AD + score_cogd + score_libra2, data=data_temp)

  
  results_temp <- bind_rows(
    tidy(model1) %>% mutate(model = "model1"),
    tidy(model2) %>% mutate(model = "model2"),
    tidy(model3) %>% mutate(model = "model3"),
    tidy(model4) %>% mutate(model = "model4"),
    tidy(model5) %>% mutate(model = "model5"),
    tidy(model6) %>% mutate(model = "model6")
  ) %>% mutate(biomarker = biomarker)
  
  summary_list[[biomarker]] <- results_temp
  
  comparison_temp <- tibble(
    biomarker = biomarker,
    model = c("model1", "model2", "model3", "model4", "model5", "model6"),
    coeffs = c(summary(model1)$)
    R2 = c(summary(model1)$r.squared, summary(model2)$r.squared, 
           summary(model3)$r.squared, summary(model4)$r.squared, 
           summary(model5)$r.squared, summary(model6)$r.squared),
    Adj_R2 = c(summary(model1)$adj.r.squared, summary(model2)$adj.r.squared,
               summary(model3)$adj.r.squared, summary(model4)$adj.r.squared,
               summary(model5)$adj.r.squared, summary(model6)$adj.r.squared),
    AIC = c(AIC(model1), AIC(model2), AIC(model3), AIC(model4), AIC(model5), AIC(model6)),
    BIC = c(BIC(model1), BIC(model2), BIC(model3), BIC(model4), BIC(model5), BIC(model6))
  ) %>%
    mutate(
      delta_R2 = R2 - lag(R2),
      delta_AIC = AIC - lag(AIC)
    )
  
  model_comparisons[[biomarker]] <- comparison_temp
  
  print(comparison_temp)
}

```
For e2+:
Model 5: genetics + family history + CogD
- ptau 181: adj r2 = 0.11, AIC = 40.96, BIC = 50.12

For e4+: 
Model 4: demographics + genetics + family history + LIBRA2
- NFL: adj r2 = 0.22, AIC = 92.92, BIC = 119.56
- ptau 217: adj r2 = 0.12, AIC = 162.87, BIC = 190.05
- ptau 231: adj r2 = 0.06, AIC = 223.15, BIC = 249.78


# Other models:
## CogDrisk (residualized version) 
```{r}
# Extract the non-demographic components of CogDRISK
cogd_nondem <- residuals(
  lm(score_cogd ~ Candidate_Age + Sex + Education_years, data = all_dat)
)

summary_list <- c()
model_comparisons <- c()

# Loop through biomarkers
for (biomarker in BIOMARKERS){
  cat("\n=== Processing:", biomarker, "===\n")
  
  data_temp <- all_dat %>%
    mutate(log_biomarker = log(all_dat[[biomarker]]))
  
  model5.1 <- lm(log_biomarker ~ Candidate_Age + Sex + Education_years + apoe_category + GRS + 
                   FDR_ext_AD + cogd_nondem, data=data_temp)
  model6.1 <- lm(log_biomarker ~ Candidate_Age + Sex + Education_years + apoe_category + GRS + FDR_ext_AD +
                   cogd_nondem + score_libra2, data=data_temp)

  
  results_temp <- bind_rows(
    tidy(model5.1) %>% mutate(model = "model5.1"),
    tidy(model6.1) %>% mutate(model = "model6.1")
  ) %>% mutate(biomarker = biomarker)
  
  summary_list[[biomarker]] <- results_temp
  
  comparison_temp <- tibble(
    biomarker = biomarker,
    model = c("model5.1", "model6.1"),
    R2 = c(summary(model5.1)$r.squared, summary(model6.1)$r.squared),
    Adj_R2 = c(summary(model5.1)$adj.r.squared, summary(model6.1)$adj.r.squared),
    AIC = c(AIC(model5.1), AIC(model6.1)),
    BIC = c(BIC(model5.1), BIC(model6.1))
  ) %>%
    mutate(
      delta_R2 = R2 - lag(R2),
      delta_AIC = AIC - lag(AIC)
    )
  
  model_comparisons[[biomarker]] <- comparison_temp
  
  print(comparison_temp)
}
```


## Interactions
```{r}
summary_list <- c()
model_comparisons <- c()

# Loop through biomarkers
for (biomarker in BIOMARKERS){
  cat("\n=== Processing:", biomarker, "===\n")
  
  data_temp <- all_dat %>%
    mutate(log_biomarker = log(all_dat[[biomarker]]))
  
  # Models 8: LIBRA2/CogD effect varying by APOE status
  model8.1 <- lm(log_biomarker ~ Candidate_Age + Sex + Education_years + apoe_category * score_libra2 + 
         GRS + FDR_ext_AD, data=data_temp)
  model8.2 <- lm(log_biomarker ~ Candidate_Age + Sex + Education_years + GRS + FDR_ext_AD + 
                   apoe_category * cogd_nondem + score_libra2, data=data_temp)

  # Model 9: LIBRA2 effect varies by age
  model9 <- lm(log_biomarker ~ Candidate_Age * score_libra2 + Sex + Education_years + apoe_category + 
                 GRS + FDR_ext_AD, data=data_temp)

  # Models 10: Clinical risk matters more in high genetic risk
  model10.1 <- lm(log_biomarker ~ score_libra2 * GRS + Candidate_Age + Sex + Education_years + 
                    apoe_category + FDR_ext_AD, data=data_temp)
  model10.2 <- lm(log_biomarker ~ score_cogd * GRS + apoe_category + FDR_ext_AD, data=data_temp)
  model10.3 <- lm(log_biomarker ~ cogd_nondem * GRS + Candidate_Age + Sex + Education_years + 
                    apoe_category + FDR_ext_AD, data=data_temp)

  
  results_temp <- bind_rows(
    tidy(model8.1) %>% mutate(model = "model8.1"),
    tidy(model8.2) %>% mutate(model = "model8.2"),
    tidy(model9) %>% mutate(model = "model9"),
    tidy(model10.1) %>% mutate(model = "model10.1"),
    tidy(model10.2) %>% mutate(model = "model10.2"),
    tidy(model10.3) %>% mutate(model = "model10.3")
  ) %>% mutate(biomarker = biomarker)
  
  summary_list[[biomarker]] <- results_temp
  
  comparison_temp <- tibble(
    biomarker = biomarker,
    model = c("model8.1", "model8.2", "model9", "model10.1", "model10.2", "model10.3"),
    R2 = c(summary(model8.1)$r.squared, summary(model8.2)$r.squared, 
           summary(model9)$r.squared, summary(model10.1)$r.squared, 
           summary(model10.2)$r.squared, summary(model10.3)$r.squared),
    Adj_R2 = c(summary(model8.1)$adj.r.squared, summary(model8.2)$adj.r.squared,
               summary(model9)$adj.r.squared, summary(model10.1)$adj.r.squared,
               summary(model10.2)$adj.r.squared, summary(model10.3)$adj.r.squared),
    AIC = c(AIC(model8.1), AIC(model8.2), AIC(model9), AIC(model10.1), AIC(model10.2), AIC(model10.3)),
    BIC = c(BIC(model8.1), BIC(model8.2), BIC(model9), BIC(model10.1), BIC(model10.2), BIC(model10.3)),
    beta = c(summary(model8.1)$Estimate, summary(model8.2)$Estimate,
               summary(model9)$Estimate, summary(model10.1)$Estimate,
               summary(model10.2)$Estimate, summary(model10.3)$Estimate)
  ) %>%
    mutate(
      delta_R2 = R2 - lag(R2),
      delta_AIC = AIC - lag(AIC)
    )
  
  model_comparisons[[biomarker]] <- comparison_temp
  
  print(comparison_temp)
}
```

Models 8: LIBRA2 (1)/CogD(2) effect varying by APOE status
Model 9: LIBRA2 effect varying by age
Models 10: LIBRA2 (1)/CogD(2) effect varying by GRS