---
title: "Untitled"
output: html_document
---

## Setup:
```{r}
source('workflow/scripts/config.R')
source('workflow/scripts/exploratory_data_analysis/load_PREVENT-AD_data.R')

cogdrisk_social <- lifestyle_dataset_cogdrisk %>%
  select(CONP_ID, starts_with("social_")) %>%
  rename(activities = social_life_frequency_activities,
         visitors = social_life_frequency_visitors,
         visits = social_life_frequency_visits,
         calls = social_life_frequency_phone_calls)
```


## Correlation statistics
```{r}
cor(cogdrisk_social[, c("activities", "visitors", "visits", "calls")], use="complete.obs", method="spearman")
```


## Barplots of social engagement questions
```{r}
df <- cogdrisk_social %>%
  pivot_longer(
    cols = activities:calls,
    names_to = "question",
    values_to = "answer"
  )

ggplot(df, aes(x = factor(answer))) +
  geom_bar(fill = "steelblue") +
  facet_wrap(
    ~ question, 
    scales="free_x",
    labeller=labeller(question=c("activities" = "Social activity frequency",
                                 "visitors" = "Home visitors",
                                 "visits" = "Visiting friends",
                                 "calls" = "Phone calls"))) +
            
  labs(
    title = "PREVENT-AD Social Engagement Response Frequency by Question",
    x = "Answer",
    y = "Frequency"
  ) +
  scale_x_discrete()
  
```


## Testing recoding strategies  
### Recoding by aggregates of questions
```{r}
cogdrisk_social_recoded_1 <- cogdrisk_social %>%
  mutate(
    # V1: Average of all social engagement questions
    social_engagement_v1 = rowMeans(select(., activities:calls)),
    social_engagement_v1_binary = as.factor(if_else(
      social_engagement_v1 > mean(cogdrisk_social_recoded$social_engagement_v1, na.rm=TRUE), 
      0, 1)
      ),
    
    # V2: Average of visit questions separately, then average across all questions
    holder = rowMeans(select(., visitors:visits))) %>%
  mutate(
    social_engagement_v2 = rowMeans(subset(., select=c(activities, calls, holder))),
    social_engagement_v2_binary = as.factor(if_else(
      social_engagement_v2 > mean(cogdrisk_social_recoded$social_engagement_v2, na.rm=TRUE), 
      0, 1)
      )
    ) 
```


### Recoding by question (weekly social engagement as threshold)
```{r}
# 0 = not lonely
# 1 = lonely
cogdrisk_social_recoded_2 <- cogdrisk_social %>%
  mutate(
    activities_binary = if_else(activities >= 5, 0, 1),
    visitors_binary = if_else(visitors >= 4, 0, 1),
    visits_binary = if_else(visits >= 4, 0, 1),
    calls_binary = if_else(calls >= 4, 0, 1)
  ) %>%
  mutate(social_engagement = if_else(rowSums(select(., contains("_binary"))) >= 2, 1, 0))

df <- cogdrisk_social_recoded_2 %>%
  pivot_longer(
    cols = activities_binary:calls_binary,
    names_to = "question",
    values_to = "answer"
  )

ggplot(df, aes(x = factor(answer))) +
  geom_bar(fill = "steelblue") +
  facet_wrap(
    ~ question, 
    scales="free_x",
    labeller=labeller(question=c("activities_binary" = "Social activity frequency",
                                 "visitors_binary" = "Home visitors",
                                 "visits_binary" = "Visiting friends",
                                 "calls_binary" = "Phone calls"))) +
            
  labs(
    title = "PREVENT-AD Social Engagement Response Frequency by Question",
    x = "Answer",
    y = "Frequency"
  ) +
  scale_x_discrete()

```

## Visualizing distributions of recoded variable 
### Version 1: Average of all social engagement questions
```{r}
mean_value_v1 = mean(cogdrisk_social_recoded_1$social_engagement_v1, na.rm=TRUE)
#median_value_v1 = median(cogdrisk_social_recoded$social_engagement_v1, na.rm=TRUE)

ggplot(cogdrisk_social_recoded_1, aes(x = social_engagement_v1)) +
  geom_histogram(fill="steelblue", binwidth=0.25) +
  geom_vline(xintercept = mean_value_v1,
             linetype = "dashed",
             color = "red") +
  annotate("text",
           x = mean_value_v1 - 0.1,
           y = 15, 
           label = paste("Mean = ", round(mean_value_v1, 2)),
           color = "red",
           angle = 90) +
  labs(title = "Histogram of social_engagement_v1",
       subtitle = "social_engagement_v1 = average of all social engagement questions",
       x = "social_engagement_v1 score",
       y = "Count")

ggplot(cogdrisk_social_recoded_1, aes(x = social_engagement_v1_binary)) +
  geom_bar(fill="steelblue") +
  geom_text(stat = "count",
            aes(label=..count..),
            vjust = 2) +
  labs(title = "Barplot of social_engagement_v1_binary",
       subtitle = "Using mean of social_engagement_v1 as threshold score",
       x = "Label",
       y = "Count") %>%
  scale_x_discrete(labels=c("0" = "Not lonely",
                            "1" = "Lonely"))
```

### Version 2: Average of visit questions separately, then average across all questions
```{r}
mean_value_v2 = mean(cogdrisk_social_recoded_1$social_engagement_v2, na.rm=TRUE)
mean_value_holder = mean(cogdrisk_social_recoded_1$holder, na.rm=TRUE)

ggplot(cogdrisk_social_recoded_1, aes(x = holder)) +
  geom_histogram(fill="steelblue", binwidth=0.5) +
  geom_vline(xintercept = mean_value_holder,
             linetype = "dashed",
             color = "red") +
  annotate("text",
           x = mean_value_holder - 0.1,
           y = 15, 
           label = paste("Mean = ", round(mean_value_holder, 2)),
           color = "red",
           angle = 90) +
  labs(title = "Histogram of average of visit questions",
       #subtitle = "social_engagement_v2 = average of visit questions separately, then average across all questions",
       x = "Average of visit questions",
       y = "Count")

ggplot(cogdrisk_social_recoded_1, aes(x = social_engagement_v2)) +
  geom_histogram(fill="steelblue", binwidth=0.25) +
  geom_vline(xintercept = mean_value_v2,
             linetype = "dashed",
             color = "red") +
  annotate("text",
           x = mean_value_v2 - 0.1,
           y = 15, 
           label = paste("Mean = ", round(mean_value_v2, 2)),
           color = "red",
           angle = 90) +
  labs(title = "Histogram of social_engagement_v2",
       subtitle = "social_engagement_v2 = average of visit questions separately, then average across all questions",
       x = "social_engagement_v2 score",
       y = "Count")

ggplot(cogdrisk_social_recoded_1, aes(x = social_engagement_v2_binary)) +
  geom_bar(fill="steelblue") +
  geom_text(stat = "count",
            aes(label=..count..),
            vjust = 2) +
  labs(title = "Barplot of social_engagement_v2_binary",
       subtitle = "Using mean of social_engagement_v2 as threshold score",
       x = "Label",
       y = "Count") %>%
  scale_x_discrete(labels=c("0" = "Not lonely",
                            "1" = "Lonely"))
```

