---
title: "Aim 1: Longitudinal Risk Burden Analysis"
format:
  html:
    embed-resources: true
    self-contained: true
    toc: true
    toc-depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = here::here())
```

```{r load-packages}
library(broom.mixed)
library(lme4)
library(patchwork)

source('workflow/scripts/config.R')
```

# Load and prepare data

```{r load-data}
# Load longitudinal data
all_dat <- readRDS(file.path(DATA_CLEANED_PATH, "PREVENTAD_longitudinal.rds"))
baseline_dat <- readRDS(file.path(DATA_CLEANED_PATH, "PREVENTAD_baseline.rds"))

# Check data structure
cat("Data dimensions:", nrow(all_dat), "rows,", ncol(all_dat), "columns\n")
cat("Number of unique participants:", length(unique(all_dat$CONP_ID)), "\n")
cat("Time range:", min(all_dat$years, na.rm = TRUE), "to",
    max(all_dat$years, na.rm = TRUE), "years\n")

# Grand-mean center the years variable
# This centers time around the average follow-up time across all observations
mean_years <- mean(all_dat$years, na.rm = TRUE)
all_dat <- all_dat %>%
  mutate(years_centered = years - mean_years)

cat("\nGrand-mean centering applied:\n")
cat("  Original years mean:", round(mean_years, 2), "years\n")
cat("  Centered years mean:", round(mean(all_dat$years_centered, na.rm = TRUE), 4), "years\n")
```

## Create risk burden variables

```{r create-risk-burden}
# Calculate cutoffs based on baseline data only (to avoid data leakage)
baseline_dat <- all_dat %>%
  group_by(CONP_ID) %>%
  slice_min(years, n = 1) %>%
  ungroup()

# Calculate cutoffs from baseline data
cogdrisk_cutoff_1sd <- mean(baseline_dat$score_cogd, na.rm = TRUE) +
                       sd(baseline_dat$score_cogd, na.rm = TRUE)
libra2_cutoff_1sd <- mean(baseline_dat$score_libra2, na.rm = TRUE) +
                     sd(baseline_dat$score_libra2, na.rm = TRUE)

cat("CogDrisk 1 SD cutoff:", round(cogdrisk_cutoff_1sd, 2), "\n")
cat("LIBRA2 1 SD cutoff:", round(libra2_cutoff_1sd, 2), "\n\n")

# Apply risk burden classification to all longitudinal data
risk_dat_long <- all_dat %>%
  mutate(
    # GRS burden - top decile
    grs_quartile = ntile(GRS, 4),
    grs_burden = if_else(grs_quartile == 4, 1, 0),

    # Family history burden - 2+ FDR
    fhx_burden = case_when(
      str_detect(family_history, "2\\+") ~ 1,
      TRUE ~ 0
    ),
    # fhx_burden = if_else(family_history == "1 FDR", 0, 1),
    
    # APOE burden - e4+
    apoe_burden = if_else(apoe == "e4+", 1, 0),

    # CogDrisk burden - 1 SD above mean (using baseline cutoff)
    cogdrisk_binary_1sd = ifelse(score_cogd > cogdrisk_cutoff_1sd, 1, 0),

    # LIBRA2 burden - 1 SD above mean (using baseline cutoff)
    libra2_binary_1sd = ifelse(score_libra2 > libra2_cutoff_1sd, 1, 0),

    # Composite risk scores
    risk_score_cogd = rowSums(across(c(fhx_burden, apoe_burden,
                                       cogdrisk_binary_1sd, grs_burden)),
                             na.rm = TRUE),
    risk_score_libra2 = rowSums(across(c(fhx_burden, apoe_burden,
                                         libra2_binary_1sd, grs_burden)),
                               na.rm = TRUE)
  )

# Print risk burden distribution
cat("Risk burden distribution:\n")
cat("\nCogDrisk risk score:\n")
table(risk_dat_long$risk_score_cogd) %>% print()
cat("\nLIBRA2 risk score:\n")
table(risk_dat_long$risk_score_libra2) %>% print()
```

## Risk burden distribution

```{r risk-burden-distribution}
#| fig-width: 12
#| fig-height: 5

p1 <- ggplot(risk_dat_long, aes(x = factor(risk_score_cogd))) +
  geom_bar(fill = "steelblue", alpha = 0.7) +
  labs(title = "CogDrisk Composite Risk Score Distribution",
       x = "Risk Score (0-4)", y = "Count") +
  theme_minimal()

p2 <- ggplot(risk_dat_long, aes(x = factor(risk_score_libra2))) +
  geom_bar(fill = "coral", alpha = 0.7) +
  labs(title = "LIBRA2 Composite Risk Score Distribution",
       x = "Risk Score (0-4)", y = "Count") +
  theme_minimal()

p1 + p2
```

# Dose-response models

## Model 1: Main effects (risk score as continuous)

**Interpretation:** These models test whether composite risk score (0-4 scale) predicts biomarker levels in a dose-response manner. The beta coefficient represents the change in biomarker level for each 1-unit increase in risk score (e.g., going from 2 to 3 risk factors). CogDrisk models are adjusted for age, sex, and BMI only. LIBRA2 models are adjusted for age, sex, BMI, and education.

```{r dose-main-models}
models_dose_main <- tribble(
  ~model, ~predictors,
  "CogDrisk_dose", "years_centered + risk_score_cogd + (1 | CONP_ID)",
  "LIBRA2_dose", "years_centered + baseline_age + sex + educ + risk_score_libra2 + (1 | CONP_ID)"
)

lmm_dose_main <- expand_grid(
  outcome = BIOMARKERS_TRANS,
  models_dose_main
) %>%
  mutate(
    eq = glue("{outcome} ~ {predictors}"),
    res = map(eq, ~lmer(as.formula(.), data = risk_dat_long, REML = FALSE)),
    dataf = map(res, tidy),
    mod = map(res, glance)
  )
```

## Model 2: Time interactions (risk score × time)

**Interpretation:** These models test whether the rate of biomarker change over time differs by risk score level and other covariates. The interaction term (years_centered:risk_score) shows how much faster (or slower) biomarkers change for each additional risk factor. For LIBRA2, we also include interactions between time and covariates (baseline_age, sex, educ) to account for differential rates of change by demographic factors. A significant positive interaction suggests that higher-risk individuals (or older individuals, etc.) show accelerated biomarker accumulation over time.

```{r dose-interact-models}
models_dose_interact <- tribble(
  ~model, ~predictors,
  "CogDrisk_interact", "years_centered * risk_score_cogd + (1 | CONP_ID)",
  "LIBRA2_interact", "years_centered * (risk_score_libra2 + baseline_age + sex + educ) + (1 | CONP_ID)"
)

lmm_dose_interact <- expand_grid(
  outcome = BIOMARKERS_TRANS,
  models_dose_interact
) %>%
  mutate(
    eq = glue("{outcome} ~ {predictors}"),
    res = map(eq, ~lmer(as.formula(.), data = risk_dat_long, REML = FALSE)),
    dataf = map(res, tidy),
    mod = map(res, glance)
  )
```

## Extract and format results

```{r format-results}
# Combine all dose-response results
dose_response_results <- bind_rows(
  lmm_dose_main %>%
    select(model, outcome, dataf) %>%
    mutate(model_type = "Main Effect"),
  lmm_dose_interact %>%
    select(model, outcome, dataf) %>%
    mutate(model_type = "Time Interaction")
) %>%
  unnest(dataf) %>%
  filter(str_detect(term, "^risk_score|^years_centered:risk_score")) %>%
  mutate(
    # Confidence intervals
    conf.low = estimate - (std.error * 1.96),
    conf.high = estimate + (std.error * 1.96),

    # Significance
    sig = ifelse(p.value < 0.05, TRUE, FALSE),
    stars = case_when(
      p.value >= 0.05 ~ "",
      p.value >= 0.01 & p.value < 0.05 ~ "*",
      p.value >= 0.001 & p.value < 0.01 ~ "**",
      p.value < 0.001 ~ "***"
    ),

    # Exponentiated values for log-transformed outcomes
    exp_est = case_when(
      str_detect(outcome, "_log$") ~ exp(estimate),
      TRUE ~ NA_real_
    ),
    exp_lci = case_when(
      str_detect(outcome, "_log$") ~ exp(conf.low),
      TRUE ~ NA_real_
    ),
    exp_uci = case_when(
      str_detect(outcome, "_log$") ~ exp(conf.high),
      TRUE ~ NA_real_
    ),

    # Percentage change per unit for all biomarkers
    pct_change_per_unit = case_when(
      str_detect(outcome, "_log$") ~ (exp_est - 1) * 100,
      TRUE ~ NA_real_
    ),

    # Risk score type
    risk_type = if_else(str_detect(model, "CogDrisk"), "CogDrisk", "LIBRA2")
  ) %>%
  mutate(across(c(estimate, std.error, statistic, p.value,
                  conf.low, conf.high, exp_est, exp_lci, exp_uci, pct_change_per_unit),
                ~round(., digits = 3)))
```

# Results Tables

## CogDrisk: Main effects

```{r table-cogdrisk-main}
dose_response_cogdrisk_main <- dose_response_results %>%
  filter(model_type == "Main Effect") %>%
  filter(risk_type == "CogDrisk") %>%
  filter(!str_detect(term, "years:")) %>%
  select(outcome, estimate, conf.low, conf.high,
         pct_change_per_unit, p.value, stars, sig) %>%
  mutate(
    outcome = case_when(
      outcome == "ab_ratio" ~ "Aβ42/40",
      outcome == "gfap_log" ~ "GFAP",
      outcome == "nfl_log" ~ "NfL",
      outcome == "ptau181_log" ~ "ptau-181",
      outcome == "ptau217_log" ~ "ptau-217",
      outcome == "ptau217_ab42_ratio_log" ~ "ptau-217/Aβ42 ratio",
      TRUE ~ outcome
    )
  ) %>%
  arrange(outcome)

dose_response_cogdrisk_main %>%
  gt() %>%
  tab_header(
    title = md("**CogDrisk: Change per Unit Increase in Risk Score**"),
    subtitle = "Main effects model"
  ) %>%
  fmt_number(
    columns = c(estimate, conf.low, conf.high, pct_change_per_unit, p.value),
    decimals = 3
  ) %>%
  cols_label(
    outcome = "Biomarker",
    estimate = "Beta",
    conf.low = "Lower CI",
    conf.high = "Upper CI",
    pct_change_per_unit = "% Change*",
    p.value = "p-value",
    stars = ""
  ) %>%
  cols_hide(columns = c(sig)) %>%
  tab_footnote(
    footnote = "% Change per unit risk score. For log-transformed biomarkers, calculated as (exp(beta)-1)*100.",
    locations = cells_column_labels(columns = pct_change_per_unit)
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(
      columns = everything(),
      rows = sig == TRUE
    )
  )
```

## LIBRA2: Main effects

```{r table-libra2-main}
dose_response_libra2_main <- dose_response_results %>%
  filter(model_type == "Main Effect") %>%
  filter(risk_type == "LIBRA2") %>%
  filter(!str_detect(term, "years:")) %>%
  select(outcome, estimate, conf.low, conf.high,
         pct_change_per_unit, p.value, stars, sig) %>%
  mutate(
    outcome = case_when(
      outcome == "ab_ratio" ~ "Aβ42/40",
      outcome == "gfap_log" ~ "GFAP",
      outcome == "nfl_log" ~ "NfL",
      outcome == "ptau181_log" ~ "ptau-181",
      outcome == "ptau217_log" ~ "ptau-217",
      outcome == "ptau217_ab42_ratio_log" ~ "ptau-217/Aβ42 ratio",
      TRUE ~ outcome
    )
  ) %>%
  arrange(outcome)

dose_response_libra2_main %>%
  gt() %>%
  tab_header(
    title = md("**LIBRA2: Change per Unit Increase in Risk Score**"),
    subtitle = "Main effects model"
  ) %>%
  fmt_number(
    columns = c(estimate, conf.low, conf.high, pct_change_per_unit, p.value),
    decimals = 3
  ) %>%
  cols_label(
    outcome = "Biomarker",
    estimate = "Beta",
    conf.low = "Lower CI",
    conf.high = "Upper CI",
    pct_change_per_unit = "% Change*",
    p.value = "p-value",
    stars = ""
  ) %>%
  cols_hide(columns = c(sig)) %>%
  tab_footnote(
    footnote = "% Change per unit risk score. For log-transformed biomarkers, calculated as (exp(beta)-1)*100.",
    locations = cells_column_labels(columns = pct_change_per_unit)
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(
      columns = everything(),
      rows = sig == TRUE
    )
  )
```

## CogDrisk: Time interaction

```{r table-cogdrisk-interact}
dose_response_interact_cogdrisk <- dose_response_results %>%
  filter(model_type == "Time Interaction") %>%
  filter(risk_type == "CogDrisk") %>%
  filter(str_detect(term, "years_centered:risk_score")) %>%
  select(outcome, estimate, conf.low, conf.high, p.value, stars, sig) %>%
  mutate(
    outcome = case_when(
      outcome == "ab_ratio" ~ "Aβ42/40",
      outcome == "gfap_log" ~ "GFAP",
      outcome == "nfl_log" ~ "NfL",
      outcome == "ptau181_log" ~ "ptau-181",
      outcome == "ptau217_log" ~ "ptau-217",
      outcome == "ptau217_ab42_ratio_log" ~ "ptau-217/Aβ42 ratio",
      TRUE ~ outcome
    )
  ) %>%
  arrange(outcome)

dose_response_interact_cogdrisk %>%
  gt() %>%
  tab_header(
    title = md("**CogDrisk: Time × Risk Score Interaction**"),
    subtitle = "Change in slope per unit increase in risk score"
  ) %>%
  fmt_number(
    columns = c(estimate, conf.low, conf.high, p.value),
    decimals = 3
  ) %>%
  cols_label(
    outcome = "Biomarker",
    estimate = "Beta",
    conf.low = "Lower CI",
    conf.high = "Upper CI",
    p.value = "p-value",
    stars = ""
  ) %>%
  cols_hide(columns = c(sig)) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(
      columns = everything(),
      rows = sig == TRUE
    )
  )
```

## LIBRA2: Time interaction

```{r table-libra2-interact}
dose_response_interact_libra2 <- dose_response_results %>%
  filter(model_type == "Time Interaction") %>%
  filter(risk_type == "LIBRA2") %>%
  filter(str_detect(term, "years_centered:risk_score")) %>%
  select(outcome, estimate, conf.low, conf.high, p.value, stars, sig) %>%
  mutate(
    outcome = case_when(
      outcome == "ab_ratio" ~ "Aβ42/40",
      outcome == "gfap_log" ~ "GFAP",
      outcome == "nfl_log" ~ "NfL",
      outcome == "ptau181_log" ~ "ptau-181",
      outcome == "ptau217_log" ~ "ptau-217",
      outcome == "ptau217_ab42_ratio_log" ~ "ptau-217/Aβ42 ratio",
      TRUE ~ outcome
    )
  ) %>%
  arrange(outcome)

dose_response_interact_libra2 %>%
  gt() %>%
  tab_header(
    title = md("**LIBRA2: Time × Risk Score Interaction**"),
    subtitle = "Change in slope per unit increase in risk score"
  ) %>%
  fmt_number(
    columns = c(estimate, conf.low, conf.high, p.value),
    decimals = 3
  ) %>%
  cols_label(
    outcome = "Biomarker",
    estimate = "Beta",
    conf.low = "Lower CI",
    conf.high = "Upper CI",
    p.value = "p-value",
    stars = ""
  ) %>%
  cols_hide(columns = c(sig)) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(
      columns = everything(),
      rows = sig == TRUE
    )
  )
```

# Dose-response plots

## CogDrisk: Main effect

```{r plot-cogdrisk-main}
#| fig-width: 10
#| fig-height: 7

plot_data_cogdrisk_main <- dose_response_results %>%
  filter(model_type == "Main Effect") %>%
  filter(risk_type == "CogDrisk") %>%
  filter(!str_detect(term, "years:")) %>%
  mutate(
    sig = factor(sig, levels = c(TRUE, FALSE)),
    outcome_label = case_when(
      outcome == "ab_ratio" ~ "Aβ42/40",
      outcome == "gfap_log" ~ "GFAP",
      outcome == "nfl_log" ~ "NfL",
      outcome == "ptau181_log" ~ "ptau-181",
      outcome == "ptau217_log" ~ "ptau-217",
      outcome == "ptau217_ab42_ratio_log" ~ "ptau-217/Aβ42",
      TRUE ~ outcome
    )
  )

p_cogd_main <- ggplot(plot_data_cogdrisk_main,
       aes(x = estimate, y = fct_reorder(outcome_label, estimate),
           xmin = conf.low, xmax = conf.high, alpha = sig)) +
  geom_point(size = 4, color = "steelblue") +
  geom_errorbarh(linewidth = 1, height = 0.3, color = "steelblue") +
  geom_vline(xintercept = 0, linetype = 2, color = "gray40") +
  scale_alpha_manual(
    values = c(`FALSE` = 0.3, `TRUE` = 1),
    labels = c(`FALSE` = "Not Significant", `TRUE` = "Significant"),
    name = "Significance"
  ) +
  labs(
    x = "Change in Biomarker per Unit Increase in Risk Score (95% CI)",
    y = NULL,
    title = "CogDrisk Dose-Response: Main Effects",
    subtitle = "Change per additional risk factor"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 15)
  )

print(p_cogd_main)

ggsave(file.path(AIM1_OUTPUT_PATH$figures, "PREVENTAD_CogDrisk_dose_main.png"),
       p_cogd_main, width = 10, height = 7, dpi = 300)
```

## LIBRA2: Main effect

```{r plot-libra2-main}
#| fig-width: 10
#| fig-height: 7

plot_data_libra2_main <- dose_response_results %>%
  filter(model_type == "Main Effect") %>%
  filter(risk_type == "LIBRA2") %>%
  filter(!str_detect(term, "years:")) %>%
  mutate(
    sig = factor(sig, levels = c(TRUE, FALSE)),
    outcome_label = case_when(
      outcome == "ab_ratio" ~ "Aβ42/40",
      outcome == "gfap_log" ~ "GFAP",
      outcome == "nfl_log" ~ "NfL",
      outcome == "ptau181_log" ~ "ptau-181",
      outcome == "ptau217_log" ~ "ptau-217",
      outcome == "ptau217_ab42_ratio_log" ~ "ptau-217/Aβ42",
      TRUE ~ outcome
    )
  )

p_libra_main <- ggplot(plot_data_libra2_main,
       aes(x = estimate, y = fct_reorder(outcome_label, estimate),
           xmin = conf.low, xmax = conf.high, alpha = sig)) +
  geom_point(size = 4, color = "coral") +
  geom_errorbarh(linewidth = 1, height = 0.3, color = "coral") +
  geom_vline(xintercept = 0, linetype = 2, color = "gray40") +
  scale_alpha_manual(
    values = c(`FALSE` = 0.3, `TRUE` = 1),
    labels = c(`FALSE` = "Not Significant", `TRUE` = "Significant"),
    name = "Significance"
  ) +
  labs(
    x = "Change in Biomarker per Unit Increase in Risk Score (95% CI)",
    y = NULL,
    title = "LIBRA2 Dose-Response: Main Effects",
    subtitle = "Change per additional risk factor"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 15)
  )

print(p_libra_main)

ggsave(file.path(AIM1_OUTPUT_PATH$figures, "PREVENTAD_LIBRA2_dose_main.png"),
       p_libra_main, width = 10, height = 7, dpi = 300)
```

## CogDrisk: Time interaction

```{r plot-cogdrisk-interact}
#| fig-width: 10
#| fig-height: 7

plot_data_cogdrisk_interact <- dose_response_results %>%
  filter(model_type == "Time Interaction") %>%
  filter(risk_type == "CogDrisk") %>%
  filter(str_detect(term, "years_centered:risk_score")) %>%
  mutate(
    sig = factor(sig, levels = c(TRUE, FALSE)),
    outcome_label = case_when(
      outcome == "ab_ratio" ~ "Aβ42/40",
      outcome == "gfap_log" ~ "GFAP",
      outcome == "nfl_log" ~ "NfL",
      outcome == "ptau181_log" ~ "ptau-181",
      outcome == "ptau217_log" ~ "ptau-217",
      outcome == "ptau217_ab42_ratio_log" ~ "ptau-217/Aβ42",
      TRUE ~ outcome
    )
  )

p_cogd_interact <- ggplot(plot_data_cogdrisk_interact,
       aes(x = estimate, y = fct_reorder(outcome_label, estimate),
           xmin = conf.low, xmax = conf.high, alpha = sig)) +
  geom_point(size = 4, color = "steelblue") +
  geom_errorbarh(linewidth = 1, height = 0.3, color = "steelblue") +
  geom_vline(xintercept = 0, linetype = 2, color = "gray40") +
  scale_alpha_manual(
    values = c(`FALSE` = 0.3, `TRUE` = 1),
    labels = c(`FALSE` = "Not Significant", `TRUE` = "Significant"),
    name = "Significance"
  ) +
  labs(
    x = "Years × Risk Score Interaction (95% CI)",
    y = NULL,
    title = "CogDrisk Dose-Response: Time Interaction",
    subtitle = "Additional change in slope per risk factor"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 15)
  )

print(p_cogd_interact)

ggsave(file.path(AIM1_OUTPUT_PATH$figures, "PREVENTAD_CogDrisk_dose_interact.png"),
       p_cogd_interact, width = 10, height = 7, dpi = 300)
```

## LIBRA2: Time interaction

```{r plot-libra2-interact}
#| fig-width: 10
#| fig-height: 7

plot_data_libra2_interact <- dose_response_results %>%
  filter(model_type == "Time Interaction") %>%
  filter(risk_type == "LIBRA2") %>%
  filter(str_detect(term, "years_centered:risk_score")) %>%
  mutate(
    sig = factor(sig, levels = c(TRUE, FALSE)),
    outcome_label = case_when(
      outcome == "ab_ratio" ~ "Aβ42/40",
      outcome == "gfap_log" ~ "GFAP",
      outcome == "nfl_log" ~ "NfL",
      outcome == "ptau181_log" ~ "ptau-181",
      outcome == "ptau217_log" ~ "ptau-217",
      outcome == "ptau217_ab42_ratio_log" ~ "ptau-217/Aβ42",
      TRUE ~ outcome
    )
  )

p_libra_interact <- ggplot(plot_data_libra2_interact,
       aes(x = estimate, y = fct_reorder(outcome_label, estimate),
           xmin = conf.low, xmax = conf.high, alpha = sig)) +
  geom_point(size = 4, color = "coral") +
  geom_errorbarh(linewidth = 1, height = 0.3, color = "coral") +
  geom_vline(xintercept = 0, linetype = 2, color = "gray40") +
  scale_alpha_manual(
    values = c(`FALSE` = 0.3, `TRUE` = 1),
    labels = c(`FALSE` = "Not Significant", `TRUE` = "Significant"),
    name = "Significance"
  ) +
  labs(
    x = "Years × Risk Score Interaction (95% CI)",
    y = NULL,
    title = "LIBRA2 Dose-Response: Time Interaction",
    subtitle = "Additional change in slope per risk factor"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 15)
  )

print(p_libra_interact)

ggsave(file.path(AIM1_OUTPUT_PATH$figures, "PREVENTAD_LIBRA2_dose_interact.png"),
       p_libra_interact, width = 10, height = 7, dpi = 300)
```

# Trajectory plots

**Methodology: Generating Predicted Trajectories**

The predicted biomarker trajectories are generated from the linear mixed-effects models (LMMs) fitted above using the following approach:

**1. Model Specification:**
- **Base model**: `biomarker ~ years_centered × risk_score + (1 | participant_ID)`
- The time × risk score interaction allows trajectories to have different slopes for each risk level
- Random intercepts account for between-participant variability in baseline biomarker levels

**2. Prediction Grid Construction:**
- For each biomarker, a prediction grid is created spanning:
  - **Time range**: From baseline (0 years) to the maximum observed follow-up for that biomarker
  - **Risk scores**: All possible composite risk levels (0, 1, 2, 3, 4)
  - **Time resolution**: Predictions at 0.5-year intervals for smooth trajectory lines
- Time values are grand-mean centered before prediction (where 0 represents ~2.6 years of follow-up), then transformed back to original scale for plotting

**3. Population-Level Predictions:**
- Predictions use `predict(..., re.form = NA)` to obtain **marginal** (population-average) predictions
- This marginalizes over random effects, showing the expected trajectory for an "average" participant at each risk level
- Individual-level predictions would include random intercepts, but would obscure population-level patterns

**4. Biomarker-Specific Time Ranges:**
- ptau-217: predictions extend to ~11 years (maximum observed follow-up)
- Other biomarkers (ptau-181, GFAP, NfL, Aβ42/40, ptau-217/Aβ42): predictions extend to ~4.6 years
- This prevents extrapolation beyond observed data ranges

**5. Covariate Values (LIBRA2 only):**
- LIBRA2 models include demographic covariates with time interactions
- Predictions are generated for a reference profile:
  - Baseline age: sample mean
  - Sex: Female
  - Education: sample mean years
- Trajectories for other demographic profiles would differ due to covariate × time interactions

## Predicted trajectories: CogDrisk

```{r trajectories-cogdrisk}
#| fig-width: 14
#| fig-height: 10

# Calculate maximum follow-up time per biomarker
biomarker_max_years <- risk_dat_long %>%
  pivot_longer(
    cols = all_of(BIOMARKERS_TRANS),
    names_to = "outcome",
    values_to = "value"
  ) %>%
  filter(!is.na(value)) %>%
  group_by(outcome) %>%
  summarise(max_years = max(years, na.rm = TRUE))

# Generate predictions from interaction model
cogdrisk_interact_models <- lmm_dose_interact %>%
  filter(str_detect(model, "CogDrisk"))

# Create biomarker-specific prediction data and generate predictions
cogdrisk_predictions <- cogdrisk_interact_models %>%
  left_join(biomarker_max_years, by = "outcome") %>%
  mutate(
    predictions = map2(res, max_years, ~{
      # Create prediction grid specific to this biomarker's time range
      pred_data <- expand_grid(
        years_centered = seq(-mean_years, .y - mean_years, by = 0.5),
        risk_score_cogd = 0:4
      ) %>%
        mutate(years = years_centered + mean_years)

      # Generate predictions
      pred_data$predicted <- predict(.x, newdata = pred_data, re.form = NA)
      pred_data
    })
  ) %>%
  select(outcome, predictions) %>%
  unnest(predictions)

# Prepare observed data
risk_dat_long_plot <- risk_dat_long %>%
  pivot_longer(
    cols = all_of(BIOMARKERS_TRANS),
    names_to = "outcome",
    values_to = "observed"
  )

# Plot trajectories
p_cogd_traj <- ggplot() +
  geom_point(data = risk_dat_long_plot,
             aes(x = years, y = observed, color = factor(risk_score_cogd)),
             alpha = 0.2, size = 0.8) +
  geom_line(data = cogdrisk_predictions,
            aes(x = years, y = predicted,
                color = factor(risk_score_cogd),
                group = risk_score_cogd),
            linewidth = 1.2) +
  scale_color_viridis_d(option = "plasma", begin = 0.2, end = 0.9) +
  facet_wrap(~outcome, scales = "free", ncol = 3) +
  labs(
    x = "Years from Baseline",
    y = "Biomarker Level",
    color = "Risk Score",
    title = "CogDrisk: Predicted Biomarker Trajectories by Risk Score",
    caption = "Note: X-axis scales differ by biomarker to reflect actual data availability"
  ) +
  theme_bw(base_size = 12) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold"),
    strip.background = element_rect(fill = "grey90")
  )

print(p_cogd_traj)

ggsave(file.path(AIM1_OUTPUT_PATH$figures, "PREVENTAD_CogDrisk_trajectories.png"),
       p_cogd_traj, width = 14, height = 10, dpi = 300)
```

**CogDrisk Results Summary:**

Key observations:

- Vertical spacing between trajectories at baseline show that individuals with higher composite risk scores already show elevated biomarker levels at study entry
- Higher-risk individuals show steeper slopes compared to lower-risk individuals for ptau181, ptau217, and AB ratio (which would indicate accelerated pathology progression)



## Predicted trajectories: LIBRA2

```{r trajectories-libra2}
#| fig-width: 14
#| fig-height: 10

# Generate predictions from interaction model
libra2_interact_models <- lmm_dose_interact %>%
  filter(str_detect(model, "LIBRA2"))

# Create biomarker-specific prediction data and generate predictions
libra2_predictions <- libra2_interact_models %>%
  left_join(biomarker_max_years, by = "outcome") %>%
  mutate(
    predictions = map2(res, max_years, ~{
      # Create prediction grid specific to this biomarker's time range
      pred_data <- expand_grid(
        years_centered = seq(-mean_years, .y - mean_years, by = 0.5),
        risk_score_libra2 = 0:4,
        baseline_age = mean(risk_dat_long$baseline_age, na.rm = TRUE),
        sex = "Female",
        educ = mean(risk_dat_long$educ, na.rm = TRUE)
      ) %>%
        mutate(years = years_centered + mean_years)

      # Generate predictions
      pred_data$predicted <- predict(.x, newdata = pred_data, re.form = NA)
      pred_data
    })
  ) %>%
  select(outcome, predictions) %>%
  unnest(predictions)

# Plot trajectories
p_libra_traj <- ggplot() +
  geom_point(data = risk_dat_long_plot,
             aes(x = years, y = observed, color = factor(risk_score_libra2)),
             alpha = 0.2, size = 0.8) +
  geom_line(data = libra2_predictions,
            aes(x = years, y = predicted,
                color = factor(risk_score_libra2),
                group = risk_score_libra2),
            linewidth = 1.2) +
  scale_color_viridis_d(option = "plasma", begin = 0.2, end = 0.9) +
  facet_wrap(~outcome, scales = "free", ncol = 3) +
  labs(
    x = "Years from Baseline",
    y = "Biomarker Level",
    color = "Risk Score",
    title = "LIBRA2: Predicted Biomarker Trajectories by Risk Score",
    caption = "Note: X-axis scales differ by biomarker to reflect actual data availability"
  ) +
  theme_bw(base_size = 12) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold"),
    strip.background = element_rect(fill = "grey90")
  )

print(p_libra_traj)

ggsave(file.path(AIM1_OUTPUT_PATH$figures, "PREVENTAD_LIBRA2_trajectories.png"),
       p_libra_traj, width = 14, height = 10, dpi = 300)
```

## Focused Plot: CogDrisk ptau Trajectories

```{r ptau-cogdrisk-focused-plot}
#| fig-width: 12
#| fig-height: 8

# Filter data for ptau217 and ptau181
ptau217_obs <- risk_dat_long_plot %>%
  filter(outcome == "ptau217_log")

ptau181_obs <- risk_dat_long_plot %>%
  filter(outcome == "ptau181_log")

ptau217_pred <- cogdrisk_predictions %>%
  filter(outcome == "ptau217_log")

ptau181_pred <- cogdrisk_predictions %>%
  filter(outcome == "ptau181_log")

# Create separate plots for each biomarker with appropriate x-axis limits
p_ptau217 <- ggplot() +
  geom_point(data = ptau217_obs,
             aes(x = years, y = observed, color = factor(risk_score_cogd)),
             alpha = 0.3, size = 1.5) +
  geom_line(data = ptau217_pred,
            aes(x = years, y = predicted,
                color = factor(risk_score_cogd),
                group = risk_score_cogd),
            linewidth = 1.5) +
  scale_color_viridis_d(option = "plasma", begin = 0.2, end = 0.9,
                        name = "Risk Score") +
  scale_x_continuous(limits = c(0, 12), breaks = seq(0, 12, 2)) +
  labs(
    x = "Years from Baseline",
    y = "ptau-217 (log-transformed)",
    title = "ptau-217 (11 years follow-up)"
  ) +
  theme_bw(base_size = 13) +
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold", size = 12)
  )

p_ptau181 <- ggplot() +
  geom_point(data = ptau181_obs,
             aes(x = years, y = observed, color = factor(risk_score_cogd)),
             alpha = 0.3, size = 1.5) +
  geom_line(data = ptau181_pred,
            aes(x = years, y = predicted,
                color = factor(risk_score_cogd),
                group = risk_score_cogd),
            linewidth = 1.5) +
  scale_color_viridis_d(option = "plasma", begin = 0.2, end = 0.9,
                        name = "# of Risk Indicators") +
  scale_x_continuous(limits = c(0, 5), breaks = seq(0, 5, 1)) +
  labs(
    x = "Years from Baseline",
    y = "ptau-181 (log-transformed)",
    title = "ptau-181"
  ) +
  theme_bw(base_size = 13) +
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold", size = 12)
  )

# Create annotation text with cohort and risk indicator information
n_total <- length(unique(risk_dat_long$CONP_ID))
n_obs <- nrow(risk_dat_long)

annotation_text <- paste0(
  "Cohort: PREVENT-AD. N = ", n_total, "; Observations = ", n_obs, "\n",
  "Risk indicators: FHx: 2+ first-degree relatives with dementia; APOE: e4+ carrier; ",
  "GRS: Genetic risk score > 1 SD above mean; CogDrisk: Modifiable risk score > 1 SD above mean\n",
  "Composite risk score: 0-4 scale (sum of binary indicators)"
)

# Combine plots using patchwork
p_ptau_cogd <- p_ptau217 / p_ptau181 +
  plot_annotation(
    title = "CogDrisk: Phosphorylated Tau Trajectories by Risk Score",
    caption = annotation_text,
    theme = theme(
      plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
      plot.caption = element_text(size = 9, hjust = 0, lineheight = 1.2, margin = margin(t = 10))
    )
  )

print(p_ptau_cogd)
print(p_ptau181)

# Save plot
ggsave(file.path(AIM1_OUTPUT_PATH$figures, "PREVENTAD_ptau_CogDrisk_trajectories.png"),
       p_ptau_cogd, width = 12, height = 9, dpi = 300)
ggsave(file.path(AIM1_OUTPUT_PATH$figures, "PREVENTAD_ptau181_trajectories.png"),
       p_ptau181, width = 12, height = 9, dpi = 300)
```

## Focused Plot: CogDrisk + LIBRA2 ptau Trajectories (combined)

```{r ptau-combined-focused-plot}
#| fig-width: 14
#| fig-height: 10

# LIBRA2 observed data — risk_score_libra2 is already in risk_dat_long_plot
ptau217_obs_libra2 <- risk_dat_long_plot %>%
  filter(outcome == "ptau217_log")

ptau181_obs_libra2 <- risk_dat_long_plot %>%
  filter(outcome == "ptau181_log")

ptau217_pred_libra2 <- libra2_predictions %>%
  filter(outcome == "ptau217_log")

ptau181_pred_libra2 <- libra2_predictions %>%
  filter(outcome == "ptau181_log")

# Shared colour scale
viridis_scale <- scale_color_viridis_d(option = "plasma", begin = 0.2, end = 0.9,
                                       name = "Risk Score")

# ptau-217: CogDrisk
p_ptau217_cogd <- ggplot() +
  geom_point(data = ptau217_obs,
             aes(x = years, y = observed, color = factor(risk_score_cogd)),
             alpha = 0.3, size = 1.2) +
  geom_line(data = ptau217_pred,
            aes(x = years, y = predicted,
                color = factor(risk_score_cogd), group = risk_score_cogd),
            linewidth = 1.5) +
  viridis_scale +
  scale_x_continuous(limits = c(0, 12), breaks = seq(0, 12, 2)) +
  labs(x = "Years from Baseline", y = "ptau-217 (log)",
       title = "ptau-217 — CogDrisk") +
  theme_bw(base_size = 12) +
  theme(plot.title = element_text(face = "bold", size = 12),
        legend.position = "none")

# ptau-217: LIBRA2
p_ptau217_libra2 <- ggplot() +
  geom_point(data = ptau217_obs_libra2,
             aes(x = years, y = observed, color = factor(risk_score_libra2)),
             alpha = 0.3, size = 1.2) +
  geom_line(data = ptau217_pred_libra2,
            aes(x = years, y = predicted,
                color = factor(risk_score_libra2), group = risk_score_libra2),
            linewidth = 1.5) +
  viridis_scale +
  scale_x_continuous(limits = c(0, 12), breaks = seq(0, 12, 2)) +
  labs(x = "Years from Baseline", y = NULL,
       title = "ptau-217 — LIBRA2") +
  theme_bw(base_size = 12) +
  theme(plot.title = element_text(face = "bold", size = 12),
        legend.position = "none")

# ptau-181: CogDrisk
p_ptau181_cogd <- ggplot() +
  geom_point(data = ptau181_obs,
             aes(x = years, y = observed, color = factor(risk_score_cogd)),
             alpha = 0.3, size = 1.2) +
  geom_line(data = ptau181_pred,
            aes(x = years, y = predicted,
                color = factor(risk_score_cogd), group = risk_score_cogd),
            linewidth = 1.5) +
  viridis_scale +
  scale_x_continuous(limits = c(0, 5), breaks = seq(0, 5, 1)) +
  labs(x = "Years from Baseline", y = "ptau-181 (log)",
       title = "ptau-181 — CogDrisk") +
  theme_bw(base_size = 12) +
  theme(plot.title = element_text(face = "bold", size = 12),
        legend.position = "bottom")

# ptau-181: LIBRA2
p_ptau181_libra2 <- ggplot() +
  geom_point(data = ptau181_obs_libra2,
             aes(x = years, y = observed, color = factor(risk_score_libra2)),
             alpha = 0.3, size = 1.2) +
  geom_line(data = ptau181_pred_libra2,
            aes(x = years, y = predicted,
                color = factor(risk_score_libra2), group = risk_score_libra2),
            linewidth = 1.5) +
  viridis_scale +
  scale_x_continuous(limits = c(0, 5), breaks = seq(0, 5, 1)) +
  labs(x = "Years from Baseline", y = NULL,
       title = "ptau-181 — LIBRA2") +
  theme_bw(base_size = 12) +
  theme(plot.title = element_text(face = "bold", size = 12),
        legend.position = "bottom")

# 2×2 layout: rows = biomarker, cols = score type
p_ptau_combined <- (p_ptau217_cogd | p_ptau217_libra2) /
                   (p_ptau181_cogd  | p_ptau181_libra2) +
  plot_annotation(
    title = "Phosphorylated Tau Trajectories by Composite Risk Score",
    caption = paste0(
      "Cohort: PREVENT-AD. N = ", length(unique(risk_dat_long$CONP_ID)),
      "; Observations = ", nrow(risk_dat_long), "\n",
      "Risk indicators: FHx ≥2 FDR; APOE e4+; GRS top quartile; ",
      "CogDrisk/LIBRA2 > 1 SD above mean. Composite score: 0–4 scale.\n",
      "LIBRA2 predictions at mean age, female sex, mean education."
    ),
    theme = theme(
      plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
      plot.caption = element_text(size = 9, hjust = 0, lineheight = 1.2,
                                  margin = margin(t = 10))
    )
  )

print(p_ptau_combined)

ggsave(file.path(AIM1_OUTPUT_PATH$figures, "PREVENTAD_ptau217_combined_trajectory.png"),
       p_ptau_combined, width = 14, height = 10, dpi = 300)
```

## Summary

1. Higher-risk individuals should show elevated biomarker levels at baseline (vertical separation of lines)
2. Lines diverge over time for ptau181 and 217, with higher-risk individuals showing steeper slopes


# Save results

```{r save-results}
# Save main results table
write_csv(dose_response_results,
          file.path(AIM1_OUTPUT_PATH$tables, "longitudinal_doseresponse_results.csv"))

# Save model fit statistics
model_fit <- bind_rows(
  lmm_dose_main %>%
    select(model, outcome, mod) %>%
    unnest(mod) %>%
    mutate(model_type = "Main Effect"),
  lmm_dose_interact %>%
    select(model, outcome, mod) %>%
    unnest(mod) %>%
    mutate(model_type = "Time Interaction")
) %>%
  select(model_type, model, outcome, AIC, BIC, logLik)

write_csv(model_fit,
          file.path(AIM1_OUTPUT_PATH$tables, "longitudinal_doseresponse_model_fit.csv"))

# Save interaction models and risk data for use in sensitivity analysis
saveRDS(lmm_dose_interact,
        file.path(DATA_INTERMEDIATE_PATH, "PREVENTAD_lmm_dose_interact.rds"))

# Save baseline risk scores and covariates per participant
risk_dat_baseline <- risk_dat_long %>%
  group_by(CONP_ID) %>%
  slice_min(years, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(CONP_ID, risk_score_cogd, risk_score_libra2, baseline_age, sex, educ)

saveRDS(risk_dat_baseline,
        file.path(DATA_INTERMEDIATE_PATH, "PREVENTAD_risk_dat_baseline.rds"))

cat("\nResults saved to:", AIM1_OUTPUT_PATH$tables, "\n")
cat("Figures saved to:", AIM1_OUTPUT_PATH$figures, "\n")
cat("Models saved to:", DATA_INTERMEDIATE_PATH, "\n")
```
