---
title: "Risk Burden Cutoff"
format:
  html:
    embed-resources: true
    self-contained: true
    toc: true
    toc-depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warnings = FALSE)
knitr::opts_knit$set(root.dir = here::here())
```

```{r setting}
library(broom)
library(car)
library(patchwork)

source('workflow/scripts/config.R')

lastvisit_dat <- readRDS(file.path(DATA_CLEANED_PATH, "PREVENTAD_lastvisit.rds"))
```

## Overview

Construction of a composite risk burden score to assess whether cumulative AD risk accumulation is associated with blood-based biomarker levels will involve the binarization of various risk components: APOE ε4 carrier status, family history (≥2 first-degree relatives with dementia), genetic risk score, and elevated modifiable lifestyle risk (CogDrisk or LIBRA2 score above a defined threshold). Since there is no defined CogDrisk or LIBRA2 threshold for the PREVENT-AD cohort, this analysis will test three binarization thresholds for the CRS component (top tertile, top quartile, and 1 SD above mean) and will compare them to assess sensitivity of the chosen cutoff.
  
## Risk burden model 

```{r risk-burden-model-data}
risk_dat <- lastvisit_dat %>%
  mutate(
    grs_quartile = ntile(GRS, 4),
    grs_burden = if_else(grs_quartile == 4, 1, 0),
    fhx_burden = if_else(family_history == "1 FDR", 0, 1),
    apoe_burden = if_else(apoe == "e4+", 1, 0),
    
    # Strategy 1: Top tertile (67th percentile)
    cogdrisk_binary_tertile = ifelse(score_cogd >= quantile(score_cogd, 2/3, na.rm = TRUE), 
                                     1, 0),
    risk_score_cogd_1 = rowSums(across(c(fhx_burden, apoe_burden, cogdrisk_binary_tertile, grs_burden))),
    libra2_binary_tertile = ifelse(score_libra2 >= quantile(score_libra2, 2/3, na.rm = TRUE), 
                                   1, 0),
    risk_score_libra2_1 = rowSums(across(c(fhx_burden, apoe_burden, libra2_binary_tertile, grs_burden))),
    
    # Strategy 2: Top quartile (75th percentile)
    cogdrisk_binary_quartile = ifelse(score_cogd >= quantile(score_cogd, 0.75, na.rm = TRUE), 
                                      1, 0),
    libra2_binary_quartile = ifelse(score_libra2 >= quantile(score_libra2, 0.75, na.rm = TRUE), 
                                      1, 0),
    risk_score_cogd_2 = rowSums(across(c(fhx_burden, apoe_burden, cogdrisk_binary_quartile,
                                    grs_burden))),
    risk_score_libra2_2 = rowSums(across(c(fhx_burden, apoe_burden, libra2_binary_quartile,
                                    grs_burden))),
    
    # Strategy 3: 1 SD above mean
    cogdrisk_binary_1sd = ifelse(score_cogd > mean(score_cogd, na.rm = TRUE) + 
                                    sd(score_cogd, na.rm = TRUE), 
                                 1, 0),
    libra2_binary_1sd = ifelse(score_libra2 > mean(score_libra2, na.rm = TRUE) + 
                                    sd(score_libra2, na.rm = TRUE), 
                                 1, 0),
    risk_score_cogd_3 = rowSums(across(c(fhx_burden, apoe_burden, cogdrisk_binary_1sd,
                                    grs_burden))),
    risk_score_libra2_3 = rowSums(across(c(fhx_burden, apoe_burden, libra2_binary_1sd,
                                    grs_burden))),
  ) 

risk_dat_longer <- risk_dat %>%
  pivot_longer(
    cols = all_of(BIOMARKERS_TRANS),
    names_to = "biomarker",
    values_to = "measure"
  ) 
```

### Check cutoff values

```{r risk-burden-model-hist}
#| fig-width: 12
#| fig-height: 18
#| out-width: "100%"
plot_cutoffs <- function(data, score_var, score_name) {
  cutoff_tertile <- quantile(data[[score_var]], 2/3, na.rm = TRUE)
  cutoff_quartile <- quantile(data[[score_var]], 0.75, na.rm = TRUE)
  cutoff_1sd <- mean(data[[score_var]], na.rm = TRUE) + sd(data[[score_var]], na.rm = TRUE)

  
  ggplot(data, aes(x = .data[[score_var]], y=after_stat(density))) +
    geom_histogram(bins = 25, fill = "steelblue", alpha = 0.7) +
    geom_density(col="red") +
    geom_vline(aes(xintercept = cutoff_tertile, color = "Top Tertile"), 
               linewidth = 1, linetype = "dashed") +
    geom_vline(aes(xintercept = cutoff_quartile, color = "Top Quartile"), 
               linewidth = 1, linetype = "dashed") +
    geom_vline(aes(xintercept = cutoff_1sd, color = "Mean + 1SD"), 
               linewidth = 1, linetype = "dashed") +
    scale_color_manual(
      name = "Cutoff Strategy",
      values = c("Top Tertile" = "blue", "Top Quartile" = "green", 
                 "Mean + 1SD" = "purple")
    ) +
    labs(title = paste(score_name, "Distribution with Cutoff Strategies"),
         x = score_name, y = "Density") +
    theme_minimal() +
    theme(legend.position = "bottom")
}

# Create plots
p1 <- plot_cutoffs(risk_dat, "score_cogd", "CogDrisk")
p2 <- plot_cutoffs(risk_dat, "score_libra2", "LIBRA2")

# Combine
combined_plot <- p1 / p2 
combined_plot
```

```{r risk-burden-model-table}
# Calculate sample sizes for each cutoff method
cutoff_sample_sizes <- tibble(
  Score = rep(c("CogDRISK", "LIBRA2"), each = 3),
  Cutoff_Method = rep(c("Top Tertile", "Top Quartile", "1 SD above mean"), 2),
  Cutoff_Value = c(
    # CogDrisk
    quantile(risk_dat$score_cogd, 2/3, na.rm = TRUE),
    quantile(risk_dat$score_cogd, 0.75, na.rm = TRUE),
    mean(risk_dat$score_cogd, na.rm = TRUE) + sd(risk_dat$score_cogd, na.rm = TRUE),
    # LIBRA2
    quantile(risk_dat$score_libra2, 2/3, na.rm = TRUE),
    quantile(risk_dat$score_libra2, 0.75, na.rm = TRUE),
    mean(risk_dat$score_libra2, na.rm = TRUE) + sd(risk_dat$score_libra2, na.rm = TRUE)
  ),
  N_High = c(
    # CogDrisk
    sum(risk_dat$cogdrisk_binary_tertile == 1, na.rm = TRUE),
    sum(risk_dat$cogdrisk_binary_quartile == 1, na.rm = TRUE),
    sum(risk_dat$cogdrisk_binary_1sd == 1, na.rm = TRUE),
    # LIBRA2
    sum(risk_dat$libra2_binary_tertile == 1, na.rm = TRUE),
    sum(risk_dat$libra2_binary_quartile == 1, na.rm = TRUE),
    sum(risk_dat$libra2_binary_1sd == 1, na.rm = TRUE)
  )
) %>%
  mutate(
    Cutoff_Value = round(Cutoff_Value, 2),
    N_Total = nrow(risk_dat),
    Percent_High = round((N_High / N_Total) * 100, 1)
  )

# Create gt table
cutoff_sample_sizes %>%
  gt() %>%
  tab_header(
    title = md("**High-Risk Classification by Cutoff Method**")
  ) %>%
  cols_label(
    Score = "Risk Score",
    Cutoff_Method = "Cutoff Method",
    Cutoff_Value = "Cutoff Value",
    N_High = "N High Risk",
    N_Total = "Total N",
    Percent_High = "% High Risk"
  ) %>%
  tab_spanner(
    label = md("**Sample Size**"),
    columns = c(N_High, N_Total, Percent_High)
  )
```

## Burden associations with biomarker levels

```{r risk-burden-assoc, warning=FALSE}
#| fig-width: 12
#| fig-height: 10
#| out-width: "100%"
risk_scores_longer <- risk_dat_longer %>%
  select(CONP_ID, risk_score_cogd_1, risk_score_libra2_1,
         risk_score_cogd_2, risk_score_libra2_2,
         risk_score_cogd_3, risk_score_libra2_3,
         biomarker, measure) %>%
  pivot_longer(cols = c(risk_score_cogd_1, risk_score_libra2_1,
                        risk_score_cogd_2, risk_score_libra2_2,
                        risk_score_cogd_3, risk_score_libra2_3),
               names_to = "risk_score_type",
               values_to = "risk_score_value") %>%
  mutate(
    # Extract model type (CogDrisk or LIBRA2)
    model_type = case_when(
      str_detect(risk_score_type, "cogd") ~ "CogDrisk",
      str_detect(risk_score_type, "libra2") ~ "LIBRA2"
    ),
    
    # Extract strategy number
    strategy_label = case_when(
      endsWith(risk_score_type, "1") ~ "Top Tertile (67th %ile)",
      endsWith(risk_score_type, "2") ~ "Top Quartile (75th %ile)",
      endsWith(risk_score_type, "3") ~ "1 SD Above Mean"
    ),
    
    risk_score_factor = factor(risk_score_value)
  )

# CogDrisk only
cogdrisk_plot <- risk_scores_longer %>%
  filter(model_type == "CogDrisk") %>%
  ggplot(aes(x = risk_score_factor, y = measure, fill = risk_score_factor)) +
  geom_violin(alpha = 0.7) +
  geom_boxplot(width = 0.1, fill = "white") +
  facet_wrap(~biomarker, scales = "free") +
  stat_compare_means(method = "kruskal.test",
                     label.y.npc = 0.8,
                     geom = "label",
                     size = 5) +
  labs(x = "CogDrisk Risk Score (0-4)",
       y = "Biomarker Level",
       title = "Biomarker Levels by CogDrisk Composite Score") +
  theme_bw() +
  theme(legend.position = "none",
        axis.text.x = element_text(size = 8),
        strip.text = element_text(size = 8))
cogdrisk_plot

# LIBRA2 only
libra2_plot <- risk_scores_longer %>%
  filter(model_type == "LIBRA2") %>%
  ggplot(aes(x = risk_score_factor, y = measure, fill = risk_score_factor)) +
  geom_violin(alpha = 0.7) +
  geom_boxplot(width = 0.1, fill = "white") +
  facet_wrap(~biomarker, scales = "free") +
  stat_compare_means(method = "kruskal.test",
                     label.y.npc = 0.8,
                     geom = "label",
                     size = 5) +
  labs(x = "LIBRA2 Risk Score (0-4)",
       y = "Biomarker Level",
       title = "Biomarker Levels by LIBRA2 Composite Score") +
  theme_bw() +
  theme(legend.position = "none",
        axis.text.x = element_text(size = 8),
        strip.text = element_text(size = 8))
libra2_plot
```


```{r risk-burden-model}
models <- tribble(~model, ~predictors,
                  "CogDrisk_1", "apoe_burden + cogdrisk_binary_tertile + grs_burden + fhx_burden",
                  "LIBRA2_1", "age + sex + educ + apoe_burden + libra2_binary_tertile + grs_burden + fhx_burden",
                  
                  "CogDrisk_2", "apoe_burden + cogdrisk_binary_quartile + grs_burden + fhx_burden",
                  "LIBRA2_2", "age + sex + educ + apoe_burden + libra2_binary_quartile + grs_burden + fhx_burden",
                  
                  "CogDrisk_3", "apoe_burden + cogdrisk_binary_1sd + grs_burden + fhx_burden",
                  "LIBRA2_3", "age + sex + educ + apoe_burden + libra2_binary_1sd + grs_burden + fhx_burden") 

# Testing cutoffs
risk_preventad <- expand_grid(
  outcome = BIOMARKERS_TRANS,
  models
  ) %>%
  mutate(
    eq = glue("{outcome} ~ {predictors}"),
    res = map(eq, ~lm(as.formula(.), data = risk_dat)),
    dataf = map(res, tidy),
    mod = map(res, glance)
  )
```

```{r risk-burden-model-results}
res_risk_preventad <- bind_rows(select(risk_preventad, model, outcome, dataf)) %>%
  unnest(dataf) %>%
  filter(
    str_detect(term, '^apoe_burden|^cogdrisk_|^libra2_|^grs_|^fhx_burden')
  ) %>%
  arrange(outcome, term, estimate) %>%
  mutate(
    # Significance flag
    sig = ifelse(p.value < 0.05, TRUE, FALSE),
    
    # Confidence intervals
    conf.low = estimate - (std.error * 1.96),
    conf.high = estimate + (std.error * 1.96),
    
    # Exponentiated values for log-transformed outcomes 
    exp_est = case_when(
      str_detect(outcome, "_log") ~ exp(estimate),
      TRUE ~ NA_real_
    ),
    exp_lci = case_when(
      str_detect(outcome, "_log") ~ exp(conf.low),
      TRUE ~ NA_real_
    ),
    exp_uci = case_when(
      str_detect(outcome, "_log") ~ exp(conf.high),
      TRUE ~ NA_real_
    ),
    pct_change = case_when(
      str_detect(outcome, "_log") ~ (exp_est - 1) * 100,
      TRUE ~ NA_real_
    ),
    
    # Result category
    res = case_when(
      sig == TRUE ~ "Significant",
      sig == FALSE ~ "Not Significant"
    ),
    stars = case_when(
      p.value > 0.05 ~ "", 
      between(p.value,  0.01, 0.05) ~ "*",
      between(p.value, 0.001, 0.01) ~ "**",
      p.value < 0.001 ~ "***"
    )
  ) %>%
  # Round numeric values
  mutate(across(c(estimate, std.error, statistic, p.value, 
                  conf.low, conf.high, exp_est, exp_lci, exp_uci, pct_change), 
                ~round(., digits = 3)))
```

## Risk burden model results - Plot

```{r risk-burden-model-plot}
#| fig-width: 14
#| fig-height: 10
#| out-width: "100%"

model_order <- c(
  # CogDRISK models - all strategies together
  "CogDrisk_1", "CogDrisk_2", "CogDrisk_3",
  # LIBRA2 models - all strategies together
  "LIBRA2_1", "LIBRA2_2", "LIBRA2_3"
)

for (i in model_order){
  model_title <- case_when(
    i == "CogDrisk_1" ~ "CogDrisk Risk Burden Model (Top tertile)",
    i == "LIBRA2_1" ~ "LIBRA2 Risk Burden Model (Top tertile)",
    
    i == "CogDrisk_2" ~ "CogDrisk Risk Burden Model (Top quartile)",
    i == "LIBRA2_2" ~ "LIBRA2 Risk Burden Model (Top quartile)",
    
    i == "CogDrisk_3" ~ "CogDrisk Risk Burden Model (1 SD above mean)",
    i == "LIBRA2_3" ~ "LIBRA2 Risk Burden Model (1 SD above mean)",
    TRUE ~ paste("Model:", i)
  )
  
  plot_data <- res_risk_preventad %>%
    filter(model == i) %>%  
    mutate(sig = factor(sig, levels = c(TRUE, FALSE)))
  
  res_plot <- ggplot(plot_data,
    aes(y = term, x = estimate, group = term, xmin = conf.low, xmax = conf.high, 
        color = term, alpha = sig)  
  ) +
  geom_point(size = 3, position = position_dodge(width = 0.5)) + 
  geom_errorbarh(linewidth = 0.8, position = position_dodge(width = 0.5), height = 0.2) + 
  geom_vline(xintercept = 0, linetype = 2) +
  scale_alpha_manual(
    values = c(`FALSE` = 0.3, `TRUE` = 1),  # Not sig = transparent, sig = opaque
    labels = c(`FALSE` = "Not Significant", `TRUE` = "Significant"),
    name = "Significance"
  ) +
  scale_color_jama() +
  facet_wrap(~outcome, scales = "free_x", ncol = 2) +
  labs(
    x = "Beta (95% CI)",
    y = "Predictor",
    title = model_title
  ) +
  theme_bw() +
  theme(
    axis.text.y = element_text(size = 10, angle = 0, hjust = 1),
    axis.text.x = element_text(size = 10),
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 14, hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(size = 11, hjust = 0.5),
    strip.background = element_rect(fill = "grey90"),
    strip.text = element_text(face = "bold", size = 11),
    legend.position = "bottom",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9)
  )
  
  print(res_plot)
}
```

## Diagnostics

```{r diagnostics-plot}
#| fig-width: 14
#| fig-height: 14
#| out-width: "100%"

for (i in models$model){
  model_title <- case_when(
    i == "CogDrisk_1" ~ "CogDrisk Risk Burden Model (Top tertile)",
    i == "LIBRA2_1" ~ "LIBRA2 Risk Burden Model (Top tertile)",
    
    i == "CogDrisk_2" ~ "CogDrisk Risk Burden Model (Top quartile)",
    i == "LIBRA2_2" ~ "LIBRA2 Risk Burden Model (Top quartile)",
    
    i == "CogDrisk_3" ~ "CogDrisk Risk Burden Model (1 SD above mean)",
    i == "LIBRA2_3" ~ "LIBRA2 Risk Burden Model (1 SD above mean)",
    TRUE ~ paste("Model:", i)
  )
  
  model_data <- risk_preventad %>%
    filter(model == i) %>%
    mutate(
      residuals = map(res, residuals),
      fitted = map(res, fitted)
    ) %>%
    select(outcome, residuals, fitted) %>%
    unnest(c(residuals, fitted))
    
  # Histogram
  p_hist <- ggplot(model_data, aes(x = residuals)) +
    geom_histogram(bins = 20, fill = "steelblue", alpha = 0.7, color = "white") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
    facet_wrap(~outcome, scales = "free_x", ncol = 3) +
    labs(title = "Histogram of Residuals",
         x = "Residuals", y = "Count") +
    theme_minimal() +
    theme(
      strip.text = element_text(face = "bold", size = 10),
      strip.background = element_rect(fill = "grey90", color = NA),
      plot.title = element_text(size = 12, face = "bold", hjust = 0.5)
    )
    
  # Q-Q plot
  p_qq <- ggplot(model_data, aes(sample = residuals)) +
    stat_qq(size = 0.8, alpha = 0.6) +
    stat_qq_line(color = "red", linewidth = 1) +
    facet_wrap(~outcome, scales = "free", ncol = 3) +
    labs(title = "Q-Q Plot",
         x = "Theoretical Quantiles", y = "Sample Quantiles") +
    theme_minimal() +
    theme(
      strip.text = element_text(face = "bold", size = 10),
      strip.background = element_rect(fill = "grey90", color = NA),
      plot.title = element_text(size = 12, face = "bold", hjust = 0.5)
    )
    
  # Residuals vs Fitted
  p_fitted <- ggplot(model_data, aes(x = fitted, y = residuals)) +
    geom_point(alpha = 0.5, size = 1) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
    geom_smooth(se = FALSE, color = "blue", linewidth = 0.8, method = "loess") +
    facet_wrap(~outcome, scales = "free", ncol = 3) +
    labs(title = "Residuals vs Fitted Values",
         x = "Fitted Values", y = "Residuals") +
    theme_minimal() +
    theme(
      strip.text = element_text(face = "bold", size = 10),
      strip.background = element_rect(fill = "grey90", color = NA),
      plot.title = element_text(size = 12, face = "bold", hjust = 0.5)
    )
  
  # Stack vertically
  combined <- p_hist / p_qq / p_fitted +
    plot_annotation(
      title = model_title,
      theme = theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5))
    )
  print(combined)
}
```



