---
title: "Aim 1: Longitudinal Risk Burden Analysis"
format:
  html:
    embed-resources: true
    self-contained: true
    toc: true
    toc-depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = here::here())
```

```{r load-packages}
library(broom.mixed)
library(lme4)
library(patchwork)
library(splines)

source('workflow/scripts/config.R')
```

```{r load-data}
# Pre-built by 00_load_analysis_data.R
all_dat <- readRDS(file.path(DATA_CLEANED_PATH, "PREVENTAD_longitudinal.rds")) %>%
  mutate(
    score_cogd_z   = as.numeric(scale(score_cogd)),
    score_libra2_z = as.numeric(scale(score_libra2))
  )

cat("Data dimensions:", nrow(all_dat), "rows,", ncol(all_dat), "columns\n")
cat("Number of unique participants:", length(unique(all_dat$CONP_ID)), "\n")
cat("Time range:", min(all_dat$years, na.rm = TRUE), "to",
    max(all_dat$years, na.rm = TRUE), "years\n")
```

```{r define-vars}
# Outcome labels (defined here for use across growth-model plots)
outcome_labels <- c(
  ab_ratio               = "A\u03b242/40",
  gfap_log               = "GFAP",
  nfl_log                = "NfL",
  ptau181_log            = "ptau-181",
  ptau217_log            = "ptau-217",
  ptau217_ab42_ratio_log = "ptau-217/A\u03b242 ratio"
)

# subjects with 3+ ptau217 observations (with all covariates)
ptau217_subjects <- all_dat %>%
  filter(!is.na(ptau217), !is.na(risk_score_cogd), 
         !is.na(risk_score_libra2), !is.na(baseline_age), 
         !is.na(sex), !is.na(educ)) %>%
  count(CONP_ID) %>%
  filter(n >= 3) %>%
  pull(CONP_ID)
```

## Biomarker unconditional growth models

```{r growth-models}
# Unconditional growth: compare linear, ns(df=2), ns(df=3)
# Fit all three specifications per biomarker to inform choice of time term.
uncond_specs <- tribble(
  ~spec,        ~time_term,
  "linear",     "years",
  "ns_df2",     "ns(years, df = 2)",
  "ns_df3",     "ns(years, df = 3)"
)

fit_uncond <- function(o, e) {
  dat_sub <- all_dat[!is.na(all_dat[[o]]), ]
  tryCatch(
    lmer(as.formula(e), data = dat_sub, REML = FALSE),
    error = function(err) {
      message("Failed to fit: ", e, "\n  Reason: ", conditionMessage(err))
      NULL
    }
  )
}

uncond_models <- expand_grid(outcome = BIOMARKERS_TRANS, uncond_specs) %>%
  mutate(
    # Random slopes for outcomes with enough longitudinal data;
    # random intercept only for sparse outcomes
    re_term = if_else(outcome %in% sparse_outcomes,
                      "(1 | CONP_ID)",
                      "(1 + years | CONP_ID)"),
    eq  = glue("{outcome} ~ {time_term} + {re_term}"),
    res = map2(outcome, as.character(eq), fit_uncond),
    mod = map(res, ~if (is.null(.x)) tibble() else glance(.x))
  )

# AIC comparison table across specs (drops failed fits)
uncond_aic <- uncond_models %>%
  filter(!map_lgl(res, is.null)) %>%
  select(outcome, spec, mod) %>%
  unnest(mod) %>%
  select(outcome, spec, AIC, BIC, logLik) %>%
  group_by(outcome) %>%
  mutate(delta_AIC = AIC - min(AIC),
         best      = AIC == min(AIC)) %>%
  ungroup()

# Sequential likelihood ratio tests:
#   ns_df2 vs linear (does adding curvature help?)
#   ns_df3 vs ns_df2 (does extra df help further?)
uncond_lrt <- uncond_models %>%
  filter(!map_lgl(res, is.null)) %>%
  select(outcome, spec, res) %>%
  pivot_wider(names_from = spec, values_from = res) %>%
  rowwise() %>%
  mutate(
    p_vs_linear = if (!is.null(linear) && !is.null(ns_df2))
      anova(linear, ns_df2)$"Pr(>Chisq)"[2] else NA_real_,
    p_vs_df2    = if (!is.null(ns_df2) && !is.null(ns_df3))
      anova(ns_df2, ns_df3)$"Pr(>Chisq)"[2] else NA_real_
  ) %>%
  ungroup() %>%
  transmute(
    outcome,
    linear = NA_real_,
    ns_df2 = p_vs_linear,
    ns_df3 = p_vs_df2
  ) %>%
  pivot_longer(c(linear, ns_df2, ns_df3),
               names_to = "spec", values_to = "lrt_p_vs_simpler")

uncond_aic <- uncond_aic %>%
  left_join(uncond_lrt, by = c("outcome", "spec")) %>%
  arrange(outcome, AIC)

print(uncond_aic)

# Keep df=3 as the default for downstream plotting (retain original object name)
uncond_spline <- uncond_models %>% filter(spec == "ns_df3") %>% select(outcome, eq, res, mod)

# Predicted trajectories — use biomarker-specific time range to avoid extrapolation
biomarker_max_years_uncond <- all_dat %>%
  pivot_longer(cols = all_of(BIOMARKERS_TRANS), names_to = "outcome", values_to = "value") %>%
  filter(!is.na(value)) %>%
  group_by(outcome) %>%
  summarise(max_years = max(years, na.rm = TRUE), .groups = "drop")

spline_preds <- uncond_spline %>%
  left_join(biomarker_max_years_uncond, by = "outcome") %>%
  mutate(preds = map2(res, max_years, ~{
    d <- tibble(years = seq(0, .y, by = 0.2))
    d$pred <- predict(.x, newdata = d, re.form = NA)
    d
  })) %>%
  select(outcome, preds) %>%
  unnest(preds) %>%
  mutate(outcome_label = factor(outcome_labels[outcome], levels = outcome_labels))

# Long observed data
obs_long <- all_dat %>%
  pivot_longer(cols = all_of(BIOMARKERS_TRANS),
               names_to = "outcome", values_to = "observed") %>%
  filter(!is.na(observed)) %>%
  mutate(outcome_label = factor(outcome_labels[outcome], levels = outcome_labels))

p_uncond <- ggplot() +
  geom_point(data = obs_long,
             aes(x = years, y = observed),
             alpha = 0.12, size = 0.6, color = "grey50") +
  geom_line(data = spline_preds,
            aes(x = years, y = pred),
            color = "#DF8F44FF", linewidth = 1.1) +
  scale_x_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.03))) +
  facet_wrap(~outcome_label, scales = "free", ncol = 3) +
  labs(
    x        = "Years from Baseline",
    y        = "Biomarker Level",
    title    = "Unconditional Biomarker Trajectories Over Follow-Up",
    subtitle = "Population-average natural spline trajectory, df=3 (LMM, random intercept)"
  ) +
  theme_bw(base_size = 12) +
  theme(
    strip.text       = element_text(face = "bold"),
    strip.background = element_rect(fill = "grey90"),
    plot.title       = element_text(face = "bold", size = 14, hjust = 0.5),
    plot.subtitle    = element_text(size = 11, hjust = 0.5, color = "grey40"),
    panel.grid.minor = element_blank()
  )

print(p_uncond)

# focused plot of ptau217 to zoom in
# AIC of 2 vs 3 dfs
```

## LME Models

### Model 1: Main effect

These models test whether composite risk score (0-4 scale) predicts biomarker levels in a dose-response manner. The beta coefficient represents the change in biomarker level for each 1-unit increase in risk score (e.g., going from 2 to 3 risk factors). LIBRA2 models are adjusted for age, sex, and education.

```{r main-models}
models_main <- tribble(
  ~model, ~predictors,
  "CogDrisk_main", "years_centered + risk_score_cogd + (1 + years_centered | CONP_ID)",
  "LIBRA2_main", "years_centered + baseline_age + sex + educ + risk_score_libra2 + (1 + years_centered | CONP_ID)"
)

models_main_sparse <- tribble(
  ~model, ~predictors,
  "CogDrisk_main", "years_centered + risk_score_cogd + (1 | CONP_ID)",
  "LIBRA2_main",   "years_centered + baseline_age + sex + educ + risk_score_libra2 + (1 | CONP_ID)"
)

lme_main <- bind_rows(
  # all other outcomes — full random slopes, full data
  expand_grid(
    outcome = setdiff(BIOMARKERS_TRANS, "ptau217_ab42_ratio_log"),
    models_main
  ),
  # ptau217_ab42 — intercept only, full data
  expand_grid(
    outcome = "ptau217_ab42_ratio_log",
    models_main_sparse
  )
) %>%
  mutate(
    dat = map(outcome, ~case_when(
      . == "ptau217_log"      ~ list(filter(all_dat, CONP_ID %in% ptau217_subjects)),
      . == "ptau217_ab42_ratio_log" ~ list(all_dat),
      TRUE                ~ list(all_dat)
    ) %>% pluck(1)),
    eq    = glue("{outcome} ~ {predictors}"),
    res   = map2(eq, dat, ~lmer(as.formula(.x), data = .y)),
    dataf = map(res, ~tidy(., conf.int = TRUE)),
    mod   = map(res, glance)
  )
```

### Model 2: Time interactions (risk score × time)

These models test whether the rate of biomarker change over time differs by risk score level and other covariates. The interaction term (years_centered:risk_score) shows how much faster (or slower) biomarkers change for each additional risk factor. For LIBRA2, interactions between time and demographic covariates (baseline_age, sex, educ) are also included. 

```{r interact-models}
models_time_interact <- tribble(
  ~model, ~predictors,
  "CogDrisk_interact", "years_centered * risk_score_cogd + (1 + years_centered | CONP_ID)",
  "LIBRA2_interact", "years_centered * (risk_score_libra2 + baseline_age + sex + educ) + (1 + years_centered | CONP_ID)"
)

models_time_interact_sparse <- tribble(
  ~model, ~predictors,
  "CogDrisk_interact", "years_centered * risk_score_cogd + (1 | CONP_ID)",
  "LIBRA2_interact",   "years_centered * (baseline_age + sex + educ + risk_score_libra2) + (1 | CONP_ID)"
)

lme_time_interact <- bind_rows(
  # all other outcomes — full random slopes, full data
  expand_grid(
    outcome = setdiff(BIOMARKERS_TRANS, "ptau217_ab42_ratio_log"),
    models_time_interact
  ),
  # ptau217_ab42 — intercept only, full data
  expand_grid(
    outcome = "ptau217_ab42_ratio_log",
    models_time_interact_sparse
  )
) %>%
  mutate(
    dat = map(outcome, ~case_when(
      . == "ptau217_log"      ~ list(filter(all_dat, CONP_ID %in% ptau217_subjects)),
      . == "ptau217_ab42_ratio_log" ~ list(all_dat),
      TRUE                ~ list(all_dat)
    ) %>% pluck(1)),
    eq    = glue("{outcome} ~ {predictors}"),
    res   = map2(eq, dat, ~lmer(as.formula(.x), data = .y)),
    dataf = map(res, ~tidy(., conf.int = TRUE)),
    mod   = map(res, glance)
  )
```

### Model 3: Multivariable

These models decompose the composite risk score into its constituent predictors: APOE ε4 burden, z-scored composite risk score (PXS), polygenic risk score (GRS), and family history burden (FHx). The time interaction model tests whether individual risk factors modify the rate of biomarker change over time.

```{r multivar-models}
# Time interaction
models_multivar_interact <- tribble(
  ~model, ~predictors,
  "CogDrisk_multivar_interact", "years_centered * (apoe_burden + score_cogd_z + GRS + fhx_burden) + (1 + years_centered | CONP_ID)",
  "LIBRA2_multivar_interact", "years_centered * (baseline_age + sex + educ + apoe_burden + score_libra2_z + GRS + fhx_burden) + (1 + years_centered | CONP_ID)"
)

models_multivar_interact_sparse <- tribble(
  ~model, ~predictors,
  "CogDrisk_multivar_interact", "years_centered * (apoe_burden + score_cogd_z + GRS + fhx_burden) + (1 | CONP_ID)",
  "LIBRA2_multivar_interact", "years_centered * (baseline_age + sex + educ + apoe_burden + score_libra2_z + GRS + fhx_burden) + (1 | CONP_ID)"
)

lme_multivar_interact <- bind_rows(
  expand_grid(outcome = setdiff(BIOMARKERS_TRANS, "ptau217_ab42_ratio_log"), models_multivar_interact),
  expand_grid(outcome = "ptau217_ab42_ratio_log", models_multivar_interact_sparse)
) %>%
  mutate(
    dat   = map(outcome, ~case_when(
      . == "ptau217_log" ~ list(filter(all_dat, CONP_ID %in% ptau217_subjects)),
      TRUE ~ list(all_dat)
    ) %>% pluck(1)),
    eq    = glue("{outcome} ~ {predictors}"),
    res   = map2(eq, dat, ~lmer(as.formula(.x), data = .y)),
    dataf = map(res, ~tidy(., conf.int = TRUE)),
    mod   = map(res, glance)
  )
```

## LME results

```{r lme-results}
# Combine all LME results
lme_results <- bind_rows(
  lme_main %>%
    select(model, outcome, dataf) %>%
    mutate(model_type = "Main Effect"),
  lme_time_interact %>%
    select(model, outcome, dataf) %>%
    mutate(model_type = "Time Interaction")
) %>%
  unnest(dataf) %>%
  filter(str_detect(term, "^risk_score|^years_centered:risk_score")) %>%
  # FDR family = 12 per section (6 biomarkers × 2 models); group by both
  # model_type and term type so the level effect and interaction term from the
  # interaction model are corrected in separate families of 12, not pooled
  group_by(model_type, is_interaction = str_detect(term, ":")) %>%
  mutate(
    # Confidence intervals
    conf.low = estimate - (std.error * 1.96),
    conf.high = estimate + (std.error * 1.96),

    # Significance — 12 tests per section
    p_adj       = p.adjust(p.value, method = "fdr"),
    sig = case_when(
      p_adj < 0.05   ~ "Significant (FDR-adjusted)",
      p.value < 0.05 ~ "Significant (non-adjusted)",
      TRUE           ~ "Not significant"
    ) %>% fct_relevel("Significant (FDR-adjusted)", "Significant (non-adjusted)", "Not significant"),
    stars = case_when(
      p.value >= 0.05 ~ "",
      p.value >= 0.01 & p.value < 0.05 ~ "*",
      p.value >= 0.001 & p.value < 0.01 ~ "**",
      p.value < 0.001 ~ "***"
    ),
    p_adj_fmt   = if_else(p_adj < 0.001, "<0.001", sprintf("%.3f", p_adj)),
    p_adj_stars = if_else(p_adj < 0.05, "sig", ""),

    # Exponentiated values for log-transformed outcomes
    exp_est = case_when(
      str_detect(outcome, "_log$") ~ exp(estimate),
      TRUE ~ NA_real_
    ),
    exp_lci = case_when(
      str_detect(outcome, "_log$") ~ exp(conf.low),
      TRUE ~ NA_real_
    ),
    exp_uci = case_when(
      str_detect(outcome, "_log$") ~ exp(conf.high),
      TRUE ~ NA_real_
    ),

    # Percentage change per unit for all biomarkers
    pct_change_per_unit = case_when(
      str_detect(outcome, "_log$") ~ (exp_est - 1) * 100,
      TRUE ~ NA_real_
    ),

    # Risk score type
    risk_type = if_else(str_detect(model, "CogDrisk"), "CogDrisk", "LIBRA2")
  ) %>%
  ungroup() %>%
  mutate(across(c(estimate, std.error, statistic, p.value,
                  conf.low, conf.high, exp_est, exp_lci, exp_uci, pct_change_per_unit),
                ~round(., digits = 3)))
```

### Main effect results

```{r main-models-plot}
#| fig-width: 10
#| fig-height: 6

plot_data_main <- lme_results %>%
  filter(str_detect(model, "main")) %>%
  mutate(
    risk_type  = factor(risk_type, levels = c("LIBRA2", "CogDrisk")),
    outcome_label = case_when(
      outcome == "ab_ratio"               ~ "A\u03b242/40",
      outcome == "gfap_log"               ~ "GFAP (log)",
      outcome == "nfl_log"                ~ "NfL (log)",
      outcome == "ptau181_log"            ~ "p-tau181 (log)",
      outcome == "ptau217_log"            ~ "p-tau217 (log)",
      outcome == "ptau217_ab42_ratio_log" ~ "p-tau217/A\u03b242 (log)",
      TRUE ~ outcome
    ),
    outcome_label = factor(outcome_label, levels = rev(c(
      "A\u03b242/40", "GFAP (log)", "NfL (log)",
      "p-tau181 (log)", "p-tau217 (log)", "p-tau217/A\u03b242 (log)"
    )))
  )

p_main_combined <- ggplot(plot_data_main,
  aes(x = estimate, y = outcome_label,
      xmin = conf.low, xmax = conf.high,
      color = risk_type, alpha = sig,
      group = risk_type)
) +
  geom_point(size = 3, position = position_dodge(width = 0.6)) +
  geom_errorbarh(aes(linewidth = sig), position = position_dodge(width = 0.6), height = 0.3) +
  geom_vline(xintercept = 0, linetype = 2, color = "grey40") +
  scale_alpha_manual(
    values = c(
      "Significant (FDR-adjusted)" = 1,
      "Significant (non-adjusted)" = 1,
      "Not significant"            = 0.25
    ),
    name = "Significance"
  ) +
  scale_linewidth_manual(
    values = c(
      "Significant (FDR-adjusted)" = 1.5,
      "Significant (non-adjusted)" = 0.8,
      "Not significant"            = 0.5
    ),
    name = "Significance"
  ) +
  scale_color_manual(
    values = c("CogDrisk" = "#374E55FF", "LIBRA2" = "#DF8F44FF"),
    breaks = c("CogDrisk", "LIBRA2"),
    name = "Model"
  ) +
  labs(
    x     = "Beta (95% CI)",
    y     = NULL#,
    #title = "Figure 3. Baseline Association with Risk Score"
  ) +
  theme_bw(base_size = 13) +
  theme(
    legend.position  = "bottom",
    legend.title     = element_text(size = 10, face = "bold"),
    legend.text      = element_text(size = 9),
    panel.grid.minor = element_blank(),
    plot.title       = element_text(face = "bold", size = 14, hjust = 0.5),
    plot.subtitle    = element_text(size = 11, hjust = 0.5, color = "grey40"),
    axis.text        = element_text(size = 11)
  )

print(p_main_combined)

# ggsave(file.path(AIM1_OUTPUT_PATH$figures, "fig4_lme_main_effects.png"),
#        p_main_combined, width = 10, height = 6, dpi = 300)
```

```{r main-models-results}
biomarker_order_lme <- c(
  "A\u03b242/40", "GFAP", "NfL", "ptau-181", "ptau-217", "ptau-217/A\u03b242 ratio"
)

term_order_lme_main <- c(
  "Composite Risk Score",
  "Covariates",
  "Age", "Sex (Male)", "Education (years)"
)

covariate_terms_lme <- c("Age", "Sex", "Education (years)")

main_all_raw <- lme_main %>%
  select(model, outcome, dataf) %>%
  unnest(dataf) %>%
  filter(effect == "fixed",
         !term %in% c("(Intercept)", "years_centered")) %>%
  mutate(
    conf.low  = estimate - 1.96 * std.error,
    conf.high = estimate + 1.96 * std.error,
    beta_se   = sprintf("%.3f (%.3f)", estimate, std.error),
    ci        = sprintf("[%.3f, %.3f]", conf.low, conf.high),
    p_fmt     = if_else(p.value < 0.001, "<0.001", sprintf("%.3f", p.value)),
    # FDR: 12 tests (6 biomarkers × 2 models) on the composite risk score only;
    # covariates (age, sex, education) are nuisance variables — excluded from pool
    p_adj = {
      adj      <- rep(NA_real_, n())
      idx      <- str_detect(term, "^risk_score")
      adj[idx] <- p.adjust(p.value[idx], method = "fdr")
      adj
    },
    p_adj_fmt   = if_else(!is.na(p_adj),
                          if_else(p_adj < 0.001, "<0.001", sprintf("%.3f", p_adj)),
                          "—"),
    p_adj_stars = if_else(!is.na(p_adj) & p_adj < 0.05, "sig", ""),
    stars     = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      TRUE            ~ ""
    ),
    model_short = if_else(str_detect(model, "CogDrisk"), "CogDrisk", "LIBRA2"),
    biomarker   = case_when(
      outcome == "ab_ratio"               ~ "A\u03b242/40",
      outcome == "gfap_log"               ~ "GFAP",
      outcome == "nfl_log"                ~ "NfL",
      outcome == "ptau181_log"            ~ "ptau-181",
      outcome == "ptau217_log"            ~ "ptau-217",
      outcome == "ptau217_ab42_ratio_log" ~ "ptau-217/A\u03b242 ratio"
    ),
    term_label  = case_when(
      str_detect(term, "^risk_score") ~ "Composite Risk Score",
      term == "baseline_age"          ~ "Age",
      term == "sexMale"               ~ "Sex",
      term == "educ"                  ~ "Education (years)",
      TRUE ~ term
    )
  )

main_wide <- main_all_raw %>%
  select(model_short, biomarker, term_label, beta_se, ci, p_fmt, p_adj_fmt, p_adj_stars, stars) %>%
  pivot_wider(
    names_from  = model_short,
    values_from = c(beta_se, ci, p_fmt, p_adj_fmt, p_adj_stars, stars),
    names_glue  = "{model_short}_{.value}"
  ) %>%
  mutate(
    across(c(CogDrisk_beta_se, CogDrisk_ci, CogDrisk_p_fmt, CogDrisk_p_adj_fmt), ~replace_na(., "\u2014")),
    CogDrisk_stars       = replace_na(CogDrisk_stars, ""),
    CogDrisk_p_adj_stars = replace_na(CogDrisk_p_adj_stars, "")
  )

# "Covariates" separator row per biomarker
covariate_header_lme <- main_wide %>%
  distinct(biomarker) %>%
  mutate(
    term_label            = "Covariates",
    CogDrisk_beta_se      = "", CogDrisk_ci = "", CogDrisk_p_fmt = "", CogDrisk_p_adj_fmt = "", CogDrisk_p_adj_stars = "", CogDrisk_stars = "",
    LIBRA2_beta_se        = "", LIBRA2_ci   = "", LIBRA2_p_fmt   = "", LIBRA2_p_adj_fmt   = "", LIBRA2_p_adj_stars   = "", LIBRA2_stars   = ""
  )

main_combined <- bind_rows(main_wide, covariate_header_lme) %>%
  mutate(
    biomarker  = factor(biomarker,  levels = biomarker_order_lme),
    term_label = factor(term_label, levels = term_order_lme_main)
  ) %>%
  arrange(biomarker, term_label) %>%
  select(biomarker, term_label,
         CogDrisk_beta_se, CogDrisk_ci, CogDrisk_p_fmt, CogDrisk_p_adj_fmt, CogDrisk_p_adj_stars, CogDrisk_stars,
         LIBRA2_beta_se,   LIBRA2_ci,   LIBRA2_p_fmt,   LIBRA2_p_adj_fmt,   LIBRA2_p_adj_stars,   LIBRA2_stars)

main_gt <- main_combined %>%
  gt(groupname_col = "biomarker") %>%
  tab_header(
    title    = md("**Table S4.** Main Effect of Composite Risk Score on Plasma Biomarker Levels"),
    subtitle = md("Linear mixed-effects models. LIBRA2 additionally adjusted for age, sex, and education.")
  ) %>%
  tab_spanner(label = md("**CogDrisk**"), columns = starts_with("CogDrisk_")) %>%
  tab_spanner(label = md("**LIBRA2**"),   columns = starts_with("LIBRA2_")) %>%
  cols_label(
    term_label         = md("**Predictor**"),
    CogDrisk_beta_se   = md("**\u03b2 (SE)**"),
    CogDrisk_ci        = md("**95% CI**"),
    CogDrisk_p_fmt     = md("**p**"),
    CogDrisk_p_adj_fmt = md("**q**"),
    LIBRA2_beta_se     = md("**\u03b2 (SE)**"),
    LIBRA2_ci          = md("**95% CI**"),
    LIBRA2_p_fmt       = md("**p**"),
    LIBRA2_p_adj_fmt   = md("**q**")
  ) %>%
  cols_hide(columns = c(CogDrisk_stars, CogDrisk_p_adj_stars, LIBRA2_stars, LIBRA2_p_adj_stars)) %>%
  tab_style(
    style     = cell_fill(color = "#e8f4e8"),
    locations = cells_body(columns = CogDrisk_p_fmt,
                           rows    = CogDrisk_stars %in% c("*", "**", "***"))
  ) %>%
  tab_style(
    style     = cell_fill(color = "#e8f4e8"),
    locations = cells_body(columns = LIBRA2_p_fmt,
                           rows    = LIBRA2_stars %in% c("*", "**", "***"))
  ) %>%
  tab_style(
    style     = cell_fill(color = "#c8e6c9"),
    locations = cells_body(columns = CogDrisk_p_adj_fmt,
                           rows    = CogDrisk_p_adj_stars == "sig")
  ) %>%
  tab_style(
    style     = cell_fill(color = "#c8e6c9"),
    locations = cells_body(columns = LIBRA2_p_adj_fmt,
                           rows    = LIBRA2_p_adj_stars == "sig")
  ) %>%
  tab_style(
    style     = list(cell_text(weight = "bold"), cell_fill(color = "#f8f9fa")),
    locations = cells_row_groups()
  ) %>%
  tab_style(
    style     = list(cell_text(weight = "bold", style = "italic"),
                     cell_fill(color = "#f0f2f4")),
    locations = cells_body(rows = term_label == "Covariates")
  ) %>%
  tab_style(
    style     = cell_text(indent = px(16)),
    locations = cells_body(
      columns = term_label,
      rows    = term_label %in% covariate_terms_lme
    )
  ) %>%
  tab_footnote(
    footnote  = md("p: unadjusted; q: Benjamini-Hochberg FDR-adjusted. \u03b2 = change in biomarker per 1-unit increase in composite risk score. Light green: p<0.05; dark green: q<0.05. \u2014 = not applicable (CogDrisk not adjusted for these covariates)."),
    locations = cells_column_spanners()
  ) %>%
  cols_align(align = "left",   columns = term_label) %>%
  cols_align(align = "center", columns = c(starts_with("CogDrisk_"), starts_with("LIBRA2_"))) %>%
  row_group_order(groups = biomarker_order_lme)

main_gt

gtsave(main_gt, filename = file.path(AIM1_OUTPUT_PATH$tables, "s_table4_lme_main.docx"))
```

### Time interaction results

```{r interact-models-plot}
#| fig-width: 16
#| fig-height: 6

panel_labels <- c(
  "level"       = "Risk score main effect",
  "interaction" = "Time \u00d7 risk score"
)

plot_data_interact <- lme_results %>%
  filter(str_detect(model, "interact")) %>%
  mutate(
    term_type = if_else(!str_detect(term, ":"), "level", "interaction"),
    term_type = factor(term_type, levels = c("level", "interaction"),
                       labels = panel_labels),
    risk_type  = factor(risk_type, levels = c("LIBRA2", "CogDrisk")),
    outcome_label = case_when(
      outcome == "ab_ratio"               ~ "A\u03b242/40",
      outcome == "gfap_log"               ~ "GFAP (log)",
      outcome == "nfl_log"                ~ "NfL (log)",
      outcome == "ptau181_log"            ~ "p-tau181 (log)",
      outcome == "ptau217_log"            ~ "p-tau217 (log)",
      outcome == "ptau217_ab42_ratio_log" ~ "p-tau217/A\u03b242 (log)",
      TRUE ~ outcome
    ),
    outcome_label = factor(outcome_label, levels = rev(c(
      "A\u03b242/40", "GFAP (log)", "NfL (log)",
      "p-tau181 (log)", "p-tau217 (log)", "p-tau217/A\u03b242 (log)"
    )))
  )

p_interact_combined <- ggplot(plot_data_interact,
  aes(x = estimate, y = outcome_label,
      xmin = conf.low, xmax = conf.high,
      color = risk_type, alpha = sig,
      group = risk_type)
) +
  geom_point(size = 3, position = position_dodge(width = 0.6)) +
  geom_errorbarh(aes(linewidth = sig), position = position_dodge(width = 0.6), height = 0.3) +
  geom_vline(xintercept = 0, linetype = 2, color = "grey40") +
  scale_alpha_manual(
    values = c(
      "Significant (FDR-adjusted)" = 1,
      "Significant (non-adjusted)" = 1,
      "Not significant"            = 0.25
    ),
    name = "Significance"
  ) +
  scale_linewidth_manual(
    values = c(
      "Significant (FDR-adjusted)" = 1.5,
      "Significant (non-adjusted)" = 0.8,
      "Not significant"            = 0.5
    ),
    name = "Significance"
  ) +
  scale_color_manual(
    values = c("CogDrisk" = "#374E55FF", "LIBRA2" = "#DF8F44FF"),
    breaks = c("CogDrisk", "LIBRA2"),
    name = "Model"
  ) +
  facet_wrap(~term_type, scales = "free_x") +
  labs(
    x     = "\u03b2 (95% CI)",
    y     = NULL
  ) +
  theme_bw(base_size = 13) +
  theme(
    legend.position  = "bottom",
    legend.title     = element_text(size = 10, face = "bold"),
    legend.text      = element_text(size = 9),
    panel.grid.minor = element_blank(),
    plot.title       = element_text(face = "bold", size = 14, hjust = 0.5),
    strip.background = element_rect(fill = "grey92"),
    strip.text       = element_text(face = "bold", size = 11),
    axis.text        = element_text(size = 11)
  )

print(p_interact_combined)

# ggsave(file.path(AIM1_OUTPUT_PATH$figures, "fig4_risk_lme_res.png"),
#        p_interact_combined, width = 10, height = 6, dpi = 300)
```

```{r interact-models-results}
term_order_lme_interact <- c(
  "Time (years)",
  "Composite Risk Score",
  "Time \u00d7 Composite Risk Score",
  "Covariates",
  "Age", "Sex (Male)", "Education (years)",
  "Time \u00d7 Covariates",
  "Time \u00d7 Age", "Time \u00d7 Sex (Male)", "Time \u00d7 Education"
)

covariate_terms_interact  <- c("Age", "Sex (Male)", "Education (years)")
covariate_interact_terms  <- c("Time \u00d7 Age", "Time \u00d7 Sex (Male)", "Time \u00d7 Education")

interact_combined_raw <- lme_time_interact %>%
  select(model, outcome, dataf) %>%
  unnest(dataf) %>%
  filter(effect == "fixed", term != "(Intercept)") %>%
  mutate(
    conf.low    = estimate - 1.96 * std.error,
    conf.high   = estimate + 1.96 * std.error,
    beta_se     = sprintf("%.3f (%.3f)", estimate, std.error),
    ci          = sprintf("[%.3f, %.3f]", conf.low, conf.high),
    p_fmt       = if_else(p.value < 0.001, "<0.001", sprintf("%.3f", p.value)),
    # FDR: two separate 12-test families (6 biomarkers \u00d7 2 models each):
    #   1. Conditional level effect (risk_score main effect at mean follow-up)
    #   2. Time \u00d7 risk score interaction term
    # Covariates, time, and intercept are excluded from correction.
    p_adj = {
      adj <- rep(NA_real_, n())
      # Family 1: level effect (risk_score*, no colon)
      idx_level <- str_detect(term, "^risk_score") & !str_detect(term, ":")
      adj[idx_level] <- p.adjust(p.value[idx_level], method = "fdr")
      # Family 2: time \u00d7 risk score interaction (12 tests)
      idx_inter <- str_detect(term, "years_centered:risk_score")
      adj[idx_inter] <- p.adjust(p.value[idx_inter], method = "fdr")
      adj
    },
    p_adj_fmt   = if_else(!is.na(p_adj),
                          if_else(p_adj < 0.001, "<0.001", sprintf("%.3f", p_adj)),
                          "\u2014"),
    p_adj_stars = if_else(!is.na(p_adj) & p_adj < 0.05, "sig", ""),
    stars       = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      TRUE            ~ ""
    ),
    model_short = if_else(str_detect(model, "CogDrisk"), "CogDrisk", "LIBRA2"),
    biomarker   = case_when(
      outcome == "ab_ratio"               ~ "A\u03b242/40",
      outcome == "gfap_log"               ~ "GFAP",
      outcome == "nfl_log"                ~ "NfL",
      outcome == "ptau181_log"            ~ "ptau-181",
      outcome == "ptau217_log"            ~ "ptau-217",
      outcome == "ptau217_ab42_ratio_log" ~ "ptau-217/A\u03b242 ratio"
    ),
    term_label  = case_when(
      term == "(Intercept)"                                         ~ "Intercept",
      term == "years_centered"                                      ~ "Time (years)",
      str_detect(term, "^risk_score") & !str_detect(term, ":")     ~ "Composite Risk Score",
      str_detect(term, "years_centered:risk_score")                 ~ "Time \u00d7 Composite Risk Score",
      term == "baseline_age"                                        ~ "Age",
      term == "sexMale"                                             ~ "Sex (Male)",
      term == "educ"                                                ~ "Education (years)",
      str_detect(term, "years_centered:baseline_age")               ~ "Time \u00d7 Age",
      str_detect(term, "years_centered:sexMale")                    ~ "Time \u00d7 Sex (Male)",
      str_detect(term, "years_centered:educ")                       ~ "Time \u00d7 Education",
      TRUE ~ term
    )
  )

interact_combined_wide <- interact_combined_raw %>%
  select(model_short, biomarker, term_label, beta_se, ci, p_fmt, p_adj_fmt, p_adj_stars, stars) %>%
  pivot_wider(
    id_cols     = c(biomarker, term_label),
    names_from  = model_short,
    values_from = c(beta_se, ci, p_fmt, p_adj_fmt, p_adj_stars, stars),
    names_glue  = "{model_short}_{.value}"
  ) %>%
  mutate(
    across(c(CogDrisk_beta_se, CogDrisk_ci, CogDrisk_p_fmt, CogDrisk_p_adj_fmt,
             LIBRA2_beta_se,   LIBRA2_ci,   LIBRA2_p_fmt,   LIBRA2_p_adj_fmt),
           ~replace_na(., "\u2014")),
    CogDrisk_stars       = replace_na(CogDrisk_stars, ""),
    CogDrisk_p_adj_stars = replace_na(CogDrisk_p_adj_stars, ""),
    LIBRA2_stars         = replace_na(LIBRA2_stars, ""),
    LIBRA2_p_adj_stars   = replace_na(LIBRA2_p_adj_stars, "")
  )

# Section separator rows per biomarker
empty_interact <- function(df) {
  mutate(df,
    CogDrisk_beta_se = "", CogDrisk_ci = "", CogDrisk_p_fmt = "",
    CogDrisk_p_adj_fmt = "", CogDrisk_p_adj_stars = "", CogDrisk_stars = "",
    LIBRA2_beta_se   = "", LIBRA2_ci   = "", LIBRA2_p_fmt   = "",
    LIBRA2_p_adj_fmt = "", LIBRA2_p_adj_stars   = "", LIBRA2_stars   = ""
  )
}

covariate_header_interact <- interact_combined_wide %>%
  distinct(biomarker) %>%
  mutate(term_label = "Covariates") %>%
  empty_interact()

covariate_interact_header <- interact_combined_wide %>%
  distinct(biomarker) %>%
  mutate(term_label = "Time \u00d7 Covariates") %>%
  empty_interact()

interact_combined <- bind_rows(
    interact_combined_wide,
    covariate_header_interact,
    covariate_interact_header
  ) %>%
  mutate(
    biomarker  = factor(biomarker,  levels = biomarker_order_lme),
    term_label = factor(term_label, levels = term_order_lme_interact)
  ) %>%
  arrange(biomarker, term_label) %>%
  select(biomarker, term_label,
         CogDrisk_beta_se, CogDrisk_ci, CogDrisk_p_fmt, CogDrisk_p_adj_fmt,
         CogDrisk_p_adj_stars, CogDrisk_stars,
         LIBRA2_beta_se,   LIBRA2_ci,   LIBRA2_p_fmt,   LIBRA2_p_adj_fmt,
         LIBRA2_p_adj_stars, LIBRA2_stars)

interact_gt <- interact_combined %>%
  gt(groupname_col = "biomarker") %>%
  tab_header(
    title    = md("**Table S5.** Full Time Interaction Model: Composite Risk Score and Covariate Effects on Plasma Biomarker Trajectories"),
    subtitle = md("Linear mixed-effects models. LIBRA2 additionally adjusted for time interactions with age, sex, and education.")
  ) %>%
  tab_spanner(label = md("**CogDrisk**"), columns = starts_with("CogDrisk_")) %>%
  tab_spanner(label = md("**LIBRA2**"),   columns = starts_with("LIBRA2_")) %>%
  cols_label(
    term_label           = md("**Predictor**"),
    CogDrisk_beta_se     = md("**\u03b2 (SE)**"),
    CogDrisk_ci          = md("**95% CI**"),
    CogDrisk_p_fmt       = md("**p**"),
    CogDrisk_p_adj_fmt   = md("**q**"),
    LIBRA2_beta_se       = md("**\u03b2 (SE)**"),
    LIBRA2_ci            = md("**95% CI**"),
    LIBRA2_p_fmt         = md("**p**"),
    LIBRA2_p_adj_fmt     = md("**q**")
  ) %>%
  cols_hide(columns = c(CogDrisk_stars, CogDrisk_p_adj_stars,
                        LIBRA2_stars,   LIBRA2_p_adj_stars)) %>%
  # Biomarker row-group headers
  tab_style(
    style     = list(cell_text(weight = "bold"), cell_fill(color = "#f8f9fa")),
    locations = cells_row_groups()
  ) %>%
  # Section header rows ("Covariates", "Time × Covariates") — bold italic, grey
  tab_style(
    style     = list(cell_text(weight = "bold", style = "italic"),
                     cell_fill(color = "#f0f2f4")),
    locations = cells_body(
      rows = term_label %in% c("Covariates", "Time \u00d7 Covariates")
    )
  ) %>%
  # Indent covariate terms
  tab_style(
    style     = cell_text(indent = px(16)),
    locations = cells_body(
      columns = term_label,
      rows    = term_label %in% c(covariate_terms_interact, covariate_interact_terms)
    )
  ) %>%
  # Unadjusted p significance — light green
  tab_style(
    style     = cell_fill(color = "#e8f4e8"),
    locations = cells_body(columns = CogDrisk_p_fmt,
                           rows    = CogDrisk_stars %in% c("*", "**", "***"))
  ) %>%
  tab_style(
    style     = cell_fill(color = "#e8f4e8"),
    locations = cells_body(columns = LIBRA2_p_fmt,
                           rows    = LIBRA2_stars %in% c("*", "**", "***"))
  ) %>%
  # FDR-adjusted q significance — darker green
  tab_style(
    style     = cell_fill(color = "#c8e6c9"),
    locations = cells_body(columns = CogDrisk_p_adj_fmt,
                           rows    = CogDrisk_p_adj_stars == "sig")
  ) %>%
  tab_style(
    style     = cell_fill(color = "#c8e6c9"),
    locations = cells_body(columns = LIBRA2_p_adj_fmt,
                           rows    = LIBRA2_p_adj_stars == "sig")
  ) %>%
  tab_footnote(
    footnote  = md("p: unadjusted; q: Benjamini-Hochberg FDR-adjusted across 12 tests (6 biomarkers \u00d7 2 models) within each term family. q is reported for: (1) the composite risk score level effect and (2) the time \u00d7 composite risk score interaction, each corrected in a separate family of 12. Covariate terms are not corrected (\u2014). Light green: p\u202f<\u202f0.05; dark green: q\u202f<\u202f0.05."),
    locations = cells_column_spanners()
  ) %>%
  cols_align(align = "left",   columns = term_label) %>%
  cols_align(align = "center", columns = c(starts_with("CogDrisk_"), starts_with("LIBRA2_"))) %>%
  row_group_order(groups = biomarker_order_lme)

interact_gt

# gtsave(interact_gt, filename = file.path(AIM1_OUTPUT_PATH$tables, "s_table5_lme_interact.docx"))
```

### Multivariable model results

```{r multivar-plot}
#| fig-width: 12
#| fig-height: 10

# ── Shared helpers ─────────────────────────────────────────────────────────────
mv_bm_label <- function(outcome) {
  case_when(
    outcome == "ab_ratio"               ~ "Aβ42/40",
    outcome == "gfap_log"               ~ "GFAP (log)",
    outcome == "nfl_log"                ~ "NfL (log)",
    outcome == "ptau181_log"            ~ "p-tau181 (log)",
    outcome == "ptau217_log"            ~ "p-tau217 (log)",
    outcome == "ptau217_ab42_ratio_log" ~ "p-tau217/Aβ42 (log)"
  )
}

mv_bm_order <- c(
  "Aβ42/40", "GFAP (log)", "NfL (log)",
  "p-tau181 (log)", "p-tau217 (log)", "p-tau217/Aβ42 (log)"
)

add_mv_plot_cols <- function(df, primary_labels) {
  df %>%
    mutate(
      conf.low    = estimate - 1.96 * std.error,
      conf.high   = estimate + 1.96 * std.error,
      # FDR on primary predictor terms only; covariates left as NA
      p_adj = {
        adj      <- rep(NA_real_, n())
        idx      <- as.character(display_label) %in% primary_labels
        adj[idx] <- p.adjust(p.value[idx], method = "fdr")
        adj
      },
      sig = factor(
        case_when(
          !is.na(p_adj) & p_adj < 0.05 ~ "Significant (FDR-adjusted)",
          p.value < 0.05               ~ "Significant (non-adjusted)",
          TRUE                         ~ "Not significant"
        ),
        levels = c("Significant (FDR-adjusted)", "Significant (non-adjusted)", "Not significant")
      ),
      biomarker   = factor(mv_bm_label(outcome), levels = mv_bm_order),
      model_short = factor(
        if_else(str_detect(model, "CogDrisk"), "CogDrisk", "LIBRA2"),
        levels = c("LIBRA2", "CogDrisk")
      )
    )
}

mv_coef_layers <- list(
  geom_vline(xintercept = 0, linetype = 2, color = "grey40"),
  scale_alpha_manual(
    values = c(
      "Significant (FDR-adjusted)" = 1,
      "Significant (non-adjusted)" = 1,
      "Not significant"            = 0.25
    ),
    name = "Significance"
  ),
  scale_linewidth_manual(
    values = c(
      "Significant (FDR-adjusted)" = 1.5,
      "Significant (non-adjusted)" = 0.8,
      "Not significant"            = 0.5
    ),
    name = "Significance"
  ),
  scale_color_manual(
    values = c("CogDrisk" = "#374E55FF", "LIBRA2" = "#DF8F44FF"),
    breaks = c("CogDrisk", "LIBRA2"),
    name   = "Model"
  ),
  facet_wrap(~biomarker, ncol = 3, scales = "free_x"),
  theme_bw(base_size = 12),
  theme(
    legend.position  = "bottom",
    legend.title     = element_text(size = 10, face = "bold"),
    legend.text      = element_text(size = 9),
    panel.grid.minor = element_blank(),
    strip.background = element_rect(fill = "grey92"),
    strip.text       = element_text(face = "bold"),
    axis.text        = element_text(size = 10)
  )
)

# ── Multivariable time interaction coefficient plot ────────────────────────────
inter_term_levels <- rev(c(
  "Time × APOE ε4+", "Time × PXS",
  "Time × GRS",           "Time × FHx",
  "Time × Age", "Time × Sex (Male)", "Time × Education"
))

mv_inter_plot_dat <- lme_multivar_interact %>%
  select(model, outcome, dataf) %>% unnest(dataf) %>%
  filter(effect == "fixed", str_detect(term, ":")) %>%
  mutate(
    display_label = factor(case_when(
      str_detect(term, "apoe_burden")                 ~ "Time × APOE ε4+",
      str_detect(term, "score_cogd_z|score_libra2_z") ~ "Time × PXS",
      str_detect(term, "\\bGRS\\b")                   ~ "Time × GRS",
      str_detect(term, "fhx_burden")                  ~ "Time × FHx",
      str_detect(term, "baseline_age")                ~ "Time × Age",
      str_detect(term, "sexMale")                     ~ "Time × Sex (Male)",
      str_detect(term, "\\beduc\\b")                  ~ "Time × Education",
      TRUE                                            ~ NA_character_
    ), levels = inter_term_levels)
  ) %>%
  filter(!is.na(display_label)) %>%
  add_mv_plot_cols(primary_labels = c(
    "Time × APOE ε4+", "Time × PXS",
    "Time × GRS",           "Time × FHx"
  ))

p_mv_inter <- ggplot(mv_inter_plot_dat,
  aes(x = estimate, y = display_label,
      xmin = conf.low, xmax = conf.high,
      color = model_short, alpha = sig, group = model_short)
) +
  geom_point(size = 2.5, position = position_dodge(width = 0.6)) +
  geom_errorbarh(aes(linewidth = sig), position = position_dodge(width = 0.6), height = 0.3) +
  labs(x = "β (95% CI)", y = NULL,
       title = "Multivariable LME: Time × Risk Interactions",
       subtitle = "Additional change in biomarker slope per unit increase in predictor") +
  mv_coef_layers

print(p_mv_inter)
```

```{r multivar-results}
# ── Row ordering ───────────────────────────────────────────────────────────────
# row_key: unique sort key; display_label: text shown in table
mv_row_order <- c(
  "inter_APOE \u03b54+", "inter_PXS", "inter_GRS", "inter_FHx",
  "inter_Covariates",
  "inter_Age", "inter_Sex (Male)", "inter_Education (years)",
  "inter_Time \u00d7",
  "inter_Time \u00d7 APOE \u03b54+", "inter_Time \u00d7 PXS",
  "inter_Time \u00d7 GRS",           "inter_Time \u00d7 FHx",
  "inter_Time \u00d7 Covariates",
  "inter_Time \u00d7 Age", "inter_Time \u00d7 Sex (Male)", "inter_Time \u00d7 Education"
)

mv_section_row_keys <- c(
  "inter_Covariates",
  "inter_Time \u00d7", "inter_Time \u00d7 Covariates"
)

# display_label values that receive section-header styling
mv_section_labels <- c(
  "Covariates", "Time \u00d7", "Time \u00d7 Covariates"
)

# display_label values that get indented
mv_indent_labels <- c(
  "Age", "Sex (Male)", "Education (years)",
  "Time \u00d7 Age", "Time \u00d7 Sex (Male)", "Time \u00d7 Education"
)

# ── Helper: outcome → biomarker label ─────────────────────────────────────────
bm_label <- function(outcome) {
  case_when(
    outcome == "ab_ratio"               ~ "A\u03b242/40",
    outcome == "gfap_log"               ~ "GFAP",
    outcome == "nfl_log"                ~ "NfL",
    outcome == "ptau181_log"            ~ "ptau-181",
    outcome == "ptau217_log"            ~ "ptau-217",
    outcome == "ptau217_ab42_ratio_log" ~ "ptau-217/A\u03b242 ratio"
  )
}

# ── Interaction model: ALL fixed effects (main effects + interaction terms) ────
mv_inter_raw <- lme_multivar_interact %>%
  select(model, outcome, dataf) %>% unnest(dataf) %>%
  filter(effect == "fixed", !term %in% c("(Intercept)", "years_centered")) %>%
  mutate(
    model_short   = if_else(str_detect(model, "CogDrisk"), "CogDrisk", "LIBRA2"),
    biomarker     = bm_label(outcome),
    display_label = case_when(
      # Main effect terms within interaction model (no colon)
      !str_detect(term, ":") & term == "apoe_burden"                          ~ "APOE \u03b54+",
      !str_detect(term, ":") & term %in% c("score_cogd_z", "score_libra2_z") ~ "PXS",
      !str_detect(term, ":") & term == "GRS"                                  ~ "GRS",
      !str_detect(term, ":") & term == "fhx_burden"                           ~ "FHx",
      !str_detect(term, ":") & term == "baseline_age"                         ~ "Age",
      !str_detect(term, ":") & term == "sexMale"                              ~ "Sex (Male)",
      !str_detect(term, ":") & term == "educ"                                 ~ "Education (years)",
      # Interaction terms (has colon)
      str_detect(term, ":") & str_detect(term, "apoe_burden")                 ~ "Time \u00d7 APOE \u03b54+",
      str_detect(term, ":") & str_detect(term, "score_cogd_z|score_libra2_z") ~ "Time \u00d7 PXS",
      str_detect(term, ":") & str_detect(term, "\\bGRS\\b")                   ~ "Time \u00d7 GRS",
      str_detect(term, ":") & str_detect(term, "fhx_burden")                  ~ "Time \u00d7 FHx",
      str_detect(term, ":") & str_detect(term, "baseline_age")                ~ "Time \u00d7 Age",
      str_detect(term, ":") & str_detect(term, "sexMale")                     ~ "Time \u00d7 Sex (Male)",
      str_detect(term, ":") & str_detect(term, "\\beduc\\b")                  ~ "Time \u00d7 Education",
      TRUE ~ NA_character_
    ),
    row_key = paste0("inter_", display_label)
  ) %>%
  filter(!is.na(display_label))

# ── Compute statistics ────────────────────────────────────────────────────────
mv_raw <- mv_inter_raw %>%
  mutate(
    conf.low      = estimate - 1.96 * std.error,
    conf.high     = estimate + 1.96 * std.error,
    beta_se       = sprintf("%.3f (%.3f)", estimate, std.error),
    ci            = sprintf("[%.3f, %.3f]", conf.low, conf.high),
    p_fmt         = if_else(p.value < 0.001, "<0.001", sprintf("%.3f", p.value)),
    stars         = case_when(
      p.value < 0.001 ~ "***", p.value < 0.01 ~ "**",
      p.value < 0.05  ~ "*",   TRUE            ~ ""
    ),
    # FDR: two separate families — level effects (APOE, PXS, GRS, FHx) and
    # time × interaction terms; covariates excluded from correction.
    p_adj = {
      adj            <- rep(NA_real_, n())
      idx_level      <- display_label %in% c("APOE \u03b54+", "PXS", "GRS", "FHx")
      idx_inter      <- display_label %in% c("Time \u00d7 APOE \u03b54+", "Time \u00d7 PXS",
                                             "Time \u00d7 GRS",             "Time \u00d7 FHx")
      adj[idx_level] <- p.adjust(p.value[idx_level], method = "fdr")
      adj[idx_inter] <- p.adjust(p.value[idx_inter], method = "fdr")
      adj
    },
    p_adj_fmt   = if_else(!is.na(p_adj),
                          if_else(p_adj < 0.001, "<0.001", sprintf("%.3f", p_adj)),
                          "\u2014"),
    p_adj_stars = if_else(!is.na(p_adj) & p_adj < 0.05, "sig", "")
  )

# ── Pivot: CogDrisk vs LIBRA2 as columns ─────────────────────────────────────
mv_wide <- mv_raw %>%
  select(row_key, display_label, biomarker, model_short,
         beta_se, ci, p_fmt, p_adj_fmt, p_adj_stars, stars) %>%
  pivot_wider(
    id_cols     = c(biomarker, row_key, display_label),
    names_from  = model_short,
    values_from = c(beta_se, ci, p_fmt, p_adj_fmt, p_adj_stars, stars),
    names_glue  = "{model_short}_{.value}",
    values_fn   = first
  ) %>%
  mutate(
    across(c(CogDrisk_beta_se, CogDrisk_ci, CogDrisk_p_fmt, CogDrisk_p_adj_fmt,
             LIBRA2_beta_se,   LIBRA2_ci,   LIBRA2_p_fmt,   LIBRA2_p_adj_fmt),
           ~replace_na(., "\u2014")),
    across(c(CogDrisk_stars, CogDrisk_p_adj_stars, LIBRA2_stars, LIBRA2_p_adj_stars),
           ~replace_na(., ""))
  )

# ── Section header rows ────────────────────────────────────────────────────────
mv_section_hdr <- expand_grid(
    row_key   = mv_section_row_keys,
    biomarker = unique(mv_wide$biomarker)
  ) %>%
  mutate(
    display_label = case_when(
      row_key == "inter_Covariates"             ~ "Covariates",
      row_key == "inter_Time \u00d7"            ~ "Time \u00d7",
      row_key == "inter_Time \u00d7 Covariates" ~ "Time \u00d7 Covariates"
    ),
    CogDrisk_beta_se = "", CogDrisk_ci = "", CogDrisk_p_fmt = "", CogDrisk_p_adj_fmt = "",
    CogDrisk_stars   = "", CogDrisk_p_adj_stars = "",
    LIBRA2_beta_se   = "", LIBRA2_ci   = "", LIBRA2_p_fmt   = "", LIBRA2_p_adj_fmt   = "",
    LIBRA2_stars     = "", LIBRA2_p_adj_stars   = ""
  )

# ── Sort and combine ──────────────────────────────────────────────────────────
mv_combined <- bind_rows(mv_wide, mv_section_hdr) %>%
  mutate(
    biomarker = factor(biomarker, levels = biomarker_order_lme),
    row_key   = factor(row_key,   levels = mv_row_order)
  ) %>%
  arrange(biomarker, row_key) %>%
  select(biomarker, display_label,
         CogDrisk_beta_se, CogDrisk_ci, CogDrisk_p_fmt, CogDrisk_p_adj_fmt,
         CogDrisk_p_adj_stars, CogDrisk_stars,
         LIBRA2_beta_se,   LIBRA2_ci,   LIBRA2_p_fmt,   LIBRA2_p_adj_fmt,
         LIBRA2_p_adj_stars,  LIBRA2_stars)

# ── gt table ──────────────────────────────────────────────────────────────────
mv_gt <- mv_combined %>%
  gt(groupname_col = "biomarker") %>%
  tab_header(
    title    = md("**Table S6.** Multivariable LME Time Interaction Model: Individual Risk Predictors and Their Association with Plasma Biomarker Trajectories"),
    subtitle = md("APOE \u03b54 burden, z-scored composite risk score (PXS), polygenic risk score (GRS), and family history burden (FHx). LIBRA2 models additionally adjusted for age, sex, and education.")
  ) %>%
  tab_spanner(label = md("**CogDrisk**"), columns = starts_with("CogDrisk_")) %>%
  tab_spanner(label = md("**LIBRA2**"),   columns = starts_with("LIBRA2_")) %>%
  cols_label(
    display_label      = md("**Predictor**"),
    CogDrisk_beta_se   = md("**\u03b2 (SE)**"),
    CogDrisk_ci        = md("**95% CI**"),
    CogDrisk_p_fmt     = md("**p**"),
    CogDrisk_p_adj_fmt = md("**q**"),
    LIBRA2_beta_se     = md("**\u03b2 (SE)**"),
    LIBRA2_ci          = md("**95% CI**"),
    LIBRA2_p_fmt       = md("**p**"),
    LIBRA2_p_adj_fmt   = md("**q**")
  ) %>%
  cols_hide(columns = c(CogDrisk_stars, CogDrisk_p_adj_stars,
                        LIBRA2_stars,   LIBRA2_p_adj_stars)) %>%
  # Biomarker row-group headers
  tab_style(style = list(cell_text(weight = "bold"), cell_fill(color = "#f8f9fa")),
            locations = cells_row_groups()) %>%
  # Section header rows — bold italic grey
  tab_style(
    style     = list(cell_text(weight = "bold", style = "italic"),
                     cell_fill(color = "#f0f2f4")),
    locations = cells_body(rows = display_label %in% mv_section_labels)
  ) %>%
  # Indent covariate / time-interaction-covariate rows
  tab_style(
    style     = cell_text(indent = px(16)),
    locations = cells_body(columns = display_label,
                           rows    = display_label %in% mv_indent_labels)
  ) %>%
  # Unadjusted p — light green
  tab_style(style = cell_fill(color = "#e8f4e8"),
            locations = cells_body(columns = CogDrisk_p_fmt,
                                   rows    = CogDrisk_stars %in% c("*","**","***"))) %>%
  tab_style(style = cell_fill(color = "#e8f4e8"),
            locations = cells_body(columns = LIBRA2_p_fmt,
                                   rows    = LIBRA2_stars %in% c("*","**","***"))) %>%
  # FDR q — darker green
  tab_style(style = cell_fill(color = "#c8e6c9"),
            locations = cells_body(columns = CogDrisk_p_adj_fmt,
                                   rows    = CogDrisk_p_adj_stars == "sig")) %>%
  tab_style(style = cell_fill(color = "#c8e6c9"),
            locations = cells_body(columns = LIBRA2_p_adj_fmt,
                                   rows    = LIBRA2_p_adj_stars == "sig")) %>%
  tab_footnote(
    footnote = md(paste0(
      "p: unadjusted; q: Benjamini-Hochberg FDR-adjusted. ",
      "PXS = z-scored composite risk score. GRS = polygenic risk score. FHx = family history burden. ",
      "\u03b2: change in biomarker per 1-unit increase in predictor. ",
      "Time \u00d7 \u03b2: additional change in biomarker slope per 1-unit increase in predictor. ",
      "Light green: p\u202f<\u202f0.05; dark green: q\u202f<\u202f0.05. ",
      "\u2014 = not applicable (CogDrisk model does not adjust for demographic covariates)."
    )),
    locations = cells_column_spanners()
  ) %>%
  cols_align(align = "left",   columns = display_label) %>%
  cols_align(align = "center", columns = -display_label) %>%
  row_group_order(groups = biomarker_order_lme)

mv_gt

# gtsave(mv_gt, filename = file.path(AIM1_OUTPUT_PATH$tables, "s_table6_lme_multivar.docx"))
```

# Trajectory plots

**Methodology: Generating Predicted Trajectories**

The predicted biomarker trajectories are generated from the LME models fitted above using the following approach:

**1. Model Specification:**
- **Base model**: `biomarker ~ years_centered × risk_score + (1 | participant_ID)`
- The time × risk score interaction allows trajectories to have different slopes for each risk level
- Random intercepts account for between-participant variability in baseline biomarker levels

**2. Prediction Grid Construction:**
- For each biomarker, a prediction grid is created spanning:
  - **Time range**: From baseline (0 years) to the maximum observed follow-up for that biomarker
  - **Risk scores**: All possible composite risk levels (0, 1, 2, 3, 4)
  - **Time resolution**: Predictions at 0.5-year intervals for smooth trajectory lines
- Time values are grand-mean centered before prediction (where 0 represents ~2.6 years of follow-up), then transformed back to original scale for plotting

**3. Population-Level Predictions:**
- Predictions use `predict(..., re.form = NA)` to obtain **marginal** (population-average) predictions
- This marginalizes over random effects, showing the expected trajectory for an "average" participant at each risk level
- Individual-level predictions would include random intercepts, but would obscure population-level patterns

**4. Biomarker-Specific Time Ranges:**
- ptau-217: predictions extend to ~11 years (maximum observed follow-up)
- Other biomarkers (ptau-181, GFAP, NfL, Aβ42/40, ptau-217/Aβ42): predictions extend to ~4.6 years
- This prevents extrapolation beyond observed data ranges

**5. Covariate Values (LIBRA2 only):**
- LIBRA2 models include demographic covariates with time interactions
- Predictions are generated for a reference profile:
  - Baseline age: sample mean
  - Sex: Female
  - Education: sample mean years
- Trajectories for other demographic profiles would differ due to covariate × time interactions

## CogDrisk

```{r trajectories-cogdrisk}
#| fig-width: 14
#| fig-height: 10
#| fig-dpi: 96

# Calculate maximum follow-up time per biomarker
biomarker_max_years <- all_dat %>%
  pivot_longer(
    cols = all_of(BIOMARKERS_TRANS),
    names_to = "outcome",
    values_to = "value"
  ) %>%
  filter(!is.na(value)) %>%
  group_by(outcome) %>%
  summarise(max_years = max(years, na.rm = TRUE))

# Grand mean of years (used to center time in the models)
mean_years <- mean(all_dat$years, na.rm = TRUE)

# Generate predictions from interaction model
cogdrisk_interact_models <- lme_time_interact %>%
  filter(str_detect(model, "CogDrisk"))

# Create biomarker-specific prediction data and generate predictions
cogdrisk_predictions <- cogdrisk_interact_models %>%
  left_join(biomarker_max_years, by = "outcome") %>%
  mutate(
    predictions = map2(res, max_years, ~{
      # Create prediction grid specific to this biomarker's time range
      pred_data <- expand_grid(
        years_centered = seq(-mean_years, .y - mean_years, by = 0.5),
        risk_score_cogd = 0:4
      ) %>%
        mutate(years = years_centered + mean_years)

      # Generate predictions
      pred_data$predicted <- predict(.x, newdata = pred_data, re.form = NA)
      pred_data
    })
  ) %>%
  select(outcome, predictions) %>%
  unnest(predictions)

# Prepare observed data (kept for downstream chunks)
all_dat_long <- all_dat %>%
  pivot_longer(
    cols = all_of(BIOMARKERS_TRANS),
    names_to = "outcome",
    values_to = "observed"
  )

# Readable outcome labels (ordered for consistent facet display)
outcome_labels <- c(
  ab_ratio               = "A\u03b242/40",
  gfap_log               = "GFAP",
  nfl_log                = "NfL",
  ptau181_log            = "ptau-181",
  ptau217_log            = "ptau-217",
  ptau217_ab42_ratio_log = "ptau-217/A\u03b242 ratio"
)

# Add readable labels to plotting datasets (separate vars; risk_dat_long_plot unchanged)
# Drop NA observed values so late-visit rows (e.g. GFAP=NA at year 10) don't
# expand the x-scale for panels where that biomarker wasn't measured.
risk_dat_long_plot_lbl <- all_dat_long %>%
  filter(!is.na(observed)) %>%
  mutate(outcome_label = factor(outcome_labels[outcome], levels = outcome_labels))

cogdrisk_pred_lbl <- cogdrisk_predictions %>%
  mutate(outcome_label = factor(outcome_labels[outcome], levels = outcome_labels))

# Interaction: time × risk score (divergence in slope)
cogd_annot_inter <- lme_results %>%
  filter(model_type == "Time Interaction",
         risk_type  == "CogDrisk",
         str_detect(term, "years_centered:risk_score")) %>%
  mutate(
    outcome_label = factor(outcome_labels[outcome], levels = outcome_labels),
    p_fmt = if_else(p.value < 0.001, "< 0.001", as.character(p.value)),
    annot = paste0("Time \u00d7 Risk: \u03b2 = ", estimate,
                   " [", conf.low, ", ", conf.high, "], \n  p = ", p_fmt, stars)
  )

# Build descriptive caption
n_pts <- length(unique(all_dat_long$CONP_ID))
n_obs <- nrow(all_dat_long)

caption_text <- paste0(
  "Cohort: PREVENT-AD. N = ", n_pts, "; Observations = ", n_obs, "\n",
  "Risk indicators: APOE e4+; Family history (\u22652 first-degree relatives);  Genetic risk score (top quartile); ",
  "CogDrisk > 1 SD above mean. Composite score: 0\u20134 scale (sum of binary indicators).\n",
  "Lines = model-predicted population trajectories (LMM: biomarker ~ years \u00d7 risk_score + (1 | ID)). ",
  "Axis scales differ by biomarker to reflect observed data range."
)

# Plot trajectories
p_cogd_traj <- ggplot() +
  geom_point(data = risk_dat_long_plot_lbl,
             aes(x = years, y = observed, color = factor(risk_score_cogd)),
             alpha = 0.2, size = 0.8) +
  geom_line(data = cogdrisk_pred_lbl,
            aes(x = years, y = predicted,
                color = factor(risk_score_cogd),
                group = risk_score_cogd),
            linewidth = 1.2) +
  geom_label(
    data = cogd_annot_inter,
    aes(x = -Inf, y = Inf, label = annot),
    inherit.aes = FALSE,
    hjust = -0.04, vjust = 1.75,
    size = 5, color = "grey20",
    label.padding = unit(0.2, "lines")
  ) +
  scale_color_viridis_d(option = "plasma", begin = 0.2, end = 0.9,
                        name = "# Risk Indicators") +
  scale_x_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.05))) +
  scale_y_continuous(expand = expansion(mult = c(0.02, 0.05))) +
  facet_wrap(~outcome_label, scales = "free", ncol = 3) +
  labs(
    x = "Years from Baseline",
    y = "Biomarker Level"
  ) +
  theme_bw(base_size = 12) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold"),
    strip.background = element_rect(fill = "grey90"),
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    # plot.subtitle = element_text(size = 12, hjust = 0.5),
    # plot.caption = element_text(size = 9, hjust = 0, lineheight = 1.2,
    legend.title     = element_text(size = 10, face = "bold"),
    legend.text      = element_text(size = 9),
    panel.grid.minor = element_blank(),
    plot.subtitle    = element_text(size = 11, hjust = 0.5, color = "grey40"),
    axis.text        = element_text(size = 11)
  )

print(p_cogd_traj)

ggsave(file.path(AIM1_OUTPUT_PATH$figures, "fig5_longitudinal_CogDrisk_trajectories.png"),
       p_cogd_traj, width = 14, height = 10, dpi = 250)
```

## LIBRA2

```{r trajectories-libra2}
#| fig-width: 14
#| fig-height: 10
#| fig-dpi: 96

# Generate predictions from interaction model
libra2_interact_models <- lme_time_interact %>%
  filter(str_detect(model, "LIBRA2"))

# Create biomarker-specific prediction data and generate predictions
libra2_predictions <- libra2_interact_models %>%
  left_join(biomarker_max_years, by = "outcome") %>%
  mutate(
    predictions = map2(res, max_years, ~{
      # Create prediction grid specific to this biomarker's time range
      pred_data <- expand_grid(
        years_centered = seq(-mean_years, .y - mean_years, by = 0.5),
        risk_score_libra2 = 0:4,
        baseline_age = mean(all_dat$baseline_age, na.rm = TRUE),
        sex = "Female",
        educ = mean(all_dat$educ, na.rm = TRUE)
      ) %>%
        mutate(years = years_centered + mean_years)

      # Generate predictions
      pred_data$predicted <- predict(.x, newdata = pred_data, re.form = NA)
      pred_data
    })
  ) %>%
  select(outcome, predictions) %>%
  unnest(predictions)

# Add readable labels; drop NA observed to avoid x-scale inflation
libra2_obs_lbl <- all_dat_long %>%
  filter(!is.na(observed)) %>%
  mutate(outcome_label = factor(outcome_labels[outcome], levels = outcome_labels))

libra2_pred_lbl <- libra2_predictions %>%
  mutate(outcome_label = factor(outcome_labels[outcome], levels = outcome_labels))

# Per-panel annotation: time × risk score interaction
libra2_annot_inter <- lme_results %>%
  filter(model_type == "Time Interaction",
         risk_type  == "LIBRA2",
         str_detect(term, "years_centered:risk_score")) %>%
  mutate(
    outcome_label = factor(outcome_labels[outcome], levels = outcome_labels),
    p_fmt = if_else(p.value < 0.001, "< 0.001", as.character(p.value)),
    annot = paste0("Time \u00d7 Risk: \u03b2 = ", estimate,
                   " [", conf.low, ", ", conf.high, "], \n  p = ", p_fmt, stars)
  )

caption_text_libra2 <- paste0(
  "Cohort: PREVENT-AD. N = ", n_pts, "; Observations = ", n_obs, "\n",
  "LIBRA2 composite risk score (0\u20134). Reference profile: female, mean age, mean education.\n",
  "Lines = model-predicted population trajectories (LMM: biomarker ~ years \u00d7 risk_score + covariates + (1 | ID)). ",
  "Axis scales differ by biomarker to reflect observed data range."
)

# Plot trajectories
p_libra_traj <- ggplot() +
  geom_point(data = libra2_obs_lbl,
             aes(x = years, y = observed, color = factor(risk_score_libra2)),
             alpha = 0.2, size = 0.8) +
  geom_line(data = libra2_pred_lbl,
            aes(x = years, y = predicted,
                color = factor(risk_score_libra2),
                group = risk_score_libra2),
            linewidth = 1.2) +
  geom_label(
    data        = libra2_annot_inter,
    aes(x = -Inf, y = Inf, label = annot),
    inherit.aes = FALSE,
    hjust = -0.04, vjust = 1.75,
    size = 5, color = "grey20",
    label.padding = unit(0.2, "lines")
  ) +
  scale_color_viridis_d(option = "plasma", begin = 0.2, end = 0.9,
                        name = "# Risk Indicators") +
  scale_x_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.05))) +
  scale_y_continuous(expand = expansion(mult = c(0.02, 0.05))) +
  facet_wrap(~outcome_label, scales = "free", ncol = 3) +
  labs(
    x = "Years from Baseline",
    y = "Biomarker Level"#,
    #title = "Figure 5. Predicted Plasma Biomarker Trajectories by LIBRA2-based Composite Risk Score"
    # caption = caption_text_libra2
  ) +
  theme_bw(base_size = 12) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold"),
    strip.background = element_rect(fill = "grey90"),
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    # plot.subtitle = element_text(size = 12, hjust = 0.5),
    # plot.caption = element_text(size = 9, hjust = 0, lineheight = 1.2,
    legend.title     = element_text(size = 10, face = "bold"),
    legend.text      = element_text(size = 9),
    panel.grid.minor = element_blank(),
    plot.subtitle    = element_text(size = 11, hjust = 0.5, color = "grey40"),
    axis.text        = element_text(size = 11)
  )

print(p_libra_traj)

ggsave(file.path(AIM1_OUTPUT_PATH$figures, "fig6_longitudinal_LIBRA2_trajectories.png"),
       p_libra_traj, width = 14, height = 10, dpi = 250)
```

## Focused Plot: CogDrisk ptau Trajectories

```{r ptau-cogdrisk-focused-plot}
#| fig-width: 12
#| fig-height: 8

# Filter data for ptau217 and ptau181
ptau217_obs <- risk_dat_long_plot %>%
  filter(outcome == "ptau217_log")

ptau181_obs <- risk_dat_long_plot %>%
  filter(outcome == "ptau181_log")

ptau217_pred <- cogdrisk_predictions %>%
  filter(outcome == "ptau217_log")

ptau181_pred <- cogdrisk_predictions %>%
  filter(outcome == "ptau181_log")

# Create separate plots for each biomarker with appropriate x-axis limits
p_ptau217 <- ggplot() +
  geom_point(data = ptau217_obs,
             aes(x = years, y = observed, color = factor(risk_score_cogd)),
             alpha = 0.3, size = 1.5) +
  geom_line(data = ptau217_pred,
            aes(x = years, y = predicted,
                color = factor(risk_score_cogd),
                group = risk_score_cogd),
            linewidth = 1.5) +
  scale_color_viridis_d(option = "plasma", begin = 0.2, end = 0.9,
                        name = "Risk Score") +
  scale_x_continuous(limits = c(0, 12), breaks = seq(0, 12, 2)) +
  labs(
    x = "Years from Baseline",
    y = "ptau-217 (log-transformed)",
    title = "ptau-217 (11 years follow-up)"
  ) +
  theme_bw(base_size = 13) +
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold", size = 12)
  )

# Create annotation text with cohort and risk indicator information
n_total <- length(unique(risk_dat_long$CONP_ID))
n_obs <- nrow(risk_dat_long)

annotation_text <- paste0(
  "Cohort: PREVENT-AD. N = ", n_total, "; Observations = ", n_obs, "\n",
  "Risk indicators: FHx: 2+ first-degree relatives with dementia; APOE: e4+ carrier; ",
  "GRS: Genetic risk score > 1 SD above mean; CogDrisk: Modifiable risk score > 1 SD above mean\n",
  "Composite risk score: 0-4 scale (sum of binary indicators)"
)

# Combine plots using patchwork
p_ptau_cogd <- p_ptau217 / p_ptau181 +
  plot_annotation(
    title = "CogDrisk: Phosphorylated Tau Trajectories by Risk Score",
    caption = annotation_text,
    theme = theme(
      plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
      plot.caption = element_text(size = 9, hjust = 0, lineheight = 1.2, margin = margin(t = 10))
    )
  )

print(p_ptau_cogd)
print(p_ptau181)

# Save plot
ggsave(file.path(AIM1_OUTPUT_PATH$figures, "PREVENTAD_ptau_CogDrisk_trajectories.png"),
       p_ptau_cogd, width = 12, height = 9, dpi = 300)
ggsave(file.path(AIM1_OUTPUT_PATH$figures, "PREVENTAD_ptau181_trajectories.png"),
       p_ptau181, width = 12, height = 9, dpi = 300)
```

## Focused Plot: LIBRA2 ptau217 Trajectories 

```{r ptau-combined-focused-plot}
ptau217_obs <- risk_dat_long_plot %>%
  filter(outcome == "ptau217_log")

ptau217_pred_libra2 <- libra2_predictions %>%
  filter(outcome == "ptau217_log")

# Interaction term annotations for ptau217 — LIBRA2
ptau217_annot_libra2 <- dose_response_results %>%
  filter(model_type == "Time Interaction",
         risk_type  == "LIBRA2",
         outcome    == "ptau217_log",
         str_detect(term, "years_centered:risk_score")) %>%
  mutate(
    p_fmt = if_else(p.value < 0.001, "< 0.001", as.character(p.value)),
    annot = paste0("Time\u00d7Risk: \u03b2 = ", estimate,
                   " [", conf.low, ", ", conf.high, "], p = ", p_fmt, stars)
  )

n_ptau217_obs <- nrow(risk_dat_long %>% filter(!is.na(ptau217_log)))

# ptau-217: LIBRA2
p217_libra2 <- ggplot() +
  geom_point(data = ptau217_obs,
             aes(x = years, y = observed, color = factor(risk_score_libra2)),
             alpha = 0.3, size = 1.2) +
  geom_line(data = ptau217_pred_libra2,
            aes(x = years, y = predicted,
                color = factor(risk_score_libra2), group = risk_score_libra2),
            linewidth = 1.5) +
  geom_label(data        = ptau217_annot_libra2,
             aes(x = -Inf, y = Inf, label = annot),
             inherit.aes = FALSE,
             hjust = -0.04, vjust = 2,
             size = 3, color = "grey20",
             label.padding = unit(0.2, "lines")) +
  scale_color_viridis_d(option = "plasma", begin = 0.2, end = 0.9,
                        name = "# Risk Indicators") +
  scale_x_continuous(limits = c(0, 12), breaks = seq(0, 12, 2)) +
  scale_y_continuous(expand = expansion(mult = c(0.02, 0.05))) +
  labs(x = "Years from Baseline", y = NULL,
       title = "p-tau217 Trajectories by Composite Risk Score") +
  theme_bw(base_size = 12) +
  theme(plot.title    = element_text(face = "bold", size = 10, hjust = 0.5),
        axis.title=element_text(size = 8),
        legend.position = "bottom",
        legend.key.size = unit(0.5, "cm"),      
        legend.text = element_text(size = 5),  
        legend.title = element_text(size = 6),   
        legend.box.margin = margin(-10, 0, 0, 0),
        strip.background = element_rect(fill = "grey90"),
        strip.text       = element_text(face = "bold"))

print(p217_libra2)

ggsave(file.path(AIM1_OUTPUT_PATH$figures, "longitudinal_ptau217_LIBRA2_trajectory.png"),
       p217_libra2, width = 4, height = 3.3, units = "in",dpi = 300)
```


# Save results

```{r save-results}
# Save main results table
write_csv(lme_results,
          file.path(AIM1_OUTPUT_PATH$tables, "longitudinal_doseresponse_results.csv"))

# Save model fit statistics
model_fit <- bind_rows(
  lmm_dose_main %>%
    select(model, outcome, mod) %>%
    unnest(mod) %>%
    mutate(model_type = "Main Effect"),
  lmm_dose_interact %>%
    select(model, outcome, mod) %>%
    unnest(mod) %>%
    mutate(model_type = "Time Interaction")
) %>%
  select(model_type, model, outcome, AIC, BIC, logLik)

write_csv(model_fit,
          file.path(AIM1_OUTPUT_PATH$tables, "longitudinal_doseresponse_model_fit.csv"))

# Save interaction models and risk data for use in sensitivity analysis
saveRDS(lmm_dose_interact,
        file.path(DATA_INTERMEDIATE_PATH, "PREVENTAD_lmm_dose_interact.rds"))

# Save baseline risk scores and covariates per participant
risk_dat_baseline <- risk_dat_long %>%
  group_by(CONP_ID) %>%
  slice_min(years, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(CONP_ID, risk_score_cogd, risk_score_libra2, baseline_age, sex, educ)

saveRDS(risk_dat_baseline,
        file.path(DATA_INTERMEDIATE_PATH, "PREVENTAD_risk_dat_baseline.rds"))

cat("\nResults saved to:", AIM1_OUTPUT_PATH$tables, "\n")
cat("Figures saved to:", AIM1_OUTPUT_PATH$figures, "\n")
cat("Models saved to:", DATA_INTERMEDIATE_PATH, "\n")
```
