---
title: "Aim 1: Longitudinal Risk Burden Sensitivity Analysis"
format:
  html:
    embed-resources: true
    self-contained: true
    toc: true
    toc-depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = here::here())
```

```{r load-packages}
library(broom.mixed)
library(lme4)
library(patchwork)
library(splines)

source('workflow/scripts/config.R')
```

```{r load-data}
# Pre-built by 00_load_analysis_data.R
all_nomci_dat <- readRDS(file.path(DATA_CLEANED_PATH, "PREVENTAD_longitudinal_nomci.rds"))

cat("Data dimensions:", nrow(all_nomci_dat), "rows,", ncol(all_nomci_dat), "columns\n")
cat("Number of unique participants:", length(unique(all_nomci_dat$CONP_ID)), "\n")
cat("Time range:", min(all_nomci_dat$years, na.rm = TRUE), "to",
    max(all_nomci_dat$years, na.rm = TRUE), "years\n")
```

```{r define-vars}
# Outcome labels (defined here for use across growth-model plots)
outcome_labels <- c(
  ab_ratio               = "A\u03b242/40",
  gfap_log               = "GFAP",
  nfl_log                = "NfL",
  ptau181_log            = "ptau-181",
  ptau217_log            = "ptau-217",
  ptau217_ab42_ratio_log = "ptau-217/A\u03b242 ratio"
)

# subjects with 3+ ptau217 observations (with all covariates)
ptau217_subjects <- all_nomci_dat %>%
  filter(!is.na(ptau217), !is.na(risk_score_cogd), 
         !is.na(risk_score_libra2), !is.na(baseline_age), 
         !is.na(sex), !is.na(educ)) %>%
  count(CONP_ID) %>%
  filter(n >= 3) %>%
  pull(CONP_ID)
```

## Biomarker unconditional growth models

```{r growth-models}
# Unconditional growth: compare linear, ns(df=2), ns(df=3)
# Fit all three specifications per biomarker to inform choice of time term.
uncond_specs <- tribble(
  ~spec,        ~time_term,
  "linear",     "years",
  "ns_df2",     "ns(years, df = 2)",
  "ns_df3",     "ns(years, df = 3)"
)

fit_uncond <- function(o, e) {
  dat_sub <- all_nomci_dat[!is.na(all_nomci_dat[[o]]), ]
  tryCatch(
    lmer(as.formula(e), data = dat_sub, REML = FALSE),
    error = function(err) {
      message("Failed to fit: ", e, "\n  Reason: ", conditionMessage(err))
      NULL
    }
  )
}

uncond_models <- expand_grid(outcome = BIOMARKERS_TRANS, uncond_specs) %>%
  mutate(
    # Random slopes for outcomes with enough longitudinal data;
    # random intercept only for sparse outcomes
    re_term = if_else(outcome %in% sparse_outcomes,
                      "(1 | CONP_ID)",
                      "(1 + years | CONP_ID)"),
    eq  = glue("{outcome} ~ {time_term} + {re_term}"),
    res = map2(outcome, as.character(eq), fit_uncond),
    mod = map(res, ~if (is.null(.x)) tibble() else glance(.x))
  )

# AIC comparison table across specs (drops failed fits)
uncond_aic <- uncond_models %>%
  filter(!map_lgl(res, is.null)) %>%
  select(outcome, spec, mod) %>%
  unnest(mod) %>%
  select(outcome, spec, AIC, BIC, logLik) %>%
  group_by(outcome) %>%
  mutate(delta_AIC = AIC - min(AIC),
         best      = AIC == min(AIC)) %>%
  ungroup()

# Sequential likelihood ratio tests:
#   ns_df2 vs linear (does adding curvature help?)
#   ns_df3 vs ns_df2 (does extra df help further?)
uncond_lrt <- uncond_models %>%
  filter(!map_lgl(res, is.null)) %>%
  select(outcome, spec, res) %>%
  pivot_wider(names_from = spec, values_from = res) %>%
  rowwise() %>%
  mutate(
    p_vs_linear = if (!is.null(linear) && !is.null(ns_df2))
      anova(linear, ns_df2)$"Pr(>Chisq)"[2] else NA_real_,
    p_vs_df2    = if (!is.null(ns_df2) && !is.null(ns_df3))
      anova(ns_df2, ns_df3)$"Pr(>Chisq)"[2] else NA_real_
  ) %>%
  ungroup() %>%
  transmute(
    outcome,
    linear = NA_real_,
    ns_df2 = p_vs_linear,
    ns_df3 = p_vs_df2
  ) %>%
  pivot_longer(c(linear, ns_df2, ns_df3),
               names_to = "spec", values_to = "lrt_p_vs_simpler")

uncond_aic <- uncond_aic %>%
  left_join(uncond_lrt, by = c("outcome", "spec")) %>%
  arrange(outcome, AIC)

print(uncond_aic)

# Keep df=3 as the default for downstream plotting (retain original object name)
uncond_spline <- uncond_models %>% filter(spec == "ns_df3") %>% select(outcome, eq, res, mod)

# Predicted trajectories — use biomarker-specific time range to avoid extrapolation
biomarker_max_years_uncond <- all_nomci_dat %>%
  pivot_longer(cols = all_of(BIOMARKERS_TRANS), names_to = "outcome", values_to = "value") %>%
  filter(!is.na(value)) %>%
  group_by(outcome) %>%
  summarise(max_years = max(years, na.rm = TRUE), .groups = "drop")

spline_preds <- uncond_spline %>%
  left_join(biomarker_max_years_uncond, by = "outcome") %>%
  mutate(preds = map2(res, max_years, ~{
    d <- tibble(years = seq(0, .y, by = 0.2))
    d$pred <- predict(.x, newdata = d, re.form = NA)
    d
  })) %>%
  select(outcome, preds) %>%
  unnest(preds) %>%
  mutate(outcome_label = factor(outcome_labels[outcome], levels = outcome_labels))

# Long observed data
obs_long <- all_nomci_dat %>%
  pivot_longer(cols = all_of(BIOMARKERS_TRANS),
               names_to = "outcome", values_to = "observed") %>%
  filter(!is.na(observed)) %>%
  mutate(outcome_label = factor(outcome_labels[outcome], levels = outcome_labels))

p_uncond <- ggplot() +
  geom_point(data = obs_long,
             aes(x = years, y = observed),
             alpha = 0.12, size = 0.6, color = "grey50") +
  geom_line(data = spline_preds,
            aes(x = years, y = pred),
            color = "#DF8F44FF", linewidth = 1.1) +
  scale_x_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.03))) +
  facet_wrap(~outcome_label, scales = "free", ncol = 3) +
  labs(
    x        = "Years from Baseline",
    y        = "Biomarker Level",
    title    = "Unconditional Biomarker Trajectories Over Follow-Up",
    subtitle = "Population-average natural spline trajectory, df=3 (LMM, random intercept)"
  ) +
  theme_bw(base_size = 12) +
  theme(
    strip.text       = element_text(face = "bold"),
    strip.background = element_rect(fill = "grey90"),
    plot.title       = element_text(face = "bold", size = 14, hjust = 0.5),
    plot.subtitle    = element_text(size = 11, hjust = 0.5, color = "grey40"),
    panel.grid.minor = element_blank()
  )

print(p_uncond)

# focused plot of ptau217 to zoom in
# AIC of 2 vs 3 dfs
```

## LME Models

### Model 1: Main effect

These models test whether composite risk score (0-4 scale) predicts biomarker levels in a dose-response manner. The beta coefficient represents the change in biomarker level for each 1-unit increase in risk score (e.g., going from 2 to 3 risk factors). LIBRA2 models are adjusted for age, sex, and education.

```{r main-models}
models_main <- tribble(
  ~model, ~predictors,
  "CogDrisk_main", "years_centered + risk_score_cogd + (1 + years_centered | CONP_ID)",
  "LIBRA2_main", "years_centered + baseline_age + sex + educ + risk_score_libra2 + (1 + years_centered | CONP_ID)"
)

models_main_sparse <- tribble(
  ~model, ~predictors,
  "CogDrisk_main", "years_centered + risk_score_cogd + (1 | CONP_ID)",
  "LIBRA2_main",   "years_centered + baseline_age + sex + educ + risk_score_libra2 + (1 | CONP_ID)"
)

lme_main <- bind_rows(
  # all other outcomes — full random slopes, full data
  expand_grid(
    outcome = setdiff(BIOMARKERS_TRANS, "ptau217_ab42_ratio_log"),
    models_main
  ),
  # ptau217_ab42 — intercept only, full data
  expand_grid(
    outcome = "ptau217_ab42_ratio_log",
    models_main_sparse
  )
) %>%
  mutate(
    dat = map(outcome, ~case_when(
      . == "ptau217_log"      ~ list(filter(all_nomci_dat, CONP_ID %in% ptau217_subjects)),
      . == "ptau217_ab42_ratio_log" ~ list(all_nomci_dat),
      TRUE                ~ list(all_nomci_dat)
    ) %>% pluck(1)),
    eq    = glue("{outcome} ~ {predictors}"),
    res   = map2(eq, dat, ~lmer(as.formula(.x), data = .y)),
    dataf = map(res, ~tidy(., conf.int = TRUE)),
    mod   = map(res, glance)
  )
```

### Model 2: Time interactions (risk score × time)

These models test whether the rate of biomarker change over time differs by risk score level and other covariates. The interaction term (years_centered:risk_score) shows how much faster (or slower) biomarkers change for each additional risk factor. For LIBRA2, interactions between time and demographic covariates (baseline_age, sex, educ) are also included. 

```{r interact-models}
models_time_interact <- tribble(
  ~model, ~predictors,
  "CogDrisk_interact", "years_centered * risk_score_cogd + (1 + years_centered | CONP_ID)",
  "LIBRA2_interact", "years_centered * (risk_score_libra2 + baseline_age + sex + educ) + (1 + years_centered | CONP_ID)"
)

models_time_interact_sparse <- tribble(
  ~model, ~predictors,
  "CogDrisk_interact", "years_centered * risk_score_cogd + (1 | CONP_ID)",
  "LIBRA2_interact",   "years_centered * (baseline_age + sex + educ + risk_score_libra2) + (1 | CONP_ID)"
)

lme_time_interact <- bind_rows(
  # all other outcomes — full random slopes, full data
  expand_grid(
    outcome = setdiff(BIOMARKERS_TRANS, "ptau217_ab42_ratio_log"),
    models_time_interact
  ),
  # ptau217_ab42 — intercept only, full data
  expand_grid(
    outcome = "ptau217_ab42_ratio_log",
    models_time_interact_sparse
  )
) %>%
  mutate(
    dat = map(outcome, ~case_when(
      . == "ptau217_log"      ~ list(filter(all_nomci_dat, CONP_ID %in% ptau217_subjects)),
      . == "ptau217_ab42_ratio_log" ~ list(all_nomci_dat),
      TRUE                ~ list(all_nomci_dat)
    ) %>% pluck(1)),
    eq    = glue("{outcome} ~ {predictors}"),
    res   = map2(eq, dat, ~lmer(as.formula(.x), data = .y)),
    dataf = map(res, ~tidy(., conf.int = TRUE)),
    mod   = map(res, glance)
  )
```

## LME results

```{r lme-results}
# Combine all LME results
lme_results <- bind_rows(
  lme_main %>%
    select(model, outcome, dataf) %>%
    mutate(model_type = "Main Effect"),
  lme_time_interact %>%
    select(model, outcome, dataf) %>%
    mutate(model_type = "Time Interaction")
) %>%
  unnest(dataf) %>%
  filter(str_detect(term, "^risk_score|^years_centered:risk_score")) %>%
  mutate(
    # Confidence intervals
    conf.low = estimate - (std.error * 1.96),
    conf.high = estimate + (std.error * 1.96),

    # Significance
    sig = ifelse(p.value < 0.05, TRUE, FALSE),
    stars = case_when(
      p.value >= 0.05 ~ "",
      p.value >= 0.01 & p.value < 0.05 ~ "*",
      p.value >= 0.001 & p.value < 0.01 ~ "**",
      p.value < 0.001 ~ "***"
    ),

    # Exponentiated values for log-transformed outcomes
    exp_est = case_when(
      str_detect(outcome, "_log$") ~ exp(estimate),
      TRUE ~ NA_real_
    ),
    exp_lci = case_when(
      str_detect(outcome, "_log$") ~ exp(conf.low),
      TRUE ~ NA_real_
    ),
    exp_uci = case_when(
      str_detect(outcome, "_log$") ~ exp(conf.high),
      TRUE ~ NA_real_
    ),

    # Percentage change per unit for all biomarkers
    pct_change_per_unit = case_when(
      str_detect(outcome, "_log$") ~ (exp_est - 1) * 100,
      TRUE ~ NA_real_
    ),

    # Risk score type
    risk_type = if_else(str_detect(model, "CogDrisk"), "CogDrisk", "LIBRA2")
  ) %>%
  mutate(across(c(estimate, std.error, statistic, p.value,
                  conf.low, conf.high, exp_est, exp_lci, exp_uci, pct_change_per_unit),
                ~round(., digits = 3)))
```

### Main effect results

```{r main-models-plot}
#| fig-width: 10
#| fig-height: 6

plot_data_main <- lme_results %>%
  filter(str_detect(model, "main")) %>%
  mutate(
    sig        = factor(sig, levels = c(TRUE, FALSE)),
    risk_type  = factor(risk_type, levels = c("LIBRA2", "CogDrisk")),
    outcome_label = case_when(
      outcome == "ab_ratio"               ~ "A\u03b242/40",
      outcome == "gfap_log"               ~ "GFAP (log)",
      outcome == "nfl_log"                ~ "NfL (log)",
      outcome == "ptau181_log"            ~ "p-tau181 (log)",
      outcome == "ptau217_log"            ~ "p-tau217 (log)",
      outcome == "ptau217_ab42_ratio_log" ~ "p-tau217/A\u03b242 (log)",
      TRUE ~ outcome
    ),
    outcome_label = factor(outcome_label, levels = rev(c(
      "A\u03b242/40", "GFAP (log)", "NfL (log)",
      "p-tau181 (log)", "p-tau217 (log)", "p-tau217/A\u03b242 (log)"
    )))
  )

p_main_combined <- ggplot(plot_data_main,
  aes(x = estimate, y = outcome_label,
      xmin = conf.low, xmax = conf.high,
      color = risk_type, alpha = sig,
      group = risk_type)
) +
  geom_point(size = 3, position = position_dodge(width = 0.6)) +
  geom_errorbarh(linewidth = 0.8, position = position_dodge(width = 0.6), height = 0.3) +
  geom_vline(xintercept = 0, linetype = 2, color = "grey40") +
  scale_alpha_manual(
    values = c(`FALSE` = 0.25, `TRUE` = 1),
    labels = c(`FALSE` = "Not Significant", `TRUE` = "Significant"),
    name = "Significance"
  ) +
  scale_color_manual(
    values = c("CogDrisk" = "#374E55FF", "LIBRA2" = "#DF8F44FF"),
    breaks = c("CogDrisk", "LIBRA2"),
    name = "Model"
  ) +
  labs(
    x     = "Beta (95% CI)",
    y     = NULL,
    title = "Figure 3. Baseline Association with Risk Score"
  ) +
  theme_bw(base_size = 13) +
  theme(
    legend.position  = "bottom",
    legend.title     = element_text(size = 10, face = "bold"),
    legend.text      = element_text(size = 9),
    panel.grid.minor = element_blank(),
    plot.title       = element_text(face = "bold", size = 14, hjust = 0.5),
    plot.subtitle    = element_text(size = 11, hjust = 0.5, color = "grey40"),
    axis.text        = element_text(size = 11)
  )

print(p_main_combined)

# ggsave(file.path(AIM1_OUTPUT_PATH$figures, "fig3_lme_main_effects.png"),
#        p_main_combined, width = 10, height = 6, dpi = 300)
```

```{r main-models-results}
biomarker_order_lme <- c(
  "A\u03b242/40", "GFAP", "NfL", "ptau-181", "ptau-217", "ptau-217/A\u03b242 ratio"
)

term_order_lme_main <- c(
  "Composite Risk Score",
  "Covariates",
  "Age", "Sex (Male)", "Education (years)"
)

covariate_terms_lme <- c("Age", "Sex (Male)", "Education (years)")

main_all_raw <- lme_main %>%
  select(model, outcome, dataf) %>%
  unnest(dataf) %>%
  filter(effect == "fixed",
         !term %in% c("(Intercept)", "years_centered")) %>%
  mutate(
    conf.low  = estimate - 1.96 * std.error,
    conf.high = estimate + 1.96 * std.error,
    beta_se   = sprintf("%.3f (%.3f)", estimate, std.error),
    ci        = sprintf("[%.3f, %.3f]", conf.low, conf.high),
    p_fmt     = if_else(p.value < 0.001, "<0.001", sprintf("%.3f", p.value)),
    stars     = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      TRUE            ~ ""
    ),
    model_short = if_else(str_detect(model, "CogDrisk"), "CogDrisk", "LIBRA2"),
    biomarker   = case_when(
      outcome == "ab_ratio"               ~ "A\u03b242/40",
      outcome == "gfap_log"               ~ "GFAP",
      outcome == "nfl_log"                ~ "NfL",
      outcome == "ptau181_log"            ~ "ptau-181",
      outcome == "ptau217_log"            ~ "ptau-217",
      outcome == "ptau217_ab42_ratio_log" ~ "ptau-217/A\u03b242 ratio"
    ),
    term_label  = case_when(
      str_detect(term, "^risk_score") ~ "Composite Risk Score",
      term == "baseline_age"          ~ "Age",
      term == "sexMale"               ~ "Sex (Male)",
      term == "educ"                  ~ "Education (years)",
      TRUE ~ term
    )
  )

main_wide <- main_all_raw %>%
  select(model_short, biomarker, term_label, beta_se, ci, p_fmt, stars) %>%
  pivot_wider(
    names_from  = model_short,
    values_from = c(beta_se, ci, p_fmt, stars),
    names_glue  = "{model_short}_{.value}"
  ) %>%
  mutate(
    across(c(CogDrisk_beta_se, CogDrisk_ci, CogDrisk_p_fmt), ~replace_na(., "\u2014")),
    CogDrisk_stars = replace_na(CogDrisk_stars, "")
  )

# "Covariates" separator row per biomarker
covariate_header_lme <- main_wide %>%
  distinct(biomarker) %>%
  mutate(
    term_label       = "Covariates",
    CogDrisk_beta_se = "", CogDrisk_ci = "", CogDrisk_p_fmt = "", CogDrisk_stars = "",
    LIBRA2_beta_se   = "", LIBRA2_ci   = "", LIBRA2_p_fmt   = "", LIBRA2_stars   = ""
  )

main_combined <- bind_rows(main_wide, covariate_header_lme) %>%
  mutate(
    biomarker  = factor(biomarker,  levels = biomarker_order_lme),
    term_label = factor(term_label, levels = term_order_lme_main)
  ) %>%
  arrange(biomarker, term_label) %>%
  select(biomarker, term_label,
         CogDrisk_beta_se, CogDrisk_ci, CogDrisk_p_fmt, CogDrisk_stars,
         LIBRA2_beta_se,   LIBRA2_ci,   LIBRA2_p_fmt,   LIBRA2_stars)

main_gt <- main_combined %>%
  gt(groupname_col = "biomarker") %>%
  tab_header(
    title    = md("**Table S4.** Main Effect of Composite Risk Score on Plasma Biomarker Levels"),
    subtitle = md("Linear mixed-effects models. LIBRA2 additionally adjusted for age, sex, and education.")
  ) %>%
  tab_spanner(label = md("**CogDrisk**"), columns = starts_with("CogDrisk_")) %>%
  tab_spanner(label = md("**LIBRA2**"),   columns = starts_with("LIBRA2_")) %>%
  cols_label(
    term_label       = md("**Predictor**"),
    CogDrisk_beta_se = md("**\u03b2 (SE)**"),
    CogDrisk_ci      = md("**95% CI**"),
    CogDrisk_p_fmt   = md("**p**"),
    CogDrisk_stars   = md(""),
    LIBRA2_beta_se   = md("**\u03b2 (SE)**"),
    LIBRA2_ci        = md("**95% CI**"),
    LIBRA2_p_fmt     = md("**p**"),
    LIBRA2_stars     = md("")
  ) %>%
  tab_style(
    style     = cell_fill(color = "#e8f4e8"),
    locations = cells_body(columns = CogDrisk_p_fmt,
                           rows    = CogDrisk_stars %in% c("*", "**", "***"))
  ) %>%
  tab_style(
    style     = cell_fill(color = "#e8f4e8"),
    locations = cells_body(columns = LIBRA2_p_fmt,
                           rows    = LIBRA2_stars %in% c("*", "**", "***"))
  ) %>%
  tab_style(
    style     = list(cell_text(weight = "bold"), cell_fill(color = "#f8f9fa")),
    locations = cells_row_groups()
  ) %>%
  tab_style(
    style     = list(cell_text(weight = "bold", style = "italic"),
                     cell_fill(color = "#f0f2f4")),
    locations = cells_body(rows = term_label == "Covariates")
  ) %>%
  tab_style(
    style     = cell_text(indent = px(16)),
    locations = cells_body(
      columns = term_label,
      rows    = term_label %in% covariate_terms_lme
    )
  ) %>%
  tab_footnote(
    footnote  = md("*** p<0.001; ** p<0.01; * p<0.05. \u03b2 = change in biomarker per 1-unit increase in composite risk score. Highlighted cells indicate p<0.05. \u2014 = not applicable (CogDrisk not adjusted for these covariates)."),
    locations = cells_column_spanners()
  ) %>%
  cols_align(align = "left",   columns = term_label) %>%
  cols_align(align = "center", columns = c(starts_with("CogDrisk_"), starts_with("LIBRA2_"))) %>%
  row_group_order(groups = biomarker_order_lme)

main_gt

# gtsave(main_gt, filename = file.path(AIM1_OUTPUT_PATH$tables, "s_table4_lme_main.docx"))
```

### Time interaction results

```{r interact-models-plot}
#| fig-width: 10
#| fig-height: 6

plot_data_interact <- lme_results %>%
  filter(str_detect(model, "interact")) %>%
  filter(str_detect(term, "years_centered:")) %>%
  mutate(
    sig        = factor(sig, levels = c(TRUE, FALSE)),
    risk_type  = factor(risk_type, levels = c("LIBRA2", "CogDrisk")),
    outcome_label = case_when(
      outcome == "ab_ratio"               ~ "A\u03b242/40",
      outcome == "gfap_log"               ~ "GFAP (log)",
      outcome == "nfl_log"                ~ "NfL (log)",
      outcome == "ptau181_log"            ~ "p-tau181 (log)",
      outcome == "ptau217_log"            ~ "p-tau217 (log)",
      outcome == "ptau217_ab42_ratio_log" ~ "p-tau217/A\u03b242 (log)",
      TRUE ~ outcome
    ),
    outcome_label = factor(outcome_label, levels = rev(c(
      "A\u03b242/40", "GFAP (log)", "NfL (log)",
      "p-tau181 (log)", "p-tau217 (log)", "p-tau217/A\u03b242 (log)"
    )))
  )

p_interact_combined <- ggplot(plot_data_interact,
  aes(x = estimate, y = outcome_label,
      xmin = conf.low, xmax = conf.high,
      color = risk_type, alpha = sig,
      group = risk_type)
) +
  geom_point(size = 3, position = position_dodge(width = 0.6)) +
  geom_errorbarh(linewidth = 0.8, position = position_dodge(width = 0.6), height = 0.3) +
  geom_vline(xintercept = 0, linetype = 2, color = "grey40") +
  scale_alpha_manual(
    values = c(`FALSE` = 0.25, `TRUE` = 1),
    labels = c(`FALSE` = "Not Significant", `TRUE` = "Significant"),
    name = "Significance"
  ) +
  scale_color_manual(
    values = c("CogDrisk" = "#374E55FF", "LIBRA2" = "#DF8F44FF"),
    breaks = c("CogDrisk", "LIBRA2"),
    name = "Model"
  ) +
  labs(
    x     = "Change in Biomarker per Unit Increase in Risk Score (95% CI)",
    y     = NULL,
    title = "Longitudinal Risk Burden: Time Interaction",
    subtitle = "Change in slope per unit increase in composite risk score (LME)"
  ) +
  theme_bw(base_size = 13) +
  theme(
    legend.position  = "bottom",
    legend.title     = element_text(size = 10, face = "bold"),
    legend.text      = element_text(size = 9),
    panel.grid.minor = element_blank(),
    plot.title       = element_text(face = "bold", size = 14, hjust = 0.5),
    plot.subtitle    = element_text(size = 11, hjust = 0.5, color = "grey40"),
    axis.text        = element_text(size = 11)
  )

print(p_interact_combined)
```

```{r interact-models-results}
interact_combined_raw <- lme_results %>%
  filter(model_type == "Time Interaction",
         str_detect(term, "years_centered:risk_score")) %>%
  mutate(
    beta_se    = sprintf("%.3f (%.3f)", estimate, std.error),
    ci         = sprintf("[%.3f, %.3f]", conf.low, conf.high),
    p_fmt      = if_else(p.value < 0.001, "<0.001", sprintf("%.3f", p.value)),
    biomarker  = case_when(
      outcome == "ab_ratio"               ~ "A\u03b242/40",
      outcome == "gfap_log"               ~ "GFAP",
      outcome == "nfl_log"                ~ "NfL",
      outcome == "ptau181_log"            ~ "ptau-181",
      outcome == "ptau217_log"            ~ "ptau-217",
      outcome == "ptau217_ab42_ratio_log" ~ "ptau-217/A\u03b242 ratio"
    ),
    term_label  = "Time \u00d7 Composite Risk Score",
    model_short = risk_type
  )

interact_combined_wide <- interact_combined_raw %>%
  select(model_short, biomarker, term_label, beta_se, ci, p_fmt, stars) %>%
  pivot_wider(
    names_from  = model_short,
    values_from = c(beta_se, ci, p_fmt, stars),
    names_glue  = "{model_short}_{.value}"
  ) %>%
  mutate(biomarker = factor(biomarker, levels = biomarker_order_lme)) %>%
  arrange(biomarker)

interact_gt <- interact_combined_wide %>%
  gt(groupname_col = "biomarker") %>%
  tab_header(
    title    = md("**Table S5.** Time \u00d7 Composite Risk Score Interaction on Plasma Biomarker Trajectories"),
    subtitle = md("Linear mixed-effects models. LIBRA2 additionally adjusted for time interactions with age, sex, and education.")
  ) %>%
  tab_spanner(label = md("**CogDrisk**"), columns = starts_with("CogDrisk_")) %>%
  tab_spanner(label = md("**LIBRA2**"),   columns = starts_with("LIBRA2_")) %>%
  cols_label(
    term_label       = md("**Predictor**"),
    CogDrisk_beta_se = md("**\u03b2 (SE)**"),
    CogDrisk_ci      = md("**95% CI**"),
    CogDrisk_p_fmt   = md("**p**"),
    CogDrisk_stars   = md(""),
    LIBRA2_beta_se   = md("**\u03b2 (SE)**"),
    LIBRA2_ci        = md("**95% CI**"),
    LIBRA2_p_fmt     = md("**p**"),
    LIBRA2_stars     = md("")
  ) %>%
  tab_style(
    style     = cell_fill(color = "#e8f4e8"),
    locations = cells_body(columns = CogDrisk_p_fmt,
                           rows    = CogDrisk_stars %in% c("*", "**", "***"))
  ) %>%
  tab_style(
    style     = cell_fill(color = "#e8f4e8"),
    locations = cells_body(columns = LIBRA2_p_fmt,
                           rows    = LIBRA2_stars %in% c("*", "**", "***"))
  ) %>%
  tab_style(
    style     = list(cell_text(weight = "bold"), cell_fill(color = "#f8f9fa")),
    locations = cells_row_groups()
  ) %>%
  tab_footnote(
    footnote  = md("*** p<0.001; ** p<0.01; * p<0.05. \u03b2 = additional change in biomarker slope per 1-unit increase in composite risk score. Highlighted cells indicate p<0.05."),
    locations = cells_column_spanners()
  ) %>%
  cols_align(align = "left",   columns = term_label) %>%
  cols_align(align = "center", columns = c(starts_with("CogDrisk_"), starts_with("LIBRA2_"))) %>%
  row_group_order(groups = biomarker_order_lme)

interact_gt

# gtsave(interact_gt, filename = file.path(AIM1_OUTPUT_PATH$tables, "s_table5_lme_interact.docx"))
```

# Trajectory plots

**Methodology: Generating Predicted Trajectories**

The predicted biomarker trajectories are generated from the LME models fitted above using the following approach:

**1. Model Specification:**
- **Base model**: `biomarker ~ years_centered × risk_score + (1 | participant_ID)`
- The time × risk score interaction allows trajectories to have different slopes for each risk level
- Random intercepts account for between-participant variability in baseline biomarker levels

**2. Prediction Grid Construction:**
- For each biomarker, a prediction grid is created spanning:
  - **Time range**: From baseline (0 years) to the maximum observed follow-up for that biomarker
  - **Risk scores**: All possible composite risk levels (0, 1, 2, 3, 4)
  - **Time resolution**: Predictions at 0.5-year intervals for smooth trajectory lines
- Time values are grand-mean centered before prediction (where 0 represents ~2.6 years of follow-up), then transformed back to original scale for plotting

**3. Population-Level Predictions:**
- Predictions use `predict(..., re.form = NA)` to obtain **marginal** (population-average) predictions
- This marginalizes over random effects, showing the expected trajectory for an "average" participant at each risk level
- Individual-level predictions would include random intercepts, but would obscure population-level patterns

**4. Biomarker-Specific Time Ranges:**
- ptau-217: predictions extend to ~11 years (maximum observed follow-up)
- Other biomarkers (ptau-181, GFAP, NfL, Aβ42/40, ptau-217/Aβ42): predictions extend to ~4.6 years
- This prevents extrapolation beyond observed data ranges

**5. Covariate Values (LIBRA2 only):**
- LIBRA2 models include demographic covariates with time interactions
- Predictions are generated for a reference profile:
  - Baseline age: sample mean
  - Sex: Female
  - Education: sample mean years
- Trajectories for other demographic profiles would differ due to covariate × time interactions

## CogDrisk

```{r trajectories-cogdrisk}
#| fig-width: 14
#| fig-height: 10
#| fig-dpi: 96

# Calculate maximum follow-up time per biomarker
biomarker_max_years <- all_nomci_dat %>%
  pivot_longer(
    cols = all_of(BIOMARKERS_TRANS),
    names_to = "outcome",
    values_to = "value"
  ) %>%
  filter(!is.na(value)) %>%
  group_by(outcome) %>%
  summarise(max_years = max(years, na.rm = TRUE))

# Grand mean of years (used to center time in the models)
mean_years <- mean(all_nomci_dat$years, na.rm = TRUE)

# Generate predictions from interaction model
cogdrisk_interact_models <- lme_time_interact %>%
  filter(str_detect(model, "CogDrisk"))

# Create biomarker-specific prediction data and generate predictions
cogdrisk_predictions <- cogdrisk_interact_models %>%
  left_join(biomarker_max_years, by = "outcome") %>%
  mutate(
    predictions = map2(res, max_years, ~{
      # Create prediction grid specific to this biomarker's time range
      pred_data <- expand_grid(
        years_centered = seq(-mean_years, .y - mean_years, by = 0.5),
        risk_score_cogd = 0:4
      ) %>%
        mutate(years = years_centered + mean_years)

      # Generate predictions
      pred_data$predicted <- predict(.x, newdata = pred_data, re.form = NA)
      pred_data
    })
  ) %>%
  select(outcome, predictions) %>%
  unnest(predictions)

# Prepare observed data (kept for downstream chunks)
all_nomci_dat_long <- all_nomci_dat %>%
  pivot_longer(
    cols = all_of(BIOMARKERS_TRANS),
    names_to = "outcome",
    values_to = "observed"
  )

# Readable outcome labels (ordered for consistent facet display)
outcome_labels <- c(
  ab_ratio               = "A\u03b242/40",
  gfap_log               = "GFAP",
  nfl_log                = "NfL",
  ptau181_log            = "ptau-181",
  ptau217_log            = "ptau-217",
  ptau217_ab42_ratio_log = "ptau-217/A\u03b242 ratio"
)

# Add readable labels to plotting datasets (separate vars; risk_dat_long_plot unchanged)
# Drop NA observed values so late-visit rows (e.g. GFAP=NA at year 10) don't
# expand the x-scale for panels where that biomarker wasn't measured.
risk_dat_long_plot_lbl <- all_nomci_dat_long %>%
  filter(!is.na(observed)) %>%
  mutate(outcome_label = factor(outcome_labels[outcome], levels = outcome_labels))

cogdrisk_pred_lbl <- cogdrisk_predictions %>%
  mutate(outcome_label = factor(outcome_labels[outcome], levels = outcome_labels))

# Interaction: time × risk score (divergence in slope)
cogd_annot_inter <- lme_results %>%
  filter(model_type == "Time Interaction",
         risk_type  == "CogDrisk",
         str_detect(term, "years_centered:risk_score")) %>%
  mutate(
    outcome_label = factor(outcome_labels[outcome], levels = outcome_labels),
    p_fmt = if_else(p.value < 0.001, "< 0.001", as.character(p.value)),
    annot = paste0("Time \u00d7 Risk: \u03b2 = ", estimate,
                   " [", conf.low, ", ", conf.high, "], \n  p = ", p_fmt, stars)
  )

# Build descriptive caption
n_pts <- length(unique(all_nomci_dat_long$CONP_ID))
n_obs <- nrow(all_nomci_dat_long)

caption_text <- paste0(
  "Cohort: PREVENT-AD. N = ", n_pts, "; Observations = ", n_obs, "\n",
  "Risk indicators: APOE e4+; Family history (\u22652 first-degree relatives);  Genetic risk score (top quartile); ",
  "CogDrisk > 1 SD above mean. Composite score: 0\u20134 scale (sum of binary indicators).\n",
  "Lines = model-predicted population trajectories (LMM: biomarker ~ years \u00d7 risk_score + (1 | ID)). ",
  "Axis scales differ by biomarker to reflect observed data range."
)

# Plot trajectories
p_cogd_traj <- ggplot() +
  geom_point(data = risk_dat_long_plot_lbl,
             aes(x = years, y = observed, color = factor(risk_score_cogd)),
             alpha = 0.2, size = 0.8) +
  geom_line(data = cogdrisk_pred_lbl,
            aes(x = years, y = predicted,
                color = factor(risk_score_cogd),
                group = risk_score_cogd),
            linewidth = 1.2) +
  geom_label(
    data = cogd_annot_inter,
    aes(x = -Inf, y = Inf, label = annot),
    inherit.aes = FALSE,
    hjust = -0.04, vjust = 1.75,
    size = 5, color = "grey20",
    label.padding = unit(0.2, "lines")
  ) +
  scale_color_viridis_d(option = "plasma", begin = 0.2, end = 0.9,
                        name = "# Risk Indicators") +
  scale_x_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.05))) +
  scale_y_continuous(expand = expansion(mult = c(0.02, 0.05))) +
  facet_wrap(~outcome_label, scales = "free", ncol = 3) +
  labs(
    x = "Years from Baseline",
    y = "Biomarker Level",
    title = "Figure 4. Predicted Plasma Biomarker Trajectories by CogDrisk-based Composite Risk Score"
  ) +
  theme_bw(base_size = 12) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold"),
    strip.background = element_rect(fill = "grey90"),
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5)#,
    # plot.subtitle = element_text(size = 12, hjust = 0.5),
    # plot.caption = element_text(size = 9, hjust = 0, lineheight = 1.2,
    #                             margin = margin(t = 10))
  )

print(p_cogd_traj)

# ggsave(file.path(AIM1_OUTPUT_PATH$figures, "fig4_longitudinal_CogDrisk_trajectories.png"),
#        p_cogd_traj, width = 14, height = 10, dpi = 250)
```

## LIBRA2

```{r trajectories-libra2}
#| fig-width: 14
#| fig-height: 10
#| fig-dpi: 96

# Generate predictions from interaction model
libra2_interact_models <- lme_time_interact %>%
  filter(str_detect(model, "LIBRA2"))

# Create biomarker-specific prediction data and generate predictions
libra2_predictions <- libra2_interact_models %>%
  left_join(biomarker_max_years, by = "outcome") %>%
  mutate(
    predictions = map2(res, max_years, ~{
      # Create prediction grid specific to this biomarker's time range
      pred_data <- expand_grid(
        years_centered = seq(-mean_years, .y - mean_years, by = 0.5),
        risk_score_libra2 = 0:4,
        baseline_age = mean(all_nomci_dat$baseline_age, na.rm = TRUE),
        sex = "Female",
        educ = mean(all_nomci_dat$educ, na.rm = TRUE)
      ) %>%
        mutate(years = years_centered + mean_years)

      # Generate predictions
      pred_data$predicted <- predict(.x, newdata = pred_data, re.form = NA)
      pred_data
    })
  ) %>%
  select(outcome, predictions) %>%
  unnest(predictions)

# Add readable labels; drop NA observed to avoid x-scale inflation
libra2_obs_lbl <- all_nomci_dat_long %>%
  filter(!is.na(observed)) %>%
  mutate(outcome_label = factor(outcome_labels[outcome], levels = outcome_labels))

libra2_pred_lbl <- libra2_predictions %>%
  mutate(outcome_label = factor(outcome_labels[outcome], levels = outcome_labels))

# Per-panel annotation: time × risk score interaction
libra2_annot_inter <- lme_results %>%
  filter(model_type == "Time Interaction",
         risk_type  == "LIBRA2",
         str_detect(term, "years_centered:risk_score")) %>%
  mutate(
    outcome_label = factor(outcome_labels[outcome], levels = outcome_labels),
    p_fmt = if_else(p.value < 0.001, "< 0.001", as.character(p.value)),
    annot = paste0("Time \u00d7 Risk: \u03b2 = ", estimate,
                   " [", conf.low, ", ", conf.high, "], \n  p = ", p_fmt, stars)
  )

caption_text_libra2 <- paste0(
  "Cohort: PREVENT-AD. N = ", n_pts, "; Observations = ", n_obs, "\n",
  "LIBRA2 composite risk score (0\u20134). Reference profile: female, mean age, mean education.\n",
  "Lines = model-predicted population trajectories (LMM: biomarker ~ years \u00d7 risk_score + covariates + (1 | ID)). ",
  "Axis scales differ by biomarker to reflect observed data range."
)

# Plot trajectories
p_libra_traj <- ggplot() +
  geom_point(data = libra2_obs_lbl,
             aes(x = years, y = observed, color = factor(risk_score_libra2)),
             alpha = 0.2, size = 0.8) +
  geom_line(data = libra2_pred_lbl,
            aes(x = years, y = predicted,
                color = factor(risk_score_libra2),
                group = risk_score_libra2),
            linewidth = 1.2) +
  geom_label(
    data        = libra2_annot_inter,
    aes(x = -Inf, y = Inf, label = annot),
    inherit.aes = FALSE,
    hjust = -0.04, vjust = 1.75,
    size = 5, color = "grey20",
    label.padding = unit(0.2, "lines")
  ) +
  scale_color_viridis_d(option = "plasma", begin = 0.2, end = 0.9,
                        name = "# Risk Indicators") +
  scale_x_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.05))) +
  scale_y_continuous(expand = expansion(mult = c(0.02, 0.05))) +
  facet_wrap(~outcome_label, scales = "free", ncol = 3) +
  labs(
    x = "Years from Baseline",
    y = "Biomarker Level",
    title = "Figure 5. Predicted Plasma Biomarker Trajectories by LIBRA2-based Composite Risk Score"
    # caption = caption_text_libra2
  ) +
  theme_bw(base_size = 12) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold"),
    strip.background = element_rect(fill = "grey90"),
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5)#,
    # plot.caption = element_text(size = 9, hjust = 0, lineheight = 1.2,
    #                             margin = margin(t = 10))
  )

print(p_libra_traj)

# ggsave(file.path(AIM1_OUTPUT_PATH$figures, "fig5_longitudinal_LIBRA2_trajectories.png"),
#        p_libra_traj, width = 14, height = 10, dpi = 250)
```

## Focused Plot: CogDrisk ptau Trajectories

```{r ptau-cogdrisk-focused-plot}
#| fig-width: 12
#| fig-height: 8

# Filter data for ptau217 and ptau181
ptau217_obs <- risk_dat_long_plot %>%
  filter(outcome == "ptau217_log")

ptau181_obs <- risk_dat_long_plot %>%
  filter(outcome == "ptau181_log")

ptau217_pred <- cogdrisk_predictions %>%
  filter(outcome == "ptau217_log")

ptau181_pred <- cogdrisk_predictions %>%
  filter(outcome == "ptau181_log")

# Create separate plots for each biomarker with appropriate x-axis limits
p_ptau217 <- ggplot() +
  geom_point(data = ptau217_obs,
             aes(x = years, y = observed, color = factor(risk_score_cogd)),
             alpha = 0.3, size = 1.5) +
  geom_line(data = ptau217_pred,
            aes(x = years, y = predicted,
                color = factor(risk_score_cogd),
                group = risk_score_cogd),
            linewidth = 1.5) +
  scale_color_viridis_d(option = "plasma", begin = 0.2, end = 0.9,
                        name = "Risk Score") +
  scale_x_continuous(limits = c(0, 12), breaks = seq(0, 12, 2)) +
  labs(
    x = "Years from Baseline",
    y = "ptau-217 (log-transformed)",
    title = "ptau-217 (11 years follow-up)"
  ) +
  theme_bw(base_size = 13) +
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold", size = 12)
  )

# Create annotation text with cohort and risk indicator information
n_total <- length(unique(risk_dat_long$CONP_ID))
n_obs <- nrow(risk_dat_long)

annotation_text <- paste0(
  "Cohort: PREVENT-AD. N = ", n_total, "; Observations = ", n_obs, "\n",
  "Risk indicators: FHx: 2+ first-degree relatives with dementia; APOE: e4+ carrier; ",
  "GRS: Genetic risk score > 1 SD above mean; CogDrisk: Modifiable risk score > 1 SD above mean\n",
  "Composite risk score: 0-4 scale (sum of binary indicators)"
)

# Combine plots using patchwork
p_ptau_cogd <- p_ptau217 / p_ptau181 +
  plot_annotation(
    title = "CogDrisk: Phosphorylated Tau Trajectories by Risk Score",
    caption = annotation_text,
    theme = theme(
      plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
      plot.caption = element_text(size = 9, hjust = 0, lineheight = 1.2, margin = margin(t = 10))
    )
  )

print(p_ptau_cogd)
print(p_ptau181)

# Save plot
ggsave(file.path(AIM1_OUTPUT_PATH$figures, "PREVENTAD_ptau_CogDrisk_trajectories.png"),
       p_ptau_cogd, width = 12, height = 9, dpi = 300)
ggsave(file.path(AIM1_OUTPUT_PATH$figures, "PREVENTAD_ptau181_trajectories.png"),
       p_ptau181, width = 12, height = 9, dpi = 300)
```

## Focused Plot: LIBRA2 ptau217 Trajectories 

```{r ptau-combined-focused-plot}
ptau217_obs <- risk_dat_long_plot %>%
  filter(outcome == "ptau217_log")

ptau217_pred_libra2 <- libra2_predictions %>%
  filter(outcome == "ptau217_log")

# Interaction term annotations for ptau217 — LIBRA2
ptau217_annot_libra2 <- dose_response_results %>%
  filter(model_type == "Time Interaction",
         risk_type  == "LIBRA2",
         outcome    == "ptau217_log",
         str_detect(term, "years_centered:risk_score")) %>%
  mutate(
    p_fmt = if_else(p.value < 0.001, "< 0.001", as.character(p.value)),
    annot = paste0("Time\u00d7Risk: \u03b2 = ", estimate,
                   " [", conf.low, ", ", conf.high, "], p = ", p_fmt, stars)
  )

n_ptau217_obs <- nrow(risk_dat_long %>% filter(!is.na(ptau217_log)))

# ptau-217: LIBRA2
p217_libra2 <- ggplot() +
  geom_point(data = ptau217_obs,
             aes(x = years, y = observed, color = factor(risk_score_libra2)),
             alpha = 0.3, size = 1.2) +
  geom_line(data = ptau217_pred_libra2,
            aes(x = years, y = predicted,
                color = factor(risk_score_libra2), group = risk_score_libra2),
            linewidth = 1.5) +
  geom_label(data        = ptau217_annot_libra2,
             aes(x = -Inf, y = Inf, label = annot),
             inherit.aes = FALSE,
             hjust = -0.04, vjust = 2,
             size = 3, color = "grey20",
             label.padding = unit(0.2, "lines")) +
  scale_color_viridis_d(option = "plasma", begin = 0.2, end = 0.9,
                        name = "# Risk Indicators") +
  scale_x_continuous(limits = c(0, 12), breaks = seq(0, 12, 2)) +
  scale_y_continuous(expand = expansion(mult = c(0.02, 0.05))) +
  labs(x = "Years from Baseline", y = NULL,
       title = "p-tau217 Trajectories by Composite Risk Score") +
  theme_bw(base_size = 12) +
  theme(plot.title    = element_text(face = "bold", size = 10, hjust = 0.5),
        axis.title=element_text(size = 8),
        legend.position = "bottom",
        legend.key.size = unit(0.5, "cm"),      
        legend.text = element_text(size = 5),  
        legend.title = element_text(size = 6),   
        legend.box.margin = margin(-10, 0, 0, 0),
        strip.background = element_rect(fill = "grey90"),
        strip.text       = element_text(face = "bold"))

print(p217_libra2)

ggsave(file.path(AIM1_OUTPUT_PATH$figures, "longitudinal_ptau217_LIBRA2_trajectory.png"),
       p217_libra2, width = 4, height = 3.3, units = "in",dpi = 300)
```


# Save results

```{r save-results}
# Save main results table
write_csv(lme_results,
          file.path(AIM1_OUTPUT_PATH$tables, "longitudinal_doseresponse_results.csv"))

# Save model fit statistics
model_fit <- bind_rows(
  lmm_dose_main %>%
    select(model, outcome, mod) %>%
    unnest(mod) %>%
    mutate(model_type = "Main Effect"),
  lmm_dose_interact %>%
    select(model, outcome, mod) %>%
    unnest(mod) %>%
    mutate(model_type = "Time Interaction")
) %>%
  select(model_type, model, outcome, AIC, BIC, logLik)

write_csv(model_fit,
          file.path(AIM1_OUTPUT_PATH$tables, "longitudinal_doseresponse_model_fit.csv"))

# Save interaction models and risk data for use in sensitivity analysis
saveRDS(lmm_dose_interact,
        file.path(DATA_INTERMEDIATE_PATH, "PREVENTAD_lmm_dose_interact.rds"))

# Save baseline risk scores and covariates per participant
risk_dat_baseline <- risk_dat_long %>%
  group_by(CONP_ID) %>%
  slice_min(years, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(CONP_ID, risk_score_cogd, risk_score_libra2, baseline_age, sex, educ)

saveRDS(risk_dat_baseline,
        file.path(DATA_INTERMEDIATE_PATH, "PREVENTAD_risk_dat_baseline.rds"))

cat("\nResults saved to:", AIM1_OUTPUT_PATH$tables, "\n")
cat("Figures saved to:", AIM1_OUTPUT_PATH$figures, "\n")
cat("Models saved to:", DATA_INTERMEDIATE_PATH, "\n")
```
